### 2022-02-20 example_draft.Rmd
first version of rmarkdown draft

---
title: "example_draft"
author: "Jon Minton"
date: "20/02/2022"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# A Special title

This is my introduction to what I'm doing.

This is something that's *quite* important.

This is something that's **really** important.

## R Markdown

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using [R Markdown](http://rmarkdown.rstudio.com).

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2022-02-25 draft_paper.Rmd
add analyses and example rmd uses

---
title: "How do changes in ex vary between countries?"
author: "Jon Minton"
date: "22/02/2022"
output:
  word_document: default
  html_document: default
---

# Introduction

This notebook will compare annual changes in life expectancy at both birth ($e_0$) and at age 65 ($e_{65}$) in the UK and UK nations, with other select European nations, including a 'Synthetic Germany' created from a weighted average of East and West German data for common years. The method for creating this 'Synthetic Germany' is also described in this and linked documents.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Methods

## Data

Data were extracted from the [Human Mortality Database](https://mortality.org/) via the R package [`HMDHFDplus`](http://www.demogr.mpg.de/en/projects_publications/publications_1904/mpidr_technical_reports/reading_human_fertility_database_and_human_mortality_database_data_into_r_5438.htm).

The following countries are included in the comparison:

- England & Wales (`GBRTENW`)
- Scotland (`GBR_SCO`)
- UK as a whole (`GBR_UK`)
- France (`FRANTP`)
- Spain (`ESP`)
- Italy (`ITA`)
- Germany
   - Total Germany (`DEUTNP`)
   - East Germany (`DEUTE`)
   - West Germany (`DEUTW`)
   - Simulated/Synthetic Germany

```{r, echo=FALSE}
# load packages

pacman::p_load(here, tidyverse)

# load data

hmd_lt <- read_rds(here("data", "lifetables.rds"))

```

## Analysis

For each of these countries, and for males and females separately, we are interested in the annual changes in life expectancy at birth ($e_0$) and life expectancy at age 65 ($e_{65}$), from 1980 to the last available year for each country.

```{r, echo = FALSE}

countries_of_interest <- c(
  "GBRTENW",
  "GBR_SCO",
  "GBR_UK",
  "FRATNP",
  "ESP",
  "ITA",
  "DEUTNP",
  "DEUTE", 
  "DEUTW"
)

hmd_ex_selected_countries <- 
  hmd_lt %>% 
    filter(code %in% countries_of_interest) %>% 
    filter(Age %in% c(0, 65)) %>% 
    select(code, Year, Age, sex, ex) %>% 
    rename(x = Age, year = Year)

hmd_ex_selected_countries

```

# Results

## Descriptive Results /Starter Eyeballing

### Life expectancy trends

The following shows life expectancy at birth and at age 65 for selected nations:

Here it is at birth...

```{r, echo = FALSE, fig.cap = "Life Expectancy at birth"}

hmd_ex_selected_countries %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 0) %>% 
  ggplot(aes(x = year, y = ex, group = sex, colour = sex, linetype = sex)) + 
  geom_line() +
  facet_wrap(~code) +
  labs(
    x = "Year",
    y = "Life Expectancy at birth",
    title = "Life expectancies at birth for selected nations",
    subtitle = "1980 to 2020 or latest available year",
    caption = "Source: Human Mortality Database"
  )

```

Here it is at age 65

```{r, echo = FALSE, fig.cap = "Life Expectancy at age 65"}
hmd_ex_selected_countries %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 65) %>% 
  ggplot(aes(x = year, y = ex, group = sex, colour = sex, linetype = sex)) + 
  geom_line() +
  facet_wrap(~code) +
  labs(
    x = "Year",
    y = "Life Expectancy at age 65",
    title = "Life expectancies at age 65 for selected nations",
    subtitle = "1980 to 2020 or latest available year",
    caption = "Source: Human Mortality Database"
  )

```

Some insightful comment about what's going on.

Scotland looks pretty bad compared with the other countries.

## Change in life expectancy in these nations

```{r}
change_in_ex_selected_countries <- 
  hmd_ex_selected_countries %>% 
    group_by(code, x, sex) %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
    ungroup()

change_in_ex_selected_countries
```

```{r}
change_in_ex_selected_countries %>% 
  filter(between(year, 1980, 2020)) %>% 
  ggplot(aes(x = year, y = delta_ex)) + 
  geom_point(alpha = 0.2) + 
  stat_smooth() + 
  facet_grid(code~sex) + 
  geom_hline(yintercept = 0)

```

# Discussion

# Bibliography

### 2022-02-25 presentation_example.Rmd
add analyses and example rmd uses

--- 2022-02-25 draft_paper.Rmd
@@
 ---
-title: "How do changes in ex vary between countries?"
+title: "results_so_far"
 author: "Jon Minton"
-date: "22/02/2022"
-output:
-  word_document: default
-  html_document: default
+date: "24/02/2022"
+output: ioslides_presentation
 ---
-
-# Introduction
-
-This notebook will compare annual changes in life expectancy at both birth ($e_0$) and at age 65 ($e_{65}$) in the UK and UK nations, with other select European nations, including a 'Synthetic Germany' created from a weighted average of East and West German data for common years.
-The method for creating this 'Synthetic Germany' is also described in this and linked documents.
 
@@
 
-# Methods
-
 ## Data
-
-Data were extracted from the [Human Mortality Database](https://mortality.org/) via the R package [`HMDHFDplus`](http://www.demogr.mpg.de/en/projects_publications/publications_1904/mpidr_technical_reports/reading_human_fertility_database_and_human_mortality_database_data_into_r_5438.htm).
-
-The following countries are included in the comparison:
 
@@
 
-# Results
-
-## Descriptive Results /Starter Eyeballing
-
-### Life expectancy trends
-
-The following shows life expectancy at birth and at age 65 for selected nations:
-
-Here it is at birth...
+## Life expectancy at birth
 
@@
 
-Here it is at age 65
+## life expectancy at age 65
 
@@
 ```
-
-Some insightful comment about what's going on.
-
-Scotland looks pretty bad compared with the other countries.
-
-## Change in life expectancy in these nations
-
-```{r}
-change_in_ex_selected_countries <- 
-  hmd_ex_selected_countries %>% 
-    group_by(code, x, sex) %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex)) %>% 
-    ungroup()
-
-change_in_ex_selected_countries
-```
-
-```{r}
-change_in_ex_selected_countries %>% 
-  filter(between(year, 1980, 2020)) %>% 
-  ggplot(aes(x = year, y = delta_ex)) + 
-  geom_point(alpha = 0.2) + 
-  stat_smooth() + 
-  facet_grid(code~sex) + 
-  geom_hline(yintercept = 0)
-
-```
-
-# Discussion
-
-# Bibliography

### 2022-02-25 R/get_data_from_hmd.R

@@ -26,7 +26,8 @@ if (!file.exists(here("data", "e0.rds"))){
     ) %>% 
     mutate(data = map(code, ~readHMDweb(.x, "E0per", my_username, my_password))) %>% 
     unnest(data) %>% 
-    pivot_longer(c("Female", "Male", "Total"), names_to = "sex", values_to = "e0") %>% 
+    pivot_longer(c("Female", "Male", "Total"), names_to = "sex", values_to = "e0") %>%
+    rename(year = Year) %>% 
     mutate(sex = tolower(sex))
   hmd_e0
   

### 2022-02-25 R/make_synthetic_germany.R

@@ -1,12 +1,29 @@
+#  Code for using numerical optimisation for producing a 'Synthetic Germany' which has similar life expectancy 
+# at birth and at age 65 to Germany as a whole (for estimating life expetancies in Germany as a whole prior to 
+# reunification )
 
-pacman::p_load(HMDHFDplus, tidyverse, here)
-# Get and calculate changes in e0
+# An application of the optim function! 
 
-source(here("R", "get_data_from_hmd.R"))
+# Author: Jon Minton
+# Date: 23 Feb 2022
 
-rm(hmd_mx)
+# First we need the packages
+pacman::p_load(tidyverse, here)
+# If getting from HMD directly we can use a script 
+# get_data_from_hmd.R
+# as follows
+# pacman::p_load(HMDHFDplus)
+# source(here("R", "get_data_from_hmd.R"))
 
-# Let's just keep e0 and e65, and for Germany
+# We need life expectancy at age 0 and 65, which we can get from the lifetables 
+
+# We will load these lifetables directly (rather than using the script) 
+
+# To allow this script to be self-contained we will refer to the github file location directly
+
+hmd_lt <- read_rds("https://github.com/JonMinton/change-in-ex/blob/main/data/lifetables.rds?raw=true")
+
+# We will now just look for German data, and for ages 0 and 65 only 
 
 e0_e65 <- 
   hmd_lt  %>% 
@@ -30,92 +47,273 @@ e0_e65 <-
 # We want to create a synthetic Germany for 1980-89 which is a 
 # mixture of West and East German populations 
 
-make_synthetic_population <- function(east_series, west_series, p_east){
-  stopifnot("Series are of different lengths" = length(east_series) == length(west_series))
-  stopifnot("Proportion out of possible bounds" = between(p_east, 0, 1))
-  
-  east_series * p_east + west_series * (1 - p_east)
-}
+# We can load the various functions created for this purpose:
 
-compare_synthetic_to_reference <- function(synthetic, reference, what = c("RMSE", "abs", "rel")){
-  stopifnot("Synthetic and reference are different lengths" = length(synthetic) == length(reference))
-  
-  what <- match.arg(what)
-  
-  if (what == "RMSE"){
-    out <- (synthetic - reference)^2 %>% 
-      mean() %>% 
-      .^(1/2)
-    return(out)
-  } else if (what == "abs"){
-    return(synthetic - reference)
-  } else if (what == "rel"){
-    out <- (synthetic - reference)/reference
[…]

### 2022-03-01 draft_paper.Rmd
add synthetic germany to series and nice labels

@@
 
+### Synthetic Germany
+
+- To maybe be moved to the appendix later
+
+To allow UK national trends to be compared with a single German population, we attempted to produce a 'Synthetic German' population with data for East and West Germany for years prior to reunification.
+We estimated that this 'Synthetic Germany' could be produced by using a weighted average of 20% East Germany, and 80% West German life expectancy trends.
+More precise estimates can be produced, and the methods used to reach this conclusion are detailed in Appendix X.
+
 ```{r, echo = FALSE}
@@
 
+# Code for producing synthetic Germany
+
+source(here("R", "make_synthetic_germany_functions.R"))
+
+series_east_f0 <- hmd_ex_selected_countries %>% 
+  filter(code == "DEUTE") %>% 
+  filter(between(year, 1980, 2020)) %>% 
+  filter(x == 0) %>% 
+  filter(sex == "female") %>% 
+  arrange(year) %>% 
+  pull(ex)
+
+series_west_f0 <- hmd_ex_selected_countries %>% 
+  filter(code == "DEUTW") %>% 
+  filter(between(year, 1980, 2020)) %>% 
+  filter(x == 0) %>% 
+  filter(sex == "female") %>% 
+  arrange(year) %>% 
+  pull(ex)
+
+tmp_df_f0 <- tibble(
+  code = "DEUTSYNTH",
+  year = 1980:2017,
+  x = 0,
+  sex = "female", 
+  ex = make_synthetic_population(series_east_f0, series_west_f0, p_east = 0.2)
+)
+
+###
+series_east_f65 <- hmd_ex_selected_countries %>% 
+  filter(code == "DEUTE") %>% 
+  filter(between(year, 1980, 2020)) %>% 
+  filter(x == 65) %>% 
+  filter(sex == "female") %>% 
+  arrange(year) %>% 
+  pull(ex)
+
+series_west_f65 <- hmd_ex_selected_countries %>% 
+  filter(code == "DEUTW") %>% 
+  filter(between(year, 1980, 2020)) %>% 
+  filter(x == 65) %>% 
+  filter(sex == "female") %>% 
+  arrange(year) %>% 
+  pull(ex)
+
+tmp_df_f65 <- tibble(
+  code = "DEUTSYNTH",
+  year = 1980:2017,
+  x = 65,
+  sex = "female", 
+  ex = make_synthetic_population(series_east_f65, series_west_f65, p_east = 0.2)
+)
+##
+
+series_east_m0 <- hmd_ex_selected_countries %>% 
+  filter(code == "DEUTE") %>% 
+  filter(between(year, 1980, 2020)) %>% 
+  filter(x == 0) %>% 
+  filter(sex == "male") %>% 
+  arrange(year) %>% 
+  pull(ex)
+
+series_west_m0 <- hmd_ex_selected_countries %>% 
+  filter(code == "DEUTW") %>% 
+  filter(between(year, 1980, 2020)) %>% 
+  filter(x == 0) %>% 
+  filter(sex == "male") %>% 
+  arrange(year) %>% 
+  pull(ex)
+
+tmp_df_m0 <- tibble(
+  code = "DEUTSYNTH",
+  year = 1980:2017,
+  x = 0,
+  sex = "male", 
+  ex = make_synthetic_population(series_east_m0, series_west_m0, p_east = 0.2)
+)
+
+###
+
+series_east_m65 <- hmd_ex_selected_countries %>% 
+  filter(code == "DEUTE") %>% 
+  filter(between(year, 1980, 2020)) %>% 
+  filter(x == 65) %>% 
+  filter(sex == "male") %>% 
+  arrange(year) %>% 
+  pull(ex)
+
+series_west_m65 <- hmd_ex_selected_countries %>% 
+  filter(code == "DEUTW") %>% 
+  filter(between(year, 1980, 2020)) %>% 
+  filter(x == 65) %>% 
+  filter(sex == "male") %>% 
+  arrange(year) %>% 
+  pull(ex)
+
+tmp_df_m65 <- tibble(
+  code = "DEUTSYNTH",
+  year = 1980:2017,
+  x = 65,
+  sex = "male", 
+  ex = make_synthetic_population(series_east_m65, series_west_m65, p_east = 0.2)
+)
+
+hmd_ex_selected_countries_with_synth <- bind_rows(
+  hmd_ex_selected_countries,
+  tmp_df_m0,
+  tmp_df_m65,
+  tmp_df_f0,
+  tmp_df_f65
+)
+
+# Labels for codes
+
+country_code_lookup <- 
+  tribble(
+    ~code, ~country,
+    "DEUTNP", "Germany",
+    "DEUTE", "East Germany",
+    "DEUTW", "West Germany",
+    "ESP", "Spain",
+    "FRATNP", "France", 
+    "ITA", "Italy",
+    "GBRTENW", "England & Wales",
+    "GBR_SCO", "Scotland",
+    "DEUTSYNTH", "Synthetic Germany"
+  )
+
 ```
@@
 
-hmd_ex_selected_countries %>% 
+hmd_ex_selected_countries_with_synth %>% 
+  left_join(country_code_lookup) %>% 
+  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy"))) %>% 
   filter(between(year, 1980, 2020)) %>% 
@@
   geom_line() +
-  facet_wrap(~code) +
+  facet_wrap(~country) +
   labs(
@@
 ```{r, echo = FALSE, fig.cap = "Life Expectancy at age 65"}
-hmd_ex_selected_countries %>% 
+hmd_ex_selected_countries_with_synth %>% 
+  left_join(country_code_lookup) %>% 
+  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy"))) %>% 
   filter(between(year, 1980, 2020)) %>% 
@@
 change_in_ex_selected_countries <- 
-  hmd_ex_selected_countries %>% 
+  hmd_ex_selected_countries_with_synth %>% 
     group_by(code, x, sex) %>% 
@@
 change_in_ex_selected_countries %>% 
+  filter(x == 0) %>% 
+  left_join(country_code_lookup) %>% 
+  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy"))) %>% 
   filter(between(year, 1980, 2020)) %>% 
@@
   stat_smooth() + 
-  facet_grid(code~sex) + 
-  geom_hline(yintercept = 0)
+  facet_grid(country~sex) + 
+  geom_hline(yintercept = 0) + 
+  theme(strip.text.y = element_text(angle = 0)) + 
+  labs(
+    x = "Year",
+    y = "Year-on-year change in life expectancy",
+    title = "Annual change in life expectancy at birth, selected countries",
+    subtitle = "Blue line: nonlinear smoother",
+    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
+  )
 

### 2022-03-01 R/make_synthetic_germany.R

@@ -350,4 +350,3 @@ synth_germany_best_shares_via_optim %>%
 # when no unified Germany existed), and how to move towards more automated and standardised ways of 
 # solving such problems in more general cases through functions and functional programming.
 
-

### 2022-03-01 draft_paper.Rmd
looked at changes in ex by decade

@@
 output:
+  html_document: default
   word_document: default
-  html_document: default
 ---
@@
 ```
+### Changes in life expectancy at birth
 
-```{r}
+```{r, echo = FALSE, warning=FALSE, message = FALSE, fig.height=8, fig.width=8}
 change_in_ex_selected_countries %>% 
@@
 
+### Changes in life expectancy at age 65
+
+```{r, echo = FALSE, warning=FALSE, message = FALSE, fig.height=8, fig.width=8}
+change_in_ex_selected_countries %>% 
+  filter(x == 65) %>% 
+  left_join(country_code_lookup) %>% 
+  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy"))) %>% 
+  filter(between(year, 1980, 2020)) %>% 
+  ggplot(aes(x = year, y = delta_ex, colour=sex)) + 
+  geom_point(alpha = 0.2) + 
+  stat_smooth() + 
+  facet_grid(country~sex) + 
+  geom_hline(yintercept = 0) + 
+  theme(strip.text.y = element_text(angle = 0)) + 
+  labs(
+    x = "Year",
+    y = "Year-on-year change in life expectancy",
+    title = "Annual change in life expectancy at age 65, selected countries",
+    subtitle = "Blue line: nonlinear smoother",
+    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
+  )
+
+```
+
 # Discussion
 
+# Time trends
+
+```{r}
+
+hmd_ex_selected_countries_with_synth %>% 
+    filter(year >= 1979) %>% 
+    group_by(code, x, sex) %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+    nest() %>% 
+    mutate(
+      rwd_model = map(data, ~lm(delta_ex ~ 1, data = . )),
+      ar_model  = map(data, ~lm(delta_ex ~ lag(delta_ex), data = .))
+    ) %>% 
+    mutate(
+      aic_rwd = map_dbl(rwd_model, AIC),
+      aic_ar  = map_dbl(ar_model, AIC)
+    ) %>% 
+    mutate(
+      diff_aic = aic_ar - aic_rwd
+    )  %>% 
+  # filter(diff_aic > 0)
+  mutate(
+    tidied_ar_model = map(ar_model, broom::tidy)
+  ) %>% 
+  unnest(tidied_ar_model) %>% 
+  filter(term == "(Intercept)") %>% 
+  arrange(desc(estimate)) %>% 
+  select(code, x, sex, estimate, std.error) %>% 
+  mutate(display_label = glue::glue("{code}_{x}_{sex}")) %>% 
+  ggplot(aes(estimate, fct_reorder(display_label, estimate))) + 
+  geom_point()
+
+```
+
+Look at improvements by decade
+
+```{r}
+
+hmd_ex_selected_countries_with_synth %>% 
+    filter(year >= 1979) %>% 
+    group_by(code, x, sex) %>% 
+    arrange(year) %>% 
+    mutate(decade = year - (year %% 10)) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>%
+    nest() %>%
+    mutate(
+      rwd_model = map(data, ~lm(delta_ex ~ factor(decade), data = . ))
+     )  %>%
+  mutate(
+    tidied_rwd_model = map(rwd_model, broom::tidy)
+  ) %>%
+  unnest(tidied_rwd_model)
+```
+
 # Bibliography

### 2022-03-01 draft_paper.Rmd
added line to analysis
Additional_text
Added some additional text on data, methods and analyses inc. time-trend

@@
 For each of these countries, and for males and females separately, we are interested in the annual changes in life expectancy at birth ($e_0$) and life expectancy at age 65 ($e_{65}$), from 1980 to the last available year for each country.
+In addition, we compare average life expectancy improvements over the 4 decades: 1980, 1990, 2000, and 2010.
 
@@
 
-The following shows life expectancy at birth and at age 65 for selected nations:
+The following figures show life expectancy at birth and at age 65 for selected nations from 1980 to the latest available year.
 
@@
 ## Change in life expectancy in these nations 
+Next, we compare the annual change in life expectancy at birth and at aged 65 years for the 6 nations from 1980 to the latest available year, by sex.
 
@@
 # Time trends 
+Finally, we used time-trend analysis to explore the changes in life expectancy over the 4 decades: 1980s, 1990s, 2000s, and 2010s.
+This produces the average change for the decade, whether negative or positive, the standard error and its statistical significance for each country by sex.
 
@@
 
+# Discussion
+Limitations: fewer years for comparison in 2010s; comparing different time periods will provide different results.
+
 # Bibliography

### 2022-03-02 draft_paper.Rmd
add flextable to library and main function to table at bottom. Other refinements still to do.

--- 2022-03-01 draft_paper.Rmd
@@
 
-pacman::p_load(here, tidyverse)
+pacman::p_load(here, tidyverse, flextable) # using flextable instead of gt. It seems to be better 
+# developed, more commonly used, and has more clearly labelled function names.
 
@@
     arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>%
     mutate(decade = year - (year %% 10)) %>% 
-    mutate(delta_ex = ex - lag(ex)) %>%
-    nest() %>%
-    mutate(
-      rwd_model = map(data, ~lm(delta_ex ~ factor(decade), data = . ))
+    filter(!is.na(delta_ex)) %>% 
+  group_by(code, x, sex, decade) %>% 
+  summarise(
+    mean_delta_ex = mean(delta_ex),
+    sd_delta_ex   = sd(delta_ex)
     ) %>% 
-  mutate(
-    tidied_rwd_model = map(rwd_model, broom::tidy)
+  ungroup() %>% 
+  mutate(cell_contents = glue::glue("{round(mean_delta_ex, 3)} ({round(sd_delta_ex, 3)})")
   ) %>% 
-  unnest(tidied_rwd_model)
+  filter(code == "GBRTENW") %>% 
+  filter(x == 0) %>% 
+  select(sex, decade, cell_contents) %>% 
+  spread(decade, cell_contents) %>% 
+  flextable()
+
 ```

### 2022-03-02 draft_paper.Rmd
add placeholder for appendix
Update draft_paper.Rmd

@@
 The method for creating this 'Synthetic Germany' is also described in this and linked documents.
+Adding a line to try contribute
 
@@
 
+# Discussion
+
+- **Limitations**: fewer years for comparison in 2010s; comparing different time periods will provide different results.
+- Add description of what this needs
+
+# Appendix A: Construction of Synthetic Germany
+
+- Add further details here.
+
 # Bibliography

### 2022-03-02 draft_paper.Rmd

@@
 
-This notebook will compare annual changes in life expectancy at both birth ($e_0$) and at age 65 ($e_{65}$) in the UK and UK nations, with other select European nations, including a 'Synthetic Germany' created from a weighted average of East and West German data for common years.
-The method for creating this 'Synthetic Germany' is also described in this and linked documents.
-Adding a line to try contribute
+This notebook will compare annual changes in life expectancy at both birth ($e_0$) and at age 65 ($e_{65}$) from 1980 to 2019 (or the latest available year) in England and Wales, Scotland, France, Italy, Spains and Germany.
+Germany includes a 'Synthetic Germany' created from a weighted average of East and West German data for common years.
+The method for creating this 'Synthetic Germany' is covered in the methods section in brief and in Appendix X for more detail.
 
@@
    - Simulated/Synthetic Germany
+
+In order to explore the hypothesis that the recent stallin in life expectancy, most marked in England and Wales, was an expected 'correction' after a decade of unusually fast life expectancy improvements, we have selected the years from 1980 onwards.
 

### 2022-03-03 draft_paper.Rmd
fail to make decent progress with tables; do lots of linear regression models on decadal differences

@@
 
+# To dos
+
+- [ ] Create flextable by decade
+
 # Introduction
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE)
+knitr::opts_chunk$set(
+  echo = FALSE,
+  tab.id = NULL,
+  tab.cap = NULL,
+  tab.topcaption = TRUE,
+  tab.lp = "tab:"
+
+  )
 ```
@@
 
-pacman::p_load(here, tidyverse, flextable) # using flextable instead of gt. It seems to be better 
+pacman::p_load(knitr, here, tidyverse, flextable) # using flextable instead of gt. It seems to be better 
 # developed, more commonly used, and has more clearly labelled function names.
@@
 
-Some insightful comment about what's going on.
+- [ ] Some insightful comment about what's going on.
 
@@
 ## Change in life expectancy in these nations
+
 Next, we compare the annual change in life expectancy at birth and at aged 65 years for the 6 nations from 1980 to the latest available year, by sex.
@@
 ```
+
 ### Changes in life expectancy at birth
+
+- [ ] make smoother lines coloured by sex
+- [ ] use consistent colours for all graphs showing separate sex results
 
@@
 
+- [ ] Use slightly stronger (and consistent) colours for sex smoother lines
+
 ```{r, echo = FALSE, warning=FALSE, message = FALSE, fig.height=8, fig.width=8}
@@
 
-# Discussion
+## Time trends
 
-# Time trends 
 Finally, we used time-trend analysis to explore the changes in life expectancy over the 4 decades: 1980s, 1990s, 2000s, and 2010s.
@@
 
-```{r}
+- [X] Give this a title and further info (lower priority)
+- [X] Look at splitting by starting age 
+    - [ ] Look at further refining starting age split
+
+```{r, echo = FALSE, warning = FALSE, message = FALSE}
 
@@
     filter(year >= 1979) %>% 
+    filter(!(code %in% c("DEUTE", "DEUTW", "DEUTNP"))) %>%  # Using only synthetic germany for longer time period
     group_by(code, x, sex) %>% 
@@
   arrange(desc(estimate)) %>%
-  select(code, x, sex, estimate, std.error) %>% 
-  mutate(display_label = glue::glue("{code}_{x}_{sex}")) %>% 
+  left_join(country_code_lookup) %>% 
+  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy"))) %>% 
+  select(code, country, x, sex, estimate, std.error) %>% 
+  mutate(display_label = glue::glue("{country}_{x}_{sex}")) %>% 
   ggplot(aes(estimate, fct_reorder(display_label, estimate))) + 
-  geom_point()
+  geom_point() +
+  expand_limits( x = 0) + 
+  geom_vline(xintercept = 0) + 
+  labs(
+    x = "Average annual life expectancy change",
+    y = "Country-sex combination",
+    title = "Average annual improvements by country and sex",
+    subtitle = "Range: 1980 to 2020 or nearest available years",
+    caption = "Source: Human Mortality Database"
+  ) + 
+  facet_wrap(~x)
 
@@
 
-Look at improvements by decade
+### Average improvements by decade
 
-```{r}
+- [ ] Add further introduction 
+- [ ] Add statistical significance (though maybe this is a different analysis)
+
+```{r, tab.cap = "Source: HMD"}
 
@@
   spread(decade, cell_contents) %>% 
-  flextable()
+#  write.csv("clipboard")
+  flextable() %>%
+  # Want to make the header row bold
+  vline(j = 1, part = "all") %>%
+  set_caption(caption = "Average improvement in life expectancy by decade")
+
+```
+
+Are the differences by decade statistically significant?
+
+```{r}
+
+dta_for_decade_comparison <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 0) %>% 
+      filter(sex == "female") %>% 
+      filter(code == "GBRTENW") %>% 
+      filter(year >= 1979) %>% 
+      # ggplot(aes(x = year, y = ex)) +
+      #   geom_line()
+      arrange(year) %>% 
+      mutate(delta_ex = ex - lag(ex)) %>% 
+    #   ggplot(aes(x = year, y = delta_ex)) + 
+    # geom_point() + geom_line() +
+    # geom_hline(yintercept = 0)
+      mutate(
+        decade = year - (year %% 10)
+      ) %>% 
+  filter(!is.na(delta_ex))
+
+model_null <- 
+  lm(delta_ex ~ 1, data = dta_for_decade_comparison)
+
+model_alt <- 
+  lm(delta_ex ~ factor(decade), data= dta_for_decade_comparison)
+
+summary(model_null)
+
+summary(model_alt)
+
+anova(model_null, model_alt)
+
+# to get the value of interest do tmp[2,6]
+
+# Generalising
+
+tidy_dataset <- function(df){
+  df %>% 
+      filter(year >= 1979) %>% 
+      arrange(year) %>% 
+      mutate(delta_ex = ex - lag(ex)) %>% 
+      mutate(
+        decade = year - (year %% 10)
+      ) %>% 
+  filter(!is.na(delta_ex))
+}
+
+make_null_model <- function(df){
+  lm(delta_ex ~ 1, data = df)
+}
+
+all_null_alt_models <-
+  hmd_ex_selected_countries_with_synth %>%
+    group_by(x, sex, code) %>% 
+    nest() %>% 
+    mutate(
+      prep_data = map(data, tidy_dataset) 
+    ) %>% 
+    mutate(
+      model_null = map(prep_data, ~lm(delta_ex ~ 1, data = .)),
+      model_alt  = map(prep_data, ~lm(delta_ex ~ factor(decade), data = .))
+  #    model_null = map(prep_data, make_null_model)
+    ) %>% 
+    mutate(
+      anova_results = map2(model_null, model_alt, anova)
+    ) %>% 
+    mutate(
+      anova_pval = map_dbl(anova_results, ~.[2,6])
+    )
+
+all_null_alt_models %>% 
+  filter(anova_pval < 0.05) -> tmp
+
+summary(tmp$model_alt[[1]])
+
+summary(tmp$model_alt[[2]])
+
+  # build null model
+  # build alt model
+  # do anova
+  # get p value from anova
 

### 2022-03-03 draft_paper.Rmd
Lu testing making edits in R stuio (line 14)
additional text please could you add more to line 467 on the difference between decades? I wonder what is easiest/most useful to include?

@@
 
+Stalling of life expectancy improvements occurred in the UK from 2010 onwards.
+It was also seen in some other Western European countries.
+While the reason and severity of this is debated, some also contend the stalling of improvements seen is artefact following an unusual decade of improvement in 2000s (Murphy 2021).
+
+-[ ] Introduction to be written once methods/results discussed and decided on
+
 This notebook will compare annual changes in life expectancy at both birth ($e_0$) and at age 65 ($e_{65}$) from 1980 to 2019 (or the latest available year) in England and Wales, Scotland, France, Italy, Spains and Germany.
@@
 The method for creating this 'Synthetic Germany' is covered in the methods section in brief and in Appendix X for more detail.
+In order to explore the hypothesis that the recent stalling in life expectancy, most marked in England and Wales, was an expected 'correction' after a decade of unusually fast life expectancy improvements, we have selected the years from 1980 onwards.
 
@@
 - Scotland (`GBR_SCO`)
-- UK as a whole (`GBR_UK`)
+- UK as a whole (`GBR_UK`) #we gaven't included UK but can do
 - France (`FRANTP`)
@@
    - Simulated/Synthetic Germany
-
-In order to explore the hypothesis that the recent stallin in life expectancy, most marked in England and Wales, was an expected 'correction' after a decade of unusually fast life expectancy improvements, we have selected the years from 1980 onwards.
 
@@
 In addition, we compare average life expectancy improvements over the 4 decades: 1980, 1990, 2000, and 2010.
+
+We then compared whether the average between the decades was significant, using a regression model.
 
@@
 
-## Descriptive Results /Starter Eyeballing
+## Descriptive Results
 
@@
 
-The following figures show life expectancy at birth and at age 65 for selected nations from 1980 to the latest available year.
-
-Here it is at birth...
+The following figures show life expectancy at birth and at age 65 for selected nations from 1980 to the latest available year for males and females.
 
@@
 ```
-
-Here it is at age 65
 
@@
 
-Scotland looks pretty bad compared with the other countries.
-
 ## Change in life expectancy in these nations
 
-Next, we compare the annual change in life expectancy at birth and at aged 65 years for the 6 nations from 1980 to the latest available year, by sex.
+Next, we compare the annual change in life expectancy at birth and at aged 65 years for the nations from 1980 to the latest available year, by sex.
 
@@
 - [ ] make smoother lines coloured by sex
-- [ ] use consistent colours for all graphs showing separate sex results
+- [X] use consistent colours for all graphs showing separate sex results
 
@@
   filter(between(year, 1980, 2020)) %>% 
-  ggplot(aes(x = year, y = delta_ex)) + 
+  ggplot(aes(x = year, y = delta_ex, colour=sex)) + 
   geom_point(alpha = 0.2) + 
@@
 
-Are the differences by decade statistically significant?
+####Are the differences by decade statistically significant? 
+To assess whether the changes by decade are statistically significant, we carried out [Jon please could you write a little more here?]
+We found the only significant change decade-on-decades was for East Germany for females at birth and at aged 65 years...
 

### 2022-03-03 draft_paper.Rmd
typo corrected

@@
 
--[ ] Introduction to be written once methods/results discussed and decided on
+- [ ] Introduction to be written once methods/results discussed and decided on
 
@@
 - Scotland (`GBR_SCO`)
-- UK as a whole (`GBR_UK`) #we gaven't included UK but can do
+- UK as a whole (`GBR_UK`) #we haven't included UK but can do
 - France (`FRANTP`)
@@
 
-####Are the differences by decade statistically significant? 
+###Are the differences by decade statistically significant? 
 To assess whether the changes by decade are statistically significant, we carried out [Jon please could you write a little more here?]

### 2022-03-04 draft_paper.Rmd
Added a description of statistical approach used, though probably too technical (We could just say something like "we did F tests comparing intercept-only against decadal fixed effects models. Further explanation and code in appendix X"?!)

@@
 In order to explore the hypothesis that the recent stalling in life expectancy, most marked in England and Wales, was an expected 'correction' after a decade of unusually fast life expectancy improvements, we have selected the years from 1980 onwards.
+
+This is a new test using R Studio on Lu's desktop in line 22
 
@@
 
-###Are the differences by decade statistically significant? 
-To assess whether the changes by decade are statistically significant, we carried out [Jon please could you write a little more here?]
+### Are the differences by decade statistically significant?
+
+Are there significant differences in average rates of change in $e_x$ by decade?
+Another way of expressing this question is to ask whether knowing the decade in which an observation has been observed is informative as to what such values will be.
+This rephrased question was addressed by, for each starting age (0 or 65 years), sex and country, fitting two competing model specifications - $M_0$ and $M_A$ - and running an F test on the residuals of these two models.
+The Null model specification, $M_0$, regresses $\delta e_x$ (observed annual changes in $e_x$) against a single intercept term $\alpha$; put another way: $M_0$ assumes that all observations are drawn from a Normal distribution with a mean value of $\alpha$, and that this is consistent across all observed decades.
+The alternative model specification, $M_A$, also includes the $\alpha$ term, but also includes dummy variables to indicate which decade the realised value of $\delta e_0$ corresponds to.
+(Because one of the decades needs to be the reference category against which other decades are compared, there are one fewer decadal dummy variables than decades in the dataset for a given country, and the $\alpha$ parameter in $M_A$ estimates the mean change in $\delta e_0$ in the first decade, and so cannot be compared directly with the $\alpha$ parameter in $M_0$.) A P-value from the F test comparing $M_0$ and $M_A$ of less than 0.05 is taken to indicate that $M_A$ should be preferred to $M_0$, and so that there are systemic differences in average rates of improvement by decade for a given country, sex, and starting age.
+
 We found the only significant change decade-on-decades was for East Germany for females at birth and at aged 65 years...

### 2022-03-04 draft_paper.Rmd
small edits and added echo=FALSE where missing

@@
 It was also seen in some other Western European countries.
-While the reason and severity of this is debated, some also contend the stalling of improvements seen is artefact following an unusual decade of improvement in 2000s (Murphy 2021).
+While the reason and severity of this is debated, some also contend the stalling of improvements seen is artefact following an unusual decade of improvement in 2000s (e.g.
+Murphy 2021).
 
@@
 In order to explore the hypothesis that the recent stalling in life expectancy, most marked in England and Wales, was an expected 'correction' after a decade of unusually fast life expectancy improvements, we have selected the years from 1980 onwards.
-
-This is a new test using R Studio on Lu's desktop in line 22
 
@@
 
-```{r}
+```{r, echo=FALSE}
 change_in_ex_selected_countries <- 
@@
 
-```{r, tab.cap = "Source: HMD"}
+The table below shows average improvements by decade for England and Wales (standard error).
+
+```{r, echo=FALSE, tab.cap = "Source: HMD"}
 
@@
 
-We found the only significant change decade-on-decades was for East Germany for females at birth and at aged 65 years...
+We found the only significant change decade-on-decades was for East Germany for females at birth and at aged 65 years.
+This is consistent with existing literature in this area, showing the convergence of life expectancy for females in East and West Germany following reunification.
 
@@
 
-- **Limitations**: fewer years for comparison in 2010s; comparing different time periods will provide different results.
-- Add description of what this needs
+- **Limitations**: fewer years for comparison in 2010s for some nations; comparing different time periods will provide different results e.g. New Labour years in E&W.
 

### 2022-03-04 draft_paper.Rmd

@@
 title: "How do changes in ex vary between countries?"
-author: "Jon Minton"
+author: "Jon Minton, Lucinda Hiam"
 date: "22/02/2022"
 output:
+  word_document: default
   html_document: default
-  word_document: default
 ---

### 2022-03-07 draft_paper.Rmd
added additional desciptive comments on modelling strategy so far

--- 2022-03-04 draft_paper.Rmd
@@
 title: "How do changes in ex vary between countries?"
-author: "Jon Minton"
+author: "Jon Minton, Lucinda Hiam"
 date: "22/02/2022"
@@
 
-```{r}
+```{r, cache=TRUE}
 
@@
 
+### Joinpoint/Segmented regression approach
+
+Though there may be systemic differences in average changes in life expectancy over time, such differences may not be patterned by decade, and so the above model specification may not have been appropriate to detect any such changes.
+Instead of looking for systemic differences by decade, we can instead look at whether there are any systemic and detectable 'breaks' in the trends over time.
+If our data were life expectancy over time, this could be achieved through joinpoint analysis, in which the slope is allowed to change over time.
+As instead are data are annual changes in life expectancy, we are interested in detecting changes in intercept over time.
+If we define our Null (no change) model as follows:
+
+$$
+M_0 : y(t) = \alpha + e
+$$
+Where $e \backsim N(0, \sigma^2)$. i.e. all observations $y$ are draws from $N(\alpha, \sigma^2)$. The alternative model specification against which this is compared is then:
+
+$$
+M_A : y(t) = \alpha_0 + \beta_1 T(t) + e 
+$$
+Where $T(t)$ is defined as $0$ where $t \lt \tau$ and $1$ where $t \geq \tau$. ($\alpha_0$ is used in $M_A$ to indicate that this parameter is not directly comparable with $\alpha$ in $M_0$).
+
+This specification shows that there is an additional parameter $\tau$ to be determined in order to produce $M_A$.
+This $\tau$ parameter is the breakpoint in the series.
+The selection of $\tau$ will be determined by numeric optimisation, by selecting the value of $\tau$ that minimised AIC.
+
+Once the best possible $M_A$, using the best value of $\tau$, is selected, $M_A$ will be compared with $M_0$ using an F-test, allowing us to determine whether there is convincing evidence of any specific break in the data.
+
+We will start with a single country, starting age, and sex, and with manual selection of $\tau$, then generalise and automate further.
+
+```{r}
+
+# Let's start with a single example population. (There are many different population subgroups, hence the 
+# benefit of being able to automate this later.) We're going to start with England & Wales, females, life 
+# expectancy at birth (x == 0)
+
+example_df <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 0) %>% 
+      filter(sex == "female") %>% 
+      filter(code == "GBRTENW") %>% 
+      filter(year >= 1979)
+
+# Now we create the change in ex variable as that's the response variable in the model
+example_df <- 
+  example_df %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) # %>% 
+  # ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+  # geom_line() +
+  # geom_hline(yintercept = 0) + 
+  # stat_smooth()
+
+# The null model is really simple: just an intercept 
+model_null <- lm(delta_ex ~ 1, data = example_df)
+summary(model_null)
+#
+
+# The code below shows how we can be even simpler, and force the regression model not to 
+# even have an intercept (i.e. have zero predictor variables)
+
+# model_nuller <- lm(delta_ex ~ 0, data = example_df)
+# summary(model_nuller)
+
+# And we can compare the intercept with the no intercept model as follows 
+# anova(model_nuller, model_null)
+# a statistically significant p value shows the intercept is definitely worth including in the 
+# model. Substantively that it's wrong to assume no improvement on average in ex over time since the 1980s
+
+# There are lots of possible alternative models, some examples of which are below:
+model_alt01 <- lm(delta_ex ~ year >= 1990, data = example_df)
+summary(model_alt01)
+
+model_alt02 <- lm(delta_ex ~ year >= 1995, data = example_df)
+summary(model_alt02)
+
+model_alt03 <- lm(delta_ex ~ year >= 1997, data = example_df)
+summary(model_alt03)
+
+model_alt04 <- lm(delta_ex ~ year >= 2008, data = example_df)
+summary(model_alt04)
+
+model_alt05 <- lm(delta_ex ~ year >= 2011, data = example_df)
+summary(model_alt05)
+
+# each of these models allows an adjustment to the horizontal line after a given year has been reached. 
+# In the examples above some possible given years are 1990, 1995, 1997, 2008 and 2011. 
+# Each of these is a candidate value for tau. But we've not looked at all possible/plausible values of this.
+
+# Hence, it would be good to automate further to search for each possible tau and see how the fit and model 
+# results change.
+
+# We'll start to automate by building a function which creates T_param given tau, and runs and returns the model
+# with this type of T_param
+# (If tau is 1990, then T_param is equivalent to model_alt01 above; if tau were 1995, then it's equivalent to model_alt02, and so on.
+The only difference is that the new coefficient created has the same name, T_param,
+# whereas in the above it would have a different name each time, because the coefficient is being 
+# created in situ, within the formula specification, rather than by pre-processing the dataframe)
+run_alt_with_given_tau <- function(tau, df){
+  df <- 
+    df %>% 
+      filter(!is.na(delta_ex)) %>% 
+      mutate(
+        T_param = ifelse(year < tau, FALSE, TRUE)
+      )
+
+  lm(delta_ex ~ T_param, data  = df)
+}
+
+# We can now run the above function, which creates a model, for a wide range of possible tau values
+# This is the sequence listed and named tau_candidate, which is every whole number from 
+# 1982 to 2015
+
+# For each of these models, we can then get a measure of model fit. We're going to use AIC, though 
+# could have used other measures. With AIC (and with BIC, and RMSE etc) better fit is indicated with 
+# a lower value
+grid_search_df <- tibble(
+  tau_candidate = 1982:2015 # column of candidate values
+) %>% 
+  mutate(
+    model_alt = map(tau_candidate, run_alt_with_given_tau, df = example_df) # run the model with the value of tau in tau_candidate
+  ) %>% 
+  mutate(
+    aic = map_dbl(model_alt, AIC)
+  )
+
+# Now we can see which proposed tau results in what fit value (AIC), initially just by graphing it.
+grid_search_df %>% 
+  ggplot(aes(tau_candidate, aic)) + 
+  geom_point()
+
+# So, in the above we found 2012 as tau minimised AIC. So, that's the best possible alternative model. 
+# Let's run and label it as such
+best_alt_model <- example_df %>% 
+  mutate( T_param = ifelse(year < 2012, FALSE, TRUE)) %>% 
+  lm(delta_ex ~ T_param, data = .)
+
+# Now we want to create the null model to compare it against (which just has an intercept term alone)
+null_model <- example_df %>% 
+  lm(delta_ex ~ 1, data = .)
+
+# And finally, we want to compare the best_alt_model against the null model using an F test
+anova(null_model, best_alt_model)
+# We are interested in the P value here. If it's statistically significant then the best_alt_model does better
+# (Is worth its additional complexity in terms of how much it makes the model fit) than the null model,
+# So... for females, from birth, in England & Wales, there's evidence that there WAS a significant change in 
+# average life expectancy improvement from 2012 onwards. (i.e. two years after austerity was introduced)
+
+# We can manually repeat the same exercise for some additional subsets of the data (population, gender, starting age combinations). The below shows the same for males in England & Wales, from birth.
+
+example_df <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 0) %>% 
+      filter(sex == "male") %>% 
+      filter(code == "GBRTENW") %>% 
+      filter(year >= 1979)
+
+example_df <- 
+  example_df %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex))
+
+grid_search_df <- tibble(
+  tau_candidate = 1982:2015
+) %>% 
+  mutate(
+    model_alt = map(tau_candidate, run_alt_with_given_tau, df = example_df)
+  ) %>% 
+  mutate(
+    aic = map_dbl(model_alt, AIC)
+  )
+
+grid_search_df %>% 
+  ggplot(aes(tau_candidate, aic)) + 
+  geom_point()
+
+best_alt_model <- example_df %>% 
+  mutate( T_param = ifelse(year < 2013, FALSE, TRUE)) %>% 
+  lm(delta_ex ~ T_param, data = .)
+
+null_model <- example_df %>% 
+  lm(delta_ex ~ 1, data = .)
+
+anova(null_model, best_alt_model)
+
+# This time 2013 is a slightly better candidate than 2012, hence 2013 being used in best_alt_model not 2012.
+
+# Another example: France, males
+###
+example_df <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 0) %>% 
+      filter(sex == "male") %>% 
+      filter(code == "FRATNP") %>% 
+      filter(year >= 1979)
+
+example_df <- 
+  example_df %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex))
+
+grid_search_df <- tibble(
+  tau_candidate = 1982:2015
+) %>% 
+  mutate(
+    model_alt = map(tau_candidate, run_alt_with_given_tau, df = example_df)
+  ) %>% 
+  mutate(
+    aic = map_dbl(model_alt, AIC)
+  )
+
+grid_search_df %>% 
+  ggplot(aes(tau_candidate, aic)) + 
+  geom_point()
+
+best_alt_model <- example_df %>% 
+  mutate( T_param = ifelse(year < 2012, FALSE, TRUE)) %>% 
+  lm(delta_ex ~ T_param, data = .)
+
+summary(best_alt_model)
+
+null_model <- example_df %>% 
+  lm(delta_ex ~ 1, data = .)
+summary(null_model)
+
+anova(null_alt_model, best_alt_model)
+
+# Here 2015 has the lowest value, but this might be an artefact of the data period covered. 
+# We can also see that 2012 is the next best candidate.
+```
+
 # Discussion

### 2022-03-07 draft_paper_sandpit.Rmd

--- 2022-03-04 draft_paper.Rmd
@@
 
+### Joinpoint/Segmented regression approach
+
+Though there may be systemic differences in average changes in life expectancy over time, such differences may not be patterned by decade, and so the above model specification may not have been appropriate to detect any such changes.
+Instead of looking for systemic differences by decade, we can instead look at whether there are any systemic and detectable 'breaks' in the trends over time.
+If our data were life expectancy over time, this could be achieved through joinpoint analysis, in which the slope is allowed to change over time.
+As instead are data are annual changes in life expectancy, we are interested in detecting changes in intercept over time.
+If we define our Null (no change) model as follows:
+
+$$
+M_0 : y(t) = \alpha + e
+$$
+Where $e \backsim N(0, \sigma^2)$. i.e. all observations $y$ are draws from $N(\alpha, \sigma^2)$. The alternative model specification against which this is compared is then:
+
+$$
+M_A : y(t) = \alpha_0 + \beta_1 T(t) + e 
+$$
+Where $T(t)$ is defined as $0$ where $t \lt \tau$ and $1$ where $t \geq \tau$. ($\alpha_0$ is used in $M_A$ to indicate that this parameter is not directly comparable with $\alpha$ in $M_0$).
+
+This specification shows that there is an additional parameter $\tau$ to be determined in order to produce $M_A$.
+This $\tau$ parameter is the breakpoint in the series.
+The selection of $\tau$ will be determined by numeric optimisation, by selecting the value of $\tau$ that minimised AIC.
+
+Once the best possible $M_A$, using the best value of $\tau$, is selected, $M_A$ will be compared with $M_0$ using an F-test, allowing us to determine whether there is convincing evidence of any specific break in the data.
+
+We will start with a single country, starting age, and sex, and with manual selection of $\tau$, then generalise and automate further.
+
+```{r}
+example_df <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 0) %>% 
+      filter(sex == "female") %>% 
+      filter(code == "GBRTENW") %>% 
+      filter(year >= 1979)
+
+example_df <- 
+  example_df %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) # %>% 
+  # ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+  # geom_line() +
+  # geom_hline(yintercept = 0) + 
+  # stat_smooth()
+
+model_null <- lm(delta_ex ~ 1, data = example_df)
+summary(model_null)
+# 
+# model_nuller <- lm(delta_ex ~ 0, data = example_df)
+# summary(model_nuller)
+# 
+# anova(model_nuller, model_null)
+
+model_alt01 <- lm(delta_ex ~ year >= 1990, data = example_df)
+summary(model_alt01)
+
+model_alt02 <- lm(delta_ex ~ year >= 1995, data = example_df)
+summary(model_alt02)
+
+model_alt03 <- lm(delta_ex ~ year >= 1997, data = example_df)
+summary(model_alt03)
+
+model_alt04 <- lm(delta_ex ~ year >= 2008, data = example_df)
+summary(model_alt04)
+
+model_alt05 <- lm(delta_ex ~ year >= 2011, data = example_df)
+summary(model_alt05)
+
+run_alt_with_given_tau <- function(tau, df){
+  df <- 
+    df %>% 
+      filter(!is.na(delta_ex)) %>% 
+      mutate(
+        T_param = ifelse(year < tau, FALSE, TRUE)
+      )
+
+  lm(delta_ex ~ T_param, data  = df)
+}
+
+grid_search_df <- tibble(
+  tau_candidate = 1982:2015
+) %>% 
+  mutate(
+    model_alt = map(tau_candidate, run_alt_with_given_tau, df = example_df)
+  ) %>% 
+  mutate(
+    bic = map_dbl(model_alt, BIC)
+  )
+
+grid_search_df %>% 
+  ggplot(aes(tau_candidate, bic)) + 
+  geom_point()
+
+best_alt_model <- example_df %>% 
+  mutate( T_param = ifelse(year < 2012, FALSE, TRUE)) %>% 
+  lm(delta_ex ~ T_param, data = .)
+
+null_alt_model <- example_df %>% 
+  lm(delta_ex ~ 1, data = .)
+
+anova(null_alt_model, best_alt_model)
+
+example_df <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 0) %>% 
+      filter(sex == "male") %>% 
+      filter(code == "GBRTENW") %>% 
+      filter(year >= 1979)
+
+example_df <- 
+  example_df %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex))
+
+grid_search_df <- tibble(
+  tau_candidate = 1982:2015
+) %>% 
+  mutate(
+    model_alt = map(tau_candidate, run_alt_with_given_tau, df = example_df)
+  ) %>% 
+  mutate(
+    aic = map_dbl(model_alt, AIC)
+  )
+
+grid_search_df %>% 
+  ggplot(aes(tau_candidate, aic)) + 
+  geom_point()
+
+best_alt_model <- example_df %>% 
+  mutate( T_param = ifelse(year < 2013, FALSE, TRUE)) %>% 
+  lm(delta_ex ~ T_param, data = .)
+
+null_alt_model <- example_df %>% 
+  lm(delta_ex ~ 1, data = .)
+
+anova(null_alt_model, best_alt_model)
+
+###
+example_df <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 0) %>% 
+      filter(sex == "male") %>% 
+      filter(code == "FRATNP") %>% 
+      filter(year >= 1979)
+
+example_df <- 
+  example_df %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex))
+
+grid_search_df <- tibble(
+  tau_candidate = 1982:2015
+) %>% 
+  mutate(
+    model_alt = map(tau_candidate, run_alt_with_given_tau, df = example_df)
+  ) %>% 
+  mutate(
+    aic = map_dbl(model_alt, AIC)
+  )
+
+grid_search_df %>% 
+  ggplot(aes(tau_candidate, aic)) + 
+  geom_point()
+
+best_alt_model <- example_df %>% 
+  mutate( T_param = ifelse(year < 2013, FALSE, TRUE)) %>% 
+  lm(delta_ex ~ T_param, data = .)
+
+null_alt_model <- example_df %>% 
+  lm(delta_ex ~ 1, data = .)
+
+anova(null_alt_model, best_alt_model)
+
+```
+
 # Discussion

### 2022-03-07 draft_paper.Rmd
changed to null_model
get optim to work; realise I probably shouldn't have tried!

@@
 # We can also see that 2012 is the next best candidate.
+
+## Generalising and automating.
+
+# We've manually found the best break points, and compared the best alt with the standard null model 
+# for each of these. What we should do now is standardise the process so it 'works' automatically 
+# for each of the population subgroups.
+
+# As with creating synthetic germany, the grid/line search approach shown above should work, 
+# but I'm going to try to use optim nonetheless for pedagogic and perfectionist(/masochistic) reasons
+
+# optim uses some starting conditions (and sometimes other conditions) to use an algorithm 
+# to find an input or group of inputs that give the best possible output from a function that takes the 
+# input/inputs.
+
+# best can either mean the lowest or the highest achievable value. When trying to minimise errors
+# (as with model specifications) that usually means trying to get the minimum value.
+
+# So, the input to the function will be the tau value. 
+# The tau value range should have some reasonable limits. These can be specified 
+# for some kinds of algorithm optim uses.
+# the limits should also be determined by the data. I don't think the tau should be 
+# set at the first or last observed year, as it doesn't make sense to speak of a 'trend' 
+# from a single observation. This thinking generalises to at least two observations too. 
+# So, that implies there should be a 'buffer' within which the optim function will look 
+# for possible tau values. This should be an adjustable parameter for passing for 
+# optim, but not the parameter for optim to determine.
+
+# There's a special type of optim algorithm for bounded optimisation like this.
+# 'L-BFGS-B'. This can be selected using the method argument, and also 
+# requires the lower and upper arguments be specified, for the lower 
+# and upper values to search within respectively.
+
+# so, we need to write a function a bit like the following
+
+get_best_tau_from_optim <- function(df, buffer = 2, what = c("tau", "optim")){
+  # We can put in some data validation checks too
+
+  what <- match.arg(what)
+
+  years <- df$year 
+  tau_lower <- min(years) + buffer # the lower bound 
+  tau_upper <- max(years) - buffer # the upper bound
+  tau_start <- (tau_upper + tau_lower) / 2 # start just in the middle 
+  # (This shouldn't matter if the algorithm is robust)
+
+  get_result_from_model <- function(par, df, what = c("AIC", "BIC", "model")){
+    # Again, can carry out more validation checks 
+    what <- match.arg(what)
+
+    tau <- par[["tau"]] # This is how parameters are packed up by optim
+
+    model <-
+      df %>% 
+        arrange(year) %>% 
+        mutate(delta_ex = ex - lag(ex)) %>% 
+        filter(!is.na(delta_ex)) %>% 
+        mutate(T_param = ifelse(year < tau, FALSE, TRUE)) %>% 
+        lm(delta_ex ~ T_param, data = .)
+
+    if (what == "model") {return(model)      }
+    if (what == "BIC")   {return(BIC(model)) }
+    if (what == "AIC")   {return(AIC(model)) }
+
+    # NULL should never be returned! If it has something's gone wrong!
+    NULL
+  }
+
+  optim_obj <- 
+    optim(
+      par = list(tau = tau_start), 
+      fn = get_result_from_model, 
+      method = "L-BFGS-B",
+      lower = tau_lower,
+      upper = tau_upper,
+      df = df
+    )
+
+  # Need to make the result dependent on the what argument
+
+  if (what == "tau") {return(optim_obj[["par"]])}
+  else if (what == "optim"){ return(optim_obj) }
+
+  # Again, the following should never be triggered
+  NULL
+}
+
+# Let's see if this now works in the functional programming context, using 
+# nest and tibble elements
+
+hmd_ex_selected_countries_with_synth %>%
+  filter(year >= 1979) %>% 
+  group_by(code, x, sex) %>% 
+  nest() %>% 
+  mutate(
+    best_tau = map_dbl(data, get_best_tau_from_optim)
+  )
+
+# I think this is getting stuck on local optima, so instead maybe 
+# the more exhaustive search should be used after all!
+
 ```

### 2022-03-08 draft_paper.Rmd
did the gridsearch approach - seem to be more sensible answers!

@@
 
-anova(null_alt_model, best_alt_model)
+anova(null_model, best_alt_model)
 
@@
 
+# There are two alternative approaches:
+
+# 1) Use an exhaustive grid search approach 
+# 2) Use an algorithm which is less likely to get stuck in local optima
+
+# The first approach is much more sensible, but the second might be a cool
+# stretch goal as well!
+
+# let's use a heavily modified version of the above function to do the grid search
+
+get_best_tau_from_gridsearch <- function(df, buffer = 2, what = c("best", "all")){
+  # We can put in some data validation checks too
+
+  what <- match.arg(what)
+
+  years <- df$year 
+  tau_lower <- min(years) + buffer # the lower bound 
+  tau_upper <- max(years) - buffer # the upper bound
+  tau_searchrange <- tau_lower:tau_upper
+
+  get_result_from_model <- function(tau, df, what = c("AIC", "BIC", "model")){
+    # Again, can carry out more validation checks 
+    what <- match.arg(what)
+
+    model <-
+      df %>% 
+        arrange(year) %>% 
+        mutate(delta_ex = ex - lag(ex)) %>% 
+        filter(!is.na(delta_ex)) %>% 
+        mutate(T_param = ifelse(year < tau, FALSE, TRUE)) %>% 
+        lm(delta_ex ~ T_param, data = .)
+
+    if (what == "model") {return(model)      }
+    if (what == "BIC")   {return(BIC(model)) }
+    if (what == "AIC")   {return(AIC(model)) }
+
+    # NULL should never be returned! If it has something's gone wrong!
+    NULL
+  }
+
+  search_df <- tibble(
+    tau = tau_searchrange
+  ) %>% 
+    mutate(aic = map_dbl(tau, get_result_from_model, df = df))
+
+  if (what == "all") {return(search_df)}
+
+  if (what == "best") {
+    out <- 
+      search_df %>% 
+        filter(aic == min(aic)) %>% 
+        pull(tau)
+    return(out)
+  }
+
+  # Again, the following should never be triggered
+  NULL
+}
+
+hmd_ex_selected_countries_with_synth %>%
+  filter(year >= 1979) %>% 
+  group_by(code, x, sex) %>% 
+  nest() %>% 
+  mutate(
+    best_tau = map_dbl(data, get_best_tau_from_gridsearch)
+  )
+
 ```

### 2022-03-08 draft_paper.Rmd
run and compare best breakpoint with no break (null) model

@@
 
+compare_best_breakmodel_with_null <- 
   hmd_ex_selected_countries_with_synth %>%
@@
       best_tau = map_dbl(data, get_best_tau_from_gridsearch)
+    ) %>% 
+    mutate(
+      model_null = map(
+        data, 
+        ~ .x %>% 
+          arrange(year) %>% 
+          mutate(delta_ex = ex - lag(ex)) %>% 
+          filter(!is.na(delta_ex)) %>% 
+          lm(delta_ex ~ 1, data = .)
+      )
+    ) %>% 
+    mutate(
+      model_alt = map2(
+        best_tau, data,
+      ~ .y %>% 
+        arrange(year) %>% 
+        mutate(delta_ex = ex - lag(ex)) %>% 
+        filter(!is.na(delta_ex)) %>% 
+        mutate(T_param = ifelse(year < .x, FALSE, TRUE)) %>% 
+        lm(delta_ex ~ T_param, data = .)
+      )
+    )
+
+compare_best_breakmodel_with_null %>% 
+  mutate(
+    aic_alt  = map_dbl(model_alt, AIC),
+    aic_null = map_dbl(model_null, AIC)
+  ) %>% 
+  mutate(
+    alt_preferred = map2_lgl(aic_alt, aic_null, ~ifelse(.x < .y, TRUE, FALSE))
   )

### 2022-03-08 draft_paper.Rmd
add f test and stat sig indicator (is the p value < 0.05?)

@@
     alt_preferred = map2_lgl(aic_alt, aic_null, ~ifelse(.x < .y, TRUE, FALSE))
+  ) %>% 
+  mutate(
+    anova_results = map2(model_alt, model_null, anova)
+  ) %>% 
+  mutate(
+    ftest_pval = map_dbl(anova_results, ~.[2,6]),
+    stat_sig_preference = ftest_pval < 0.05
   )

### 2022-03-09 draft_paper.Rmd
did simulated annealing; shouldn't have...

@@
 
+# Now let's finish with the most unneccesarily complicated variant: using a different algorithm in 
+# optim which will allow local minima to be escaped. This is the simulated annealing method 
+# (method = "SANN") which is based on physics models. Unfortunately this is not a bounded algoritm,
+# i.e. does not take values within limits, so instead the parameter passed to it has to be 
+# converted and deconverted from a bounded range using a link function. 
+# This can be done with a modification of the sigmoid function, which shifts the transformed 
+# value from the range 0-1 to the range lower_tau-upper_tau.
+
+# Let's build up this function as follows:
+
+# sigmoid_shift <- function(x, lower = 0, upper = 1){
+#   (upper - lower) / (1 + exp(-x)) + lower
+# }
+# # 
+# curve(sigmoid_shift, -3, 3) # basic sigmoid 
+# 
+# sigmoid_shift2 <- function(x){
+#   sigmoid_shift(x = x, lower = 1981, upper = 2015)
+# }
+# 
+# curve(sigmoid_shift2, -3, 3)
+
+# So this will work as a transformation from an unbounded space (x) to a 
+# bounded space (f(x))
+
+# We'll also need to reverse the operation
+
+# # After a bit of pen-and-paper algebra...
+# sigmoid_unshift <- function(z, lower = 0, upper = 1){
+#   -log((upper - lower) / (z - lower) -1)
+# }
+# 
+# Let's double check...
+
+#  
+# sigmoid_shift(-4, lower = 1982, upper = 2015) %>% 
+#   sigmoid_unshift(lower = 1982, upper = 2015)
+# 
+# 
+# 
+# get_best_tau_from_optim_sann <- function(df, buffer = 2, what = c("tau", "tau_z", "optim")){
+#   # We can put in some data validation checks too 
+#   sigmoid_shift <- function(x, lower = 0, upper = 1){
+#     (upper - lower) / (1 + exp(-x)) + lower
+#   }
+#  
+#   sigmoid_unshift <- function(z, lower = 0, upper = 1){
+#   -log((upper - lower) / (z - lower) -1)
+#   } 
+#   
+#   
+#   what <- match.arg(what)
+#   
+#   years <- df$year 
+#   tau_lower <- min(years) + buffer # the lower bound 
+#   tau_upper <- max(years) - buffer # the upper bound
+#   # tau_start <- (tau_upper + tau_lower) / 2 # start just in the middle 
+#   # (This shouldn't matter if the algorithm is robust) 
+#   
+#   get_result_from_model <- function(par, df, what = c("AIC", "BIC", "model")){
+#     # Again, can carry out more validation checks 
+#     what <- match.arg(what)
+#     
+#     tau_z <- par[["tau_z"]] # This is how parameters are packed up by optim
+#     tau   <- sigmoid_shift(tau_z, lower = tau_lower, upper = tau_upper)
+#     
+#     model <-
+#       df %>% 
+#         arrange(year) %>% 
+#         mutate(delta_ex = ex - lag(ex)) %>% 
+#         filter(!is.na(delta_ex)) %>% 
+#         mutate(T_param = ifelse(year < tau, FALSE, TRUE)) %>% 
+#         lm(delta_ex ~ T_param, data = .)
+# 
+#     if (what == "model") {return(model)      }
+#     if (what == "BIC")   {return(BIC(model)) }
+#     if (what == "AIC")   {return(AIC(model)) }
+#     
+#     # NULL should never be returned! If it has something's gone wrong!
+#     NULL
+#   }
+# 
+#   optim_obj <- 
+#     optim(
+#       par = list(tau_z = 0), 
+#       fn = get_result_from_model, 
+#       method = "SANN",
+#       df = df
+#     )
+# 
+#   # Need to make the result dependent on the what argument
+#   
+#   if (what == "tau") {return(sigmoid_shift(optim_obj[["par"]], lower = tau_lower, upper = tau_upper))}
+#   if (what == "tau_z") {return(optim_obj[["par"]])}  
+#   else if (what == "optim"){ return(optim_obj) }
+#   
+#   # Again, the following should never be triggered
+#   NULL
+# }
+# 
+# 
+# sann_estimates <- 
+#   hmd_ex_selected_countries_with_synth %>%
+#     filter(year >= 1979) %>% 
+#     group_by(code, x, sex) %>% 
+#     nest() %>% 
+#     mutate(
+#       best_tau = map_dbl(data, get_best_tau_from_optim_sann)
+#     )
+# 
+# # This takes ABSOLUTELY AGES, and really is the sledgehammer-to-crack
+# # -a-nut approach to use only as a last resort. #
+# # However if you ever do have a complex and non-finite fitness 
+# # surface to try to optimise, and it also requires searching 
+# # over a bounded area, then this is how you can do it!
+
 ```

### 2022-03-09 draft_paper.Rmd
Added a table! Eventually.

--- 2022-03-08 draft_paper.Rmd
@@
 
-compare_best_breakmodel_with_null %>% 
+tmp <- compare_best_breakmodel_with_null %>% 
   mutate(
@@
     alt_preferred = map2_lgl(aic_alt, aic_null, ~ifelse(.x < .y, TRUE, FALSE))
+  ) %>% 
+  mutate(
+    anova_results = map2(model_alt, model_null, anova)
+  ) %>% 
+  mutate(
+    ftest_pval = map_dbl(anova_results, ~.[2,6]),
+    stat_sig_preference = ftest_pval < 0.05
   )
 
+```
+
+Next, we display a table with those countries where the breakpoint shows as statistically significant compared to the null, where best_tau is the year.
+
+```{r}
+stat_sig_table <- select(tmp, code, x, sex, best_tau, stat_sig_preference) %>% filter(stat_sig_preference==TRUE)
+
+rename(stat_sig_table, age = x, country = code)
+view(stat_sig_table)
+
+#using flextable
+myft <- flextable(stat_sig_table)
+myft
+myft <- theme_vanilla(myft)
+myft
 ```

### 2022-03-09 draft_paper.Rmd

--- 2022-03-09 draft_paper.Rmd
@@
 
+Next, we display a table with those countries where the breakpoint shows as statistically significant compared to the null, where best_tau is the year.
+
+```{r}
+stat_sig_table <- select(tmp, code, x, sex, best_tau, stat_sig_preference) %>% filter(stat_sig_preference==TRUE)
+
+rename(stat_sig_table, age = x, country = code)
+view(stat_sig_table)
+
+#using flextable
+myft <- flextable(stat_sig_table)
+myft
+myft <- theme_vanilla(myft)
+myft
+```
+
 # Discussion

### 2022-03-09 draft_paper.Rmd

@@
 ```
+Next, let's look closer at France, Scotland, and E&Ws life expectancy at birth for males.
+Jon: can we somehow include these 3 in the below to produce a graph? I tried adding into the first chunk but it does not run the later parts.
+
+```{r}
+example_df <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 0) %>% 
+      filter(sex == "male") %>% 
+      filter(code == "FRATNP") %>% 
+      filter(year >= 1979)
+
+example_df <- 
+  example_df %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex))
+
+grid_search_df <- tibble(
+  tau_candidate = 1982:2015
+) %>% 
+  mutate(
+    model_alt = map(tau_candidate, run_alt_with_given_tau, df = example_df)
+  ) %>% 
+  mutate(
+    aic = map_dbl(model_alt, AIC)
+  )
+
+grid_search_df %>% 
+  ggplot(aes(tau_candidate, aic)) + 
+  geom_point()
+```
 

### 2022-03-10 draft_graphs.Rmd
new doc with graphs for each country, age, sex and labels

--- 2022-03-09 draft_paper.Rmd
@@
 
-```{r}
-
-# Let's start with a single example population. (There are many different population subgroups, hence the 
-# benefit of being able to automate this later.) We're going to start with England & Wales, females, life 
-# expectancy at birth (x == 0)
-
-example_df <- 
+## Graphs of annual change in life expectancy
+
+The below graphs show the annual change in life expectancy at birth and at aged 65 years for males and females for the nations examined. The pink and green lines are non-linear smoothed.
+
+NB: those with statistically significant 'breakpoints' in annual change in life expectancy are:
+- Germany (need to re-download from GIT)
+
+### England and Wales
+
+```{r, echo = FALSE, warning=FALSE, message = FALSE}
+
+#ENW female at birth
+f_e0_ENW <- 
   hmd_ex_selected_countries_with_synth %>% 
@@
 # Now we create the change in ex variable as that's the response variable in the model
-example_df <- 
-  example_df %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex)) # %>% 
-  # ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
-  # geom_line() +
-  # geom_hline(yintercept = 0) + 
-  # stat_smooth()
-
-# The null model is really simple: just an intercept 
-model_null <- lm(delta_ex ~ 1, data = example_df)
-summary(model_null)
+  f_e0_ENW <- f_e0_ENW %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+   geom_line() +
+   geom_hline(yintercept = 0) + 
+   stat_smooth(colour = "lightpink3")
+
+  #ENW male at birth
+m_e0_ENW <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 0) %>% 
+      filter(sex == "male") %>% 
+      filter(code == "GBRTENW") %>% 
+      filter(year >= 1979)
+
+  m_e0_ENW <- m_e0_ENW %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+   geom_line() +
+   geom_hline(yintercept = 0) + 
+   stat_smooth(colour = "aquamarine4")
+
+  #ENW female at age 65
+f_e65_ENW <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 65) %>% 
+      filter(sex == "female") %>% 
+      filter(code == "GBRTENW") %>% 
+      filter(year >= 1979)
+
+# Now we create the change in ex variable as that's the response variable in the model
+  f_e65_ENW <- f_e65_ENW %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+   geom_line() +
+   geom_hline(yintercept = 0) + 
+   stat_smooth(colour = "lightpink3")
+
+  #ENW male at aged 65
+m_e65_ENW <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 65) %>% 
+      filter(sex == "male") %>% 
+      filter(code == "GBRTENW") %>% 
+      filter(year >= 1979)
+
+ m_e65_ENW <- m_e65_ENW %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+   geom_line() +
+   geom_hline(yintercept = 0) + 
+   stat_smooth(colour = "aquamarine4")
+
+ #add labels
+f_e0_ENW + labs(y = "Annual change in e0", title = "England and Wales", subtitle = "Life expectancy changes at birth for females", caption = "Source: Human Mortality Database")
+
+m_e0_ENW + labs(y = "Annual change in e0", title = "England and Wales", subtitle = "Life expectancy changes at birth for males", caption = "Source: Human Mortality Database")
+
+f_e65_ENW + labs(y = "Annual change in e65", title = "England and Wales", subtitle = "Life expectancy changes at aged 65 for females", caption = "Source: Human Mortality Database")
+
+m_e65_ENW + labs(y = "Annual change in e65", title = "England and Wales", subtitle = "Life expectancy changes at aged 65 for males", caption = "Source: Human Mortality Database")
+
+  #to make grid, first assign to object
+# f_e0_ENW <- f_e0_ENW + labs(y = "Annual change in e0", title = "England and Wales", subtitle = "Life expectancy changes at birth for females", caption = "Source: Human Mortality Database")
 # 
-
-# The code below shows how we can be even simpler, and force the regression model not to 
-# even have an intercept (i.e. have zero predictor variables)
-
-# model_nuller <- lm(delta_ex ~ 0, data = example_df)
-# summary(model_nuller)
-
-# And we can compare the intercept with the no intercept model as follows 
-# anova(model_nuller, model_null)
-# a statistically significant p value shows the intercept is definitely worth including in the 
-# model. Substantively that it's wrong to assume no improvement on average in ex over time since the 1980s
-
-# There are lots of possible alternative models, some examples of which are below:
-model_alt01 <- lm(delta_ex ~ year >= 1990, data = example_df)
-summary(model_alt01)
-
-model_alt02 <- lm(delta_ex ~ year >= 1995, data = example_df)
-summary(model_alt02)
-
-model_alt03 <- lm(delta_ex ~ year >= 1997, data = example_df)
-summary(model_alt03)
-
-model_alt04 <- lm(delta_ex ~ year >= 2008, data = example_df)
-summary(model_alt04)
-
-model_alt05 <- lm(delta_ex ~ year >= 2011, data = example_df)
-summary(model_alt05)
-
-# each of these models allows an adjustment to the horizontal line after a given year has been reached. 
-# In the examples above some possible given years are 1990, 1995, 1997, 2008 and 2011. 
-# Each of these is a candidate value for tau. But we've not looked at all possible/plausible values of this.
-
-# Hence, it would be good to automate further to search for each possible tau and see how the fit and model 
-# results change.
-
-# We'll start to automate by building a function which creates T_param given tau, and runs and returns the model
-# with this type of T_param
-# (If tau is 1990, then T_param is equivalent to model_alt01 above; if tau were 1995, then it's equivalent to model_alt02, and so on.
-The only difference is that the new coefficient created has the same name, T_param,
-# whereas in the above it would have a different name each time, because the coefficient is being 
-# created in situ, within the formula specification, rather than by pre-processing the dataframe)
-run_alt_with_given_tau <- function(tau, df){
-  df <- 
-    df %>% 
-      filter(!is.na(delta_ex)) %>% 
-      mutate(
-        T_param = ifelse(year < tau, FALSE, TRUE)
-      )
-
-  lm(delta_ex ~ T_param, data  = df)
-}
-
-# We can now run the above function, which creates a model, for a wide range of possible tau values
-# This is the sequence listed and named tau_candidate, which is every whole number from 
-# 1982 to 2015
-
-# For each of these models, we can then get a measure of model fit. We're going to use AIC, though 
-# could have used other measures. With AIC (and with BIC, and RMSE etc) better fit is indicated with 
-# a lower value
-grid_search_df <- tibble(
-  tau_candidate = 1982:2015 # column of candidate values
-) %>% 
-  mutate(
-    model_alt = map(tau_candidate, run_alt_with_given_tau, df = example_df) # run the model with the value of tau in tau_candidate
-  ) %>% 
-  mutate(
-    aic = map_dbl(model_alt, AIC)
-  )
-
-# Now we can see which proposed tau results in what fit value (AIC), initially just by graphing it.
-grid_search_df %>% 
-  ggplot(aes(tau_candidate, aic)) + 
-  geom_point()
-
-# So, in the above we found 2012 as tau minimised AIC. So, that's the best possible alternative model. 
-# Let's run and label it as such
-best_alt_model <- example_df %>% 
-  mutate( T_param = ifelse(year < 2012, FALSE, TRUE)) %>% 
-  lm(delta_ex ~ T_param, data = .)
-
-# Now we want to create the null model to compare it against (which just has an intercept term alone)
-null_model <- example_df %>% 
-  lm(delta_ex ~ 1, data = .)
-
-# And finally, we want to compare the best_alt_model against the null model using an F test
-anova(null_model, best_alt_model)
-# We are interested in the P value here. If it's statistically significant then the best_alt_model does better
-# (Is worth its additional complexity in terms of how much it makes the model fit) than the null model,
-# So... for females, from birth, in England & Wales, there's evidence that there WAS a significant change in 
-# average life expectancy improvement from 2012 onwards. (i.e. two years after austerity was introduced)
-
-# We can manually repeat the same exercise for some additional subsets of the data (population, gender, starting age combinations). The below shows the same for males in England & Wales, from birth.
-
-example_df <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 0) %>% 
-      filter(sex == "male") %>% 
-      filter(code == "GBRTENW") %>% 
-      filter(year >= 1979)
-
-example_df <- 
-  example_df %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex))
-
-grid_search_df <- tibble(
-  tau_candidate = 1982:2015
-) %>% 
-  mutate(
-    model_alt = map(tau_candidate, run_alt_with_given_tau, df = example_df)
-  ) %>% 
-  mutate(
-    aic = map_dbl(model_alt, AIC)
-  )
-
-grid_search_df %>% 
-  ggplot(aes(tau_candidate, aic)) + 
-  geom_point()
-
-best_alt_model <- example_df %>% 
-  mutate( T_param = ifelse(year < 2013, FALSE, TRUE)) %>% 
-  lm(delta_ex ~ T_param, data = .)
-
-null_model <- example_df %>% 
-  lm(delta_ex ~ 1, data = .)
-
-anova(null_model, best_alt_model)
-
-# This time 2013 is a slightly better candidate than 2012, hence 2013 being used in best_alt_model not 2012.
-
-# Another example: France, males
-###
-example_df <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 0) %>% 
-      filter(sex == "male") %>% 
+# m_e0_ENW <- m_e0_ENW + labs(y = "Annual change in e0", title = "England and Wales", subtitle = "Life expectancy changes at birth for males", caption = "Source: Human Mortality Database")
+# 
+# f_e65_ENW <- f_e65_ENW + labs(y = "Annual change in e65", title = "England and Wales", subtitle = "Life expectancy changes at aged 65 for females", caption = "Source: Human Mortality Database")
+# 
+# m_e65_ENW <- m_e65_ENW + labs(y = "Annual change in e65", title = "England and Wales", subtitle = "Life expectancy changes at aged 65 for males", caption = "Source: Human Mortality Database")
+
+#changing labels for grid to work
+# f_e0_ENW <- f_e0_ENW + labs(y = "Annual change in e0", title = "England and Wales", subtitle = "Life expectancy changes at birth for females")
+# m_e0_ENW <- m_e0_ENW + labs(y = "Annual change in e0", subtitle = "Life expectancy changes at birth for males")
+# f_e65_ENW <- f_e65_ENW + labs(y = "Annual change in e65")
+# m_e65_ENW <- m_e65_ENW + labs(y = "Annual change in e65", caption = "Source: Human Mortality Database")
+# #put side by side, needs package gridExtra then uses grid.arrange(plot1, plot2, ncol=2)
+# grid.arrange(f_e0_ENW, m_e0_ENW, f_e65_ENW, m_e65_ENW, ncol=2, nrow=2)
+# #grid.arrange(f_e65_ENW, m_e65_ENW, ncol=2, nrow=2)
+
+```
+
+### Scotland
+```{r, echo = FALSE, warning=FALSE, message = FALSE}
+#SCO female at birth
+f_e0_SCO <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 0) %>% 
+      filter(sex == "female") %>% 
+      filter(code == "GBR_SCO") %>% 
+      filter(year >= 1979)
+
+# Now we create the change in ex variable as that's the response variable in the model
+  f_e0_SCO <- f_e0_SCO %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+   geom_line() +
+   geom_hline(yintercept = 0) + 
+   stat_smooth(colour = "lightpink3")
+
+  #SCO male at birth
+m_e0_SCO <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 0) %>% 
+      filter(sex == "male") %>% 
+      filter(code == "GBR_SCO") %>% 
+      filter(year >= 1979)
+
+  m_e0_SCO <- m_e0_SCO %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+   geom_line() +
+   geom_hline(yintercept = 0) + 
+   stat_smooth(colour = "aquamarine4")
+
+  #SCO female at age 65
+f_e65_SCO <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 65) %>% 
+      filter(sex == "female") %>% 
+      filter(code == "GBR_SCO") %>% 
+      filter(year >= 1979)
+
+# Now we create the change in ex variable as that's the response variable in the model
+  f_e65_SCO <- f_e65_SCO %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+   geom_line() +
+   geom_hline(yintercept = 0) + 
+   stat_smooth(colour = "lightpink3")
+
+  #SCO male at aged 65
+m_e65_SCO <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 65) %>% 
+      filter(sex == "male") %>% 
+      filter(code == "GBR_SCO") %>% 
+      filter(year >= 1979)
+
+ m_e65_SCO <-  m_e65_SCO %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+   geom_line() +
+   geom_hline(yintercept = 0) + 
+   stat_smooth(colour = "aquamarine4", show.legend = TRUE)
+
+  #add labels
+f_e0_SCO + labs(y = "Annual change in e0", title = "Scotland", subtitle = "Life expectancy changes at birth for females", caption = "Source: Human Mortality Database")
+ m_e0_SCO + labs(y = "Annual change in e0", title = "Scotland", subtitle = "Life expectancy changes at birth for males", caption = "Source: Human Mortality Database")
+ f_e65_SCO + labs(y = "Annual change in e65", title = "Scotland", subtitle = "Life expectancy changes at aged 65 for females", caption = "Source: Human Mortality Database")
+ m_e65_SCO + labs(y = "Annual change in e65", title = "Scotland", subtitle = "Life expectancy changes at aged 65 for males", caption = "Source: Human Mortality Database")
+```
+
+### France
+
+```{r, echo = FALSE, warning=FALSE, message = FALSE}
+#FRA female at birth
+f_e0_FRA <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 0) %>% 
+      filter(sex == "female") %>% 
       filter(code == "FRATNP") %>% 
@@
 
-example_df <- 
-  example_df %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex))
-
-grid_search_df <- tibble(
-  tau_candidate = 1982:2015
-) %>% 
-  mutate(
-    model_alt = map(tau_candidate, run_alt_with_given_tau, df = example_df)
-  ) %>% 
-  mutate(
-    aic = map_dbl(model_alt, AIC)
-  )
-
-grid_search_df %>% 
-  ggplot(aes(tau_candidate, aic)) + 
-  geom_point()
-
-best_alt_model <- example_df %>% 
-  mutate( T_param = ifelse(year < 2012, FALSE, TRUE)) %>% 
-  lm(delta_ex ~ T_param, data = .)
-
-summary(best_alt_model)
-
-null_model <- example_df %>% 
-  lm(delta_ex ~ 1, data = .)
-summary(null_model)
-
-anova(null_model, best_alt_model)
-
-# Here 2015 has the lowest value, but this might be an artefact of the data period covered. 
-# We can also see that 2012 is the next best candidate.
-
-## Generalising and automating.
-
-# We've manually found the best break points, and compared the best alt with the standard null model 
-# for each of these. What we should do now is standardise the process so it 'works' automatically 
-# for each of the population subgroups.
-
-# As with creating synthetic germany, the grid/line search approach shown above should work, 
-# but I'm going to try to use optim nonetheless for pedagogic and perfectionist(/masochistic) reasons
-
-# optim uses some starting conditions (and sometimes other conditions) to use an algorithm 
-# to find an input or group of inputs that give the best possible output from a function that takes the 
-# input/inputs.
-
-# best can either mean the lowest or the highest achievable value. When trying to minimise errors
-# (as with model specifications) that usually means trying to get the minimum value.
-
-# So, the input to the function will be the tau value. 
-# The tau value range should have some reasonable limits. These can be specified 
-# for some kinds of algorithm optim uses.
-# the limits should also be determined by the data. I don't think the tau should be 
-# set at the first or last observed year, as it doesn't make sense to speak of a 'trend' 
-# from a single observation. This thinking generalises to at least two observations too. 
-# So, that implies there should be a 'buffer' within which the optim function will look 
-# for possible tau values. This should be an adjustable parameter for passing for 
-# optim, but not the parameter for optim to determine.
-
-# There's a special type of optim algorithm for bounded optimisation like this.
-# 'L-BFGS-B'. This can be selected using the method argument, and also 
-# requires the lower and upper arguments be specified, for the lower 
-# and upper values to search within respectively.
-
-# so, we need to write a function a bit like the following
-
-get_best_tau_from_optim <- function(df, buffer = 2, what = c("tau", "optim")){
-  # We can put in some data validation checks too
-
-  what <- match.arg(what)
-
-  years <- df$year 
-  tau_lower <- min(years) + buffer # the lower bound 
-  tau_upper <- max(years) - buffer # the upper bound
-  tau_start <- (tau_upper + tau_lower) / 2 # start just in the middle 
-  # (This shouldn't matter if the algorithm is robust)
-
-  get_result_from_model <- function(par, df, what = c("AIC", "BIC", "model")){
-    # Again, can carry out more validation checks 
-    what <- match.arg(what)
-
-    tau <- par[["tau"]] # This is how parameters are packed up by optim
-
-    model <-
-      df %>% 
-        arrange(year) %>% 
-        mutate(delta_ex = ex - lag(ex)) %>% 
-        filter(!is.na(delta_ex)) %>% 
-        mutate(T_param = ifelse(year < tau, FALSE, TRUE)) %>% 
-        lm(delta_ex ~ T_param, data = .)
-
-    if (what == "model") {return(model)      }
-    if (what == "BIC")   {return(BIC(model)) }
-    if (what == "AIC")   {return(AIC(model)) }
-
-    # NULL should never be returned! If it has something's gone wrong!
-    NULL
-  }
-
-  optim_obj <- 
-    optim(
-      par = list(tau = tau_start), 
-      fn = get_result_from_model, 
-      method = "L-BFGS-B",
-      lower = tau_lower,
-      upper = tau_upper,
-      df = df
-    )
-
-  # Need to make the result dependent on the what argument
-
-  if (what == "tau") {return(optim_obj[["par"]])}
-  else if (what == "optim"){ return(optim_obj) }
-
-  # Again, the following should never be triggered
-  NULL
-}
-
-# Let's see if this now works in the functional programming context, using 
-# nest and tibble elements
-
-hmd_ex_selected_countries_with_synth %>%
-  filter(year >= 1979) %>% 
-  group_by(code, x, sex) %>% 
-  nest() %>% 
-  mutate(
-    best_tau = map_dbl(data, get_best_tau_from_optim)
-  )
-
-# I think this is getting stuck on local optima, so instead maybe 
-# the more exhaustive search should be used after all!
-
-# There are two alternative approaches:
-
-# 1) Use an exhaustive grid search approach 
-# 2) Use an algorithm which is less likely to get stuck in local optima
-
-# The first approach is much more sensible, but the second might be a cool
-# stretch goal as well!
-
-# let's use a heavily modified version of the above function to do the grid search
-
-get_best_tau_from_gridsearch <- function(df, buffer = 2, what = c("best", "all")){
-  # We can put in some data validation checks too
-
-  what <- match.arg(what)
-
-  years <- df$year 
-  tau_lower <- min(years) + buffer # the lower bound 
-  tau_upper <- max(years) - buffer # the upper bound
-  tau_searchrange <- tau_lower:tau_upper
-
-  get_result_from_model <- function(tau, df, what = c("AIC", "BIC", "model")){
-    # Again, can carry out more validation checks 
-    what <- match.arg(what)
-
-    model <-
-      df %>% 
-        arrange(year) %>% 
-        mutate(delta_ex = ex - lag(ex)) %>% 
-        filter(!is.na(delta_ex)) %>% 
-        mutate(T_param = ifelse(year < tau, FALSE, TRUE)) %>% 
-        lm(delta_ex ~ T_param, data = .)
-
-    if (what == "model") {return(model)      }
-    if (what == "BIC")   {return(BIC(model)) }
-    if (what == "AIC")   {return(AIC(model)) }
-
-    # NULL should never be returned! If it has something's gone wrong!
-    NULL
-  }
-
-  search_df <- tibble(
-    tau = tau_searchrange
-  ) %>% 
-    mutate(aic = map_dbl(tau, get_result_from_model, df = df))
-
-  if (what == "all") {return(search_df)}
-
-  if (what == "best") {
-    out <- 
-      search_df %>% 
-        filter(aic == min(aic)) %>% 
-        pull(tau)
-    return(out)
-  }
-
-  # Again, the following should never be triggered
-  NULL
-}
-
-compare_best_breakmodel_with_null <- 
-  hmd_ex_selected_countries_with_synth %>%
-    filter(year >= 1979) %>% 
-    group_by(code, x, sex) %>% 
-    nest() %>% 
-    mutate(
-      best_tau = map_dbl(data, get_best_tau_from_gridsearch)
-    ) %>% 
-    mutate(
-      model_null = map(
-        data, 
-        ~ .x %>% 
-          arrange(year) %>% 
-          mutate(delta_ex = ex - lag(ex)) %>% 
-          filter(!is.na(delta_ex)) %>% 
-          lm(delta_ex ~ 1, data = .)
-      )
-    ) %>% 
-    mutate(
-      model_alt = map2(
-        best_tau, data,
-      ~ .y %>% 
-        arrange(year) %>% 
-        mutate(delta_ex = ex - lag(ex)) %>% 
-        filter(!is.na(delta_ex)) %>% 
-        mutate(T_param = ifelse(year < .x, FALSE, TRUE)) %>% 
-        lm(delta_ex ~ T_param, data = .)
-      )
-    )
-
-tmp <- compare_best_breakmodel_with_null %>% 
-  mutate(
-    aic_alt  = map_dbl(model_alt, AIC),
-    aic_null = map_dbl(model_null, AIC)
-  ) %>% 
-  mutate(
-    alt_preferred = map2_lgl(aic_alt, aic_null, ~ifelse(.x < .y, TRUE, FALSE))
-  ) %>% 
-  mutate(
-    anova_results = map2(model_alt, model_null, anova)
-  ) %>% 
-  mutate(
-    ftest_pval = map_dbl(anova_results, ~.[2,6]),
-    stat_sig_preference = ftest_pval < 0.05
-  )
-
-```
-
-Next, we display a table with those countries where the breakpoint shows as statistically significant compared to the null, where best_tau is the year.
-
-```{r}
-stat_sig_table <- select(tmp, code, x, sex, best_tau, stat_sig_preference) %>% filter(stat_sig_preference==TRUE)
-
-rename(stat_sig_table, age = x, country = code)
-view(stat_sig_table)
-
-#using flextable
-myft <- flextable(stat_sig_table)
-myft
-myft <- theme_vanilla(myft)
-myft
+# Now we create the change in ex variable as that's the response variable in the model
+  f_e0_FRA <- f_e0_FRA %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+   geom_line() +
+   geom_hline(yintercept = 0) + 
+   stat_smooth(colour = "lightpink3")
+
+  #FRA male at birth
+m_e0_FRA <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 0) %>% 
+      filter(sex == "male") %>% 
+      filter(code == "FRATNP") %>% 
+      filter(year >= 1979)
+
+  m_e0_FRA <- m_e0_FRA %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+   geom_line() +
+   geom_hline(yintercept = 0) + 
+   stat_smooth(colour = "aquamarine4")
+
+  #FRA female at age 65
+f_e65_FRA <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 65) %>% 
+      filter(sex == "female") %>% 
+      filter(code == "FRATNP") %>% 
+      filter(year >= 1979)
+
+# Now we create the change in ex variable as that's the response variable in the model
+  f_e65_FRA <- f_e65_FRA %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+   geom_line() +
+   geom_hline(yintercept = 0) + 
+   stat_smooth(colour = "lightpink3")
+
+  #FRA male at aged 65
+m_e65_FRA <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 65) %>% 
+      filter(sex == "male") %>% 
+      filter(code == "FRATNP") %>% 
+      filter(year >= 1979)
+
+ m_e65_FRA <-  m_e65_FRA %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+   geom_line() +
+   geom_hline(yintercept = 0) + 
+   stat_smooth(colour = "aquamarine4", show.legend = TRUE)
+
+  #add labels
+f_e0_FRA + labs(y = "Annual change in e0", title = "France", subtitle = "Life expectancy changes at birth for females", caption = "Source: Human Mortality Database")
+ m_e0_FRA + labs(y = "Annual change in e0", title = "France", subtitle = "Life expectancy changes at birth for males", caption = "Source: Human Mortality Database")
+ f_e65_FRA + labs(y = "Annual change in e65", title = "France", subtitle = "Life expectancy changes at aged 65 for females", caption = "Source: Human Mortality Database")
+ m_e65_FRA + labs(y = "Annual change in e65", title = "France", subtitle = "Life expectancy changes at aged 65 for males", caption = "Source: Human Mortality Database")
+```
+
+### Spain
+
+```{r, echo = FALSE, warning=FALSE, message = FALSE}
+#ESP female at birth
+f_e0_ESP <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 0) %>% 
+      filter(sex == "female") %>% 
+      filter(code == "ESP") %>% 
+      filter(year >= 1979)
+
+# Now we create the change in ex variable as that's the response variable in the model
+  f_e0_ESP <- f_e0_ESP %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+   geom_line() +
+   geom_hline(yintercept = 0) + 
+   stat_smooth(colour = "lightpink3")
+
+  #ESP male at birth
+m_e0_ESP <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 0) %>% 
+      filter(sex == "male") %>% 
+      filter(code == "ESP") %>% 
+      filter(year >= 1979)
+
+  m_e0_ESP <- m_e0_ESP %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+   geom_line() +
+   geom_hline(yintercept = 0) + 
+   stat_smooth(colour = "aquamarine4")
+
+  #ESP female at age 65
+f_e65_ESP <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 65) %>% 
+      filter(sex == "female") %>% 
+      filter(code == "ESP") %>% 
+      filter(year >= 1979)
+
+# Now we create the change in ex variable as that's the response variable in the model
+  f_e65_ESP <- f_e65_ESP %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+   geom_line() +
+   geom_hline(yintercept = 0) + 
+   stat_smooth(colour = "lightpink3")
+
+  #ESP male at aged 65
+m_e65_ESP <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 65) %>% 
+      filter(sex == "male") %>% 
+      filter(code == "ESP") %>% 
+      filter(year >= 1979)
+
+ m_e65_ESP <-  m_e65_ESP %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+   geom_line() +
+   geom_hline(yintercept = 0) + 
+   stat_smooth(colour = "aquamarine4", show.legend = TRUE)
+
+  #add labels
+f_e0_ESP + labs(y = "Annual change in e0", title = "Spain", subtitle = "Life expectancy changes at birth for females", caption = "Source: Human Mortality Database")
+ m_e0_ESP + labs(y = "Annual change in e0", title = "Spain", subtitle = "Life expectancy changes at birth for males", caption = "Source: Human Mortality Database")
+ f_e65_ESP + labs(y = "Annual change in e65", title = "Spain", subtitle = "Life expectancy changes at aged 65 for females", caption = "Source: Human Mortality Database")
+ m_e65_ESP + labs(y = "Annual change in e65", title = "Spain", subtitle = "Life expectancy changes at aged 65 for males", caption = "Source: Human Mortality Database")
+```
+
+### Italy
+
+```{r, echo = FALSE, warning=FALSE, message = FALSE}
+#ITA female at birth
+f_e0_ITA <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 0) %>% 
+      filter(sex == "female") %>% 
+      filter(code == "ITA") %>% 
+      filter(year >= 1979)
+
+# Now we create the change in ex variable as that's the response variable in the model
+  f_e0_ITA <- f_e0_ITA %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+   geom_line() +
+   geom_hline(yintercept = 0) + 
+   stat_smooth(colour = "lightpink3")
+
+  #ITA male at birth
+m_e0_ITA <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 0) %>% 
+      filter(sex == "male") %>% 
+      filter(code == "ITA") %>% 
+      filter(year >= 1979)
+
+  m_e0_ITA <- m_e0_ITA %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+   geom_line() +
+   geom_hline(yintercept = 0) + 
+   stat_smooth(colour = "aquamarine4")
+
+  #ITA female at age 65
+f_e65_ITA <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 65) %>% 
+      filter(sex == "female") %>% 
+      filter(code == "ITA") %>% 
+      filter(year >= 1979)
+
+# Now we create the change in ex variable as that's the response variable in the model
+  f_e65_ITA <- f_e65_ITA %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+   geom_line() +
+   geom_hline(yintercept = 0) + 
+   stat_smooth(colour = "lightpink3")
+
+  #ITA male at aged 65
+m_e65_ITA <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 65) %>% 
+      filter(sex == "male") %>% 
+      filter(code == "ITA") %>% 
+      filter(year >= 1979)
+
+ m_e65_ITA <-  m_e65_ITA %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+   geom_line() +
+   geom_hline(yintercept = 0) + 
+   #Trying to add legend
+   #stat_smooth(aes(colour = "aquamarine4"), show.legend = TRUE, labels = "Smoothed")
+  stat_smooth(colour = "aquamarine4")
+
+  #add labels
+f_e0_ITA + labs(y = "Annual change in e0", title = "Italy", subtitle = "Life expectancy changes at birth for females", caption = "Source: Human Mortality Database")
+ m_e0_ITA + labs(y = "Annual change in e0", title = "Italy", subtitle = "Life expectancy changes at birth for males", caption = "Source: Human Mortality Database")
+ f_e65_ITA + labs(y = "Annual change in e65", title = "Italy", subtitle = "Life expectancy changes at aged 65 for females", caption = "Source: Human Mortality Database")
+ m_e65_ITA + labs(y = "Annual change in e65", title = "Italy", subtitle = "Life expectancy changes at aged 65 for males", caption = "Source: Human Mortality Database")
+```
+
+### Germany: Synthetic
+
+```{r, echo = FALSE, warning=FALSE, message = FALSE}
+#DEUTS female at birth
+f_e0_DEUTS <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 0) %>% 
+      filter(sex == "female") %>% 
+      filter(code == "DEUTSYNTH") %>% 
+      filter(year >= 1979)
+
+# Now we create the change in ex variable as that's the response variable in the model
+  f_e0_DEUTS <- f_e0_DEUTS %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+   geom_line() +
+   geom_hline(yintercept = 0) + 
+   stat_smooth(colour = "lightpink3")
+
+  #DEUTS male at birth
+m_e0_DEUTS <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 0) %>% 
+      filter(sex == "male") %>% 
+      filter(code == "DEUTSYNTH") %>% 
+      filter(year >= 1979)
+
+  m_e0_DEUTS <- m_e0_DEUTS %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+   geom_line() +
+   geom_hline(yintercept = 0) + 
+   stat_smooth(colour = "aquamarine4")
+
+  #DEUTS female at age 65
+f_e65_DEUTS <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 65) %>% 
+      filter(sex == "female") %>% 
+      filter(code == "DEUTSYNTH") %>% 
+      filter(year >= 1979)
+
+# Now we create the change in ex variable as that's the response variable in the model
+  f_e65_DEUTS <- f_e65_DEUTS %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+   geom_line() +
+   geom_hline(yintercept = 0) + 
+   stat_smooth(colour = "lightpink3")
+
+  #DEUTS male at aged 65
+m_e65_DEUTS <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 65) %>% 
+      filter(sex == "male") %>% 
+      filter(code == "DEUTSYNTH") %>% 
+      filter(year >= 1979)
+
+ m_e65_DEUTS <-  m_e65_DEUTS %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+   geom_line() +
+   geom_hline(yintercept = 0) + 
+   stat_smooth(colour = "aquamarine4", show.legend = TRUE)
+
+  #add labels
+f_e0_DEUTS + labs(y = "Annual change in e0", title = "Germany: Synthetic", subtitle = "Life expectancy changes at birth for females", caption = "Source: Human Mortality Database")
+ m_e0_DEUTS + labs(y = "Annual change in e0", title = "Germany: Synthetic", subtitle = "Life expectancy changes at birth for males", caption = "Source: Human Mortality Database")
+ f_e65_DEUTS + labs(y = "Annual change in e65", title = "Germany: Synthetic", subtitle = "Life expectancy changes at aged 65 for females", caption = "Source: Human Mortality Database")
+ m_e65_DEUTS + labs(y = "Annual change in e65", title = "Germany: Synthetic", subtitle = "Life expectancy changes at aged 65 for males", caption = "Source: Human Mortality Database")
+```
+
+### Germany: East
+
+```{r, echo = FALSE, warning=FALSE, message = FALSE}
+#DEUTE female at birth
+f_e0_DEUTE <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 0) %>% 
+      filter(sex == "female") %>% 
+      filter(code == "DEUTE") %>% 
+      filter(year >= 1979)
+
+# Now we create the change in ex variable as that's the response variable in the model
+  f_e0_DEUTE <- f_e0_DEUTE %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+   geom_line() +
+   geom_hline(yintercept = 0) + 
+   stat_smooth(colour = "lightpink3")
+
+  #DEUTE male at birth
+m_e0_DEUTE <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 0) %>% 
+      filter(sex == "male") %>% 
+      filter(code == "DEUTE") %>% 
+      filter(year >= 1979)
+
+  m_e0_DEUTE <- m_e0_DEUTE %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+   geom_line() +
+   geom_hline(yintercept = 0) + 
+   stat_smooth(colour = "aquamarine4")
+
+  #DEUTE female at age 65
+f_e65_DEUTE <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 65) %>% 
+      filter(sex == "female") %>% 
+      filter(code == "DEUTE") %>% 
+      filter(year >= 1979)
+
+# Now we create the change in ex variable as that's the response variable in the model
+  f_e65_DEUTE <- f_e65_DEUTE %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+   geom_line() +
+   geom_hline(yintercept = 0) + 
+   stat_smooth(colour = "lightpink3")
+
+  #DEUTE male at aged 65
+m_e65_DEUTE <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 65) %>% 
+      filter(sex == "male") %>% 
+      filter(code == "DEUTE") %>% 
+      filter(year >= 1979)
+
+ m_e65_DEUTE <-  m_e65_DEUTE %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+   geom_line() +
+   geom_hline(yintercept = 0) + 
+   stat_smooth(colour = "aquamarine4", show.legend = TRUE)
+
+  #add labels
+f_e0_DEUTE + labs(y = "Annual change in e0", title = "Germany: East", subtitle = "Life expectancy changes at birth for females", caption = "Source: Human Mortality Database")
+ m_e0_DEUTE + labs(y = "Annual change in e0", title = "Germany: East", subtitle = "Life expectancy changes at birth for males", caption = "Source: Human Mortality Database")
+ f_e65_DEUTE + labs(y = "Annual change in e65", title = "Germany: East", subtitle = "Life expectancy changes at aged 65 for females", caption = "Source: Human Mortality Database")
+ m_e65_DEUTE + labs(y = "Annual change in e65", title = "Germany: East", subtitle = "Life expectancy changes at aged 65 for males", caption = "Source: Human Mortality Database")
+```
+
+### Germany: West
+
+```{r, echo = FALSE, warning=FALSE, message = FALSE}
+#DEUTW female at birth
+f_e0_DEUTW <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 0) %>% 
+      filter(sex == "female") %>% 
+      filter(code == "DEUTW") %>% 
+      filter(year >= 1979)
+
+# Now we create the change in ex variable as that's the response variable in the model
+  f_e0_DEUTW <- f_e0_DEUTW %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+   geom_line() +
+   geom_hline(yintercept = 0) + 
+   stat_smooth(colour = "lightpink3")
+
+  #DEUTW male at birth
+m_e0_DEUTW <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 0) %>% 
+      filter(sex == "male") %>% 
+      filter(code == "DEUTW") %>% 
+      filter(year >= 1979)
+
+  m_e0_DEUTW <- m_e0_DEUTW %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+   geom_line() +
+   geom_hline(yintercept = 0) + 
+   stat_smooth(colour = "aquamarine4")
+
+  #DEUTW female at age 65
+f_e65_DEUTW <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 65) %>% 
+      filter(sex == "female") %>% 
+      filter(code == "DEUTW") %>% 
+      filter(year >= 1979)
+
+# Now we create the change in ex variable as that's the response variable in the model
+  f_e65_DEUTW <- f_e65_DEUTW %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+   geom_line() +
+   geom_hline(yintercept = 0) + 
+   stat_smooth(colour = "lightpink3")
+
+  #DEUTW male at aged 65
+m_e65_DEUTW <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 65) %>% 
+      filter(sex == "male") %>% 
+      filter(code == "DEUTW") %>% 
+      filter(year >= 1979)
+
+ m_e65_DEUTW <-  m_e65_DEUTW %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+   geom_line() +
+   geom_hline(yintercept = 0) + 
+   stat_smooth(colour = "aquamarine4", show.legend = TRUE)
+
+  #add labels
+f_e0_DEUTW + labs(y = "Annual change in e0", title = "Germany: West", subtitle = "Life expectancy changes at birth for females", caption = "Source: Human Mortality Database")
+ m_e0_DEUTW + labs(y = "Annual change in e0", title = "Germany: West", subtitle = "Life expectancy changes at birth for males", caption = "Source: Human Mortality Database")
+ f_e65_DEUTW + labs(y = "Annual change in e65", title = "Germany: West", subtitle = "Life expectancy changes at aged 65 for females", caption = "Source: Human Mortality Database")
+ m_e65_DEUTW + labs(y = "Annual change in e65", title = "Germany: West", subtitle = "Life expectancy changes at aged 65 for males", caption = "Source: Human Mortality Database")
 ```

### 2022-03-10 draft_graphs.Rmd

@@
 
-# Now we create the change in ex variable as that's the response variable in the model
+# Now we create the change in ex variable 
   f_e65_ENW <- f_e65_ENW %>% 
@@
 
+#   hmd_ex_selected_countries_with_synth %>% 
+#       filter(sex == "male") %>% 
+#       filter(code %in% c("GBRTENW", "GBR_SCO")) %>% 
+#     group_by(code, x) %>% arrange(year) %>% 
+#       mutate(delta_ex = ex - lag(ex)) %>%
+#     ungroup %>% 
+#     filter(year >= 1979) %>% 
+#     ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+#    geom_line() +
+#    geom_hline(yintercept = 0) + 
+#    stat_smooth(colour = "aquamarine4")+
+# facet_grid(x ~ code)
+
+  ex_ENW <- hmd_ex_selected_countries_with_synth %>%
+      filter(sex %in% c("female", "male")) %>%
+      filter(code == "GBRTENW") %>%
+    group_by(sex, x) %>% arrange(year) %>%
+      mutate(delta_ex = ex - lag(ex)) %>%
+    ungroup %>%
+    filter(year >= 1979) %>%
+    ggplot(aes(x = year, y = delta_ex)) + geom_point() +
+   geom_line() +
+   geom_hline(yintercept = 0) +
+    stat_smooth(mapping = aes(colour = sex)) +
+       facet_grid(x ~ sex)
+
+  ex_ENW + labs(x = "Year", y = "Annual change in life expectancy", title = "England and Wales", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
+
 ```
@@
  m_e65_SCO + labs(y = "Annual change in e65", title = "Scotland", subtitle = "Life expectancy changes at aged 65 for males", caption = "Source: Human Mortality Database")
+
+ ex_SCO <- hmd_ex_selected_countries_with_synth %>%
+      filter(sex %in% c("female", "male")) %>%
+      filter(code == "GBR_SCO") %>%
+    group_by(sex, x) %>% arrange(year) %>%
+      mutate(delta_ex = ex - lag(ex)) %>%
+    ungroup %>%
+    filter(year >= 1979) %>%
+    ggplot(aes(x = year, y = delta_ex)) + geom_point() +
+   geom_line() +
+   geom_hline(yintercept = 0) +
+    stat_smooth(mapping = aes(colour = sex)) +
+       facet_grid(x ~ sex)
+
+  ex_ENW + labs(x = "Year", y = "Annual change in life expectancy", title = "Scotland", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
 ```
@@
  m_e65_FRA + labs(y = "Annual change in e65", title = "France", subtitle = "Life expectancy changes at aged 65 for males", caption = "Source: Human Mortality Database")
+
+ ex_FRA <- hmd_ex_selected_countries_with_synth %>%
+      filter(sex %in% c("female", "male")) %>%
+      filter(code == "FRATNP") %>%
+    group_by(sex, x) %>% arrange(year) %>%
+      mutate(delta_ex = ex - lag(ex)) %>%
+    ungroup %>%
+    filter(year >= 1979) %>%
+    ggplot(aes(x = year, y = delta_ex)) + geom_point() +
+   geom_line() +
+   geom_hline(yintercept = 0) +
+    stat_smooth(mapping = aes(colour = sex)) +
+       facet_grid(x ~ sex)
+
+  ex_ENW + labs(x = "Year", y = "Annual change in life expectancy", title = "France", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
 ```
@@
  m_e65_ESP + labs(y = "Annual change in e65", title = "Spain", subtitle = "Life expectancy changes at aged 65 for males", caption = "Source: Human Mortality Database")
+
+ ex_ESP <- hmd_ex_selected_countries_with_synth %>%
+      filter(sex %in% c("female", "male")) %>%
+      filter(code == "ESP") %>%
+    group_by(sex, x) %>% arrange(year) %>%
+      mutate(delta_ex = ex - lag(ex)) %>%
+    ungroup %>%
+    filter(year >= 1979) %>%
+    ggplot(aes(x = year, y = delta_ex)) + geom_point() +
+   geom_line() +
+   geom_hline(yintercept = 0) +
+    stat_smooth(mapping = aes(colour = sex)) +
+       facet_grid(x ~ sex)
+
+  ex_ENW + labs(x = "Year", y = "Annual change in life expectancy", title = "Spain", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
 ```
@@
  m_e65_ITA + labs(y = "Annual change in e65", title = "Italy", subtitle = "Life expectancy changes at aged 65 for males", caption = "Source: Human Mortality Database")
+
+ ex_ITA <- hmd_ex_selected_countries_with_synth %>%
+      filter(sex %in% c("female", "male")) %>%
+      filter(code == "ITA") %>%
+    group_by(sex, x) %>% arrange(year) %>%
+      mutate(delta_ex = ex - lag(ex)) %>%
+    ungroup %>%
+    filter(year >= 1979) %>%
+    ggplot(aes(x = year, y = delta_ex)) + geom_point() +
+   geom_line() +
+   geom_hline(yintercept = 0) +
+    stat_smooth(mapping = aes(colour = sex)) +
+       facet_grid(x ~ sex)
+
+  ex_ENW + labs(x = "Year", y = "Annual change in life expectancy", title = "Italy", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
 ```
@@
  m_e65_DEUTS + labs(y = "Annual change in e65", title = "Germany: Synthetic", subtitle = "Life expectancy changes at aged 65 for males", caption = "Source: Human Mortality Database")
+
+ #Grid
+ ex_DEUTSYNTH <- hmd_ex_selected_countries_with_synth %>%
+      filter(sex %in% c("female", "male")) %>%
+      filter(code == "DEUTHSYNTH") %>%
+    group_by(sex, x) %>% arrange(year) %>%
+      mutate(delta_ex = ex - lag(ex)) %>%
+    ungroup %>%
+    filter(year >= 1979) %>%
+    ggplot(aes(x = year, y = delta_ex)) + geom_point() +
+   geom_line() +
+   geom_hline(yintercept = 0) +
+    stat_smooth(mapping = aes(colour = sex)) +
+       facet_grid(x ~ sex)
+
+  ex_ENW + labs(x = "Year", y = "Annual change in life expectancy", title = "Germany: Synthetic", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
 ```
@@
  m_e65_DEUTE + labs(y = "Annual change in e65", title = "Germany: East", subtitle = "Life expectancy changes at aged 65 for males", caption = "Source: Human Mortality Database")
+
+ ex_DEUTE <- hmd_ex_selected_countries_with_synth %>%
+      filter(sex %in% c("female", "male")) %>%
+      filter(code == "DEUTE") %>%
+    group_by(sex, x) %>% arrange(year) %>%
+      mutate(delta_ex = ex - lag(ex)) %>%
+    ungroup %>%
+    filter(year >= 1979) %>%
+    ggplot(aes(x = year, y = delta_ex)) + geom_point() +
+   geom_line() +
+   geom_hline(yintercept = 0) +
+    stat_smooth(mapping = aes(colour = sex)) +
+       facet_grid(x ~ sex)
+
+  ex_ENW + labs(x = "Year", y = "Annual change in life expectancy", title = "Germany: East", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
 ```
@@
  m_e65_DEUTW + labs(y = "Annual change in e65", title = "Germany: West", subtitle = "Life expectancy changes at aged 65 for males", caption = "Source: Human Mortality Database")
+
+ ex_DEUTW <- hmd_ex_selected_countries_with_synth %>%
+      filter(sex %in% c("female", "male")) %>%
+      filter(code == "DEUTW") %>%
+    group_by(sex, x) %>% arrange(year) %>%
+      mutate(delta_ex = ex - lag(ex)) %>%
+    ungroup %>%
+    filter(year >= 1979) %>%
+    ggplot(aes(x = year, y = delta_ex)) + geom_point() +
+   geom_line() +
+   geom_hline(yintercept = 0) +
+    stat_smooth(mapping = aes(colour = sex)) +
+       facet_grid(x ~ sex)
+
+  ex_ENW + labs(x = "Year", y = "Annual change in life expectancy", title = "Germany: West", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
 ```

### 2022-03-12 draft_working_paper.Rmd
outstanding commits showing - so ticking stage and pushing!

--- 2022-03-10 draft_graphs.Rmd
@@
 date: "22/02/2022"
-output:
-  html_document: default
-  word_document: default
+output: word_document
 ---
@@
 # load packages
-
 pacman::p_load(knitr, here, tidyverse, flextable) # using flextable instead of gt. It seems to be better 
 # developed, more commonly used, and has more clearly labelled function names.
-
 # load data 
-
 hmd_lt <- read_rds(here("data", "lifetables.rds"))
-
 ```
@@
 ```{r, echo = FALSE}
-
 countries_of_interest <- c(
@@
 )
-
 hmd_ex_selected_countries <- 
@@
     rename(x = Age, year = Year)
-
 hmd_ex_selected_countries
-
 # Code for producing synthetic Germany
-
 source(here("R", "make_synthetic_germany_functions.R"))
-
 series_east_f0 <- hmd_ex_selected_countries %>% 
@@
   pull(ex)
-
 series_west_f0 <- hmd_ex_selected_countries %>% 
@@
   pull(ex)
-
 tmp_df_f0 <- tibble(
@@
 )
-
 ###
@@
   pull(ex)
-
 series_west_f65 <- hmd_ex_selected_countries %>% 
@@
   pull(ex)
-
 tmp_df_f65 <- tibble(
@@
 ##
-
 series_east_m0 <- hmd_ex_selected_countries %>% 
@@
   pull(ex)
-
 series_west_m0 <- hmd_ex_selected_countries %>% 
@@
   pull(ex)
-
 tmp_df_m0 <- tibble(
@@
 )
-
 ###
-
 series_east_m65 <- hmd_ex_selected_countries %>% 
@@
   pull(ex)
-
 series_west_m65 <- hmd_ex_selected_countries %>% 
@@
   pull(ex)
-
 tmp_df_m65 <- tibble(
@@
 )
-
 hmd_ex_selected_countries_with_synth <- bind_rows(
@@
 )
-
 # Labels for codes 
-
 country_code_lookup <- 
@@
   )
-
 ```
@@
 ```{r, echo = FALSE, fig.cap = "Life Expectancy at birth"}
-
 hmd_ex_selected_countries_with_synth %>% 
@@
   )
-
-```
-
-- [ ] Some insightful comment about what's going on.
+```
 
@@
 ```{r, echo = FALSE, warning = FALSE, message = FALSE}
-
 hmd_ex_selected_countries_with_synth %>% 
@@
   ggplot(aes(estimate, fct_reorder(display_label, estimate))) + 
-  geom_point() +
+  geom_point(aes(colour = x), show.legend = FALSE) +
   expand_limits( x = 0) + 
@@
 
-### Average improvements by decade
-
-- [ ] Add further introduction 
-- [ ] Add statistical significance (though maybe this is a different analysis)
-
-The table below shows average improvements by decade for England and Wales (standard error).
-
-```{r, echo=FALSE, tab.cap = "Source: HMD"}
-
-hmd_ex_selected_countries_with_synth %>% 
-    filter(year >= 1979) %>% 
-    group_by(code, x, sex) %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex)) %>%
-    mutate(decade = year - (year %% 10)) %>% 
-    filter(!is.na(delta_ex)) %>% 
-  group_by(code, x, sex, decade) %>% 
-  summarise(
-    mean_delta_ex = mean(delta_ex),
-    sd_delta_ex   = sd(delta_ex)
-    ) %>% 
-  ungroup() %>% 
-  mutate(cell_contents = glue::glue("{round(mean_delta_ex, 3)} ({round(sd_delta_ex, 3)})")
-  ) %>% 
-  filter(code == "GBRTENW") %>% 
-  filter(x == 0) %>% 
-  select(sex, decade, cell_contents) %>% 
-  spread(decade, cell_contents) %>% 
-#  write.csv("clipboard")
-  flextable() %>%
-  # Want to make the header row bold
-  vline(j = 1, part = "all") %>%
-  set_caption(caption = "Average improvement in life expectancy by decade")
-
-```
-
-### Are the differences by decade statistically significant?
-
-Are there significant differences in average rates of change in $e_x$ by decade?
-Another way of expressing this question is to ask whether knowing the decade in which an observation has been observed is informative as to what such values will be.
-This rephrased question was addressed by, for each starting age (0 or 65 years), sex and country, fitting two competing model specifications - $M_0$ and $M_A$ - and running an F test on the residuals of these two models.
-The Null model specification, $M_0$, regresses $\delta e_x$ (observed annual changes in $e_x$) against a single intercept term $\alpha$; put another way: $M_0$ assumes that all observations are drawn from a Normal distribution with a mean value of $\alpha$, and that this is consistent across all observed decades.
-The alternative model specification, $M_A$, also includes the $\alpha$ term, but also includes dummy variables to indicate which decade the realised value of $\delta e_0$ corresponds to.
-(Because one of the decades needs to be the reference category against which other decades are compared, there are one fewer decadal dummy variables than decades in the dataset for a given country, and the $\alpha$ parameter in $M_A$ estimates the mean change in $\delta e_0$ in the first decade, and so cannot be compared directly with the $\alpha$ parameter in $M_0$.) A P-value from the F test comparing $M_0$ and $M_A$ of less than 0.05 is taken to indicate that $M_A$ should be preferred to $M_0$, and so that there are systemic differences in average rates of improvement by decade for a given country, sex, and starting age.
-
-We found the only significant change decade-on-decades was for East Germany for females at birth and at aged 65 years.
-This is consistent with existing literature in this area, showing the convergence of life expectancy for females in East and West Germany following reunification.
-
-```{r, cache=TRUE}
-
-dta_for_decade_comparison <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 0) %>% 
-      filter(sex == "female") %>% 
-      filter(code == "GBRTENW") %>% 
-      filter(year >= 1979) %>% 
-      # ggplot(aes(x = year, y = ex)) +
-      #   geom_line()
-      arrange(year) %>% 
-      mutate(delta_ex = ex - lag(ex)) %>% 
-    #   ggplot(aes(x = year, y = delta_ex)) + 
-    # geom_point() + geom_line() +
-    # geom_hline(yintercept = 0)
-      mutate(
-        decade = year - (year %% 10)
-      ) %>% 
-  filter(!is.na(delta_ex))
-
-model_null <- 
-  lm(delta_ex ~ 1, data = dta_for_decade_comparison)
-
-model_alt <- 
-  lm(delta_ex ~ factor(decade), data= dta_for_decade_comparison)
-
-summary(model_null)
-
-summary(model_alt)
-
-anova(model_null, model_alt)
-
-# to get the value of interest do tmp[2,6]
-
-# Generalising
-
-tidy_dataset <- function(df){
-  df %>% 
-      filter(year >= 1979) %>% 
-      arrange(year) %>% 
-      mutate(delta_ex = ex - lag(ex)) %>% 
-      mutate(
-        decade = year - (year %% 10)
-      ) %>% 
-  filter(!is.na(delta_ex))
-}
-
-make_null_model <- function(df){
-  lm(delta_ex ~ 1, data = df)
-}
-
-all_null_alt_models <-
-  hmd_ex_selected_countries_with_synth %>%
-    group_by(x, sex, code) %>% 
-    nest() %>% 
-    mutate(
-      prep_data = map(data, tidy_dataset) 
-    ) %>% 
-    mutate(
-      model_null = map(prep_data, ~lm(delta_ex ~ 1, data = .)),
-      model_alt  = map(prep_data, ~lm(delta_ex ~ factor(decade), data = .))
-  #    model_null = map(prep_data, make_null_model)
-    ) %>% 
-    mutate(
-      anova_results = map2(model_null, model_alt, anova)
-    ) %>% 
-    mutate(
-      anova_pval = map_dbl(anova_results, ~.[2,6])
-    )
-
-all_null_alt_models %>% 
-  filter(anova_pval < 0.05) -> tmp
-
-summary(tmp$model_alt[[1]])
-
-summary(tmp$model_alt[[2]])
-
-  # build null model
-  # build alt model
-  # do anova
-  # get p value from anova
-
-```
-
-### Joinpoint/Segmented regression approach
-
-Though there may be systemic differences in average changes in life expectancy over time, such differences may not be patterned by decade, and so the above model specification may not have been appropriate to detect any such changes.
-Instead of looking for systemic differences by decade, we can instead look at whether there are any systemic and detectable 'breaks' in the trends over time.
-If our data were life expectancy over time, this could be achieved through joinpoint analysis, in which the slope is allowed to change over time.
-As instead are data are annual changes in life expectancy, we are interested in detecting changes in intercept over time.
-If we define our Null (no change) model as follows:
-
-$$
-M_0 : y(t) = \alpha + e
-$$
-Where $e \backsim N(0, \sigma^2)$. i.e. all observations $y$ are draws from $N(\alpha, \sigma^2)$. The alternative model specification against which this is compared is then:
-
-$$
-M_A : y(t) = \alpha_0 + \beta_1 T(t) + e 
-$$
-Where $T(t)$ is defined as $0$ where $t \lt \tau$ and $1$ where $t \geq \tau$. ($\alpha_0$ is used in $M_A$ to indicate that this parameter is not directly comparable with $\alpha$ in $M_0$).
-
-This specification shows that there is an additional parameter $\tau$ to be determined in order to produce $M_A$.
-This $\tau$ parameter is the breakpoint in the series.
-The selection of $\tau$ will be determined by numeric optimisation, by selecting the value of $\tau$ that minimised AIC.
-
-Once the best possible $M_A$, using the best value of $\tau$, is selected, $M_A$ will be compared with $M_0$ using an F-test, allowing us to determine whether there is convincing evidence of any specific break in the data.
-
-We will start with a single country, starting age, and sex, and with manual selection of $\tau$, then generalise and automate further.
-
 ## Graphs of annual change in life expectancy
@@
 
-NB: those with statistically significant 'breakpoints' in annual change in life expectancy are:
-- Germany (need to re-download from GIT)
+NB: those with statistically significant 'breakpoints' in annual change in life expectancy are listed in the table above.
 
@@
 
-  ex_ENW + labs(x = "Year", y = "Annual change in life expectancy", title = "Scotland", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
+  ex_SCO + labs(x = "Year", y = "Annual change in life expectancy", title = "Scotland", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
 ```
@@
 
-  ex_ENW + labs(x = "Year", y = "Annual change in life expectancy", title = "France", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
+  ex_FRA + labs(x = "Year", y = "Annual change in life expectancy", title = "France", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
 ```
@@
 
-  ex_ENW + labs(x = "Year", y = "Annual change in life expectancy", title = "Spain", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
+  ex_ESP + labs(x = "Year", y = "Annual change in life expectancy", title = "Spain", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
 ```
@@
 
-  ex_ENW + labs(x = "Year", y = "Annual change in life expectancy", title = "Italy", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
+  ex_ITA + labs(x = "Year", y = "Annual change in life expectancy", title = "Italy", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
 ```
@@
       filter(sex %in% c("female", "male")) %>%
-      filter(code == "DEUTHSYNTH") %>%
+      filter(code == "DEUTSYNTH") %>%
     group_by(sex, x) %>% arrange(year) %>%
@@
 
-  ex_ENW + labs(x = "Year", y = "Annual change in life expectancy", title = "Germany: Synthetic", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
+  ex_DEUTSYNTH + labs(x = "Year", y = "Annual change in life expectancy", title = "Germany: Synthetic", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
 ```
@@
 
-  ex_ENW + labs(x = "Year", y = "Annual change in life expectancy", title = "Germany: East", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
+  ex_DEUTE + labs(x = "Year", y = "Annual change in life expectancy", title = "Germany: East", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
 ```
@@
 
-  ex_ENW + labs(x = "Year", y = "Annual change in life expectancy", title = "Germany: West", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
+  ex_DEUTW + labs(x = "Year", y = "Annual change in life expectancy", title = "Germany: West", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
+```
+
+## Average improvements by decade
+
+- [ ] Add further introduction 
+- [ ] Add statistical significance (though maybe this is a different analysis)
+
+The table below shows average improvements by decade for England and Wales (standard error).
+
+```{r, echo=FALSE, tab.cap = "Source: HMD"}
+hmd_ex_selected_countries_with_synth %>% 
+    filter(year >= 1979) %>% 
+    group_by(code, x, sex) %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>%
+    mutate(decade = year - (year %% 10)) %>% 
+    filter(!is.na(delta_ex)) %>% 
+  group_by(code, x, sex, decade) %>% 
+  summarise(
+    mean_delta_ex = mean(delta_ex),
+    sd_delta_ex   = sd(delta_ex)
+    ) %>% 
+  ungroup() %>% 
+  mutate(cell_contents = glue::glue("{round(mean_delta_ex, 3)} ({round(sd_delta_ex, 3)})")
+  ) %>% 
+  filter(code == "GBRTENW") %>% 
+  filter(x == 0) %>% 
+  select(sex, decade, cell_contents) %>% 
+  spread(decade, cell_contents) %>% 
+#  write.csv("clipboard")
+  flextable() %>%
+  # Want to make the header row bold
+  vline(j = 1, part = "all") %>%
+  set_caption(caption = "Average improvement in life expectancy by decade")
+
+```
+
+### Are the differences by decade statistically significant?
+
+Are there significant differences in average rates of change in $e_x$ by decade?
+Another way of expressing this question is to ask whether knowing the decade in which an observation has been observed is informative as to what such values will be.
+This rephrased question was addressed by, for each starting age (0 or 65 years), sex and country, fitting two competing model specifications - $M_0$ and $M_A$ - and running an F test on the residuals of these two models.
+The Null model specification, $M_0$, regresses $\delta e_x$ (observed annual changes in $e_x$) against a single intercept term $\alpha$; put another way: $M_0$ assumes that all observations are drawn from a Normal distribution with a mean value of $\alpha$, and that this is consistent across all observed decades.
+The alternative model specification, $M_A$, also includes the $\alpha$ term, but also includes dummy variables to indicate which decade the realised value of $\delta e_0$ corresponds to.
+(Because one of the decades needs to be the reference category against which other decades are compared, there are one fewer decadal dummy variables than decades in the dataset for a given country, and the $\alpha$ parameter in $M_A$ estimates the mean change in $\delta e_0$ in the first decade, and so cannot be compared directly with the $\alpha$ parameter in $M_0$.) A P-value from the F test comparing $M_0$ and $M_A$ of less than 0.05 is taken to indicate that $M_A$ should be preferred to $M_0$, and so that there are systemic differences in average rates of improvement by decade for a given country, sex, and starting age.
+
+We found the only significant change decade-on-decades was for East Germany for females at birth and at aged 65 years.
+This is consistent with existing literature in this area, showing the convergence of life expectancy for females in East and West Germany following reunification.
+
+```{r, cache=TRUE}
+dta_for_decade_comparison <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 0) %>% 
+      filter(sex == "female") %>% 
+      filter(code == "GBRTENW") %>% 
+      filter(year >= 1979) %>% 
+      # ggplot(aes(x = year, y = ex)) +
+      #   geom_line()
+      arrange(year) %>% 
+      mutate(delta_ex = ex - lag(ex)) %>% 
+    #   ggplot(aes(x = year, y = delta_ex)) + 
+    # geom_point() + geom_line() +
+    # geom_hline(yintercept = 0)
+      mutate(
+        decade = year - (year %% 10)
+      ) %>% 
+  filter(!is.na(delta_ex))
+model_null <- 
+  lm(delta_ex ~ 1, data = dta_for_decade_comparison)
+model_alt <- 
+  lm(delta_ex ~ factor(decade), data= dta_for_decade_comparison)
+summary(model_null)
+summary(model_alt)
+anova(model_null, model_alt)
+# to get the value of interest do tmp[2,6]
+# Generalising 
+tidy_dataset <- function(df){
+  df %>% 
+      filter(year >= 1979) %>% 
+      arrange(year) %>% 
+      mutate(delta_ex = ex - lag(ex)) %>% 
+      mutate(
+        decade = year - (year %% 10)
+      ) %>% 
+  filter(!is.na(delta_ex))
+}
+make_null_model <- function(df){
+  lm(delta_ex ~ 1, data = df)
+}
+all_null_alt_models <-
+  hmd_ex_selected_countries_with_synth %>%
+    group_by(x, sex, code) %>% 
+    nest() %>% 
+    mutate(
+      prep_data = map(data, tidy_dataset) 
+    ) %>% 
+    mutate(
+      model_null = map(prep_data, ~lm(delta_ex ~ 1, data = .)),
+      model_alt  = map(prep_data, ~lm(delta_ex ~ factor(decade), data = .))
+  #    model_null = map(prep_data, make_null_model)
+    ) %>% 
+    mutate(
+      anova_results = map2(model_null, model_alt, anova)
+    ) %>% 
+    mutate(
+      anova_pval = map_dbl(anova_results, ~.[2,6])
+    ) 
+all_null_alt_models %>% 
+  filter(anova_pval < 0.05) -> tmp
+summary(tmp$model_alt[[1]])
+summary(tmp$model_alt[[2]])
+  # build null model
+  # build alt model
+  # do anova
+  # get p value from anova
+
+```
+
+### Joinpoint/Segmented regression approach
+
+Though there may be systemic differences in average changes in life expectancy over time, such differences may not be patterned by decade, and so the above model specification may not have been appropriate to detect any such changes.
+Instead of looking for systemic differences by decade, we can instead look at whether there are any systemic and detectable 'breaks' in the trends over time.
+If our data were life expectancy over time, this could be achieved through joinpoint analysis, in which the slope is allowed to change over time.
+As instead are data are annual changes in life expectancy, we are interested in detecting changes in intercept over time.
+If we define our Null (no change) model as follows:
+
+$$
+M_0 : y(t) = \alpha + e
+$$
+Where $e \backsim N(0, \sigma^2)$. i.e. all observations $y$ are draws from $N(\alpha, \sigma^2)$. The alternative model specification against which this is compared is then:
+
+$$
+M_A : y(t) = \alpha_0 + \beta_1 T(t) + e 
+$$
+Where $T(t)$ is defined as $0$ where $t \lt \tau$ and $1$ where $t \geq \tau$. ($\alpha_0$ is used in $M_A$ to indicate that this parameter is not directly comparable with $\alpha$ in $M_0$).
+
+This specification shows that there is an additional parameter $\tau$ to be determined in order to produce $M_A$.
+This $\tau$ parameter is the breakpoint in the series.
+The selection of $\tau$ will be determined by numeric optimisation, by selecting the value of $\tau$ that minimised AIC.
+
+Once the best possible $M_A$, using the best value of $\tau$, is selected, $M_A$ will be compared with $M_0$ using an F-test, allowing us to determine whether there is convincing evidence of any specific break in the data.
+
+We will start with a single country, starting age, and sex, and with manual selection of $\tau$, then generalise and automate further.
+
+```{r}
+# Let's start with a single example population. (There are many different population subgroups, hence the 
+# benefit of being able to automate this later.) We're going to start with England & Wales, females, life 
+# expectancy at birth (x == 0)
+example_df <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 0) %>% 
+      filter(sex == "female") %>% 
+      filter(code == "GBRTENW") %>% 
+      filter(year >= 1979)
+# Now we create the change in ex variable as that's the response variable in the model
+example_df <- 
+  example_df %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) # %>% 
+  # ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+  # geom_line() +
+  # geom_hline(yintercept = 0) + 
+  # stat_smooth()
+# The null model is really simple: just an intercept 
+model_null <- lm(delta_ex ~ 1, data = example_df)
+summary(model_null)
+# 
+# The code below shows how we can be even simpler, and force the regression model not to 
+# even have an intercept (i.e. have zero predictor variables)
+# model_nuller <- lm(delta_ex ~ 0, data = example_df)
+# summary(model_nuller)
+# And we can compare the intercept with the no intercept model as follows 
+# anova(model_nuller, model_null)
+# a statistically significant p value shows the intercept is definitely worth including in the 
+# model. Substantively that it's wrong to assume no improvement on average in ex over time since the 1980s
+# There are lots of possible alternative models, some examples of which are below:
+model_alt01 <- lm(delta_ex ~ year >= 1990, data = example_df)
+summary(model_alt01)
+model_alt02 <- lm(delta_ex ~ year >= 1995, data = example_df)
+summary(model_alt02)
+model_alt03 <- lm(delta_ex ~ year >= 1997, data = example_df)
+summary(model_alt03)
+model_alt04 <- lm(delta_ex ~ year >= 2008, data = example_df)
+summary(model_alt04)
+model_alt05 <- lm(delta_ex ~ year >= 2011, data = example_df)
+summary(model_alt05)
+# each of these models allows an adjustment to the horizontal line after a given year has been reached. 
+# In the examples above some possible given years are 1990, 1995, 1997, 2008 and 2011. 
+# Each of these is a candidate value for tau. But we've not looked at all possible/plausible values of this. 
+# Hence, it would be good to automate further to search for each possible tau and see how the fit and model 
+# results change. 
+# We'll start to automate by building a function which creates T_param given tau, and runs and returns the model
+# with this type of T_param
+# (If tau is 1990, then T_param is equivalent to model_alt01 above; if tau were 1995, then it's equivalent to model_alt02, and so on.
+The only difference is that the new coefficient created has the same name, T_param,
+# whereas in the above it would have a different name each time, because the coefficient is being 
+# created in situ, within the formula specification, rather than by pre-processing the dataframe)
+run_alt_with_given_tau <- function(tau, df){
+  df <- 
+    df %>% 
+      filter(!is.na(delta_ex)) %>% 
+      mutate(
+        T_param = ifelse(year < tau, FALSE, TRUE)
+      )
+
+  lm(delta_ex ~ T_param, data  = df)
+}
+# We can now run the above function, which creates a model, for a wide range of possible tau values
+# This is the sequence listed and named tau_candidate, which is every whole number from 
+# 1982 to 2015
+# For each of these models, we can then get a measure of model fit. We're going to use AIC, though 
+# could have used other measures. With AIC (and with BIC, and RMSE etc) better fit is indicated with 
+# a lower value
+grid_search_df <- tibble(
+  tau_candidate = 1982:2015 # column of candidate values
+) %>% 
+  mutate(
+    model_alt = map(tau_candidate, run_alt_with_given_tau, df = example_df) # run the model with the value of tau in tau_candidate
+  ) %>% 
+  mutate(
+    aic = map_dbl(model_alt, AIC)
+  )
+# Now we can see which proposed tau results in what fit value (AIC), initially just by graphing it.
+grid_search_df %>% 
+  ggplot(aes(tau_candidate, aic)) + 
+  geom_point()
+# So, in the above we found 2012 as tau minimised AIC. So, that's the best possible alternative model. 
+# Let's run and label it as such
+best_alt_model <- example_df %>% 
+  mutate( T_param = ifelse(year < 2012, FALSE, TRUE)) %>% 
+  lm(delta_ex ~ T_param, data = .)
+# Now we want to create the null model to compare it against (which just has an intercept term alone)
+null_model <- example_df %>% 
+  lm(delta_ex ~ 1, data = .)
+# And finally, we want to compare the best_alt_model against the null model using an F test
+anova(null_model, best_alt_model)
+# We are interested in the P value here. If it's statistically significant then the best_alt_model does better
+# (Is worth its additional complexity in terms of how much it makes the model fit) than the null model,
+# So... for females, from birth, in England & Wales, there's evidence that there WAS a significant change in 
+# average life expectancy improvement from 2012 onwards. (i.e. two years after austerity was introduced)
+# We can manually repeat the same exercise for some additional subsets of the data (population, gender, starting age combinations). The below shows the same for males in England & Wales, from birth.
+example_df <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 0) %>% 
+      filter(sex == "male") %>% 
+      filter(code == "GBRTENW") %>% 
+      filter(year >= 1979)
+example_df <- 
+  example_df %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex))
+grid_search_df <- tibble(
+  tau_candidate = 1982:2015
+) %>% 
+  mutate(
+    model_alt = map(tau_candidate, run_alt_with_given_tau, df = example_df)
+  ) %>% 
+  mutate(
+    aic = map_dbl(model_alt, AIC)
+  )
+grid_search_df %>% 
+  ggplot(aes(tau_candidate, aic)) + 
+  geom_point()
+best_alt_model <- example_df %>% 
+  mutate( T_param = ifelse(year < 2013, FALSE, TRUE)) %>% 
+  lm(delta_ex ~ T_param, data = .)
+null_model <- example_df %>% 
+  lm(delta_ex ~ 1, data = .)
+anova(null_model, best_alt_model)
+# This time 2013 is a slightly better candidate than 2012, hence 2013 being used in best_alt_model not 2012. 
+# Another example: France, males
+###
+example_df <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 0) %>% 
+      filter(sex == "male") %>% 
+      filter(code == "FRATNP") %>% 
+      filter(year >= 1979)
+example_df <- 
+  example_df %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex))
+grid_search_df <- tibble(
+  tau_candidate = 1982:2015
+) %>% 
+  mutate(
+    model_alt = map(tau_candidate, run_alt_with_given_tau, df = example_df)
+  ) %>% 
+  mutate(
+    aic = map_dbl(model_alt, AIC)
+  )
+grid_search_df %>% 
+  ggplot(aes(tau_candidate, aic)) + 
+  geom_point()
+best_alt_model <- example_df %>% 
+  mutate( T_param = ifelse(year < 2012, FALSE, TRUE)) %>% 
+  lm(delta_ex ~ T_param, data = .)
+summary(best_alt_model)
+null_model <- example_df %>% 
+  lm(delta_ex ~ 1, data = .)
+summary(null_model)
+anova(null_model, best_alt_model)
+# Here 2015 has the lowest value, but this might be an artefact of the data period covered. 
+# We can also see that 2012 is the next best candidate.
+## Generalising and automating. 
+# We've manually found the best break points, and compared the best alt with the standard null model 
+# for each of these. What we should do now is standardise the process so it 'works' automatically 
+# for each of the population subgroups. 
+# As with creating synthetic germany, the grid/line search approach shown above should work, 
+# but I'm going to try to use optim nonetheless for pedagogic and perfectionist(/masochistic) reasons
+# optim uses some starting conditions (and sometimes other conditions) to use an algorithm 
+# to find an input or group of inputs that give the best possible output from a function that takes the 
+# input/inputs. 
+# best can either mean the lowest or the highest achievable value. When trying to minimise errors
+# (as with model specifications) that usually means trying to get the minimum value. 
+# So, the input to the function will be the tau value. 
+# The tau value range should have some reasonable limits. These can be specified 
+# for some kinds of algorithm optim uses.
+# the limits should also be determined by the data. I don't think the tau should be 
+# set at the first or last observed year, as it doesn't make sense to speak of a 'trend' 
+# from a single observation. This thinking generalises to at least two observations too. 
+# So, that implies there should be a 'buffer' within which the optim function will look 
+# for possible tau values. This should be an adjustable parameter for passing for 
+# optim, but not the parameter for optim to determine. 
+# There's a special type of optim algorithm for bounded optimisation like this.
+# 'L-BFGS-B'. This can be selected using the method argument, and also 
+# requires the lower and upper arguments be specified, for the lower 
+# and upper values to search within respectively. 
+# so, we need to write a function a bit like the following
+get_best_tau_from_optim <- function(df, buffer = 2, what = c("tau", "optim")){
+  # We can put in some data validation checks too
+
+  what <- match.arg(what)
+
+  years <- df$year 
+  tau_lower <- min(years) + buffer # the lower bound 
+  tau_upper <- max(years) - buffer # the upper bound
+  tau_start <- (tau_upper + tau_lower) / 2 # start just in the middle 
+  # (This shouldn't matter if the algorithm is robust)
+
+  get_result_from_model <- function(par, df, what = c("AIC", "BIC", "model")){
+    # Again, can carry out more validation checks 
+    what <- match.arg(what)
+
+    tau <- par[["tau"]] # This is how parameters are packed up by optim
+    model <-
+      df %>% 
+        arrange(year) %>% 
+        mutate(delta_ex = ex - lag(ex)) %>% 
+        filter(!is.na(delta_ex)) %>% 
+        mutate(T_param = ifelse(year < tau, FALSE, TRUE)) %>% 
+        lm(delta_ex ~ T_param, data = .)
+    if (what == "model") {return(model)      }
+    if (what == "BIC")   {return(BIC(model)) }
+    if (what == "AIC")   {return(AIC(model)) }
+
+    # NULL should never be returned! If it has something's gone wrong!
+    NULL
+  }
+
+  optim_obj <- 
+    optim(
+      par = list(tau = tau_start), 
+      fn = get_result_from_model, 
+      method = "L-BFGS-B",
+      lower = tau_lower,
+      upper = tau_upper,
+      df = df
+    )
+  # Need to make the result dependent on the what argument
+
+  if (what == "tau") {return(optim_obj[["par"]])}
+  else if (what == "optim"){ return(optim_obj) }
+
+  # Again, the following should never be triggered
+  NULL
+}
+# Let's see if this now works in the functional programming context, using 
+# nest and tibble elements 
+hmd_ex_selected_countries_with_synth %>%
+  filter(year >= 1979) %>% 
+  group_by(code, x, sex) %>% 
+  nest() %>% 
+  mutate(
+    best_tau = map_dbl(data, get_best_tau_from_optim)
+  )
+# I think this is getting stuck on local optima, so instead maybe 
+# the more exhaustive search should be used after all! 
+# There are two alternative approaches: 
+# 1) Use an exhaustive grid search approach 
+# 2) Use an algorithm which is less likely to get stuck in local optima
+# The first approach is much more sensible, but the second might be a cool
+# stretch goal as well! 
+# let's use a heavily modified version of the above function to do the grid search
+get_best_tau_from_gridsearch <- function(df, buffer = 2, what = c("best", "all")){
+  # We can put in some data validation checks too
+
+  what <- match.arg(what)
+
+  years <- df$year 
+  tau_lower <- min(years) + buffer # the lower bound 
+  tau_upper <- max(years) - buffer # the upper bound
+  tau_searchrange <- tau_lower:tau_upper
+  get_result_from_model <- function(tau, df, what = c("AIC", "BIC", "model")){
+    # Again, can carry out more validation checks 
+    what <- match.arg(what)
+
+    model <-
+      df %>% 
+        arrange(year) %>% 
+        mutate(delta_ex = ex - lag(ex)) %>% 
+        filter(!is.na(delta_ex)) %>% 
+        mutate(T_param = ifelse(year < tau, FALSE, TRUE)) %>% 
+        lm(delta_ex ~ T_param, data = .)
+    if (what == "model") {return(model)      }
+    if (what == "BIC")   {return(BIC(model)) }
+    if (what == "AIC")   {return(AIC(model)) }
+
+    # NULL should never be returned! If it has something's gone wrong!
+    NULL
+  }
+
+  search_df <- tibble(
+    tau = tau_searchrange
+  ) %>% 
+    mutate(aic = map_dbl(tau, get_result_from_model, df = df))
+
+  if (what == "all") {return(search_df)}
+
+  if (what == "best") {
+    out <- 
+      search_df %>% 
+        filter(aic == min(aic)) %>% 
+        pull(tau)
+    return(out)
+  }
+  # Again, the following should never be triggered
+  NULL
+}
+compare_best_breakmodel_with_null <- 
+  hmd_ex_selected_countries_with_synth %>%
+    filter(year >= 1979) %>% 
+    group_by(code, x, sex) %>% 
+    nest() %>% 
+    mutate(
+      best_tau = map_dbl(data, get_best_tau_from_gridsearch)
+    ) %>% 
+    mutate(
+      model_null = map(
+        data, 
+        ~ .x %>% 
+          arrange(year) %>% 
+          mutate(delta_ex = ex - lag(ex)) %>% 
+          filter(!is.na(delta_ex)) %>% 
+          lm(delta_ex ~ 1, data = .)
+      )
+    ) %>% 
+    mutate(
+      model_alt = map2(
+        best_tau, data,
+      ~ .y %>% 
+        arrange(year) %>% 
+        mutate(delta_ex = ex - lag(ex)) %>% 
+        filter(!is.na(delta_ex)) %>% 
+        mutate(T_param = ifelse(year < .x, FALSE, TRUE)) %>% 
+        lm(delta_ex ~ T_param, data = .)
+      )
+    )
+tmp <- compare_best_breakmodel_with_null %>% 
+  mutate(
+    aic_alt  = map_dbl(model_alt, AIC),
+    aic_null = map_dbl(model_null, AIC)
+  ) %>% 
+  mutate(
+    alt_preferred = map2_lgl(aic_alt, aic_null, ~ifelse(.x < .y, TRUE, FALSE))
+  ) %>% 
+  mutate(
+    anova_results = map2(model_alt, model_null, anova)
+  ) %>% 
+  mutate(
+    ftest_pval = map_dbl(anova_results, ~.[2,6]),
+    stat_sig_preference = ftest_pval < 0.05
+  )
+
+# Now let's finish with the most unneccesarily complicated variant: using a different algorithm in 
+# optim which will allow local minima to be escaped. This is the simulated annealing method 
+# (method = "SANN") which is based on physics models. Unfortunately this is not a bounded algoritm,
+# i.e. does not take values within limits, so instead the parameter passed to it has to be 
+# converted and deconverted from a bounded range using a link function. 
+# This can be done with a modification of the sigmoid function, which shifts the transformed 
+# value from the range 0-1 to the range lower_tau-upper_tau. 
+# Let's build up this function as follows: 
+# sigmoid_shift <- function(x, lower = 0, upper = 1){
+#   (upper - lower) / (1 + exp(-x)) + lower
+# }
+# # 
+# curve(sigmoid_shift, -3, 3) # basic sigmoid 
+# 
+# sigmoid_shift2 <- function(x){
+#   sigmoid_shift(x = x, lower = 1981, upper = 2015)
+# }
+# 
+# curve(sigmoid_shift2, -3, 3)
+# So this will work as a transformation from an unbounded space (x) to a 
+# bounded space (f(x))
+# We'll also need to reverse the operation 
+# # After a bit of pen-and-paper algebra...
+# sigmoid_unshift <- function(z, lower = 0, upper = 1){
+#   -log((upper - lower) / (z - lower) -1)
+# }
+# 
+# Let's double check...
+#  
+# sigmoid_shift(-4, lower = 1982, upper = 2015) %>% 
+#   sigmoid_unshift(lower = 1982, upper = 2015)
+# 
+# 
+# 
+# get_best_tau_from_optim_sann <- function(df, buffer = 2, what = c("tau", "tau_z", "optim")){
+#   # We can put in some data validation checks too 
+#   sigmoid_shift <- function(x, lower = 0, upper = 1){
+#     (upper - lower) / (1 + exp(-x)) + lower
+#   }
+#  
+#   sigmoid_unshift <- function(z, lower = 0, upper = 1){
+#   -log((upper - lower) / (z - lower) -1)
+#   } 
+#   
+#   
+#   what <- match.arg(what)
+#   
+#   years <- df$year 
+#   tau_lower <- min(years) + buffer # the lower bound 
+#   tau_upper <- max(years) - buffer # the upper bound
+#   # tau_start <- (tau_upper + tau_lower) / 2 # start just in the middle 
+#   # (This shouldn't matter if the algorithm is robust) 
+#   
+#   get_result_from_model <- function(par, df, what = c("AIC", "BIC", "model")){
+#     # Again, can carry out more validation checks 
+#     what <- match.arg(what)
+#     
+#     tau_z <- par[["tau_z"]] # This is how parameters are packed up by optim
+#     tau   <- sigmoid_shift(tau_z, lower = tau_lower, upper = tau_upper)
+#     
+#     model <-
+#       df %>% 
+#         arrange(year) %>% 
+#         mutate(delta_ex = ex - lag(ex)) %>% 
+#         filter(!is.na(delta_ex)) %>% 
+#         mutate(T_param = ifelse(year < tau, FALSE, TRUE)) %>% 
+#         lm(delta_ex ~ T_param, data = .)
+# 
+#     if (what == "model") {return(model)      }
+#     if (what == "BIC")   {return(BIC(model)) }
+#     if (what == "AIC")   {return(AIC(model)) }
+#     
+#     # NULL should never be returned! If it has something's gone wrong!
+#     NULL
+#   }
+# 
+#   optim_obj <- 
+#     optim(
+#       par = list(tau_z = 0), 
+#       fn = get_result_from_model, 
+#       method = "SANN",
+#       df = df
+#     )
+# 
+#   # Need to make the result dependent on the what argument
+#   
+#   if (what == "tau") {return(sigmoid_shift(optim_obj[["par"]], lower = tau_lower, upper = tau_upper))}
+#   if (what == "tau_z") {return(optim_obj[["par"]])}  
+#   else if (what == "optim"){ return(optim_obj) }
+#   
+#   # Again, the following should never be triggered
+#   NULL
+# }
+# 
+# 
+# sann_estimates <- 
+#   hmd_ex_selected_countries_with_synth %>%
+#     filter(year >= 1979) %>% 
+#     group_by(code, x, sex) %>% 
+#     nest() %>% 
+#     mutate(
+#       best_tau = map_dbl(data, get_best_tau_from_optim_sann)
+#     )
+# 
+# # This takes ABSOLUTELY AGES, and really is the sledgehammer-to-crack
+# # -a-nut approach to use only as a last resort. #
+# # However if you ever do have a complex and non-finite fitness 
+# # surface to try to optimise, and it also requires searching 
+# # over a bounded area, then this is how you can do it! 
+```
+
+### Table of countries breakpoints with statistical significace
+
+Next, we display a table with those countries where the breakpoint shows as statistically significant (at p value <0.05) compared to the null, where best_tau is the year.
+
+```{r}
+stat_sig_table <- select(tmp, code, x, sex, best_tau, ftest_pval, stat_sig_preference) %>% arrange(ftest_pval)
+
+#arrange(stat_sig_table, ftest_pval)
+
+view(stat_sig_table)
+
+colnames(stat_sig_table) = c("Country", "Age", "Sex", "Best_Tau", "F Test P Value", "Statistically significant")
+
+  #filter(stat_sig_preference==TRUE) %>%
+
+#using flextable
+myft <- flextable(stat_sig_table)
+myft <- theme_vanilla(myft)
+
+# #myft <- flextable(stat_sig_table) %>%
+#   compose(j = "Statistically significant",
+#           value = as_paragraph(
+#             as_chunk("TRUE",
+#                      props = fp_text_default(color = "#C32900", bold = TRUE))))
+
+myft
+```
+
+```{r}
+example_df <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(x == 0) %>% 
+      filter(sex == "male") %>% 
+      filter(code == "FRATNP") %>% 
+      filter(year >= 1979)
+example_df <- 
+  example_df %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex))
+grid_search_df <- tibble(
+  tau_candidate = 1982:2015
+) %>% 
+  mutate(
+    model_alt = map(tau_candidate, run_alt_with_given_tau, df = example_df)
+  ) %>% 
+  mutate(
+    aic = map_dbl(model_alt, AIC)
+  )
+grid_search_df %>% 
+  ggplot(aes(tau_candidate, aic)) + 
+  geom_point()
 ```

### 2022-03-15 draft_working_paper.Rmd

@@
     filter(year >= 1979) %>%
-    ggplot(aes(x = year, y = delta_ex)) + geom_point() +
+    ggplot(aes(x = year, y = delta_ex)) + geom_point(size=0.3) +
    geom_line() +

### 2022-03-15 draft_paper.Rmd
local commit of draft_paper

@@
 ```
+
 Next, let's look closer at France, Scotland, and E&Ws life expectancy at birth for males.
+
 Jon: can we somehow include these 3 in the below to produce a graph? I tried adding into the first chunk but it does not run the later parts.

### 2022-03-23 draft_working_paper.Rmd
pack breakpoint code into functions

@@
 ```{r}
-# Let's start with a single example population. (There are many different population subgroups, hence the 
-# benefit of being able to automate this later.) We're going to start with England & Wales, females, life 
-# expectancy at birth (x == 0)
-example_df <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 0) %>% 
-      filter(sex == "female") %>% 
-      filter(code == "GBRTENW") %>% 
-      filter(year >= 1979)
-# Now we create the change in ex variable as that's the response variable in the model
-example_df <- 
-  example_df %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex)) # %>% 
-  # ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
-  # geom_line() +
-  # geom_hline(yintercept = 0) + 
-  # stat_smooth()
-# The null model is really simple: just an intercept 
-model_null <- lm(delta_ex ~ 1, data = example_df)
-summary(model_null)
-# 
-# The code below shows how we can be even simpler, and force the regression model not to 
-# even have an intercept (i.e. have zero predictor variables)
-# model_nuller <- lm(delta_ex ~ 0, data = example_df)
-# summary(model_nuller)
-# And we can compare the intercept with the no intercept model as follows 
-# anova(model_nuller, model_null)
-# a statistically significant p value shows the intercept is definitely worth including in the 
-# model. Substantively that it's wrong to assume no improvement on average in ex over time since the 1980s
-# There are lots of possible alternative models, some examples of which are below:
-model_alt01 <- lm(delta_ex ~ year >= 1990, data = example_df)
-summary(model_alt01)
-model_alt02 <- lm(delta_ex ~ year >= 1995, data = example_df)
-summary(model_alt02)
-model_alt03 <- lm(delta_ex ~ year >= 1997, data = example_df)
-summary(model_alt03)
-model_alt04 <- lm(delta_ex ~ year >= 2008, data = example_df)
-summary(model_alt04)
-model_alt05 <- lm(delta_ex ~ year >= 2011, data = example_df)
-summary(model_alt05)
-# each of these models allows an adjustment to the horizontal line after a given year has been reached. 
-# In the examples above some possible given years are 1990, 1995, 1997, 2008 and 2011. 
-# Each of these is a candidate value for tau. But we've not looked at all possible/plausible values of this. 
-# Hence, it would be good to automate further to search for each possible tau and see how the fit and model 
-# results change. 
-# We'll start to automate by building a function which creates T_param given tau, and runs and returns the model
-# with this type of T_param
-# (If tau is 1990, then T_param is equivalent to model_alt01 above; if tau were 1995, then it's equivalent to model_alt02, and so on.
-The only difference is that the new coefficient created has the same name, T_param,
-# whereas in the above it would have a different name each time, because the coefficient is being 
-# created in situ, within the formula specification, rather than by pre-processing the dataframe)
-run_alt_with_given_tau <- function(tau, df){
-  df <- 
-    df %>% 
-      filter(!is.na(delta_ex)) %>% 
-      mutate(
-        T_param = ifelse(year < tau, FALSE, TRUE)
-      )
-
-  lm(delta_ex ~ T_param, data  = df)
-}
 # We can now run the above function, which creates a model, for a wide range of possible tau values
@@
 # a lower value
-grid_search_df <- tibble(
-  tau_candidate = 1982:2015 # column of candidate values
-) %>% 
-  mutate(
-    model_alt = map(tau_candidate, run_alt_with_given_tau, df = example_df) # run the model with the value of tau in tau_candidate
-  ) %>% 
-  mutate(
-    aic = map_dbl(model_alt, AIC)
-  )
-# Now we can see which proposed tau results in what fit value (AIC), initially just by graphing it.
-grid_search_df %>% 
-  ggplot(aes(tau_candidate, aic)) + 
-  geom_point()
-# So, in the above we found 2012 as tau minimised AIC. So, that's the best possible alternative model. 
-# Let's run and label it as such
-best_alt_model <- example_df %>% 
-  mutate( T_param = ifelse(year < 2012, FALSE, TRUE)) %>% 
-  lm(delta_ex ~ T_param, data = .)
-# Now we want to create the null model to compare it against (which just has an intercept term alone)
-null_model <- example_df %>% 
-  lm(delta_ex ~ 1, data = .)
-# And finally, we want to compare the best_alt_model against the null model using an F test
-anova(null_model, best_alt_model)
-# We are interested in the P value here. If it's statistically significant then the best_alt_model does better
-# (Is worth its additional complexity in terms of how much it makes the model fit) than the null model,
-# So... for females, from birth, in England & Wales, there's evidence that there WAS a significant change in 
-# average life expectancy improvement from 2012 onwards. (i.e. two years after austerity was introduced)
-# We can manually repeat the same exercise for some additional subsets of the data (population, gender, starting age combinations). The below shows the same for males in England & Wales, from birth.
-example_df <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 0) %>% 
-      filter(sex == "male") %>% 
-      filter(code == "GBRTENW") %>% 
-      filter(year >= 1979)
-example_df <- 
-  example_df %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex))
-grid_search_df <- tibble(
-  tau_candidate = 1982:2015
-) %>% 
-  mutate(
-    model_alt = map(tau_candidate, run_alt_with_given_tau, df = example_df)
-  ) %>% 
-  mutate(
-    aic = map_dbl(model_alt, AIC)
-  )
-grid_search_df %>% 
-  ggplot(aes(tau_candidate, aic)) + 
-  geom_point()
-best_alt_model <- example_df %>% 
-  mutate( T_param = ifelse(year < 2013, FALSE, TRUE)) %>% 
-  lm(delta_ex ~ T_param, data = .)
-null_model <- example_df %>% 
-  lm(delta_ex ~ 1, data = .)
-anova(null_model, best_alt_model)
-# This time 2013 is a slightly better candidate than 2012, hence 2013 being used in best_alt_model not 2012. 
-# Another example: France, males
-###
-example_df <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 0) %>% 
-      filter(sex == "male") %>% 
-      filter(code == "FRATNP") %>% 
-      filter(year >= 1979)
-example_df <- 
-  example_df %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex))
-grid_search_df <- tibble(
-  tau_candidate = 1982:2015
-) %>% 
-  mutate(
-    model_alt = map(tau_candidate, run_alt_with_given_tau, df = example_df)
-  ) %>% 
-  mutate(
-    aic = map_dbl(model_alt, AIC)
-  )
-grid_search_df %>% 
-  ggplot(aes(tau_candidate, aic)) + 
-  geom_point()
-best_alt_model <- example_df %>% 
-  mutate( T_param = ifelse(year < 2012, FALSE, TRUE)) %>% 
-  lm(delta_ex ~ T_param, data = .)
-summary(best_alt_model)
-null_model <- example_df %>% 
-  lm(delta_ex ~ 1, data = .)
-summary(null_model)
-anova(null_model, best_alt_model)
-# Here 2015 has the lowest value, but this might be an artefact of the data period covered. 
-# We can also see that 2012 is the next best candidate.
-## Generalising and automating. 
-# We've manually found the best break points, and compared the best alt with the standard null model 
-# for each of these. What we should do now is standardise the process so it 'works' automatically 
-# for each of the population subgroups. 
-# As with creating synthetic germany, the grid/line search approach shown above should work, 
-# but I'm going to try to use optim nonetheless for pedagogic and perfectionist(/masochistic) reasons
-# optim uses some starting conditions (and sometimes other conditions) to use an algorithm 
-# to find an input or group of inputs that give the best possible output from a function that takes the 
-# input/inputs. 
-# best can either mean the lowest or the highest achievable value. When trying to minimise errors
-# (as with model specifications) that usually means trying to get the minimum value. 
-# So, the input to the function will be the tau value. 
-# The tau value range should have some reasonable limits. These can be specified 
-# for some kinds of algorithm optim uses.
-# the limits should also be determined by the data. I don't think the tau should be 
-# set at the first or last observed year, as it doesn't make sense to speak of a 'trend' 
-# from a single observation. This thinking generalises to at least two observations too. 
-# So, that implies there should be a 'buffer' within which the optim function will look 
-# for possible tau values. This should be an adjustable parameter for passing for 
-# optim, but not the parameter for optim to determine. 
-# There's a special type of optim algorithm for bounded optimisation like this.
-# 'L-BFGS-B'. This can be selected using the method argument, and also 
-# requires the lower and upper arguments be specified, for the lower 
-# and upper values to search within respectively. 
-# so, we need to write a function a bit like the following
-get_best_tau_from_optim <- function(df, buffer = 2, what = c("tau", "optim")){
-  # We can put in some data validation checks too
-
-  what <- match.arg(what)
-
-  years <- df$year 
-  tau_lower <- min(years) + buffer # the lower bound 
-  tau_upper <- max(years) - buffer # the upper bound
-  tau_start <- (tau_upper + tau_lower) / 2 # start just in the middle 
-  # (This shouldn't matter if the algorithm is robust)
-
-  get_result_from_model <- function(par, df, what = c("AIC", "BIC", "model")){
-    # Again, can carry out more validation checks 
-    what <- match.arg(what)
-
-    tau <- par[["tau"]] # This is how parameters are packed up by optim
-    model <-
-      df %>% 
-        arrange(year) %>% 
-        mutate(delta_ex = ex - lag(ex)) %>% 
-        filter(!is.na(delta_ex)) %>% 
-        mutate(T_param = ifelse(year < tau, FALSE, TRUE)) %>% 
-        lm(delta_ex ~ T_param, data = .)
-    if (what == "model") {return(model)      }
-    if (what == "BIC")   {return(BIC(model)) }
-    if (what == "AIC")   {return(AIC(model)) }
-
-    # NULL should never be returned! If it has something's gone wrong!
-    NULL
-  }
-
-  optim_obj <- 
-    optim(
-      par = list(tau = tau_start), 
-      fn = get_result_from_model, 
-      method = "L-BFGS-B",
-      lower = tau_lower,
-      upper = tau_upper,
-      df = df
-    )
-  # Need to make the result dependent on the what argument
-
-  if (what == "tau") {return(optim_obj[["par"]])}
-  else if (what == "optim"){ return(optim_obj) }
-
-  # Again, the following should never be triggered
-  NULL
-}
+
+source(here("R", "breakpoint_functions.R"))
+
 # Let's see if this now works in the functional programming context, using 
@@
 # let's use a heavily modified version of the above function to do the grid search
-get_best_tau_from_gridsearch <- function(df, buffer = 2, what = c("best", "all")){
-  # We can put in some data validation checks too
-
-  what <- match.arg(what)
-
-  years <- df$year 
-  tau_lower <- min(years) + buffer # the lower bound 
-  tau_upper <- max(years) - buffer # the upper bound
-  tau_searchrange <- tau_lower:tau_upper
-  get_result_from_model <- function(tau, df, what = c("AIC", "BIC", "model")){
-    # Again, can carry out more validation checks 
-    what <- match.arg(what)
-
-    model <-
-      df %>% 
-        arrange(year) %>% 
-        mutate(delta_ex = ex - lag(ex)) %>% 
-        filter(!is.na(delta_ex)) %>% 
-        mutate(T_param = ifelse(year < tau, FALSE, TRUE)) %>% 
-        lm(delta_ex ~ T_param, data = .)
-    if (what == "model") {return(model)      }
-    if (what == "BIC")   {return(BIC(model)) }
-    if (what == "AIC")   {return(AIC(model)) }
-
-    # NULL should never be returned! If it has something's gone wrong!
-    NULL
-  }
-
-  search_df <- tibble(
-    tau = tau_searchrange
-  ) %>% 
-    mutate(aic = map_dbl(tau, get_result_from_model, df = df))
-
-  if (what == "all") {return(search_df)}
-
-  if (what == "best") {
-    out <- 
-      search_df %>% 
-        filter(aic == min(aic)) %>% 
-        pull(tau)
-    return(out)
-  }
-  # Again, the following should never be triggered
-  NULL
-}
+
 compare_best_breakmodel_with_null <- 
@@
 
-# Now let's finish with the most unneccesarily complicated variant: using a different algorithm in 
-# optim which will allow local minima to be escaped. This is the simulated annealing method 
-# (method = "SANN") which is based on physics models. Unfortunately this is not a bounded algoritm,
-# i.e. does not take values within limits, so instead the parameter passed to it has to be 
-# converted and deconverted from a bounded range using a link function. 
-# This can be done with a modification of the sigmoid function, which shifts the transformed 
-# value from the range 0-1 to the range lower_tau-upper_tau. 
-# Let's build up this function as follows: 
-# sigmoid_shift <- function(x, lower = 0, upper = 1){
-#   (upper - lower) / (1 + exp(-x)) + lower
-# }
-# # 
-# curve(sigmoid_shift, -3, 3) # basic sigmoid 
-# 
-# sigmoid_shift2 <- function(x){
-#   sigmoid_shift(x = x, lower = 1981, upper = 2015)
-# }
-# 
-# curve(sigmoid_shift2, -3, 3)
-# So this will work as a transformation from an unbounded space (x) to a 
-# bounded space (f(x))
-# We'll also need to reverse the operation 
-# # After a bit of pen-and-paper algebra...
-# sigmoid_unshift <- function(z, lower = 0, upper = 1){
-#   -log((upper - lower) / (z - lower) -1)
-# }
-# 
-# Let's double check...
-#  
-# sigmoid_shift(-4, lower = 1982, upper = 2015) %>% 
-#   sigmoid_unshift(lower = 1982, upper = 2015)
-# 
-# 
-# 
-# get_best_tau_from_optim_sann <- function(df, buffer = 2, what = c("tau", "tau_z", "optim")){
-#   # We can put in some data validation checks too 
-#   sigmoid_shift <- function(x, lower = 0, upper = 1){
-#     (upper - lower) / (1 + exp(-x)) + lower
-#   }
-#  
-#   sigmoid_unshift <- function(z, lower = 0, upper = 1){
-#   -log((upper - lower) / (z - lower) -1)
-#   } 
-#   
-#   
-#   what <- match.arg(what)
-#   
-#   years <- df$year 
-#   tau_lower <- min(years) + buffer # the lower bound 
-#   tau_upper <- max(years) - buffer # the upper bound
-#   # tau_start <- (tau_upper + tau_lower) / 2 # start just in the middle 
-#   # (This shouldn't matter if the algorithm is robust) 
-#   
-#   get_result_from_model <- function(par, df, what = c("AIC", "BIC", "model")){
-#     # Again, can carry out more validation checks 
-#     what <- match.arg(what)
-#     
-#     tau_z <- par[["tau_z"]] # This is how parameters are packed up by optim
-#     tau   <- sigmoid_shift(tau_z, lower = tau_lower, upper = tau_upper)
-#     
-#     model <-
-#       df %>% 
-#         arrange(year) %>% 
-#         mutate(delta_ex = ex - lag(ex)) %>% 
-#         filter(!is.na(delta_ex)) %>% 
-#         mutate(T_param = ifelse(year < tau, FALSE, TRUE)) %>% 
-#         lm(delta_ex ~ T_param, data = .)
-# 
-#     if (what == "model") {return(model)      }
-#     if (what == "BIC")   {return(BIC(model)) }
-#     if (what == "AIC")   {return(AIC(model)) }
-#     
-#     # NULL should never be returned! If it has something's gone wrong!
-#     NULL
-#   }
-# 
-#   optim_obj <- 
-#     optim(
-#       par = list(tau_z = 0), 
-#       fn = get_result_from_model, 
-#       method = "SANN",
-#       df = df
-#     )
-# 
-#   # Need to make the result dependent on the what argument
-#   
-#   if (what == "tau") {return(sigmoid_shift(optim_obj[["par"]], lower = tau_lower, upper = tau_upper))}
-#   if (what == "tau_z") {return(optim_obj[["par"]])}  
-#   else if (what == "optim"){ return(optim_obj) }
-#   
-#   # Again, the following should never be triggered
-#   NULL
-# }
-# 
-# 
-# sann_estimates <- 
-#   hmd_ex_selected_countries_with_synth %>%
-#     filter(year >= 1979) %>% 
-#     group_by(code, x, sex) %>% 
-#     nest() %>% 
-#     mutate(
-#       best_tau = map_dbl(data, get_best_tau_from_optim_sann)
-#     )
-# 
-# # This takes ABSOLUTELY AGES, and really is the sledgehammer-to-crack
-# # -a-nut approach to use only as a last resort. #
-# # However if you ever do have a complex and non-finite fitness 
-# # surface to try to optimise, and it also requires searching 
-# # over a bounded area, then this is how you can do it! 
+```
+
+Let's now try to double-check the above results by running breakpoint analysis using the `segmented` package
+
+```{r}
+# Firstly, we need the data
+
+# Let's start with one of the populations, then generalise
+
+test_dta <- 
+  hmd_ex_selected_countries_with_synth %>%
+    filter(year >= 1979) %>% 
+    filter(code == "GBRTENW") %>% 
+    filter(sex == "female") %>% 
+    filter(x == 0)
+
+# Now we create a simple model with no breakpoint for y against t
+
+test_null_mdl <- lm(ex ~ year, data = test_dta)
+
+# Then we create a segmented version of the same model
+
+test_seg_mdl <- segmented::segmented(test_null_mdl)
+
+anova(test_null_mdl, test_seg_mdl)
+
+# now to pack into a function
+
+estimate_breakpoint_and_pval <- function(df){
+  null_mdl <- lm(ex ~ year, data = df) 
+  seg_mdl <- segmented::segmented(null_mdl)   
+  anova_diff <- anova(null_mdl, seg_mdl)
+
+  list(
+    null = null_mdl, 
+    seg = seg_mdl,
+    diff = anova_diff
+  )
+}
+
+segmented_breakpoint_models <- 
+  hmd_ex_selected_countries_with_synth %>% 
+    filter(year >= 1979) %>% 
+    group_by(code, x, sex) %>% 
+    nest() %>% 
+    mutate(
+      mdl_outputs = map(data, estimate_breakpoint_and_pval)
+    ) %>% 
+    unnest_wider(mdl_outputs) %>% 
+    mutate(
+      ftest_pval = map_dbl(diff, ~.[2,6]),
+      stat_sig_preference = ftest_pval < 0.05
+    ) %>% 
+    mutate(
+      breakpoint = map_dbl(seg, ~.[["psi"]][1,2]),
+      breakpoint_se = map_dbl(seg, ~.[["psi"]][1,3])
+    )
+
+segmented_breakpoint_models
+
+```
+
+Now I'm going to plot the breakpoints + SEs on the data itself
+
+```{r}
+segmented_breakpoint_models %>% 
+  unnest(data) %>% 
+  ggplot(aes(x = year, y = ex, colour = sex)) + 
+  geom_line() + 
+  facet_grid(x ~ code, scales = "free_y") + 
+  geom_vline(aes(xintercept = breakpoint, color = sex))
+
+```
+
+And now for change in ex
+
+- [ ] Jon to figure out why this doesn't work...
+
+```{r, eval = FALSE}
+estimate_breakpoint_and_pval_change <- function(df){
+  null_mdl <- lm(delta_ex ~ year, data = df) 
+  seg_mdl <- segmented::segmented(null_mdl)   
+  anova_diff <- anova(null_mdl, seg_mdl)
+
+  list(
+    null = null_mdl, 
+    seg = seg_mdl,
+    diff = anova_diff
+  )
+}
+
+segmented_breakpoint_models_change <- 
+  hmd_ex_selected_countries_with_synth %>% 
+    filter(code != "DEUTNP") %>% 
+    filter(year >= 1979) %>% 
+    group_by(code, x, sex) %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+    filter(!is.na(delta_ex)) %>% 
+    nest() %>% 
+    mutate(
+      mdl_outputs = map(data, estimate_breakpoint_and_pval_change)
+    ) %>% 
+    unnest_wider(mdl_outputs) %>% 
+    mutate(
+      ftest_pval = map_dbl(diff, ~.[2,6]),
+      stat_sig_preference = ftest_pval < 0.05
+    ) %>% 
+    mutate(
+      breakpoint = map_dbl(seg, ~.[["psi"]][1,2]),
+      breakpoint_se = map_dbl(seg, ~.[["psi"]][1,3])
+    )
+
+segmented_breakpoint_models_change
+
 ```
@@
 
+Next, let's look closer at France, Scotland, and E&Ws life expectancy at birth for males.
+
 ```{r}

### 2022-03-23 R/breakpoint_examples.R

@@ -1,0 +1,286 @@
+# Breakpoint script 
+
+source(here("R", "breakpoint_functions.R"))
+# Below is an example explaining the intuition of the breakpoint analysis
+
+# Let's start with a single example population. (There are many different population subgroups, hence the 
+# benefit of being able to automate this later.) We're going to start with England & Wales, females, life 
+# expectancy at birth (x == 0)
+example_df <- 
+  hmd_ex_selected_countries_with_synth %>% 
+  filter(x == 0) %>% 
+  filter(sex == "female") %>% 
+  filter(code == "GBRTENW") %>% 
+  filter(year >= 1979)
+# Now we create the change in ex variable as that's the response variable in the model
+example_df <- 
+  example_df %>% 
+  arrange(year) %>% 
+  mutate(delta_ex = ex - lag(ex)) # %>% 
+# ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+# geom_line() +
+# geom_hline(yintercept = 0) + 
+# stat_smooth()
+# The null model is really simple: just an intercept 
+model_null <- lm(delta_ex ~ 1, data = example_df)
+summary(model_null)
+# 
+# The code below shows how we can be even simpler, and force the regression model not to 
+# even have an intercept (i.e. have zero predictor variables)
+# model_nuller <- lm(delta_ex ~ 0, data = example_df)
+# summary(model_nuller)
+# And we can compare the intercept with the no intercept model as follows 
+# anova(model_nuller, model_null)
+# a statistically significant p value shows the intercept is definitely worth including in the 
+# model. Substantively that it's wrong to assume no improvement on average in ex over time since the 1980s
+# There are lots of possible alternative models, some examples of which are below:
+model_alt01 <- lm(delta_ex ~ year >= 1990, data = example_df)
+summary(model_alt01)
+model_alt02 <- lm(delta_ex ~ year >= 1995, data = example_df)
+summary(model_alt02)
+model_alt03 <- lm(delta_ex ~ year >= 1997, data = example_df)
+summary(model_alt03)
+model_alt04 <- lm(delta_ex ~ year >= 2008, data = example_df)
+summary(model_alt04)
+model_alt05 <- lm(delta_ex ~ year >= 2011, data = example_df)
+summary(model_alt05)
+# each of these models allows an adjustment to the horizontal line after a given year has been reached. 
+# In the examples above some possible given years are 1990, 1995, 1997, 2008 and 2011. 
[…]

### 2022-03-23 R/breakpoint_functions.R

@@ -1,0 +1,116 @@
+
+
+# Breakpoint functions
+
+run_alt_with_given_tau <- function(tau, df){
+  df <- 
+    df %>% 
+    filter(!is.na(delta_ex)) %>% 
+    mutate(
+      T_param = ifelse(year < tau, FALSE, TRUE)
+    )
+  
+  lm(delta_ex ~ T_param, data  = df)
+}
+
+
+
+
+get_best_tau_from_optim <- function(df, buffer = 2, what = c("tau", "optim")){
+  # We can put in some data validation checks too 
+  
+  what <- match.arg(what)
+  
+  years <- df$year 
+  tau_lower <- min(years) + buffer # the lower bound 
+  tau_upper <- max(years) - buffer # the upper bound
+  tau_start <- (tau_upper + tau_lower) / 2 # start just in the middle 
+  # (This shouldn't matter if the algorithm is robust) 
+  
+  get_result_from_model <- function(par, df, what = c("AIC", "BIC", "model")){
+    # Again, can carry out more validation checks 
+    what <- match.arg(what)
+    
+    tau <- par[["tau"]] # This is how parameters are packed up by optim
+    model <-
+      df %>% 
+      arrange(year) %>% 
+      mutate(delta_ex = ex - lag(ex)) %>% 
+      filter(!is.na(delta_ex)) %>% 
+      mutate(T_param = ifelse(year < tau, FALSE, TRUE)) %>% 
+      lm(delta_ex ~ T_param, data = .)
+    if (what == "model") {return(model)      }
+    if (what == "BIC")   {return(BIC(model)) }
+    if (what == "AIC")   {return(AIC(model)) }
+    
+    # NULL should never be returned! If it has something's gone wrong!
+    NULL
+  }
+  
+  optim_obj <- 
+    optim(
+      par = list(tau = tau_start), 
+      fn = get_result_from_model, 
+      method = "L-BFGS-B",
+      lower = tau_lower,
+      upper = tau_upper,
+      df = df
+    )
+  # Need to make the result dependent on the what argument
+  
+  if (what == "tau") {return(optim_obj[["par"]])}
+  else if (what == "optim"){ return(optim_obj) }
+  
+  # Again, the following should never be triggered
+  NULL
+}
+
+
+
+get_best_tau_from_gridsearch <- function(df, buffer = 2, what = c("best", "all")){
+  # We can put in some data validation checks too 
+  
[…]

### 2022-04-11 draft_working_paper.Rmd
add diagnostic statistical plots comparing breakpoints between methods

@@
 # load packages
-pacman::p_load(knitr, here, tidyverse, flextable) # using flextable instead of gt. It seems to be better 
-# developed, more commonly used, and has more clearly labelled function names.
+pacman::p_load(knitr, here, segmented, tidyverse)
+
 # load data 
@@
 
-- [ ] make smoother lines coloured by sex
+- [X] make smoother lines coloured by sex
 - [X] use consistent colours for all graphs showing separate sex results
@@
 ### Changes in life expectancy at age 65
-
-- [ ] Use slightly stronger (and consistent) colours for sex smoother lines
 
@@
 ### Scotland
+
 ```{r, echo = FALSE, warning=FALSE, message = FALSE}
@@
   select(sex, decade, cell_contents) %>% 
-  spread(decade, cell_contents) %>% 
+  spread(decade, cell_contents) #%>% 
 #  write.csv("clipboard")
-  flextable() %>%
-  # Want to make the header row bold
-  vline(j = 1, part = "all") %>%
-  set_caption(caption = "Average improvement in life expectancy by decade")
 
@@
 # nest and tibble elements 
-hmd_ex_selected_countries_with_synth %>%
-  filter(year >= 1979) %>% 
-  group_by(code, x, sex) %>% 
-  nest() %>% 
-  mutate(
-    best_tau = map_dbl(data, get_best_tau_from_optim)
-  )
+# hmd_ex_selected_countries_with_synth %>%
+#   filter(year >= 1979) %>% 
+#   group_by(code, x, sex) %>% 
+#   nest() %>% 
+#   mutate(
+#     best_tau = map_dbl(data, get_best_tau_from_optim)
+#   )
 # I think this is getting stuck on local optima, so instead maybe 
@@
     mutate(
-      best_tau = map_dbl(data, get_best_tau_from_gridsearch)
+      best_tau = map_dbl(data, get_best_tau_from_gridsearch),
+      all_taus = map(data, get_best_tau_from_gridsearch, what = "all")
     ) %>% 
@@
       )
-    )
-tmp <- compare_best_breakmodel_with_null %>% 
+    ) %>% 
   mutate(
@@
 ```
+
+Let's now try to summarise the above results intelligently.
+
+We want to know:
+
+- [ ] Best year
+- [ ] Whether stat sig over null 
+- [ ] Estimate before break
+- [ ] Estimate after break
+
+Let's take the object above then get this info.
+
+```{r}
+compare_best_breakmodel_with_null %>% 
+  mutate(population = glue::glue("{code}_{sex}_{x}")) %>% 
+  ggplot(aes(best_tau, fct_reorder(population, best_tau), colour = stat_sig_preference)) + 
+  geom_point()
+
+```
+I'm now a bit concerned by how many picked 2015. Is this an artefact of when the data finished?
+
+To test this further I've now saved AIC by proposed year so let's look at how clear a 'winner' is for each population
+
+```{r}
+compare_best_breakmodel_with_null %>% 
+  select(code, x, sex, best_tau, all_taus, ftest_pval) %>% 
+  unnest(all_taus) %>% 
+  ungroup() %>% 
+  mutate(population = glue::glue("{code}_{x}_{sex}")) %>% 
+  mutate(
+    ftest_cat = case_when(
+      ftest_pval < 0.01 ~ "<0.01",
+      between(ftest_pval, 0.01, 0.05) ~ "0.01-0.05",
+      between(ftest_pval, 0.05, 0.10) ~ "0.05-0.10",
+      TRUE                            ~ "Not Stat. Sig."
+    ),
+    ftest_cat = factor(ftest_cat, 
+      levels = c("<0.01", "0.01-0.05", "0.05-0.10", "Not Stat. Sig."),
+      ordered = TRUE
+    )
+  ) %>% 
+  ggplot(aes(tau, aic)) +
+  geom_line() + geom_point(aes(colour = ftest_cat, alpha = tau == best_tau)) + 
+  facet_wrap(~population, scales = "free_y") +
+  scale_colour_manual(
+    values = c(
+      "<0.01" = "darkred", 
+      "0.01-0.05"  = "red", 
+      "0.05-0.10" = "orange", 
+      "Not Stat. Sig." = "lightblue"
+      )
+  ) + 
+  scale_alpha_manual(values = c(`FALSE` = 0, `TRUE` = 1), guide = "none") +
+  labs(
+    title = "Penalised fit (AIC) against proposed changepoint year, various population",
+    subtitle = "Best fit identified indicate with a point, whose colour indicates statistical significance of changepoint model compared with Null model",
+    colour = "Statistical\nsignificance", 
+    x = "Year proposed",
+    y = "AIC (lower is better)"
+  )
+```
+
+I think this is a fairly complex visualisation, but hopefully usable either in the main paper or (more likely) supplementary material.
+
+Here's a table attempting to summarise (subjectively) the above graphs:
+
+| **Population**    | **x** | **Sex** | **Main Breakpoint (Stat sig.)** | **Other possible breakpoints** |
+|-------------------|-------|---------|---------------------------------|--------------------------------|
+| East Germany      | 0     | male    | 1991 **                         | No                             |
+| East Germany      | 0     | female  | 2005 **                         | 2015                           |
+| East Germany      | 65    | male    | 1991 **                         | No                             |
+| East Germany      | 65    | female  | 1991 **                         | No                             |
+| Germany           | 0     | male    | 2015 **                         | 2007ish                        |
+| Germany           | 0     | female  | 2007 **                         | 2005-2008                      |
+| Germany           | 65    | male    | 2015 **                         | No                             |
+| Germany           | 65    | female  | 2015 *                          | 2005-2008                      |
+| Synthetic Germany | 0     | male    | 2015 **                         | Weak 2007                      |
+| Synthetic Germany | 0     | female  | 2007 **                         | 2005-2008, 2015                |
+| Synthetic Germany | 65    | male    | 2015 **                         | No                             |
+| Synthetic Germany | 65    | female  | 2015 *                          | Weak 2005-2007                 |
+| West Germany      | 0     | male    | 2015 **                         | Weak 2007                      |
+| West Germany      | 0     | female  | 1988 **                         | 2005-2008ish                   |
+| West Germany      | 65    | male    | 2015 **                         | No                             |
+| West Germany      | 65    | female  | 2015 *                          | 2008 ish (Weak)                |
+| Spain             | 0     | male    | 1983 *                          | No                             |
+| Spain             | 0     | female  | 1983 *                          | No                             |
+| Spain             | 65    | male    | 2000 (NS)                       | 2015                           |
+| Spain             | 65    | female  | 2014 (NS)                       | 1983?                          |
+| France            | 0     | male    | 2015 (NS)                       | ?                              |
+| France            | 0     | female  | 2015 (NS)                       | ?                              |
+| France            | 65    | male    | 2015 **                         | No                             |
+| France            | 65    | female  | 2015 *                          | No                             |
+| Scotland          | 0     | male    | 2015 ***                        | No                             |
+| Scotland          | 0     | female  | 1980 *                          | 2015 (Strong)                  |
+| Scotland          | 65    | male    | 1994 (NS)                       | No?                            |
+| Scotland          | 65    | female  | 1981 (NS)                       | No                             |
+| England & Wales   | 0     | male    | 2013 ***                        | 2012-2015ish                   |
+| England & Wales   | 0     | female  | 2012 **                         | Around 2015                    |
+| England & Wales   | 65    | male    | 2012 *                          | 2015                           |
+| England & Wales   | 65    | female  | 2012 (NS)                       | 2015                           |
+| Italy             | 0     | male    | 1981 (NS)                       | 2015                           |
+| Italy             | 0     | female  | 2005 (NS)                       | Weak 2015                      |
+| Italy             | 65    | male    | 1984 (NS)                       | 1981                           |
+| Italy             | 65    | female  | 2005 (NS)                       | 1984                           |
 
@@
 
+estimate_breakpoint_and_pval_v2 <- function(df){
+  null_mdl <- lm(ex ~ year, data = df) 
+  seg_mdl <- segmented::segmented(null_mdl, seg.Z= ~year, psi = 2010)   
+  anova_diff <- anova(null_mdl, seg_mdl)
+
+  list(
+    null = null_mdl, 
+    seg = seg_mdl,
+    diff = anova_diff
+  )
+}
+
+segmented_breakpoint_models_v2 <- 
+  hmd_ex_selected_countries_with_synth %>% 
+    filter(year >= 1979) %>% 
+    group_by(code, x, sex) %>% 
+    nest() %>% 
+    mutate(
+      mdl_outputs = map(data, estimate_breakpoint_and_pval_v2)
+    ) %>% 
+    unnest_wider(mdl_outputs) %>% 
+    mutate(
+      ftest_pval = map_dbl(diff, ~.[2,6]),
+      stat_sig_preference = ftest_pval < 0.05
+    ) %>% 
+    mutate(
+      breakpoint = map_dbl(seg, ~.[["psi"]][1,2]),
+      breakpoint_se = map_dbl(seg, ~.[["psi"]][1,3])
+    )
+
+segmented_breakpoint_models_v2
+
+# To check the results are the same more formally
+
+identical(
+  segmented_breakpoint_models %>% 
+    select(
+      code, x, sex, ftest_pval, stat_sig_preference, breakpoint
+    ), 
+  segmented_breakpoint_models_v2 %>% 
+    select(
+      code, x, sex, ftest_pval, stat_sig_preference, breakpoint
+    )
+  )
+
+# Returns FALSE
+
+# How are they different? 
+models_comparison <- 
+segmented_breakpoint_models %>% 
+  select(code, x, sex, ftest_pval, stat_sig_preference, breakpoint, breakpoint_se) %>% 
+  group_by(code, x, sex) %>% 
+  nest(v1 = c(ftest_pval, stat_sig_preference, breakpoint, breakpoint_se)) %>% 
+  left_join(
+    segmented_breakpoint_models_v2 %>% 
+      select(code, x, sex, ftest_pval, stat_sig_preference, breakpoint, breakpoint_se) %>% 
+      group_by(code, x, sex) %>% 
+      nest(v2 = c(ftest_pval, stat_sig_preference, breakpoint, breakpoint_se))
+  ) %>% 
+  unnest_wider(
+    v1, names_sep = "."
+  ) %>% 
+  unnest_wider(
+    v2, names_sep = "."
+  )
+
+# How do the breakpoints and SEs differ? 
+models_comparison %>% 
+  ggplot() + 
+  geom_point(aes(v1.breakpoint, v2.breakpoint))
+
+# There's just one difference. What is it? 
+models_comparison %>% 
+  mutate(bp_diff = v2.breakpoint - v1.breakpoint) %>% 
+  mutate(population = glue::glue("{code}_{x}_{sex}")) %>% 
+  ggplot(aes(bp_diff, fct_reorder(population, bp_diff))) + 
+    geom_point()
+
+# So, the different starting points make a substantive difference only 
+# for FRATNP males, at age 65
+
+# I suspect they some smaller difference for other populations however. Let's 
+# check
+
+models_comparison %>% 
+  mutate(bp_diff = v2.breakpoint - v1.breakpoint) %>% 
+  mutate(population = glue::glue("{code}_{x}_{sex}")) %>% 
+  filter(population != "FRATNP_65_male") %>% 
+  ggplot(aes(bp_diff, fct_reorder(population, bp_diff))) + 
+    geom_point()
+
+# Yes, there are differences, but they are tiny...
+
 ```
@@
 
+I'm now going to compare the breakpoints from the segmented package with the best point from the line/gridsearch approach
+
+```{r}
+compare_best_breakmodel_with_null %>% 
+  select(code, x, sex, best_tau, all_taus, ftest_pval) %>% 
+  unnest(all_taus) %>% 
+  ungroup() %>% left_join(
+    segmented_breakpoint_models %>% 
+      select(code, x, sex, bp_ftest_pval = ftest_pval, bp_breakpoint = breakpoint, bp_breakpoint_se = breakpoint_se)
+  ) %>% 
+  mutate(population = glue::glue("{code}_{x}_{sex}")) %>% 
+  mutate(
+    ftest_cat = case_when(
+      ftest_pval < 0.01 ~ "<0.01",
+      between(ftest_pval, 0.01, 0.05) ~ "0.01-0.05",
+      between(ftest_pval, 0.05, 0.10) ~ "0.05-0.10",
+      TRUE                            ~ "Not Stat. Sig."
+    ),
+    ftest_cat = factor(ftest_cat, 
+      levels = c("<0.01", "0.01-0.05", "0.05-0.10", "Not Stat. Sig."),
+      ordered = TRUE
+    )
+  ) %>% 
+  mutate(
+    bp_ftest_cat = case_when(
+      bp_ftest_pval < 0.01 ~ "<0.01",
+      between(bp_ftest_pval, 0.01, 0.05) ~ "0.01-0.05",
+      between(bp_ftest_pval, 0.05, 0.10) ~ "0.05-0.10",
+      TRUE                            ~ "Not Stat. Sig."
+    ),
+    bp_ftest_cat = factor(bp_ftest_cat, 
+      levels = c("<0.01", "0.01-0.05", "0.05-0.10", "Not Stat. Sig."),
+      ordered = TRUE
+    )
+  ) %>% 
+  ggplot(aes(tau, aic)) +
+  geom_line() + 
+  geom_point(aes(colour = ftest_cat, alpha = tau == best_tau)) + 
+  facet_wrap(~population, scales = "free_y") +
+  scale_colour_manual(
+    values = c(
+      "<0.01" = "darkred", 
+      "0.01-0.05"  = "red", 
+      "0.05-0.10" = "orange", 
+      "Not Stat. Sig." = "lightblue"
+      )
+  ) + 
+  geom_vline(
+    aes(
+      xintercept = bp_breakpoint,
+      colour = bp_ftest_cat  
+    )
+  ) + 
+  geom_rect(
+    aes(
+      xmin = bp_breakpoint - 1.96 * bp_breakpoint_se, 
+      xmax = bp_breakpoint + 1.96 * bp_breakpoint_se
+    ),
+      ymin = -Inf, ymax = Inf,
+    alpha = 0.01
+  ) + 
+  scale_alpha_manual(values = c(`FALSE` = 0, `TRUE` = 1), guide = "none") +
+  labs(
+    title = "Penalised fit (AIC) against proposed changepoint year, various population",
+    subtitle = "Best fit identified indicate with a point, whose colour indicates statistical significance of changepoint model compared with Null model",
+    colour = "Statistical\nsignificance", 
+    x = "Year proposed",
+    y = "AIC (lower is better)",
+    caption = "Vertical lines and bands show estimates and 95% CIs from segmented package"
+  )
+```
+
 And now for change in ex
 
-- [ ] Jon to figure out why this doesn't work...
+- [X] Jon to figure out why this doesn't work...
+    - I think it doesn't work because the `segmented` package functions expect the line to be joined up.
+Instead it looks like the [structchange](https://cran.r-project.org/web/packages/strucchange/index.html) package can be used for more general model specifications.
 
@@
 
-```{r}
+```{r, eval = FALSE}
 stat_sig_table <- select(tmp, code, x, sex, best_tau, ftest_pval, stat_sig_preference) %>% arrange(ftest_pval)
@@
 # Bibliography
+
+# Dream Graphs
+
+To finally complete these, without getting distracted/stuck on how to double check and improve upon the underlying data/models
+
+- [ ] I've finally remembered that I can get the breakpoints by aggregating up the changes, as I've done many times previously!
+
+## Graph of ex against year, predicted and observed.
+
+```{r}
+# Start with Scotland, female, 0
+
+test_data <- compare_best_breakmodel_with_null %>% 
+  filter(code == "GBR_SCO", x == 0, sex == "female") %>% 
+  pluck("data", 1)
+
+test_data %>% 
+  ggplot(aes(year, ex)) + 
+  geom_point()
+
+# Now to get the modelled estimates
+
+compare_best_breakmodel_with_null %>% 
+  filter(code == "GBR_SCO", x == 0, sex == "female") %>% 
+  pluck("model_alt", 1) %>% 
+  broom::tidy()
+
+# So, we want the last year before the breakpoint, then to produce a line that just integrates the (Intercept) term as many years as we want to project forward for.
+
+# So, let's think of a function that makes a graph, and what inputs it needs
+
+show_counterfactual_figure <- function(df, best_tau, alt_model){
+  # First we might want to estimate the intercept for the pre-slowdown portion 
+  # while fixing the slope at that estimated by (Intercept) in the delta_ex model
+
+  pred_slope <- alt_model %>%  broom::tidy() %>% 
+    filter(term == "(Intercept)") %>% 
+    pluck("estimate")
+  # 
+  pred_slope_after <- alt_model %>% 
+    broom::tidy() %>% 
+    filter(term == "T_paramTRUE") %>% 
+    pluck("estimate")
+
+  pred_intercept <-  mean(df$ex - pred_slope * df$year)
+
+  pred_intercept_at_tau <- pred_intercept - pred_slope_after * best_tau
+
+  # I think I should use geom_area for the after part 
+  df %>% 
+    mutate(
+      y_pred_before = pred_intercept + pred_slope * year,
+      y_pred_after  = pred_intercept_at_tau + (pred_slope + pred_slope_after) * year
+    ) %>% 
+    mutate(
+      y_pred_before = ifelse(year < best_tau, NA, y_pred_before),
+      y_pred_after  = ifelse(year < best_tau, NA, y_pred_after)
+    ) %>% 
+    ggplot() + 
+    geom_point(aes(year, ex)) + 
+    geom_vline(xintercept = best_tau) + 
+    geom_abline(intercept = pred_intercept, slope = pred_slope) + 
+    geom_ribbon(aes(ymax = y_pred_before, ymin = y_pred_after, x = year), alpha = 0.5)
+}
+
+this_df <- compare_best_breakmodel_with_null %>% 
+  filter(code == "GBR_SCO", x == 0, sex == "male") %>% 
+  pluck("data", 1)
+
+this_best_tau <- compare_best_breakmodel_with_null %>% 
+  filter(code == "GBR_SCO", x == 0, sex == "male") %>% 
+  pluck("best_tau")
+
+this_model_alt <- compare_best_breakmodel_with_null %>% 
+  filter(code == "GBR_SCO", x == 0, sex == "male") %>% 
+  pluck("model_alt", 1)
+
+show_counterfactual_figure(this_df, this_best_tau, this_model_alt)
+
+# Now to do for all
+
+# Let's also label and set axes intelligently
+
+add_labels_axes <- function(gg, country, start_age, sex, 
+                            ylims = list(
+                                birth = c(70, 95),
+                                x65 = c(0, 25)
+                              )
+                            ){
+  gg + 
+    labs(
+      x = "Year", 
+      y = case_when(
+        start_age == 0  ~ "Life expectancy at birth",
+        start_age == 65 ~ "Life expectancy at age 65",
+        TRUE            ~ NA_character_
+      ),
+      title = glue::glue("Life expectancy in {country}, at age {start_age}, for {sex}s"),
+      subtitle = "Shaded areas show deviation from projected life expectancy before any breakpoint in trend", 
+      caption = "Source: Human Mortality Database"
+    ) + 
+    expand_limits( 
+      y = case_when(
+        start_age == 0 ~ ylims[["birth"]],
+        start_age == 65 ~ ylims[["x65"]],
+        TRUE    ~ c(0, 90)
+      )
+    )
+}
+
+compare_best_breakmodel_with_null %>% 
+  left_join(country_code_lookup) %>% 
+  mutate(
+    ggsegplot = pmap(
+      .l = list(df = data, best_tau = best_tau, alt_model = model_alt),
+      .f = show_counterfactual_figure
+    ),
+    ggsegplot = pmap(.l = list(gg = ggsegplot, country = country, start_age = x, sex = sex), add_labels_axes)
+  ) %>% 
+  mutate(NULL = walk(ggsegplot, print))
+
+```

### 2022-05-05 draft_working_paper.Rmd
consider another summary graph

@@
   )
+
+ggsave("figures/aic_from_gridsearch_and_segmented_bp_estimates.png", height = 20, width = 30, units = "cm", dpi = 150)
 ```
@@
 ```
+
+# Another attempt at a simple summary
+
+It could be we don't get a clear changepoint for some populations because the annual variance as a function of change is particularly large.
+This would be more an issue for smaller populations than larger populations.
+The changepoint can therefore be expressed as a standarised unit.
+
+```{r}
+
+mean_sd_chex <- 
+  hmd_ex_selected_countries_with_synth %>% 
+    filter(code %in% c("DEUTSYNTH", "GBRTENW", "GBR_SCO", "FRATNP", "ITA", "ESP")) %>% 
+    filter(between(year, 1979, 2019)) %>% 
+    filter(x == 0) %>% 
+    group_by(code, sex) %>% 
+    arrange(year) %>% 
+    mutate(ch_ex = ex - lag(ex)) %>% 
+    filter(!is.na(ch_ex)) %>% 
+    summarise(
+      mean_chex = mean(ch_ex),
+      sd_chex   = sd(ch_ex)
+    ) %>% 
+    mutate(b = mean_chex / sd_chex) %>% 
+    ungroup() %>% 
+    arrange(desc(mean_chex))
+
+mean_sd_chex
+
+# Now to compare with the changepoint model
+
+get_chpt_coeffs <- function(x){
+  tidied_coeffs <- broom::tidy(x)
+
+  tibble(
+    mean_chex_before = tidied_coeffs$estimate[1],
+    sd_chex_before   = tidied_coeffs$std.error[1],
+    mean_chex_after  = tidied_coeffs$estimate[2],
+    sd_chex_after    = tidied_coeffs$std.error[2]
+  )
+}
+
+broke_unbroke <- 
+  mean_sd_chex %>% 
+    left_join(
+      compare_best_breakmodel_with_null %>% 
+        ungroup() %>% 
+        filter(x == 0) %>% 
+        select(code, sex, best_tau, model_alt, ftest_pval) %>% 
+        mutate(coeffs = map(model_alt, get_chpt_coeffs)) %>% 
+        unnest(coeffs) %>% 
+        select(-model_alt) 
+    ) %>% 
+    mutate(
+      ch_as_ratio_of_mean_chex = mean_chex_after / mean_chex,
+      pos_chex_after = mean_chex_before + mean_chex_after
+    )
+
+# now to visualise in a simple way
+
+expand_grid(
+  year = 1980:2019,
+  broke_unbroke
+  ) %>% 
+  mutate(
+    broke_val = if_else(year < best_tau, mean_chex_before, mean_chex_before + mean_chex_after)
+  ) %>% 
+  left_join(
+    hmd_ex_selected_countries_with_synth %>% 
+      filter(code %in% c("DEUTSYNTH", "GBRTENW", "GBR_SCO", "FRATNP", "ITA", "ESP")) %>% 
+      filter(between(year, 1979, 2019)) %>% 
+      filter(x == 0) %>% 
+      group_by(code, sex) %>% 
+      arrange(year) %>% 
+      mutate(ch_ex = ex - lag(ex)) %>% 
+      filter(!is.na(ch_ex)) %>% 
+      ungroup() %>% 
+      select(code, year, sex, ch_ex)
+  ) %>% 
+  ggplot(aes(year)) +
+  geom_hline(aes(yintercept = mean_chex)) + 
+  geom_hline(yintercept = 0, size = 1.3) + 
+  facet_grid(code ~ sex) +
+  geom_ribbon(aes(ymin = mean_chex - sd_chex, ymax = mean_chex + sd_chex), alpha = 0.2) + 
+  geom_ribbon(aes(ymin = mean_chex - 2 * sd_chex, ymax = mean_chex + 2 * sd_chex), alpha = 0.2) + 
+  expand_limits(y = 0) + 
+  geom_step(aes(y = broke_val, alpha = 1 - ftest_pval, linetype = ftest_pval < 0.05), colour = "red") + 
+  geom_point(aes(y = ch_ex)) +
+  scale_alpha_continuous(range = c(0, 1)) + 
+  scale_linetype_manual(values = c("FALSE" = "dotted", "TRUE" = "solid")) + 
+  guides(alpha = "none", linetype = "none")
+
+```

### 2022-06-02 draft_working_paper.Rmd
try to tidy main analysis; build simple app to look at how changepoint models work

--- 2022-03-23 draft_working_paper.Rmd
@@
 # load packages
-pacman::p_load(knitr, here, tidyverse, flextable) # using flextable instead of gt. It seems to be better 
-# developed, more commonly used, and has more clearly labelled function names.
+pacman::p_load(knitr, here, segmented, tidyverse, kableExtra)
+
 # load data 
 hmd_lt <- read_rds(here("data", "lifetables.rds"))
-```
-
-## Analysis
-
-For each of these countries, and for males and females separately, we are interested in the annual changes in life expectancy at birth ($e_0$) and life expectancy at age 65 ($e_{65}$), from 1980 to the last available year for each country.
-In addition, we compare average life expectancy improvements over the 4 decades: 1980, 1990, 2000, and 2010.
-
-We then compared whether the average between the decades was significant, using a regression model.
-
-### Synthetic Germany
-
-- To maybe be moved to the appendix later
-
-To allow UK national trends to be compared with a single German population, we attempted to produce a 'Synthetic German' population with data for East and West Germany for years prior to reunification.
-We estimated that this 'Synthetic Germany' could be produced by using a weighted average of 20% East Germany, and 80% West German life expectancy trends.
-More precise estimates can be produced, and the methods used to reach this conclusion are detailed in Appendix X.
-
-```{r, echo = FALSE}
-countries_of_interest <- c(
-  "GBRTENW",
-  "GBR_SCO",
-  "GBR_UK",
-  "FRATNP",
-  "ESP",
-  "ITA",
-  "DEUTNP",
-  "DEUTE", 
-  "DEUTW"
-)
-hmd_ex_selected_countries <- 
-  hmd_lt %>% 
-    filter(code %in% countries_of_interest) %>% 
-    filter(Age %in% c(0, 65)) %>% 
-    select(code, Year, Age, sex, ex) %>% 
-    rename(x = Age, year = Year)
-hmd_ex_selected_countries
-# Code for producing synthetic Germany
-source(here("R", "make_synthetic_germany_functions.R"))
-series_east_f0 <- hmd_ex_selected_countries %>% 
-  filter(code == "DEUTE") %>% 
-  filter(between(year, 1980, 2020)) %>% 
-  filter(x == 0) %>% 
-  filter(sex == "female") %>% 
-  arrange(year) %>% 
-  pull(ex)
-series_west_f0 <- hmd_ex_selected_countries %>% 
-  filter(code == "DEUTW") %>% 
-  filter(between(year, 1980, 2020)) %>% 
-  filter(x == 0) %>% 
-  filter(sex == "female") %>% 
-  arrange(year) %>% 
-  pull(ex)
-tmp_df_f0 <- tibble(
-  code = "DEUTSYNTH",
-  year = 1980:2017,
-  x = 0,
-  sex = "female", 
-  ex = make_synthetic_population(series_east_f0, series_west_f0, p_east = 0.2)
-)
-###
-series_east_f65 <- hmd_ex_selected_countries %>% 
-  filter(code == "DEUTE") %>% 
-  filter(between(year, 1980, 2020)) %>% 
-  filter(x == 65) %>% 
-  filter(sex == "female") %>% 
-  arrange(year) %>% 
-  pull(ex)
-series_west_f65 <- hmd_ex_selected_countries %>% 
-  filter(code == "DEUTW") %>% 
-  filter(between(year, 1980, 2020)) %>% 
-  filter(x == 65) %>% 
-  filter(sex == "female") %>% 
-  arrange(year) %>% 
-  pull(ex)
-tmp_df_f65 <- tibble(
-  code = "DEUTSYNTH",
-  year = 1980:2017,
-  x = 65,
-  sex = "female", 
-  ex = make_synthetic_population(series_east_f65, series_west_f65, p_east = 0.2)
-)
-##
-series_east_m0 <- hmd_ex_selected_countries %>% 
-  filter(code == "DEUTE") %>% 
-  filter(between(year, 1980, 2020)) %>% 
-  filter(x == 0) %>% 
-  filter(sex == "male") %>% 
-  arrange(year) %>% 
-  pull(ex)
-series_west_m0 <- hmd_ex_selected_countries %>% 
-  filter(code == "DEUTW") %>% 
-  filter(between(year, 1980, 2020)) %>% 
-  filter(x == 0) %>% 
-  filter(sex == "male") %>% 
-  arrange(year) %>% 
-  pull(ex)
-tmp_df_m0 <- tibble(
-  code = "DEUTSYNTH",
-  year = 1980:2017,
-  x = 0,
-  sex = "male", 
-  ex = make_synthetic_population(series_east_m0, series_west_m0, p_east = 0.2)
-)
-###
-series_east_m65 <- hmd_ex_selected_countries %>% 
-  filter(code == "DEUTE") %>% 
-  filter(between(year, 1980, 2020)) %>% 
-  filter(x == 65) %>% 
-  filter(sex == "male") %>% 
-  arrange(year) %>% 
-  pull(ex)
-series_west_m65 <- hmd_ex_selected_countries %>% 
-  filter(code == "DEUTW") %>% 
-  filter(between(year, 1980, 2020)) %>% 
-  filter(x == 65) %>% 
-  filter(sex == "male") %>% 
-  arrange(year) %>% 
-  pull(ex)
-tmp_df_m65 <- tibble(
-  code = "DEUTSYNTH",
-  year = 1980:2017,
-  x = 65,
-  sex = "male", 
-  ex = make_synthetic_population(series_east_m65, series_west_m65, p_east = 0.2)
-)
-hmd_ex_selected_countries_with_synth <- bind_rows(
-  hmd_ex_selected_countries,
-  tmp_df_m0,
-  tmp_df_m65,
-  tmp_df_f0,
-  tmp_df_f65
-)
+
 # Labels for codes 
@@
   )
+
+countries_of_interest <- c(
+  "GBRTENW",
+  "GBR_SCO",
+  "GBR_UK",
+  "FRATNP",
+  "ESP",
+  "ITA",
+  "DEUTNP",
+  "DEUTE", 
+  "DEUTW"
+)
+
+source(here("R", "make_synthetic_germany_functions.R"))
+
+```
+
+## Analysis
+
+For each of these countries, and for males and females separately, we are interested in the annual changes in life expectancy at birth ($e_0$) and life expectancy at age 65 ($e_{65}$), from 1980 to the last available year for each country.
+In addition, we compare average life expectancy improvements over the 4 decades: 1980, 1990, 2000, and 2010.
+
+We then compared whether the average between the decades was significant, using a regression model.
+
+### Synthetic Germany
+
+- To maybe be moved to the appendix later
+
+To allow UK national trends to be compared with a single German population, we attempted to produce a 'Synthetic German' population with data for East and West Germany for years prior to reunification.
+We estimated that this 'Synthetic Germany' could be produced by using a weighted average of 20% East Germany, and 80% West German life expectancy trends.
+More precise estimates can be produced, and the methods used to reach this conclusion are detailed in Appendix X.
+
+```{r, echo = FALSE}
+source(here("R", "make_pop_selection.R"))
 ```
@@
     ungroup() 
-
-change_in_ex_selected_countries
 ```
@@
 ### Changes in life expectancy at birth
-
-- [ ] make smoother lines coloured by sex
-- [X] use consistent colours for all graphs showing separate sex results
 
@@
 
-- [ ] Use slightly stronger (and consistent) colours for sex smoother lines
-
 ```{r, echo = FALSE, warning=FALSE, message = FALSE, fig.height=8, fig.width=8}
@@
 This produces the average change for the decade, whether negative or positive, the standard error and its statistical significance for each country by sex.
-
-- [X] Give this a title and further info (lower priority)
-- [X] Look at splitting by starting age 
-    - [ ] Look at further refining starting age split
 
@@
 
-### England and Wales
-
-```{r, echo = FALSE, warning=FALSE, message = FALSE}
-
-#ENW female at birth
-f_e0_ENW <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 0) %>% 
-      filter(sex == "female") %>% 
-      filter(code == "GBRTENW") %>% 
-      filter(year >= 1979)
-
-# Now we create the change in ex variable as that's the response variable in the model
-  f_e0_ENW <- f_e0_ENW %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex)) %>% 
-   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
-   geom_line() +
-   geom_hline(yintercept = 0) + 
-   stat_smooth(colour = "lightpink3")
-
-  #ENW male at birth
-m_e0_ENW <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 0) %>% 
-      filter(sex == "male") %>% 
-      filter(code == "GBRTENW") %>% 
-      filter(year >= 1979)
-
-  m_e0_ENW <- m_e0_ENW %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex)) %>% 
-   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
-   geom_line() +
-   geom_hline(yintercept = 0) + 
-   stat_smooth(colour = "aquamarine4")
-
-  #ENW female at age 65
-f_e65_ENW <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 65) %>% 
-      filter(sex == "female") %>% 
-      filter(code == "GBRTENW") %>% 
-      filter(year >= 1979)
-
-# Now we create the change in ex variable 
-  f_e65_ENW <- f_e65_ENW %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex)) %>% 
-   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
-   geom_line() +
-   geom_hline(yintercept = 0) + 
-   stat_smooth(colour = "lightpink3")
-
-  #ENW male at aged 65
-m_e65_ENW <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 65) %>% 
-      filter(sex == "male") %>% 
-      filter(code == "GBRTENW") %>% 
-      filter(year >= 1979)
-
- m_e65_ENW <- m_e65_ENW %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex)) %>% 
-   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
-   geom_line() +
-   geom_hline(yintercept = 0) + 
-   stat_smooth(colour = "aquamarine4")
-
- #add labels
-f_e0_ENW + labs(y = "Annual change in e0", title = "England and Wales", subtitle = "Life expectancy changes at birth for females", caption = "Source: Human Mortality Database")
-
-m_e0_ENW + labs(y = "Annual change in e0", title = "England and Wales", subtitle = "Life expectancy changes at birth for males", caption = "Source: Human Mortality Database")
-
-f_e65_ENW + labs(y = "Annual change in e65", title = "England and Wales", subtitle = "Life expectancy changes at aged 65 for females", caption = "Source: Human Mortality Database")
-
-m_e65_ENW + labs(y = "Annual change in e65", title = "England and Wales", subtitle = "Life expectancy changes at aged 65 for males", caption = "Source: Human Mortality Database")
-
-  #to make grid, first assign to object
-# f_e0_ENW <- f_e0_ENW + labs(y = "Annual change in e0", title = "England and Wales", subtitle = "Life expectancy changes at birth for females", caption = "Source: Human Mortality Database")
-# 
-# m_e0_ENW <- m_e0_ENW + labs(y = "Annual change in e0", title = "England and Wales", subtitle = "Life expectancy changes at birth for males", caption = "Source: Human Mortality Database")
-# 
-# f_e65_ENW <- f_e65_ENW + labs(y = "Annual change in e65", title = "England and Wales", subtitle = "Life expectancy changes at aged 65 for females", caption = "Source: Human Mortality Database")
-# 
-# m_e65_ENW <- m_e65_ENW + labs(y = "Annual change in e65", title = "England and Wales", subtitle = "Life expectancy changes at aged 65 for males", caption = "Source: Human Mortality Database")
-
-#changing labels for grid to work
-# f_e0_ENW <- f_e0_ENW + labs(y = "Annual change in e0", title = "England and Wales", subtitle = "Life expectancy changes at birth for females")
-# m_e0_ENW <- m_e0_ENW + labs(y = "Annual change in e0", subtitle = "Life expectancy changes at birth for males")
-# f_e65_ENW <- f_e65_ENW + labs(y = "Annual change in e65")
-# m_e65_ENW <- m_e65_ENW + labs(y = "Annual change in e65", caption = "Source: Human Mortality Database")
-# #put side by side, needs package gridExtra then uses grid.arrange(plot1, plot2, ncol=2)
-# grid.arrange(f_e0_ENW, m_e0_ENW, f_e65_ENW, m_e65_ENW, ncol=2, nrow=2)
-# #grid.arrange(f_e65_ENW, m_e65_ENW, ncol=2, nrow=2)
-
-#   hmd_ex_selected_countries_with_synth %>% 
-#       filter(sex == "male") %>% 
-#       filter(code %in% c("GBRTENW", "GBR_SCO")) %>% 
-#     group_by(code, x) %>% arrange(year) %>% 
-#       mutate(delta_ex = ex - lag(ex)) %>%
-#     ungroup %>% 
-#     filter(year >= 1979) %>% 
-#     ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
-#    geom_line() +
-#    geom_hline(yintercept = 0) + 
-#    stat_smooth(colour = "aquamarine4")+
-# facet_grid(x ~ code)
-
-  ex_ENW <- hmd_ex_selected_countries_with_synth %>%
-      filter(sex %in% c("female", "male")) %>%
-      filter(code == "GBRTENW") %>%
-    group_by(sex, x) %>% arrange(year) %>%
-      mutate(delta_ex = ex - lag(ex)) %>%
-    ungroup %>%
-    filter(year >= 1979) %>%
-    ggplot(aes(x = year, y = delta_ex)) + geom_point(size=0.3) +
-   geom_line() +
-   geom_hline(yintercept = 0) +
-    stat_smooth(mapping = aes(colour = sex)) +
-       facet_grid(x ~ sex)
-
-  ex_ENW + labs(x = "Year", y = "Annual change in life expectancy", title = "England and Wales", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
-
-```
-
-### Scotland
-```{r, echo = FALSE, warning=FALSE, message = FALSE}
-#SCO female at birth
-f_e0_SCO <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 0) %>% 
-      filter(sex == "female") %>% 
-      filter(code == "GBR_SCO") %>% 
-      filter(year >= 1979)
-
-# Now we create the change in ex variable as that's the response variable in the model
-  f_e0_SCO <- f_e0_SCO %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex)) %>% 
-   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
-   geom_line() +
-   geom_hline(yintercept = 0) + 
-   stat_smooth(colour = "lightpink3")
-
-  #SCO male at birth
-m_e0_SCO <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 0) %>% 
-      filter(sex == "male") %>% 
-      filter(code == "GBR_SCO") %>% 
-      filter(year >= 1979)
-
-  m_e0_SCO <- m_e0_SCO %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex)) %>% 
-   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
-   geom_line() +
-   geom_hline(yintercept = 0) + 
-   stat_smooth(colour = "aquamarine4")
-
-  #SCO female at age 65
-f_e65_SCO <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 65) %>% 
-      filter(sex == "female") %>% 
-      filter(code == "GBR_SCO") %>% 
-      filter(year >= 1979)
-
-# Now we create the change in ex variable as that's the response variable in the model
-  f_e65_SCO <- f_e65_SCO %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex)) %>% 
-   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
-   geom_line() +
-   geom_hline(yintercept = 0) + 
-   stat_smooth(colour = "lightpink3")
-
-  #SCO male at aged 65
-m_e65_SCO <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 65) %>% 
-      filter(sex == "male") %>% 
-      filter(code == "GBR_SCO") %>% 
-      filter(year >= 1979)
-
- m_e65_SCO <-  m_e65_SCO %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex)) %>% 
-   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
-   geom_line() +
-   geom_hline(yintercept = 0) + 
-   stat_smooth(colour = "aquamarine4", show.legend = TRUE)
-
-  #add labels
-f_e0_SCO + labs(y = "Annual change in e0", title = "Scotland", subtitle = "Life expectancy changes at birth for females", caption = "Source: Human Mortality Database")
- m_e0_SCO + labs(y = "Annual change in e0", title = "Scotland", subtitle = "Life expectancy changes at birth for males", caption = "Source: Human Mortality Database")
- f_e65_SCO + labs(y = "Annual change in e65", title = "Scotland", subtitle = "Life expectancy changes at aged 65 for females", caption = "Source: Human Mortality Database")
- m_e65_SCO + labs(y = "Annual change in e65", title = "Scotland", subtitle = "Life expectancy changes at aged 65 for males", caption = "Source: Human Mortality Database")
-
- ex_SCO <- hmd_ex_selected_countries_with_synth %>%
-      filter(sex %in% c("female", "male")) %>%
-      filter(code == "GBR_SCO") %>%
-    group_by(sex, x) %>% arrange(year) %>%
-      mutate(delta_ex = ex - lag(ex)) %>%
-    ungroup %>%
-    filter(year >= 1979) %>%
-    ggplot(aes(x = year, y = delta_ex)) + geom_point(size=0.3) +
-   geom_line() +
-   geom_hline(yintercept = 0) +
-    stat_smooth(mapping = aes(colour = sex)) +
-       facet_grid(x ~ sex)
-
-  ex_SCO + labs(x = "Year", y = "Annual change in life expectancy", title = "Scotland", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
-```
-
-### France
-
-```{r, echo = FALSE, warning=FALSE, message = FALSE}
-#FRA female at birth
-f_e0_FRA <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 0) %>% 
-      filter(sex == "female") %>% 
-      filter(code == "FRATNP") %>% 
-      filter(year >= 1979)
-
-# Now we create the change in ex variable as that's the response variable in the model
-  f_e0_FRA <- f_e0_FRA %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex)) %>% 
-   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
-   geom_line() +
-   geom_hline(yintercept = 0) + 
-   stat_smooth(colour = "lightpink3")
-
-  #FRA male at birth
-m_e0_FRA <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 0) %>% 
-      filter(sex == "male") %>% 
-      filter(code == "FRATNP") %>% 
-      filter(year >= 1979)
-
-  m_e0_FRA <- m_e0_FRA %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex)) %>% 
-   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
-   geom_line() +
-   geom_hline(yintercept = 0) + 
-   stat_smooth(colour = "aquamarine4")
-
-  #FRA female at age 65
-f_e65_FRA <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 65) %>% 
-      filter(sex == "female") %>% 
-      filter(code == "FRATNP") %>% 
-      filter(year >= 1979)
-
-# Now we create the change in ex variable as that's the response variable in the model
-  f_e65_FRA <- f_e65_FRA %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex)) %>% 
-   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
-   geom_line() +
-   geom_hline(yintercept = 0) + 
-   stat_smooth(colour = "lightpink3")
-
-  #FRA male at aged 65
-m_e65_FRA <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 65) %>% 
-      filter(sex == "male") %>% 
-      filter(code == "FRATNP") %>% 
-      filter(year >= 1979)
-
- m_e65_FRA <-  m_e65_FRA %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex)) %>% 
-   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
-   geom_line() +
-   geom_hline(yintercept = 0) + 
-   stat_smooth(colour = "aquamarine4", show.legend = TRUE)
-
-  #add labels
-f_e0_FRA + labs(y = "Annual change in e0", title = "France", subtitle = "Life expectancy changes at birth for females", caption = "Source: Human Mortality Database")
- m_e0_FRA + labs(y = "Annual change in e0", title = "France", subtitle = "Life expectancy changes at birth for males", caption = "Source: Human Mortality Database")
- f_e65_FRA + labs(y = "Annual change in e65", title = "France", subtitle = "Life expectancy changes at aged 65 for females", caption = "Source: Human Mortality Database")
- m_e65_FRA + labs(y = "Annual change in e65", title = "France", subtitle = "Life expectancy changes at aged 65 for males", caption = "Source: Human Mortality Database")
-
- ex_FRA <- hmd_ex_selected_countries_with_synth %>%
-      filter(sex %in% c("female", "male")) %>%
-      filter(code == "FRATNP") %>%
-    group_by(sex, x) %>% arrange(year) %>%
-      mutate(delta_ex = ex - lag(ex)) %>%
-    ungroup %>%
-    filter(year >= 1979) %>%
-    ggplot(aes(x = year, y = delta_ex)) + geom_point(size=0.3) +
-   geom_line() +
-   geom_hline(yintercept = 0) +
-    stat_smooth(mapping = aes(colour = sex)) +
-       facet_grid(x ~ sex)
-
-  ex_FRA + labs(x = "Year", y = "Annual change in life expectancy", title = "France", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
-```
-
-### Spain
-
-```{r, echo = FALSE, warning=FALSE, message = FALSE}
-#ESP female at birth
-f_e0_ESP <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 0) %>% 
-      filter(sex == "female") %>% 
-      filter(code == "ESP") %>% 
-      filter(year >= 1979)
-
-# Now we create the change in ex variable as that's the response variable in the model
-  f_e0_ESP <- f_e0_ESP %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex)) %>% 
-   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
-   geom_line() +
-   geom_hline(yintercept = 0) + 
-   stat_smooth(colour = "lightpink3")
-
-  #ESP male at birth
-m_e0_ESP <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 0) %>% 
-      filter(sex == "male") %>% 
-      filter(code == "ESP") %>% 
-      filter(year >= 1979)
-
-  m_e0_ESP <- m_e0_ESP %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex)) %>% 
-   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
-   geom_line() +
-   geom_hline(yintercept = 0) + 
-   stat_smooth(colour = "aquamarine4")
-
-  #ESP female at age 65
-f_e65_ESP <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 65) %>% 
-      filter(sex == "female") %>% 
-      filter(code == "ESP") %>% 
-      filter(year >= 1979)
-
-# Now we create the change in ex variable as that's the response variable in the model
-  f_e65_ESP <- f_e65_ESP %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex)) %>% 
-   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
-   geom_line() +
-   geom_hline(yintercept = 0) + 
-   stat_smooth(colour = "lightpink3")
-
-  #ESP male at aged 65
-m_e65_ESP <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 65) %>% 
-      filter(sex == "male") %>% 
-      filter(code == "ESP") %>% 
-      filter(year >= 1979)
-
- m_e65_ESP <-  m_e65_ESP %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex)) %>% 
-   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
-   geom_line() +
-   geom_hline(yintercept = 0) + 
-   stat_smooth(colour = "aquamarine4", show.legend = TRUE)
-
-  #add labels
-f_e0_ESP + labs(y = "Annual change in e0", title = "Spain", subtitle = "Life expectancy changes at birth for females", caption = "Source: Human Mortality Database")
- m_e0_ESP + labs(y = "Annual change in e0", title = "Spain", subtitle = "Life expectancy changes at birth for males", caption = "Source: Human Mortality Database")
- f_e65_ESP + labs(y = "Annual change in e65", title = "Spain", subtitle = "Life expectancy changes at aged 65 for females", caption = "Source: Human Mortality Database")
- m_e65_ESP + labs(y = "Annual change in e65", title = "Spain", subtitle = "Life expectancy changes at aged 65 for males", caption = "Source: Human Mortality Database")
-
- ex_ESP <- hmd_ex_selected_countries_with_synth %>%
-      filter(sex %in% c("female", "male")) %>%
-      filter(code == "ESP") %>%
-    group_by(sex, x) %>% arrange(year) %>%
-      mutate(delta_ex = ex - lag(ex)) %>%
-    ungroup %>%
-    filter(year >= 1979) %>%
-    ggplot(aes(x = year, y = delta_ex)) + geom_point(size=0.3) +
-   geom_line() +
-   geom_hline(yintercept = 0) +
-    stat_smooth(mapping = aes(colour = sex)) +
-       facet_grid(x ~ sex)
-
-  ex_ESP + labs(x = "Year", y = "Annual change in life expectancy", title = "Spain", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
-```
-
-### Italy
-
-```{r, echo = FALSE, warning=FALSE, message = FALSE}
-#ITA female at birth
-f_e0_ITA <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 0) %>% 
-      filter(sex == "female") %>% 
-      filter(code == "ITA") %>% 
-      filter(year >= 1979)
-
-# Now we create the change in ex variable as that's the response variable in the model
-  f_e0_ITA <- f_e0_ITA %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex)) %>% 
-   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
-   geom_line() +
-   geom_hline(yintercept = 0) + 
-   stat_smooth(colour = "lightpink3")
-
-  #ITA male at birth
-m_e0_ITA <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 0) %>% 
-      filter(sex == "male") %>% 
-      filter(code == "ITA") %>% 
-      filter(year >= 1979)
-
-  m_e0_ITA <- m_e0_ITA %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex)) %>% 
-   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
-   geom_line() +
-   geom_hline(yintercept = 0) + 
-   stat_smooth(colour = "aquamarine4")
-
-  #ITA female at age 65
-f_e65_ITA <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 65) %>% 
-      filter(sex == "female") %>% 
-      filter(code == "ITA") %>% 
-      filter(year >= 1979)
-
-# Now we create the change in ex variable as that's the response variable in the model
-  f_e65_ITA <- f_e65_ITA %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex)) %>% 
-   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
-   geom_line() +
-   geom_hline(yintercept = 0) + 
-   stat_smooth(colour = "lightpink3")
-
-  #ITA male at aged 65
-m_e65_ITA <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 65) %>% 
-      filter(sex == "male") %>% 
-      filter(code == "ITA") %>% 
-      filter(year >= 1979)
-
- m_e65_ITA <-  m_e65_ITA %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex)) %>% 
-   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
-   geom_line() +
-   geom_hline(yintercept = 0) + 
-   #Trying to add legend
-   #stat_smooth(aes(colour = "aquamarine4"), show.legend = TRUE, labels = "Smoothed")
-  stat_smooth(colour = "aquamarine4")
-
-  #add labels
-f_e0_ITA + labs(y = "Annual change in e0", title = "Italy", subtitle = "Life expectancy changes at birth for females", caption = "Source: Human Mortality Database")
- m_e0_ITA + labs(y = "Annual change in e0", title = "Italy", subtitle = "Life expectancy changes at birth for males", caption = "Source: Human Mortality Database")
- f_e65_ITA + labs(y = "Annual change in e65", title = "Italy", subtitle = "Life expectancy changes at aged 65 for females", caption = "Source: Human Mortality Database")
- m_e65_ITA + labs(y = "Annual change in e65", title = "Italy", subtitle = "Life expectancy changes at aged 65 for males", caption = "Source: Human Mortality Database")
-
- ex_ITA <- hmd_ex_selected_countries_with_synth %>%
-      filter(sex %in% c("female", "male")) %>%
-      filter(code == "ITA") %>%
-    group_by(sex, x) %>% arrange(year) %>%
-      mutate(delta_ex = ex - lag(ex)) %>%
-    ungroup %>%
-    filter(year >= 1979) %>%
-    ggplot(aes(x = year, y = delta_ex)) + geom_point(size=0.3) +
-   geom_line() +
-   geom_hline(yintercept = 0) +
-    stat_smooth(mapping = aes(colour = sex)) +
-       facet_grid(x ~ sex)
-
-  ex_ITA + labs(x = "Year", y = "Annual change in life expectancy", title = "Italy", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
-```
-
-### Germany: Synthetic
-
-```{r, echo = FALSE, warning=FALSE, message = FALSE}
-#DEUTS female at birth
-f_e0_DEUTS <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 0) %>% 
-      filter(sex == "female") %>% 
-      filter(code == "DEUTSYNTH") %>% 
-      filter(year >= 1979)
-
-# Now we create the change in ex variable as that's the response variable in the model
-  f_e0_DEUTS <- f_e0_DEUTS %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex)) %>% 
-   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
-   geom_line() +
-   geom_hline(yintercept = 0) + 
-   stat_smooth(colour = "lightpink3")
-
-  #DEUTS male at birth
-m_e0_DEUTS <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 0) %>% 
-      filter(sex == "male") %>% 
-      filter(code == "DEUTSYNTH") %>% 
-      filter(year >= 1979)
-
-  m_e0_DEUTS <- m_e0_DEUTS %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex)) %>% 
-   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
-   geom_line() +
-   geom_hline(yintercept = 0) + 
-   stat_smooth(colour = "aquamarine4")
-
-  #DEUTS female at age 65
-f_e65_DEUTS <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 65) %>% 
-      filter(sex == "female") %>% 
-      filter(code == "DEUTSYNTH") %>% 
-      filter(year >= 1979)
-
-# Now we create the change in ex variable as that's the response variable in the model
-  f_e65_DEUTS <- f_e65_DEUTS %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex)) %>% 
-   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
-   geom_line() +
-   geom_hline(yintercept = 0) + 
-   stat_smooth(colour = "lightpink3")
-
-  #DEUTS male at aged 65
-m_e65_DEUTS <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 65) %>% 
-      filter(sex == "male") %>% 
-      filter(code == "DEUTSYNTH") %>% 
-      filter(year >= 1979)
-
- m_e65_DEUTS <-  m_e65_DEUTS %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex)) %>% 
-   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
-   geom_line() +
-   geom_hline(yintercept = 0) + 
-   stat_smooth(colour = "aquamarine4", show.legend = TRUE)
-
-  #add labels
-f_e0_DEUTS + labs(y = "Annual change in e0", title = "Germany: Synthetic", subtitle = "Life expectancy changes at birth for females", caption = "Source: Human Mortality Database")
- m_e0_DEUTS + labs(y = "Annual change in e0", title = "Germany: Synthetic", subtitle = "Life expectancy changes at birth for males", caption = "Source: Human Mortality Database")
- f_e65_DEUTS + labs(y = "Annual change in e65", title = "Germany: Synthetic", subtitle = "Life expectancy changes at aged 65 for females", caption = "Source: Human Mortality Database")
- m_e65_DEUTS + labs(y = "Annual change in e65", title = "Germany: Synthetic", subtitle = "Life expectancy changes at aged 65 for males", caption = "Source: Human Mortality Database")
-
- #Grid
- ex_DEUTSYNTH <- hmd_ex_selected_countries_with_synth %>%
-      filter(sex %in% c("female", "male")) %>%
-      filter(code == "DEUTSYNTH") %>%
-    group_by(sex, x) %>% arrange(year) %>%
-      mutate(delta_ex = ex - lag(ex)) %>%
-    ungroup %>%
-    filter(year >= 1979) %>%
-    ggplot(aes(x = year, y = delta_ex)) + geom_point(size=0.3) +
-   geom_line() +
-   geom_hline(yintercept = 0) +
-    stat_smooth(mapping = aes(colour = sex)) +
-       facet_grid(x ~ sex)
-
-  ex_DEUTSYNTH + labs(x = "Year", y = "Annual change in life expectancy", title = "Germany: Synthetic", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
-```
-
-### Germany: East
-
-```{r, echo = FALSE, warning=FALSE, message = FALSE}
-#DEUTE female at birth
-f_e0_DEUTE <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 0) %>% 
-      filter(sex == "female") %>% 
-      filter(code == "DEUTE") %>% 
-      filter(year >= 1979)
-
-# Now we create the change in ex variable as that's the response variable in the model
-  f_e0_DEUTE <- f_e0_DEUTE %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex)) %>% 
-   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
-   geom_line() +
-   geom_hline(yintercept = 0) + 
-   stat_smooth(colour = "lightpink3")
-
-  #DEUTE male at birth
-m_e0_DEUTE <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 0) %>% 
-      filter(sex == "male") %>% 
-      filter(code == "DEUTE") %>% 
-      filter(year >= 1979)
-
-  m_e0_DEUTE <- m_e0_DEUTE %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex)) %>% 
-   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
-   geom_line() +
-   geom_hline(yintercept = 0) + 
-   stat_smooth(colour = "aquamarine4")
-
-  #DEUTE female at age 65
-f_e65_DEUTE <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 65) %>% 
-      filter(sex == "female") %>% 
-      filter(code == "DEUTE") %>% 
-      filter(year >= 1979)
-
-# Now we create the change in ex variable as that's the response variable in the model
-  f_e65_DEUTE <- f_e65_DEUTE %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex)) %>% 
-   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
-   geom_line() +
-   geom_hline(yintercept = 0) + 
-   stat_smooth(colour = "lightpink3")
-
-  #DEUTE male at aged 65
-m_e65_DEUTE <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 65) %>% 
-      filter(sex == "male") %>% 
-      filter(code == "DEUTE") %>% 
-      filter(year >= 1979)
-
- m_e65_DEUTE <-  m_e65_DEUTE %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex)) %>% 
-   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
-   geom_line() +
-   geom_hline(yintercept = 0) + 
-   stat_smooth(colour = "aquamarine4", show.legend = TRUE)
-
-  #add labels
-f_e0_DEUTE + labs(y = "Annual change in e0", title = "Germany: East", subtitle = "Life expectancy changes at birth for females", caption = "Source: Human Mortality Database")
- m_e0_DEUTE + labs(y = "Annual change in e0", title = "Germany: East", subtitle = "Life expectancy changes at birth for males", caption = "Source: Human Mortality Database")
- f_e65_DEUTE + labs(y = "Annual change in e65", title = "Germany: East", subtitle = "Life expectancy changes at aged 65 for females", caption = "Source: Human Mortality Database")
- m_e65_DEUTE + labs(y = "Annual change in e65", title = "Germany: East", subtitle = "Life expectancy changes at aged 65 for males", caption = "Source: Human Mortality Database")
-
- ex_DEUTE <- hmd_ex_selected_countries_with_synth %>%
-      filter(sex %in% c("female", "male")) %>%
-      filter(code == "DEUTE") %>%
-    group_by(sex, x) %>% arrange(year) %>%
-      mutate(delta_ex = ex - lag(ex)) %>%
-    ungroup %>%
-    filter(year >= 1979) %>%
-    ggplot(aes(x = year, y = delta_ex)) + geom_point(size=0.3) +
-   geom_line() +
-   geom_hline(yintercept = 0) +
-    stat_smooth(mapping = aes(colour = sex)) +
-       facet_grid(x ~ sex)
-
-  ex_DEUTE + labs(x = "Year", y = "Annual change in life expectancy", title = "Germany: East", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
-```
-
-### Germany: West
-
-```{r, echo = FALSE, warning=FALSE, message = FALSE}
-#DEUTW female at birth
-f_e0_DEUTW <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 0) %>% 
-      filter(sex == "female") %>% 
-      filter(code == "DEUTW") %>% 
-      filter(year >= 1979)
-
-# Now we create the change in ex variable as that's the response variable in the model
-  f_e0_DEUTW <- f_e0_DEUTW %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex)) %>% 
-   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
-   geom_line() +
-   geom_hline(yintercept = 0) + 
-   stat_smooth(colour = "lightpink3")
-
-  #DEUTW male at birth
-m_e0_DEUTW <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 0) %>% 
-      filter(sex == "male") %>% 
-      filter(code == "DEUTW") %>% 
-      filter(year >= 1979)
-
-  m_e0_DEUTW <- m_e0_DEUTW %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex)) %>% 
-   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
-   geom_line() +
-   geom_hline(yintercept = 0) + 
-   stat_smooth(colour = "aquamarine4")
-
-  #DEUTW female at age 65
-f_e65_DEUTW <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 65) %>% 
-      filter(sex == "female") %>% 
-      filter(code == "DEUTW") %>% 
-      filter(year >= 1979)
-
-# Now we create the change in ex variable as that's the response variable in the model
-  f_e65_DEUTW <- f_e65_DEUTW %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex)) %>% 
-   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
-   geom_line() +
-   geom_hline(yintercept = 0) + 
-   stat_smooth(colour = "lightpink3")
-
-  #DEUTW male at aged 65
-m_e65_DEUTW <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 65) %>% 
-      filter(sex == "male") %>% 
-      filter(code == "DEUTW") %>% 
-      filter(year >= 1979)
-
- m_e65_DEUTW <-  m_e65_DEUTW %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex)) %>% 
-   ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
-   geom_line() +
-   geom_hline(yintercept = 0) + 
-   stat_smooth(colour = "aquamarine4", show.legend = TRUE)
-
-  #add labels
-f_e0_DEUTW + labs(y = "Annual change in e0", title = "Germany: West", subtitle = "Life expectancy changes at birth for females", caption = "Source: Human Mortality Database")
- m_e0_DEUTW + labs(y = "Annual change in e0", title = "Germany: West", subtitle = "Life expectancy changes at birth for males", caption = "Source: Human Mortality Database")
- f_e65_DEUTW + labs(y = "Annual change in e65", title = "Germany: West", subtitle = "Life expectancy changes at aged 65 for females", caption = "Source: Human Mortality Database")
- m_e65_DEUTW + labs(y = "Annual change in e65", title = "Germany: West", subtitle = "Life expectancy changes at aged 65 for males", caption = "Source: Human Mortality Database")
-
- ex_DEUTW <- hmd_ex_selected_countries_with_synth %>%
-      filter(sex %in% c("female", "male")) %>%
-      filter(code == "DEUTW") %>%
-    group_by(sex, x) %>% arrange(year) %>%
-      mutate(delta_ex = ex - lag(ex)) %>%
-    ungroup %>%
-    filter(year >= 1979) %>%
-    ggplot(aes(x = year, y = delta_ex)) + geom_point(size=0.3) +
-   geom_line() +
-   geom_hline(yintercept = 0) +
-    stat_smooth(mapping = aes(colour = sex)) +
-       facet_grid(x ~ sex)
-
-  ex_DEUTW + labs(x = "Year", y = "Annual change in life expectancy", title = "Germany: West", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
+```{r}
+source(here("R", "annual_change_graphs_by_population.R"))
+source(here("R", "breakpoint_functions.R"))
 ```
@@
 
-```{r, cache=TRUE}
-dta_for_decade_comparison <- 
-  hmd_ex_selected_countries_with_synth %>% 
-      filter(x == 0) %>% 
-      filter(sex == "female") %>% 
-      filter(code == "GBRTENW") %>% 
-      filter(year >= 1979) %>% 
-      # ggplot(aes(x = year, y = ex)) +
-      #   geom_line()
-      arrange(year) %>% 
-      mutate(delta_ex = ex - lag(ex)) %>% 
-    #   ggplot(aes(x = year, y = delta_ex)) + 
-    # geom_point() + geom_line() +
-    # geom_hline(yintercept = 0)
-      mutate(
-        decade = year - (year %% 10)
-      ) %>% 
-  filter(!is.na(delta_ex))
-model_null <- 
-  lm(delta_ex ~ 1, data = dta_for_decade_comparison)
-model_alt <- 
-  lm(delta_ex ~ factor(decade), data= dta_for_decade_comparison)
-summary(model_null)
-summary(model_alt)
-anova(model_null, model_alt)
-# to get the value of interest do tmp[2,6]
-# Generalising 
-tidy_dataset <- function(df){
-  df %>% 
-      filter(year >= 1979) %>% 
-      arrange(year) %>% 
-      mutate(delta_ex = ex - lag(ex)) %>% 
-      mutate(
-        decade = year - (year %% 10)
-      ) %>% 
-  filter(!is.na(delta_ex))
-}
-make_null_model <- function(df){
-  lm(delta_ex ~ 1, data = df)
-}
-all_null_alt_models <-
-  hmd_ex_selected_countries_with_synth %>%
-    group_by(x, sex, code) %>% 
-    nest() %>% 
-    mutate(
-      prep_data = map(data, tidy_dataset) 
-    ) %>% 
-    mutate(
-      model_null = map(prep_data, ~lm(delta_ex ~ 1, data = .)),
-      model_alt  = map(prep_data, ~lm(delta_ex ~ factor(decade), data = .))
-  #    model_null = map(prep_data, make_null_model)
-    ) %>% 
-    mutate(
-      anova_results = map2(model_null, model_alt, anova)
-    ) %>% 
-    mutate(
-      anova_pval = map_dbl(anova_results, ~.[2,6])
-    ) 
-all_null_alt_models %>% 
-  filter(anova_pval < 0.05) -> tmp
-summary(tmp$model_alt[[1]])
-summary(tmp$model_alt[[2]])
-  # build null model
-  # build alt model
-  # do anova
-  # get p value from anova
-
-```
-
-### Joinpoint/Segmented regression approach
+```{r echo = FALSE}
+source(here("R", "make_decadal_change_models.R"))
+
+```
+
+The following combinations of country, starting age, and sex indicate statistically signifanct improvements in model fit if using a model which allows the annual changes to vary by decade
+
+```{r}
+all_null_decadal_models %>% 
+  filter(anova_pval < 0.05) %>% 
+  arrange(desc(anova_pval)) %>% 
+  select(code, x, sex, anova_pval)
+```
+* East Germany, Females, e0
+* East Germany, Males, e65
+
+For these two populations, the decadal model has the following coefficients:
+
+```{r}
+all_null_decadal_models %>% 
+  filter(anova_pval < 0.05) %>% 
+  arrange(desc(anova_pval)) %>% 
+  select(code, x, sex, model_alt) %>% 
+  mutate(
+    model_results = map(model_alt, broom::tidy)
+  ) %>% 
+  select(-model_alt) %>% 
+  unnest(model_results) %>% 
+  mutate(
+    term  = case_when(
+      term == "(Intercept)" ~ "Baseline (1980s)",
+      term == "factor(decade)1990" ~ "1990s",
+      term == "factor(decade)2000" ~ "2000s",
+      term == "factor(decade)2010" ~ "2010s",
+      TRUE ~ NA_character_
+    )
+  ) %>% 
+  mutate(
+    label = glue::glue("{round(estimate, 3)}{case_when(p.value < 0.01 ~ '***', p.value < 0.05 ~ '**', p.value < 0.10 ~ '*', TRUE ~ '')} ({round(std.error, 3)})")
+  ) %>% 
+  ungroup() %>% 
+  select(
+    x, Term = term, label
+  ) %>% 
+  pivot_wider(values_from = label, names_from = x) %>% 
+  kable(format = "html", table.attr = "style=\"color: black;\"",caption = "Coefficients of decadal differences, East Germany, females, by starting age (0 = life expectancy at birth; 65 = life expectancy at age 65 years") %>%
+  kableExtra::kable_styling() %>% 
+  footnote(general = "The baseline is the average annual rate of improvement in the 1980s.
+The other terms indicate increments or decrements from this baseline.
+Statistical significance of coefficients: * if p < 0.10, ** if p < 0.05, *** if p < 0.01"
+  )
+```
+
+The above table indicates that, in females in East Germany, the 1990s was associated with a substantial increase in annual life expectancy increases, at both birth ($e_0$) and age 65 years ($e_{65}$).
+These increases throughout the 1990s were greater than a fifth of a year per year on average.
+
+## Changepoint and breakpoint analysis.
 
@@
 Instead of looking for systemic differences by decade, we can instead look at whether there are any systemic and detectable 'breaks' in the trends over time.
-If our data were life expectancy over time, this could be achieved through joinpoint analysis, in which the slope is allowed to change over time.
-As instead are data are annual changes in life expectancy, we are interested in detecting changes in intercept over time.
-If we define our Null (no change) model as follows:
+We should also be mindful that, over the period under consideration, there may potentially be no, one, or more distinct breaks in the series.
+
+To investigate this, we use two approaches to identify both whether and how many breaks may exist in the life expectancy trend data.
+
+* Approach one: breakpoint analysis on life expectancy against time 
+* Approach two: changepoint analysis on life expectancy change rates against time.
+
+For both approaches, the underlying rationale is the same:
+
+* We first fit a 'null' (no change) model to the data. These are models that represent the assumption that no change occurred in trends in life expectancy over time. 
+* Then we fit a series on one-break models to the data.
+These are models that represent the assumption that a single significant change occurred to the series over time.
+From this series of one-break models, we attempt to find the best one-break model.
+* We then fit a series of two-break models to the data.
+These are models that represent the assumption that were were two significant changes to the series over time.
+From this series of two-break models, we attempt to find the best two-break models.
+* We now have three models to compare: the null model ($M_0$), the best one-break model ($M_1$), and the best two-break model ($M_2$).
+Although both $M_1$ can be compared directly^[By 'compared directly', we mean that one model specification can be expressed as a restricted/constrained version of another model, the unrestricted model, with one or more terms in the unrestricted model set to fixed values, usually zero, in the restricted model.
+Such models can be compared directly using an F-test] with $M_0$, and $M_2$ can be compared directly with $M_0$, $M_2$ cannot necessarily be compared directly with $M_1$.
+Instead, the triplet of models ($M_0$, $M_1$ and $M_2$) are compared indirectly using BIC^[BIC stands for Bayesian Information Criterion, like the similar AIC (An Information Criterion or Akaike's Information Criterion), is a penalised model fit score.
+By penalised this means that the fit of the model to the data (more specifically its log likelihood) is calculated, then a 'penalty' is applied to this score based on the complexity of the model.
+BIC and AIC differ only according to how the penalty is applied, with BIC tending to penalise more complex models more severely than AIC.
+This means BIC will tend to be more conservative in selecting models, providing some protection against overfitting.
+Both AIC and BIC can be used to compare both nested and non-nested models based on the same dataset, unlike the F-test, the Lagrange Multiplier Test, and so on.
+Measures of model fit should *not* be used to compare models fit to different datasets.^[
+For example, it would be wrong to conclude that a model with an $R^2$ or adjusted $R^2$ fit to one dataset of 0.80 is 'better' than a model with an $R^2$ or adjusted $R^2$ of 0.50 fit to a different dataset, even though $R^2$ and its variants are often (mis)interpreted in this way
+], which like AIC provides a penalised model fit score.
+Lower BIC scores indicate better fit to the data, and so, for each dataset, the model ($M_0$, $M_1$ or $M_2$) with the lowest BIC will be selected.
+* For the best of the three models for each dataset, the breakpoint or breakpoints (if any), and the model predictions, will be presented and visualised.
+
+Breaks in the data are calculated through two approaches:
+
+1.
+For life expectancy against time ($e_x \sim t$), breakpoints are calculated using the $segmented$ function in the $segmented$ package.
+Both one and two breakpoint models are calculated using this package.
+2. For changes in life expectancy against time ($ \frac{de_x}{dt} \sim t$), regressions which allow varying intercepts are calculated. More specifically, the models being compared are: 
+    * **$M_0$**: $\frac{de_x}{dt} \sim \alpha$
+    * **$M_1$**: $\frac{de_x}{dt} \sim \alpha + \beta T$, where
 
 $$
-M_0 : y(t) = \alpha + e
+T = \left\{
+    \begin{array}\\
+        1 & \mbox{if } \ t \ge \tau \\
+        0 & \mbox{otherwise}
+    \end{array}
+\right.
 $$
-Where $e \backsim N(0, \sigma^2)$. i.e. all observations $y$ are draws from $N(\alpha, \sigma^2)$. The alternative model specification against which this is compared is then:
+    * **M_2**: $\frac{de_x}{dt} \sim \alpha + \beta_1 T_1 + \beta_2 T_2$, where
 
 $$
-M_A : y(t) = \alpha_0 + \beta_1 T(t) + e 
+T_1 = \left\{
+    \begin{array}\\
+        1 & \mbox{if } \tau_1 \le t \lt \tau_2 \\
+        0 & \mbox{otherwise}
+    \end{array}
+\right.
 $$
-Where $T(t)$ is defined as $0$ where $t \lt \tau$ and $1$ where $t \geq \tau$. ($\alpha_0$ is used in $M_A$ to indicate that this parameter is not directly comparable with $\alpha$ in $M_0$).
-
-This specification shows that there is an additional parameter $\tau$ to be determined in order to produce $M_A$.
-This $\tau$ parameter is the breakpoint in the series.
-The selection of $\tau$ will be determined by numeric optimisation, by selecting the value of $\tau$ that minimised AIC.
-
-Once the best possible $M_A$, using the best value of $\tau$, is selected, $M_A$ will be compared with $M_0$ using an F-test, allowing us to determine whether there is convincing evidence of any specific break in the data.
-
-We will start with a single country, starting age, and sex, and with manual selection of $\tau$, then generalise and automate further.
-
-```{r}
-# We can now run the above function, which creates a model, for a wide range of possible tau values
-# This is the sequence listed and named tau_candidate, which is every whole number from 
-# 1982 to 2015
-# For each of these models, we can then get a measure of model fit. We're going to use AIC, though 
-# could have used other measures. With AIC (and with BIC, and RMSE etc) better fit is indicated with 
-# a lower value
-
-source(here("R", "breakpoint_functions.R"))
-
-# Let's see if this now works in the functional programming context, using 
-# nest and tibble elements 
-hmd_ex_selected_countries_with_synth %>%
+and
+
+$$
+T_2 = \left\{
+    \begin{array}\\
+        1 & \mbox{if } t \ge \tau_2 \\
+        0 & \mbox{otherwise}
+    \end{array}
+\right.
+$$
+i.e.
+the selection of $\tau_1$ and $\tau_2$ partitions the model into three sections: $\frac{de_x}{dt} \sim \alpha$ where $t \lt \tau_1$, $\frac{de_x}{dt} \sim \alpha + \beta_1$ where $\tau_1 \le t \lt \tau_2$, and $\frac{de_x}{dt} \sim \alpha + \beta_2$ where $t \ge \tau_2$.
+As $\tau_1 \lt \tau_2$, only values of $\tau_2$ which are greater than $\tau_1$ are considered.
+Otherwise all whole integer values of $\tau$ are seached through for both $M_1$ and $M_2$.
+
+For any particular population $D_i$, we should generally expect a high degree of concordance between the breakpoint or breakpoints identified by the `segmented` package, and the changepoint or changepoints identified using annual changes in life expectancy against time.
+However, we should not expect the results from the two approaches to be identical, due to differences in the algorithms used to select the best model parameters.
+
+The above analyses are performed for each population group, comprising different combinations of country, sex, and starting age.
+
+# A slight pivot
+
+I've realised a better use of the segmented package may be to identify and compare zero breakpoint, one breakpoint, and two breakpoint models. The code below will do this
+
+```{r}
+
+estimate_breakpoints_and_pval <- function(df){
+  null_mdl <- lm(ex ~ year, data = df) 
+  seg_mdl <- segmented::segmented(null_mdl, seg.Z= ~year, psi = 2010)
+  seg2_mdl <- segmented::segmented(null_mdl, seg.z= ~year, psi= c(1985, 2010)) # added to test
+
+  list(
+    null = null_mdl, 
+    seg = seg_mdl,
+    seg2 = seg2_mdl
+  )
+}
+
+segmented_breakpoints_models <- 
+  hmd_ex_selected_countries_with_synth %>% 
+  filter(code != "DEUTNP") %>%   
   filter(year >= 1979) %>% 
@@
   mutate(
-    best_tau = map_dbl(data, get_best_tau_from_optim)
-  )
-# I think this is getting stuck on local optima, so instead maybe 
-# the more exhaustive search should be used after all! 
-# There are two alternative approaches: 
-# 1) Use an exhaustive grid search approach 
-# 2) Use an algorithm which is less likely to get stuck in local optima
-# The first approach is much more sensible, but the second might be a cool
-# stretch goal as well! 
-# let's use a heavily modified version of the above function to do the grid search
-
-compare_best_breakmodel_with_null <- 
-  hmd_ex_selected_countries_with_synth %>%
+    mdl_outputs = map(data, estimate_breakpoints_and_pval)
+  ) %>% 
+  unnest_longer(mdl_outputs)
+
+# Now let's get the BIC for each model
+
+make_predictions <- function(mdl, dta = tibble(year = 1980:2020)){
+  tibble(
+    year = dta$year, 
+    ex_pred = predict(mdl, newdata = dta)
+  )
+}
+
+best_model_predictions_descriptions <- 
+segmented_breakpoints_models %>% 
+  mutate(bic = map_dbl(mdl_outputs, BIC)) %>% 
+  group_by(code, x, sex) %>% 
+  mutate(rank_bic = rank(bic)) %>%
+  filter(rank_bic == 1) %>% 
+  mutate(
+    best_model = case_when(
+      mdl_outputs_id == 'seg2' ~ "Two breakpoints",
+      mdl_outputs_id == 'seg'  ~ "One breakpoint", 
+      mdl_outputs_id == 'null' ~ "No breakpoints"
+    )
+  ) %>% 
+  mutate(
+    first_breakpoint =    map2_dbl(
+      mdl_outputs_id, mdl_outputs, 
+      function(x, y){
+        if (x == 'null'  ) {NA_real_} else {
+          y[["psi"]][1,2]
+        }
+      }                        
+      ),
+      first_breakpoint_se = map2_dbl(
+        mdl_outputs_id, mdl_outputs,
+        function(x, y){
+        if (x == 'null'  ) {NA_real_} else {
+          y[["psi"]][1,3]
+        }
+
+        }
+      ),
+    second_breakpoint = map2_dbl(
+      mdl_outputs_id, mdl_outputs,
+      function(x, y){
+        if (x == 'seg2') {
+          y[["psi"]][2,2]
+        } else  {
+            NA_real_
+          }
+      }                           
+      ),
+    second_breakpoint_se = map2_dbl(
+      mdl_outputs_id, mdl_outputs,
+      function(x, y){
+        if (x == 'seg2') {
+          y[["psi"]][2,3]
+        } else  {
+            NA_real_
+          }
+      }
+    )
+  ) %>%  # let's add predictions too 
+  mutate(
+    pred_data = map(mdl_outputs, make_predictions),
+    joined_data = map2(data, pred_data, left_join)
+  ) %>% 
+  select(code, x, sex, joined_data, first_breakpoint:second_breakpoint_se) %>% 
+  unnest(joined_data)
+
+```
+
+Let's visualise
+
+```{r}
+best_model_predictions_descriptions %>% 
+  filter(x == 0) %>% 
+  ggplot(aes(x = year, colour = sex)) + 
+  geom_point(aes(y = ex, shape = sex), alpha = 0.25) + 
+  geom_line(aes(y = ex_pred)) + 
+  facet_wrap(~code) +
+  geom_vline(aes(xintercept = first_breakpoint, colour = sex)) + 
+  geom_vline(aes(xintercept = second_breakpoint, colour = sex)) + 
+  labs(
+    x = "Year", y = "Life expectancy at birth (years)",
+    title = "Breakpoint models for selected countries"
+  )
+
+# Or monochrome
+
+best_model_predictions_descriptions %>% 
+  filter(x == 0) %>% 
+  ggplot(aes(x = year)) + 
+  geom_point(aes(y = ex), alpha = 0.25) + 
+  geom_line(aes(y = ex_pred)) + 
+  facet_grid(sex~code) +
+  geom_vline(aes(xintercept = first_breakpoint)) + 
+  geom_vline(aes(xintercept = second_breakpoint)) + 
+  labs(
+    x = "Year", y = "Life expectancy at birth (years)",
+    title = "e0, Breakpoint models for selected countries"
+  )
+
+ggsave(here("figures", "best_0to2bp_e0.png"), height = 20, width = 30, units = "cm", dpi = 150)
+
+```
+
+I think the monochrome display works better, as there's too many vertical lines otherwise.
+
+And the same for conditional life expectancy
+
+```{r}
+best_model_predictions_descriptions %>% 
+  filter(x == 65) %>% 
+  ggplot(aes(x = year)) + 
+  geom_point(aes(y = ex), alpha = 0.25) + 
+  geom_line(aes(y = ex_pred)) + 
+  facet_grid(sex~code) +
+  geom_vline(aes(xintercept = first_breakpoint)) + 
+  geom_vline(aes(xintercept = second_breakpoint)) + 
+  labs(
+    x = "Year", y = "Life expectancy at age 65 (years)",
+    title = "e65, Breakpoint models for selected countries"
+  )
+
+ggsave(here("figures", "best_0to2bp_e65.png"), height = 20, width = 30, units = "cm", dpi = 150)
+
+```
+I think we can show how the gradients are predicted to change by differencing again
+
+```{r}
+best_model_predictions_descriptions %>% 
+  filter(x == 0) %>% 
+  group_by(code, x, sex) %>% 
+  arrange(year) %>% 
+  mutate(
+    ch_ex = ex - lag(ex),
+    ch_ex_pred = ex_pred - lag(ex_pred)
+    ) %>% 
+  ungroup() %>% 
+  ggplot(aes(x = year)) + 
+  geom_point(aes(y = ch_ex), alpha = 0.25) + 
+  geom_step(aes(y = ch_ex_pred), direction = "vh") + 
+  facet_grid(sex~code) + 
+  geom_hline(yintercept = 0) + 
+  geom_ribbon(
+    aes(
+      ymin = ifelse(ch_ex_pred < 0, ch_ex_pred, 0), 
+      ymax = ifelse(ch_ex_pred < 0, 0, ch_ex_pred),
+      fill = ch_ex_pred < 0
+      ),
+    alpha = 0.2
+    ) + 
+  scale_fill_manual(values = c(`TRUE` = 'red', `FALSE` = 'blue')) +
+  guides(fill = "none") + 
+  labs(
+    x = "Year", y = "Annual change in life expectancy at birth",
+    title = "Annual change in life expectancy at birth, with implied changes from breakpoint models",
+    subtitle = "Breakpoint models calculated using segmented package and selected using BIC",
+    caption = "Lines are from breakpoint models; these are 'kinked' because slopes are for single years, not changes between years"
+  )
+
+ggsave(here("figures", "best_0to2bp_change_e0.png"), height = 20, width = 30, units = "cm", dpi = 150)
+
+```
+```{r}
+best_model_predictions_descriptions %>% 
+  filter(x == 65) %>% 
+  group_by(code, x, sex) %>% 
+  arrange(year) %>% 
+  mutate(
+    ch_ex = ex - lag(ex),
+    ch_ex_pred = ex_pred - lag(ex_pred)
+    ) %>% 
+  ungroup() %>% 
+  ggplot(aes(x = year)) + 
+  geom_point(aes(y = ch_ex), alpha = 0.25) + 
+  geom_step(aes(y = ch_ex_pred), direction = "vh") + 
+  facet_grid(sex~code) + 
+  geom_hline(yintercept = 0) + 
+  geom_ribbon(
+    aes(
+      ymin = ifelse(ch_ex_pred < 0, ch_ex_pred, 0), 
+      ymax = ifelse(ch_ex_pred < 0, 0, ch_ex_pred),
+      fill = ch_ex_pred < 0
+      ),
+    alpha = 0.2
+    ) + 
+  scale_fill_manual(values = c(`TRUE` = 'red', `FALSE` = 'blue')) +
+  guides(fill = "none") + 
+  labs(
+    x = "Year", y = "Annual change in life expectancy at age 65",
+    title = "Annual change in life expectancy at age 65, with implied changes from breakpoint models",
+    subtitle = "Breakpoint models calculated using segmented package and selected using BIC",
+    caption = "Lines are from breakpoint models; these are 'kinked' because slopes are for single years, not changes between years"
+  )
+
+ggsave(here("figures", "best_0to2bp_change_e65.png"), height = 20, width = 30, units = "cm", dpi = 150)
+
+```
+
+# Secondary confirmation
+
+I've reluctantly convinced myself that keeping two methods for estimating breakpoints, one using $y ~ t$, the other using $dy/dt ~ t$ (i.e.
+changepoint analysis) is still an important aspect of the paper.
+For the changepoint modelling I looked at each possible value of t (grid search, 1d).
+For two breakpoints I guess I should repeat the approach, but with all permutations (removing duplicates for t1 and t2).
+This might take a little while but should provide reassurance.
+
+```{r}
+changepoint_breakpoints_models <- 
+  hmd_ex_selected_countries_with_synth %>% 
+  filter(code != "DEUTNP") %>%   
   filter(year >= 1979) %>% 
@@
   mutate(
-      best_tau = map_dbl(data, get_best_tau_from_gridsearch)
-    ) %>% 
-    mutate(
-      model_null = map(
-        data, 
-        ~ .x %>% 
+    mdl_outputs = map(data, get_best_taus_from_gridsearch, what = "all")
+  )
+
+```
+
+Let's try to make sense of this
+
+```{r}
+changepoint_breakpoints_models %>% 
+  unnest(mdl_outputs) %>% 
+  group_by(code, x, sex) %>% 
+  mutate(bic_std = (bic - min(bic)) / (max(bic) - min(bic))) %>% 
+  ggplot(aes(x = tau1, y = tau2, colour = bic_std, fill = bic_std)) + 
+  geom_raster() + 
+  facet_wrap(~ code + sex + x, ncol = 8) + 
+  scale_color_viridis_c() + 
+  geom_point(colour= 'red', 
+             data = changepoint_breakpoints_models %>% 
+              unnest(mdl_outputs) %>% 
+              group_by(code, x, sex) %>% 
+              mutate(bic_std = (bic - min(bic)) / (max(bic) - min(bic))) %>% 
+              filter(bic_std == min(bic_std))
+  ) + 
+  coord_fixed() +
+  theme(axis.text.x = element_text(angle = 90))
+```
+
+For a number of populations, the red points (lowest aics) are just slightly off the diagonal, implying something like a single changepoint model.
+This can be tested further by seeing if those 2 cp models that are near the diagonal tend to get outcompleted with a 1 cp model.
+There are also models with horizontal dark lines, implying indifference to the second value. Once again these are likely (perhaps?) to be outcompleted with the 1 cp models.
+
+So, let's get the one CP models, and the associated BIC. (I have changed the default returned to BIC from AIC)
+
+```{r}
+changepoint_breakpoint_models <- 
+  hmd_ex_selected_countries_with_synth %>% 
+  filter(code != "DEUTNP") %>%   
+  filter(year >= 1979) %>% 
+  group_by(code, x, sex) %>% 
+  nest() %>% 
+  mutate(
+    mdl_outputs = map(data, get_best_tau_from_gridsearch, what = "all")
+  )
+
+```
+
+Now let's compare 1cp with 2cp
+
+```{r}
+estimate_null <- function(df){
+  df %>% 
     arrange(year) %>% 
@@
     lm(delta_ex ~ 1, data = .)
-      )
-    ) %>% 
-    mutate(
-      model_alt = map2(
-        best_tau, data,
-      ~ .y %>% 
+}
+
+compare_bic_cps <- 
+  hmd_ex_selected_countries_with_synth %>% 
+    filter(code != "DEUTNP") %>%   
+    filter(year >= 1979) %>% 
+    group_by(code, x, sex) %>% 
+    nest() %>% 
+    mutate(
+      mdl_0 = map(data, estimate_null),
+      bic_0    = map_dbl(mdl_0, BIC)
+    ) %>% 
+    select(code, x, sex, mdl_0, bic_0) %>% 
+    left_join(
+      changepoint_breakpoint_models %>% 
+        unnest(mdl_outputs) %>% 
+        group_by(code, x, sex) %>% 
+        filter(bic == min(bic)) %>% 
+        select(code, x, sex, tau1 = tau, bic_1 = bic)  
+    ) %>% 
+    left_join(
+      changepoint_breakpoints_models %>% 
+        unnest(mdl_outputs) %>% 
+        group_by(code, x, sex) %>% 
+        filter(bic == min(bic)) %>% 
+        select(code, x, sex, tau1_2 = tau1, tau2_2 = tau2, bic_2 = bic)
+    ) %>% 
+  select(-mdl_0) %>% 
+  ungroup() # %>% 
+#  write.csv("tmp.csv")
+```
+
+The results are mixed.
+Generally the 2cp model preferred to the 1cp model, and only rarely is the 0cp model preferred.
+For *most* of the populations, the 2cp and 1cp model contain a tau either identical to or one year apart from each other.
+This is reassuring.
+
+- Germany East, female, 0: no close match (2)
+- Germany East, female, 65: 1987 for both (2)
+- Germany East, male, 0: 1991 for both (2) 
+- Germany East, male, 65: 1991 for both (2)
+- Germany West, female, 0: 1988 for both (1)
+- Germany West, female, 65: no close match (2)
+- Germany West, male, 0: No close match (2)
+- Germany West, male, 65: 2015 for both (2)
+- SynthGermany, male, 0: no match (1)
+- SynthGermany, male, 65: 2015 for both (1) 
+- SynthGermany, female, 0: no match (1)
+- SynthGermany, female, 65: no match (2)
+- Spain, female, 0: 1983 for both (2 but close)
+- Spain, female, 65: 0cp preferred (0)
+- Spain, male, 0: 1983 for both (2)
+- Spain, male, 65: 0cp preferred (0)
+- France, female, 0: No close match (2)
+- France, female, 65: No close match (2)
+- France, male, 0: No clear match (2)
+- France, male, 65: No clear match (2)
+- Italy, female, 0: 2005 for both (2)
+- Italy, female, 65: 2005 for both (2)
+- Italy: male, 0: 0cp preferred (0)
+- Italy, male, 65: No close match (2)
+- England & Wales, female, 0: 2012 for both (1)
+- England & Wales, female, 65: No close match (0)
+- England & Wales, male, 0: 2012/13 (1)
+- England & Wales, male, 65: 2012 (2)
+- Scotland, male, 0: 2015 in both (1)
+- Scotland, male, 65: 1994 in both (2)
+- Scotland, female, 0: no close match (2)
+- Scotland, female, 65: no close match (0)
+
+For three of the four England & Wales population, a breakpoint is consistently identified at around 2012.
+
+Now I want to know whether the estimates are about the same whether 1cp or 2cp used.
+
+```{r}
+
+run_tau_1 <- function(tau, df){
+  df %>% 
     arrange(year) %>% 
@@
     filter(!is.na(delta_ex)) %>% 
-        mutate(T_param = ifelse(year < .x, FALSE, TRUE)) %>% 
+    mutate(T_param = ifelse(year < tau, FALSE, TRUE)) %>% 
     lm(delta_ex ~ T_param, data = .)
-      )
-    )
-tmp <- compare_best_breakmodel_with_null %>% 
-  mutate(
-    aic_alt  = map_dbl(model_alt, AIC),
-    aic_null = map_dbl(model_null, AIC)
-  ) %>% 
-  mutate(
-    alt_preferred = map2_lgl(aic_alt, aic_null, ~ifelse(.x < .y, TRUE, FALSE))
-  ) %>% 
-  mutate(
-    anova_results = map2(model_alt, model_null, anova)
-  ) %>% 
-  mutate(
-    ftest_pval = map_dbl(anova_results, ~.[2,6]),
-    stat_sig_preference = ftest_pval < 0.05
-  )
-
-```
-
-Let's now try to double-check the above results by running breakpoint analysis using the `segmented` package
-
-```{r}
-# Firstly, we need the data
-
-# Let's start with one of the populations, then generalise
-
-test_dta <- 
-  hmd_ex_selected_countries_with_synth %>%
+}
+
+run_tau_2 <- function(tau1, tau2, df){
+  df %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+    filter(!is.na(delta_ex)) %>% 
+    mutate(T_param = ifelse(
+      year < tau1, 
+      "bp0", 
+      ifelse(
+        year < tau2, "bp1", "bp2"
+        )
+      )
+    ) %>% 
+    lm(delta_ex ~ T_param, data = .)
+}
+
+make_path_coords <- function(df, taus = NULL, mdl){
+  stopifnot(
+    "wrong number of coefficients for the model type" = 
+    length(coefficients(mdl)) == length(taus) + 1
+  )
+
+  min_x = min(df$year)
+  max_x = max(df$year)
+
+  coeffs <- coefficients(mdl)
+
+  y0 <- coeffs[1]
+
+  out <- tribble(
+    ~x, ~y,
+    min_x, y0
+  )
+
+  if(length(coeffs) == 1){
+    out <- out %>%
+      bind_rows(
+        tribble(
+          ~x, ~y,
+          max_x, y0
+        )
+      )
+    return(out)
+
+  } else if (length(coeffs) == 2){
+    out <- out %>% 
+      bind_rows(
+        tribble(
+          ~x, ~y,
+          taus[1], y0,
+          taus[1], y0 + coeffs[2],
+          max_x, y0 + coeffs[2]
+        )
+      )
+    return(out)
+
+  } else if (length(coeffs) == 3){
+    out <- out %>% 
+      bind_rows(
+        tribble(
+          ~x, ~y,
+          taus[1], y0,
+          taus[1], y0 + coeffs[2],
+          taus[2], y0 + coeffs[2],
+          taus[2], y0 + coeffs[3],
+          max_x, y0 + coeffs[3]
+
+        )
+      )
+
+    return(out)
+  }
+  return(NULL) # should not be triggered  
+}
+
+# Let's try this with 0cp
+
+paths_0cp <- 
+  hmd_ex_selected_countries_with_synth %>% 
+    filter(code != "DEUTNP") %>%   
     filter(year >= 1979) %>% 
+    group_by(code, x, sex) %>% 
+    nest() %>% 
+    mutate(
+      mdl_0 = map(data, estimate_null)
+    ) %>% 
+    mutate(bic = map_dbl(mdl_0, BIC)) %>% 
+    mutate(
+      tau_paths = pmap(.l = list(df = data, mdl = mdl_0), make_path_coords)
+    ) %>% 
+  select(code, x, sex, bic_0cp = bic, paths_0cp = tau_paths)
+
+# Now 1cp
+paths_1cp <- 
+  changepoint_breakpoint_models %>% 
+    unnest(mdl_outputs) %>% 
+    group_by(code, x, sex) %>% 
+    filter(bic == min(bic)) %>% 
+    ungroup() %>% 
+    mutate(
+      mdl_with_best_1cp = map2(tau, data, run_tau_1)
+    ) %>% 
+    mutate(
+      tau_paths = pmap(
+        .l = list(
+          df = data, 
+          mdl = mdl_with_best_1cp, 
+          taus = c(tau)
+        ), 
+        make_path_coords
+      )
+    ) %>% 
+    select(
+      code, x, sex, bic_1cp = bic, paths_1cp = tau_paths
+    )
+
+# and now for 2cp
+paths_2cp <- 
+  changepoint_breakpoints_models %>% 
+    unnest(mdl_outputs) %>% 
+    group_by(code, x, sex) %>% 
+    filter(bic == min(bic)) %>% 
+    ungroup() %>% 
+    mutate(mdl_with_best_2cp = pmap(.l = list(tau1=tau1, tau2=tau2, df=data), run_tau_2)) %>%
+    mutate(taus = map2(tau1, tau2, c)) %>% 
+    mutate(
+      tau_paths = pmap(
+        .l = list(
+          df = data, 
+          mdl = mdl_with_best_2cp, 
+          taus = taus
+        ), 
+        make_path_coords
+      )
+    ) %>% 
+    select(code, x, sex, bic_2cp = bic, paths_2cp = tau_paths)
+
+```
+
+Let's now see how the three compare
+
+```{r}
+
+reduce(list(paths_0cp, paths_1cp, paths_2cp), left_join) %>% 
+  rename(start_age = x) %>% 
+  pivot_longer(cols = starts_with(c("path", "bic")), names_to = c(".value", "path_type"), names_sep = "_") %>% 
+  group_by(code, start_age, sex) %>% 
+  mutate(
+    bic_rank = as.character(rank(bic))
+  ) %>% 
+  ungroup() %>% 
+  unnest(paths) %>% 
+  filter(code == "ESP") %>% 
+#  filter(start_age == 65) %>% 
+  ggplot(aes(x, y, group = path_type, colour = path_type, linetype = bic_rank, alpha = bic_rank)) + 
+  geom_path() + 
+  facet_grid(start_age~sex) + 
+  expand_limits(y  = 0) + 
+  geom_hline(yintercept = 0) +
+  scale_alpha_manual("BIC", values = c("1" = 1.0, "2" = 0.5, "3" = 0.25)) +
+  scale_linetype_manual(values = c("1" = "solid", "2" = "dashed", "3" = "dashed")) + 
+  guides(alpha = FALSE, linetype = FALSE) + 
+    geom_point(
+    aes(x = year, y = ch_ex), 
+    inherit.aes = FALSE,
+    data =   hmd_ex_selected_countries_with_synth %>%
+      rename(start_age = x) %>% 
+      filter(code == "ESP") %>%
+      group_by(sex, start_age) %>% 
+      arrange(year) %>% 
+      mutate(ch_ex = ex - lag(ex)) %>% 
+      ungroup() %>% 
+      filter(year >= 1979)
+
+  )
+
+```
+It looks like the 2cp model is sometimes 'flexing' to pick up a distinct year with especially high or low mortality.
+This is consistent with infectious diseases but more an impulse rather than a step-change, which is what we are interested in.
+
+I think I'd like to try to make this interactive with a simple (!) shiny app.
+
+What specific files do I need to save?
+
+```{r}
+cp_paths <- 
+  reduce(list(paths_0cp, paths_1cp, paths_2cp), left_join) %>% 
+    rename(start_age = x) %>% 
+    pivot_longer(cols = starts_with(c("path", "bic")), names_to = c(".value", "path_type"), names_sep = "_") %>% 
+    group_by(code, start_age, sex) %>% 
+    mutate(
+      bic_rank = as.character(rank(bic))
+    ) %>% 
+    ungroup() %>% 
+    unnest(paths)
+
+write_rds(cp_paths, here("apps", "compare_cps", "data", "cp_paths.rds"))
+
+write_rds(hmd_ex_selected_countries_with_synth, here("apps", "compare_cps", "data", "countries_data.rds"))
+
+```
+
+I would like to plot the above with the data too, and to plot these as figures
+
+```{r}
+
+# Let's start with just a single population, and the figure created for England & Wales at age 0
+
+g_path <- 
+  reduce(list(paths_0cp, paths_1cp, paths_2cp), left_join) %>% 
+    rename(start_age = x) %>% 
+    pivot_longer(cols = starts_with("path"), names_to = "path_type", names_prefix = "paths_", values_to = "df") %>% 
+    unnest(df) %>% 
     filter(code == "GBRTENW") %>% 
-    filter(sex == "female") %>% 
-    filter(x == 0)
-
-# Now we create a simple model with no breakpoint for y against t
-
-test_null_mdl <- lm(ex ~ year, data = test_dta)
-
-# Then we create a segmented version of the same model
-
-test_seg_mdl <- segmented::segmented(test_null_mdl)
-
-anova(test_null_mdl, test_seg_mdl)
-
-# now to pack into a function
-
-estimate_breakpoint_and_pval <- function(df){
-  null_mdl <- lm(ex ~ year, data = df) 
-  seg_mdl <- segmented::segmented(null_mdl)   
-  anova_diff <- anova(null_mdl, seg_mdl)
-
-  list(
-    null = null_mdl, 
-    seg = seg_mdl,
-    diff = anova_diff
-  )
-}
-
-segmented_breakpoint_models <- 
-  hmd_ex_selected_countries_with_synth %>% 
+    filter(start_age == 0) %>% 
+    ggplot(aes(x, y, group = path_type, colour = path_type), alpha = 0.6) + 
+    geom_path() + 
+    facet_wrap(~sex) + 
+    expand_limits(y  = 0) + 
+    geom_hline(yintercept = 0)
+
+g_path
+
+# Now the points
+
+g_points <- 
+  hmd_ex_selected_countries_with_synth %>%
+  rename(start_age = x) %>% 
+  filter(code == "GBRTENW") %>%
+  filter(start_age == 0) %>%
+  group_by(sex) %>% 
+  arrange(year) %>% 
+  mutate(ch_ex = ex - lag(ex)) %>% 
+  ungroup() %>% 
   filter(year >= 1979) %>% 
+  ggplot(aes(x = year, y = ch_ex)) + 
+  facet_wrap(~sex) + 
+  geom_point()
+
+g_points
+
+g_path + g_points
+
+g_path + 
+  geom_point(
+    aes(x = year, y = ch_ex), 
+    inherit.aes = FALSE,
+    data =   hmd_ex_selected_countries_with_synth %>%
+      rename(start_age = x) %>% 
+      filter(code == "GBRTENW") %>%
+      filter(start_age == 0) %>%
+      group_by(sex) %>% 
+      arrange(year) %>% 
+      mutate(ch_ex = ex - lag(ex)) %>% 
+      ungroup() %>% 
+      filter(year >= 1979)
+
+  )
+
+# I think I also want to know which cp has the lowest BIC
+
+```
+
+```{r}
+# For 0cp
+hmd_ex_selected_countries_with_synth %>% 
+  filter(code != "DEUTNP") %>%   
+  filter(year >= 1979) %>% 
   group_by(code, x, sex) %>% 
@@
   mutate(
-      mdl_outputs = map(data, estimate_breakpoint_and_pval)
-    ) %>% 
-    unnest_wider(mdl_outputs) %>% 
-    mutate(
-      ftest_pval = map_dbl(diff, ~.[2,6]),
-      stat_sig_preference = ftest_pval < 0.05
-    ) %>% 
-    mutate(
-      breakpoint = map_dbl(seg, ~.[["psi"]][1,2]),
-      breakpoint_se = map_dbl(seg, ~.[["psi"]][1,3])
-    )
-
-segmented_breakpoint_models
-
-```
-
-Now I'm going to plot the breakpoints + SEs on the data itself
-
-```{r}
-segmented_breakpoint_models %>% 
-  unnest(data) %>% 
-  ggplot(aes(x = year, y = ex, colour = sex)) + 
-  geom_line() + 
-  facet_grid(x ~ code, scales = "free_y") + 
-  geom_vline(aes(xintercept = breakpoint, color = sex))
-
-```
-
-And now for change in ex
-
-- [ ] Jon to figure out why this doesn't work...
+    mdl_0 = map(data, estimate_null)
+  ) %>% 
+  mutate(
+    tdy_mdl_0 = map(mdl_0, broom::tidy)
+  ) %>% 
+  select(code, x, sex, tdy_mdl_0) %>% 
+  unnest(tdy_mdl_0)
+
+# For 1cp
+changepoint_breakpoint_models %>% 
+  unnest(mdl_outputs) %>% 
+  group_by(code, x, sex) %>% 
+  filter(bic == min(bic)) %>% 
+  ungroup() %>% 
+  mutate(
+    mdl_with_best_1cp = map2(tau, data, run_tau_1)
+  ) %>% 
+  mutate(
+    tdy_mdl_with_1cp = map(mdl_with_best_1cp, broom::tidy)
+  ) %>% 
+  select(code, x, sex, tau, tdy_mdl_with_1cp) %>% 
+  unnest(tdy_mdl_with_1cp)
+
+# for 2cp
+
+changepoint_breakpoints_models %>% 
+  unnest(mdl_outputs) %>% 
+  group_by(code, x, sex) %>% 
+  filter(bic == min(bic)) %>% 
+  ungroup() %>% 
+  mutate(mdl_with_best_2cp = pmap(.l = list(tau1=tau1, tau2=tau2, df=data), run_tau_2)) %>%
+  mutate(tdy_mdl_with_best_2cp = map(mdl_with_best_2cp, broom::tidy)) %>% 
+  select(code, x, sex, tau1, tau2, tdy_mdl_with_best_2cp) %>% 
+  unnest(tdy_mdl_with_best_2cp)
+
+#So, what's the best way to turn these into graphical instructions?
+
+# For the first one, we want 
+# {x, y}
+# {1980, 0.13}
+# {1991, 0.13}
+# {1991, 0.13 + 0.30}
+# {2002, 0.13 + 0.30}
+# {2002, 0.13 + 0.01}
+# {2015, 0.13 + 0.01}
+
+```
 
@@
 # Bibliography
+
+# Dream Graphs
+
+To finally complete these, without getting distracted/stuck on how to double check and improve upon the underlying data/models
+
+- [ ] I've finally remembered that I can get the breakpoints by aggregating up the changes, as I've done many times previously!
+
+## Graph of ex against year, predicted and observed.
+
+```{r}
+# Start with Scotland, female, 0
+
+test_data <- compare_best_breakmodel_with_null %>% 
+  filter(code == "GBR_SCO", x == 0, sex == "female") %>% 
+  pluck("data", 1)
+
+test_data %>% 
+  ggplot(aes(year, ex)) + 
+  geom_point()
+
+# Now to get the modelled estimates
+
+compare_best_breakmodel_with_null %>% 
+  filter(code == "GBR_SCO", x == 0, sex == "female") %>% 
+  pluck("model_alt", 1) %>% 
+  broom::tidy()
+
+# So, we want the last year before the breakpoint, then to produce a line that just integrates the (Intercept) term as many years as we want to project forward for.
+
+# So, let's think of a function that makes a graph, and what inputs it needs
+
+show_counterfactual_figure <- function(df, best_tau, alt_model){
+  # First we might want to estimate the intercept for the pre-slowdown portion 
+  # while fixing the slope at that estimated by (Intercept) in the delta_ex model
+
+  pred_slope <- alt_model %>%  broom::tidy() %>% 
+    filter(term == "(Intercept)") %>% 
+    pluck("estimate")
+  # 
+  pred_slope_after <- alt_model %>% 
+    broom::tidy() %>% 
+    filter(term == "T_paramTRUE") %>% 
+    pluck("estimate")
+
+  pred_intercept <-  mean(df$ex - pred_slope * df$year)
+
+  pred_intercept_at_tau <- pred_intercept - pred_slope_after * best_tau
+
+  # I think I should use geom_area for the after part 
+  df %>% 
+    mutate(
+      y_pred_before = pred_intercept + pred_slope * year,
+      y_pred_after  = pred_intercept_at_tau + (pred_slope + pred_slope_after) * year
+    ) %>% 
+    mutate(
+      y_pred_before = ifelse(year < best_tau, NA, y_pred_before),
+      y_pred_after  = ifelse(year < best_tau, NA, y_pred_after)
+    ) %>% 
+    ggplot() + 
+    geom_point(aes(year, ex)) + 
+    geom_vline(xintercept = best_tau) + 
+    geom_abline(intercept = pred_intercept, slope = pred_slope) + 
+    geom_ribbon(aes(ymax = y_pred_before, ymin = y_pred_after, x = year), alpha = 0.5)
+}
+
+this_df <- compare_best_breakmodel_with_null %>% 
+  filter(code == "GBR_SCO", x == 0, sex == "male") %>% 
+  pluck("data", 1)
+
+this_best_tau <- compare_best_breakmodel_with_null %>% 
+  filter(code == "GBR_SCO", x == 0, sex == "male") %>% 
+  pluck("best_tau")
+
+this_model_alt <- compare_best_breakmodel_with_null %>% 
+  filter(code == "GBR_SCO", x == 0, sex == "male") %>% 
+  pluck("model_alt", 1)
+
+show_counterfactual_figure(this_df, this_best_tau, this_model_alt)
+
+# Now to do for all
+
+# Let's also label and set axes intelligently
+
+add_labels_axes <- function(gg, country, start_age, sex, 
+                            ylims = list(
+                                birth = c(70, 95),
+                                x65 = c(0, 25)
+                              )
+                            ){
+  gg + 
+    labs(
+      x = "Year", 
+      y = case_when(
+        start_age == 0  ~ "Life expectancy at birth",
+        start_age == 65 ~ "Life expectancy at age 65",
+        TRUE            ~ NA_character_
+      ),
+      title = glue::glue("Life expectancy in {country}, at age {start_age}, for {sex}s"),
+      subtitle = "Shaded areas show deviation from projected life expectancy before any breakpoint in trend", 
+      caption = "Source: Human Mortality Database"
+    ) + 
+    expand_limits( 
+      y = case_when(
+        start_age == 0 ~ ylims[["birth"]],
+        start_age == 65 ~ ylims[["x65"]],
+        TRUE    ~ c(0, 90)
+      )
+    )
+}
+
+compare_best_breakmodel_with_null %>% 
+  left_join(country_code_lookup) %>% 
+  mutate(
+    ggsegplot = pmap(
+      .l = list(df = data, best_tau = best_tau, alt_model = model_alt),
+      .f = show_counterfactual_figure
+    ),
+    ggsegplot = pmap(.l = list(gg = ggsegplot, country = country, start_age = x, sex = sex), add_labels_axes)
+  ) %>% 
+  mutate(NULL = walk(ggsegplot, print))
+
+```
+
+# Another attempt at a simple summary
+
+It could be we don't get a clear changepoint for some populations because the annual variance as a function of change is particularly large.
+This would be more an issue for smaller populations than larger populations.
+The changepoint can therefore be expressed as a standarised unit.
+
+```{r}
+
+mean_sd_chex <- 
+  hmd_ex_selected_countries_with_synth %>% 
+    filter(code %in% c("DEUTSYNTH", "GBRTENW", "GBR_SCO", "FRATNP", "ITA", "ESP")) %>% 
+    filter(between(year, 1979, 2019)) %>% 
+    filter(x == 0) %>% 
+    group_by(code, sex) %>% 
+    arrange(year) %>% 
+    mutate(ch_ex = ex - lag(ex)) %>% 
+    filter(!is.na(ch_ex)) %>% 
+    summarise(
+      mean_chex = mean(ch_ex),
+      sd_chex   = sd(ch_ex)
+    ) %>% 
+    mutate(b = mean_chex / sd_chex) %>% 
+    ungroup() %>% 
+    arrange(desc(mean_chex))
+
+mean_sd_chex
+
+# Now to compare with the changepoint model
+
+get_chpt_coeffs <- function(x){
+  tidied_coeffs <- broom::tidy(x)
+
+  tibble(
+    mean_chex_before = tidied_coeffs$estimate[1],
+    sd_chex_before   = tidied_coeffs$std.error[1],
+    mean_chex_after  = tidied_coeffs$estimate[2],
+    sd_chex_after    = tidied_coeffs$std.error[2]
+  )
+}
+
+broke_unbroke <- 
+  mean_sd_chex %>% 
+    left_join(
+      compare_best_breakmodel_with_null %>% 
+        ungroup() %>% 
+        filter(x == 0) %>% 
+        select(code, sex, best_tau, model_alt, ftest_pval) %>% 
+        mutate(coeffs = map(model_alt, get_chpt_coeffs)) %>% 
+        unnest(coeffs) %>% 
+        select(-model_alt) 
+    ) %>% 
+    mutate(
+      ch_as_ratio_of_mean_chex = mean_chex_after / mean_chex,
+      pos_chex_after = mean_chex_before + mean_chex_after
+    )
+
+# now to visualise in a simple way
+
+expand_grid(
+  year = 1980:2019,
+  broke_unbroke
+  ) %>% 
+  mutate(
+    broke_val = if_else(year < best_tau, mean_chex_before, mean_chex_before + mean_chex_after)
+  ) %>% 
+  left_join(
+    hmd_ex_selected_countries_with_synth %>% 
+      filter(code %in% c("DEUTSYNTH", "GBRTENW", "GBR_SCO", "FRATNP", "ITA", "ESP")) %>% 
+      filter(between(year, 1979, 2019)) %>% 
+      filter(x == 0) %>% 
+      group_by(code, sex) %>% 
+      arrange(year) %>% 
+      mutate(ch_ex = ex - lag(ex)) %>% 
+      filter(!is.na(ch_ex)) %>% 
+      ungroup() %>% 
+      select(code, year, sex, ch_ex)
+  ) %>% 
+  ggplot(aes(year)) +
+  geom_hline(aes(yintercept = mean_chex)) + 
+  geom_hline(yintercept = 0, size = 1.3) + 
+  facet_grid(code ~ sex) +
+  geom_ribbon(aes(ymin = mean_chex - sd_chex, ymax = mean_chex + sd_chex), alpha = 0.2) + 
+  geom_ribbon(aes(ymin = mean_chex - 2 * sd_chex, ymax = mean_chex + 2 * sd_chex), alpha = 0.2) + 
+  expand_limits(y = 0) + 
+  geom_step(aes(y = broke_val, alpha = 1 - ftest_pval, linetype = ftest_pval < 0.05), colour = "red") + 
+  geom_point(aes(y = ch_ex)) +
+  scale_alpha_continuous(range = c(0, 1)) + 
+  scale_linetype_manual(values = c("FALSE" = "dotted", "TRUE" = "solid")) + 
+  guides(alpha = "none", linetype = "none")
+
+```

### 2022-06-02 notebooks/old_bp_cp_comparison.Rmd
try to tidy main analysis; build simple app to look at how changepoint models work

---
title: "R Notebook"
output: html_notebook
---

# Code and notes on breakpoint and changepoint

Below is a chunk of code trying to make sense of the changepoint and breakpoint analyses when only comparing one changepoint/breakpoint against zero changepoints/breakpoints

# Start

```{r eval = FALSE}
# We can now run the above function, which creates a model, for a wide range of possible tau values
# This is the sequence listed and named tau_candidate, which is every whole number from 
# 1982 to 2015
# For each of these models, we can then get a measure of model fit. We're going to use AIC, though 
# could have used other measures. With AIC (and with BIC, and RMSE etc) better fit is indicated with 
# a lower value

source(here("R", "breakpoint_functions.R"))

# Let's see if this now works in the functional programming context, using 
# nest and tibble elements 
# hmd_ex_selected_countries_with_synth %>%
#   filter(year >= 1979) %>% 
#   group_by(code, x, sex) %>% 
#   nest() %>% 
#   mutate(
#     best_tau = map_dbl(data, get_best_tau_from_optim)
#   )
# I think this is getting stuck on local optima, so instead maybe 
# the more exhaustive search should be used after all! 
# There are two alternative approaches: 
# 1) Use an exhaustive grid search approach 
# 2) Use an algorithm which is less likely to get stuck in local optima
# The first approach is much more sensible, but the second might be a cool
# stretch goal as well! 
# let's use a heavily modified version of the above function to do the grid search

compare_best_breakmodel_with_null <- 
  hmd_ex_selected_countries_with_synth %>%
    filter(year >= 1979) %>% 
    group_by(code, x, sex) %>% 
    nest() %>% 
    mutate(
      best_tau = map_dbl(data, get_best_tau_from_gridsearch),
      all_taus = map(data, get_best_tau_from_gridsearch, what = "all")
    ) %>% 
    mutate(
      model_null = map(
        data, 
        ~ .x %>% 
          arrange(year) %>% 
          mutate(delta_ex = ex - lag(ex)) %>% 
          filter(!is.na(delta_ex)) %>% 
          lm(delta_ex ~ 1, data = .)
      )
    ) %>% 
    mutate(
      model_alt = map2(
        best_tau, data,
      ~ .y %>% 
        arrange(year) %>% 
        mutate(delta_ex = ex - lag(ex)) %>% 
        filter(!is.na(delta_ex)) %>% 
        mutate(T_param = ifelse(year < .x, FALSE, TRUE)) %>% 
        lm(delta_ex ~ T_param, data = .)
      )
    ) %>% 
  mutate(
    aic_alt  = map_dbl(model_alt, AIC),
    aic_null = map_dbl(model_null, AIC)
  ) %>% 
  mutate(
    alt_preferred = map2_lgl(aic_alt, aic_null, ~ifelse(.x < .y, TRUE, FALSE))
  ) %>% 
  mutate(
    anova_results = map2(model_alt, model_null, anova)
  ) %>% 
  mutate(
    ftest_pval = map_dbl(anova_results, ~.[2,6]),
    stat_sig_preference = ftest_pval < 0.05
  )

```

Let's now try to summarise the above results intelligently.

We want to know:

- [ ] Best year
- [ ] Whether stat sig over null 
- [ ] Estimate before break
- [ ] Estimate after break

Let's take the object above then get this info.

```{r}
compare_best_breakmodel_with_null %>% 
  mutate(population = glue::glue("{code}_{sex}_{x}")) %>% 
  ggplot(aes(best_tau, fct_reorder(population, best_tau), colour = stat_sig_preference)) + 
  geom_point()

```

I'm now a bit concerned by how many picked 2015. Is this an artefact of when the data finished?

To test this further I've now saved AIC by proposed year so let's look at how clear a 'winner' is for each population

```{r}
compare_best_breakmodel_with_null %>% 
  select(code, x, sex, best_tau, all_taus, ftest_pval) %>% 
  unnest(all_taus) %>% 
  ungroup() %>% 
  mutate(population = glue::glue("{code}_{x}_{sex}")) %>% 
  mutate(
    ftest_cat = case_when(
      ftest_pval < 0.01 ~ "<0.01",
      between(ftest_pval, 0.01, 0.05) ~ "0.01-0.05",
      between(ftest_pval, 0.05, 0.10) ~ "0.05-0.10",
      TRUE                            ~ "Not Stat. Sig."
    ),
    ftest_cat = factor(ftest_cat, 
      levels = c("<0.01", "0.01-0.05", "0.05-0.10", "Not Stat. Sig."),
      ordered = TRUE
    )
  ) %>% 
  ggplot(aes(tau, aic)) +
  geom_line() + geom_point(aes(colour = ftest_cat, alpha = tau == best_tau)) + 
  facet_wrap(~population, scales = "free_y") +
  scale_colour_manual(
    values = c(
      "<0.01" = "darkred", 
      "0.01-0.05"  = "red", 
      "0.05-0.10" = "orange", 
      "Not Stat. Sig." = "lightblue"
      )
  ) + 
  scale_alpha_manual(values = c(`FALSE` = 0, `TRUE` = 1), guide = "none") +
  labs(
    title = "Penalised fit (AIC) against proposed changepoint year, various population",
    subtitle = "Best fit identified indicate with a point, whose colour indicates statistical significance of changepoint model compared with Null model",
    colour = "Statistical\nsignificance", 
    x = "Year proposed",
    y = "AIC (lower is better)"
  )
```

I think this is a fairly complex visualisation, but hopefully usable either in the main paper or (more likely) supplementary material.

Here's a table attempting to summarise (subjectively) the above graphs:

| **Population**    | **x** | **Sex** | **Main Breakpoint (Stat sig.)** | **Other possible breakpoints** |
|-------------------|-------|---------|---------------------------------|--------------------------------|
| East Germany      | 0     | male    | 1991 **                         | No                             |
| East Germany      | 0     | female  | 2005 **                         | 2015                           |
| East Germany      | 65    | male    | 1991 **                         | No                             |
| East Germany      | 65    | female  | 1991 **                         | No                             |
| Germany           | 0     | male    | 2015 **                         | 2007ish                        |
| Germany           | 0     | female  | 2007 **                         | 2005-2008                      |
| Germany           | 65    | male    | 2015 **                         | No                             |
| Germany           | 65    | female  | 2015 *                          | 2005-2008                      |
| Synthetic Germany | 0     | male    | 2015 **                         | Weak 2007                      |
| Synthetic Germany | 0     | female  | 2007 **                         | 2005-2008, 2015                |
| Synthetic Germany | 65    | male    | 2015 **                         | No                             |
| Synthetic Germany | 65    | female  | 2015 *                          | Weak 2005-2007                 |
| West Germany      | 0     | male    | 2015 **                         | Weak 2007                      |
| West Germany      | 0     | female  | 1988 **                         | 2005-2008ish                   |
| West Germany      | 65    | male    | 2015 **                         | No                             |
| West Germany      | 65    | female  | 2015 *                          | 2008 ish (Weak)                |
| Spain             | 0     | male    | 1983 *                          | No                             |
| Spain             | 0     | female  | 1983 *                          | No                             |
| Spain             | 65    | male    | 2000 (NS)                       | 2015                           |
| Spain             | 65    | female  | 2014 (NS)                       | 1983?                          |
| France            | 0     | male    | 2015 (NS)                       | ?                              |
| France            | 0     | female  | 2015 (NS)                       | ?                              |
| France            | 65    | male    | 2015 **                         | No                             |
| France            | 65    | female  | 2015 *                          | No                             |
| Scotland          | 0     | male    | 2015 ***                        | No                             |
| Scotland          | 0     | female  | 1980 *                          | 2015 (Strong)                  |
| Scotland          | 65    | male    | 1994 (NS)                       | No?                            |
| Scotland          | 65    | female  | 1981 (NS)                       | No                             |
| England & Wales   | 0     | male    | 2013 ***                        | 2012-2015ish                   |
| England & Wales   | 0     | female  | 2012 **                         | Around 2015                    |
| England & Wales   | 65    | male    | 2012 *                          | 2015                           |
| England & Wales   | 65    | female  | 2012 (NS)                       | 2015                           |
| Italy             | 0     | male    | 1981 (NS)                       | 2015                           |
| Italy             | 0     | female  | 2005 (NS)                       | Weak 2015                      |
| Italy             | 65    | male    | 1984 (NS)                       | 1981                           |
| Italy             | 65    | female  | 2005 (NS)                       | 1984                           |

Let's now try to double-check the above results by running breakpoint analysis using the `segmented` package

```{r}
source(here("R", "segmented_script.R"))

```

Now I'm going to plot the breakpoints + SEs on the data itself

```{r}
segmented_breakpoint_models %>% 
  unnest(data) %>% 
  ggplot(aes(x = year, y = ex, colour = sex)) + 
  geom_line() + 
  facet_grid(x ~ code, scales = "free_y") + 
  geom_vline(aes(xintercept = breakpoint, color = sex))

```

I'm now going to compare the breakpoints from the segmented package with the best point from the line/gridsearch approach

```{r}
compare_best_breakmodel_with_null %>% 
  select(code, x, sex, best_tau, all_taus, ftest_pval) %>% 
  unnest(all_taus) %>% 
  ungroup() %>% left_join(
    segmented_breakpoint_models %>% 
      select(code, x, sex, bp_ftest_pval = ftest_pval, bp_breakpoint = breakpoint, bp_breakpoint_se = breakpoint_se)
  ) %>% 
  mutate(population = glue::glue("{code}_{x}_{sex}")) %>% 
  mutate(
    ftest_cat = case_when(
      ftest_pval < 0.01 ~ "<0.01",
      between(ftest_pval, 0.01, 0.05) ~ "0.01-0.05",
      between(ftest_pval, 0.05, 0.10) ~ "0.05-0.10",
      TRUE                            ~ "Not Stat. Sig."
    ),
    ftest_cat = factor(ftest_cat, 
      levels = c("<0.01", "0.01-0.05", "0.05-0.10", "Not Stat. Sig."),
      ordered = TRUE
    )
  ) %>% 
  mutate(
    bp_ftest_cat = case_when(
      bp_ftest_pval < 0.01 ~ "<0.01",
      between(bp_ftest_pval, 0.01, 0.05) ~ "0.01-0.05",
      between(bp_ftest_pval, 0.05, 0.10) ~ "0.05-0.10",
      TRUE                            ~ "Not Stat. Sig."
    ),
    bp_ftest_cat = factor(bp_ftest_cat, 
      levels = c("<0.01", "0.01-0.05", "0.05-0.10", "Not Stat. Sig."),
      ordered = TRUE
    )
  ) %>% 
  ggplot(aes(tau, aic)) +
  geom_line() + 
  geom_point(aes(colour = ftest_cat, alpha = tau == best_tau)) + 
  facet_wrap(~population, scales = "free_y") +
  scale_colour_manual(
    values = c(
      "<0.01" = "darkred", 
      "0.01-0.05"  = "red", 
      "0.05-0.10" = "orange", 
      "Not Stat. Sig." = "lightblue"
      )
  ) + 
  geom_vline(
    aes(
      xintercept = bp_breakpoint,
      colour = bp_ftest_cat  
    )
  ) + 
  geom_rect(
    aes(
      xmin = bp_breakpoint - 1.96 * bp_breakpoint_se, 
      xmax = bp_breakpoint + 1.96 * bp_breakpoint_se
    ),
      ymin = -Inf, ymax = Inf,
    alpha = 0.01
  ) + 
  scale_alpha_manual(values = c(`FALSE` = 0, `TRUE` = 1), guide = "none") +
  labs(
    title = "Penalised fit (AIC) against proposed changepoint year, various population",
    subtitle = "Best fit identified indicate with a point, whose colour indicates statistical significance of changepoint model compared with Null model",
    colour = "Statistical\nsignificance", 
    x = "Year proposed",
    y = "AIC (lower is better)",
    caption = "Vertical lines and bands show estimates and 95% CIs from segmented package"
  )

ggsave("figures/aic_from_gridsearch_and_segmented_bp_estimates.png", height = 20, width = 30, units = "cm", dpi = 150)
```

And now for change in ex

- [X] Jon to figure out why this doesn't work...
    - I think it doesn't work because the `segmented` package functions expect the line to be joined up. Instead it looks like the [structchange](https://cran.r-project.org/web/packages/strucchange/index.html) package can be used for more general model specifications.

### 2022-06-02 R/annual_change_graphs_by_population.R

@@ -1,0 +1,757 @@
+### England and Wales
+
+#ENW female at birth
+f_e0_ENW <- 
+  hmd_ex_selected_countries_with_synth %>% 
+  filter(x == 0) %>% 
+  filter(sex == "female") %>% 
+  filter(code == "GBRTENW") %>% 
+  filter(year >= 1979)
+
+# Now we create the change in ex variable as that's the response variable in the model
+f_e0_ENW <- f_e0_ENW %>% 
+  arrange(year) %>% 
+  mutate(delta_ex = ex - lag(ex)) %>% 
+  ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+  geom_line() +
+  geom_hline(yintercept = 0) + 
+  stat_smooth(colour = "lightpink3") 
+
+#ENW male at birth
+m_e0_ENW <- 
+  hmd_ex_selected_countries_with_synth %>% 
+  filter(x == 0) %>% 
+  filter(sex == "male") %>% 
+  filter(code == "GBRTENW") %>% 
+  filter(year >= 1979)
+
+m_e0_ENW <- m_e0_ENW %>% 
+  arrange(year) %>% 
+  mutate(delta_ex = ex - lag(ex)) %>% 
+  ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+  geom_line() +
+  geom_hline(yintercept = 0) + 
+  stat_smooth(colour = "aquamarine4") 
+
+#ENW female at age 65
+f_e65_ENW <- 
+  hmd_ex_selected_countries_with_synth %>% 
+  filter(x == 65) %>% 
+  filter(sex == "female") %>% 
+  filter(code == "GBRTENW") %>% 
+  filter(year >= 1979)
+
+# Now we create the change in ex variable 
+f_e65_ENW <- f_e65_ENW %>% 
+  arrange(year) %>% 
+  mutate(delta_ex = ex - lag(ex)) %>% 
+  ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
+  geom_line() +
+  geom_hline(yintercept = 0) + 
+  stat_smooth(colour = "lightpink3") 
+
+#ENW male at aged 65
+m_e65_ENW <- 
+  hmd_ex_selected_countries_with_synth %>% 
+  filter(x == 65) %>% 
+  filter(sex == "male") %>% 
+  filter(code == "GBRTENW") %>% 
[…]

### 2022-06-02 R/breakpoint_functions.R

@@ -76,7 +76,7 @@ get_best_tau_from_gridsearch <- function(df, buffer = 2, what = c("best", "all")
   tau_lower <- min(years) + buffer # the lower bound 
   tau_upper <- max(years) - buffer # the upper bound
   tau_searchrange <- tau_lower:tau_upper
-  get_result_from_model <- function(tau, df, what = c("AIC", "BIC", "model")){
+  get_result_from_model <- function(tau, df, what = c("BIC", "AIC", "model")){ #I've changed the order to default to BIC (as more comparable with segmented)
     # Again, can carry out more validation checks 
     what <- match.arg(what)
     
@@ -98,19 +98,79 @@ get_best_tau_from_gridsearch <- function(df, buffer = 2, what = c("best", "all")
   search_df <- tibble(
     tau = tau_searchrange
   ) %>% 
-    mutate(aic = map_dbl(tau, get_result_from_model, df = df))
+    mutate(bic = map_dbl(tau, get_result_from_model, df = df))
   
   if (what == "all") {return(search_df)}
   
   if (what == "best") {
     out <- 
       search_df %>% 
-      filter(aic == min(aic)) %>% 
-      pull(tau)
-    return(out)
+      filter(bic == min(bic)) %>% 
+      select(tau, bic)
+    return(out) # I've changed this to compare both tau and bic more easily with 2 cp models
   }
   # Again, the following should never be triggered
   NULL
 }
 
+# This is a modification of get_best_tau_from_gridsearch, for two taus
+get_best_taus_from_gridsearch <- function(df, buffer = 2, what = c("best", "all")){
+  # We can put in some data validation checks too 
+  
+  what <- match.arg(what)
+  
+  years <- df$year 
+  tau_lower <- min(years) + buffer # the lower bound 
+  tau_upper <- max(years) - buffer # the upper bound
+  
+  # Now need to list the permutations of taus to consider
+  
+  tau_grid <- expand_grid(
+    tau1 = tau_lower:tau_upper,
+    tau2 = tau_lower:tau_upper
+  ) %>% 
+    filter(tau2 > tau1) # tau2 should be greater than tau1
+  
+  get_result_from_model <- function(tau1, tau2, df, what = c("BIC", "AIC", "model")){ #I've changed the order to default to BIC (as more comparable with segmented)
+    # Again, can carry out more validation checks 
[…]

### 2022-06-02 R/make_decadal_change_models.R

@@ -1,0 +1,67 @@
+# dta_for_decade_comparison <- 
+#   hmd_ex_selected_countries_with_synth %>% 
+#   filter(x == 0) %>% 
+#   filter(sex == "female") %>% 
+#   filter(code == "GBRTENW") %>% 
+#   filter(year >= 1979) %>% 
+#   # ggplot(aes(x = year, y = ex)) +
+#   #   geom_line()
+#   arrange(year) %>% 
+#   mutate(delta_ex = ex - lag(ex)) %>% 
+#   #   ggplot(aes(x = year, y = delta_ex)) + 
+#   # geom_point() + geom_line() +
+#   # geom_hline(yintercept = 0)
+#   mutate(
+#     decade = year - (year %% 10)
+#   ) %>% 
+#   filter(!is.na(delta_ex))
+# model_null <- 
+#   lm(delta_ex ~ 1, data = dta_for_decade_comparison)
+# model_alt <- 
+#   lm(delta_ex ~ factor(decade), data= dta_for_decade_comparison)
+# summary(model_null)
+# summary(model_alt)
+# anova(model_null, model_alt)
+# to get the value of interest do tmp[2,6]
+# Generalising 
+tidy_dataset <- function(df){
+  df %>% 
+    filter(year >= 1979) %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+    mutate(
+      decade = year - (year %% 10)
+    ) %>% 
+    filter(!is.na(delta_ex))
+}
+
+make_null_model <- function(df){
+  lm(delta_ex ~ 1, data = df)
+}
+all_null_decadal_models <-
+  hmd_ex_selected_countries_with_synth %>%
+  group_by(x, sex, code) %>% 
+  nest() %>% 
+  mutate(
+    prep_data = map(data, tidy_dataset) 
+  ) %>% 
+  mutate(
+    model_null = map(prep_data, ~lm(delta_ex ~ 1, data = .)),
+    model_alt  = map(prep_data, ~lm(delta_ex ~ factor(decade), data = .))
+    #    model_null = map(prep_data, make_null_model)
+  ) %>% 
+  mutate(
+    anova_results = map2(model_null, model_alt, anova)
+  ) %>% 
+  mutate(
+    anova_pval = map_dbl(anova_results, ~.[2,6])
+  ) 
+
+# all_null_alt_models %>% 
+#   filter(anova_pval < 0.05) -> tmp
[…]

### 2022-06-02 R/make_pop_selection.R

@@ -1,0 +1,107 @@
+hmd_ex_selected_countries <- 
+  hmd_lt %>% 
+  filter(code %in% countries_of_interest) %>% 
+  filter(Age %in% c(0, 65)) %>% 
+  select(code, Year, Age, sex, ex) %>% 
+  rename(x = Age, year = Year)
+#hmd_ex_selected_countries
+# Code for producing synthetic Germany
+
+series_east_f0 <- hmd_ex_selected_countries %>% 
+  filter(code == "DEUTE") %>% 
+  filter(between(year, 1980, 2020)) %>% 
+  filter(x == 0) %>% 
+  filter(sex == "female") %>% 
+  arrange(year) %>% 
+  pull(ex)
+series_west_f0 <- hmd_ex_selected_countries %>% 
+  filter(code == "DEUTW") %>% 
+  filter(between(year, 1980, 2020)) %>% 
+  filter(x == 0) %>% 
+  filter(sex == "female") %>% 
+  arrange(year) %>% 
+  pull(ex)
+tmp_df_f0 <- tibble(
+  code = "DEUTSYNTH",
+  year = 1980:2017,
+  x = 0,
+  sex = "female", 
+  ex = make_synthetic_population(series_east_f0, series_west_f0, p_east = 0.2)
+)
+###
+series_east_f65 <- hmd_ex_selected_countries %>% 
+  filter(code == "DEUTE") %>% 
+  filter(between(year, 1980, 2020)) %>% 
+  filter(x == 65) %>% 
+  filter(sex == "female") %>% 
+  arrange(year) %>% 
+  pull(ex)
+series_west_f65 <- hmd_ex_selected_countries %>% 
+  filter(code == "DEUTW") %>% 
+  filter(between(year, 1980, 2020)) %>% 
+  filter(x == 65) %>% 
+  filter(sex == "female") %>% 
+  arrange(year) %>% 
+  pull(ex)
+tmp_df_f65 <- tibble(
+  code = "DEUTSYNTH",
+  year = 1980:2017,
+  x = 65,
+  sex = "female", 
+  ex = make_synthetic_population(series_east_f65, series_west_f65, p_east = 0.2)
+)
+##
+series_east_m0 <- hmd_ex_selected_countries %>% 
+  filter(code == "DEUTE") %>% 
+  filter(between(year, 1980, 2020)) %>% 
+  filter(x == 0) %>% 
[…]

### 2022-06-02 R/segmented_script.R

@@ -1,0 +1,156 @@
+# Segmented regression code 
+
+# This is a long chunk of code which makes used of the segmented package to try to identify if there are 
+# breakpoings, and whether they correspond to changepoints in
+# Let's start with one of the populations, then generalise 
+
+test_dta <- 
+  hmd_ex_selected_countries_with_synth %>%
+  filter(year >= 1979) %>% 
+  filter(code == "GBRTENW") %>% 
+  filter(sex == "female") %>% 
+  filter(x == 0)
+
+# Now we create a simple model with no breakpoint for y against t 
+
+test_null_mdl <- lm(ex ~ year, data = test_dta) 
+
+# Then we create a segmented version of the same model 
+
+
+test_seg_mdl <- segmented::segmented(test_null_mdl)
+
+anova(test_null_mdl, test_seg_mdl)
+
+# now to pack into a function 
+
+estimate_breakpoint_and_pval <- function(df){
+  null_mdl <- lm(ex ~ year, data = df) 
+  seg_mdl <- segmented::segmented(null_mdl)   
+  anova_diff <- anova(null_mdl, seg_mdl)
+  
+  list(
+    null = null_mdl, 
+    seg = seg_mdl,
+    diff = anova_diff
+  )
+}
+
+segmented_breakpoint_models <- 
+  hmd_ex_selected_countries_with_synth %>% 
+  filter(year >= 1979) %>% 
+  group_by(code, x, sex) %>% 
+  nest() %>% 
+  mutate(
+    mdl_outputs = map(data, estimate_breakpoint_and_pval)
+  ) %>% 
+  unnest_wider(mdl_outputs) %>% 
+  mutate(
+    ftest_pval = map_dbl(diff, ~.[2,6]),
+    stat_sig_preference = ftest_pval < 0.05
+  ) %>% 
+  mutate(
+    breakpoint = map_dbl(seg, ~.[["psi"]][1,2]),
+    breakpoint_se = map_dbl(seg, ~.[["psi"]][1,3])
+  )
+
+segmented_breakpoint_models
+
+
+
+estimate_breakpoint_and_pval_v2 <- function(df){
+  null_mdl <- lm(ex ~ year, data = df) 
+  seg_mdl <- segmented::segmented(null_mdl, seg.Z= ~year, psi = 2010)
+  seg2_mdl <- segmented::segmented(null_mdl, seg.z= ~year, psi= c(1985, 2010)) # added to test
[…]

### 2022-06-02 apps/compare_cps/compare_changepoints/app.R

@@ -1,0 +1,95 @@
+library(tidyverse, here)
+library(shiny)
+
+projdir <- here("apps", "compare_cps")
+
+country_selection <- c(
+  "England & Wales" = "GBRTENW", 
+  "Scotland"        = "GBR_SCO",
+  "Italy"           = "ITA",
+  "Spain"           = "ESP",
+  "Germany"         = "DEUTSYNTH",
+  "France"          = "FRATNP",
+  "Germany (East)"  = "DEUTE",
+  "Germany (West)"  = "DEUTW"
+)
+
+cp_paths <- read_rds(here(projdir, "data", "cp_paths.rds"))
+country_data <- read_rds(here(projdir, "data", "countries_data.rds"))
+
+
+# Define UI for application that draws a histogram
+ui <- fluidPage(
+
+    # Application title
+    titlePanel("Compare changepoint models"),
+
+    # Sidebar with a slider input for number of bins 
+    sidebarLayout(
+        sidebarPanel(
+          selectInput("code_select", "Select a country", choices = country_selection)
+        ),
+
+        # Show a plot of the generated distribution
+        mainPanel(
+          plotOutput("mainplot")
+        )
+    )
+)
+
+# Define server logic required to draw a histogram
+server <- function(input, output) {
+  # reactives
+  get_country_data <- reactive({
+    country_data %>% 
+      filter(code == input$code_select) %>%
+      rename(start_age = x) %>% 
+      group_by(sex, start_age) %>% 
+      arrange(year) %>% 
+      mutate(ex_prime = ex - lag(ex)) %>% 
+      ungroup() %>% 
+      filter(year >= 1980)
+  })
+  
+  get_relevant_cppaths <- reactive({
+    cp_paths %>% 
+      filter(code == input$code_select) %>% 
+      rename(year = x) 
+  })
+  
+  make_title <- reactive({
+    glue::glue("Change in ex for {names(country_selection[country_selection==input$code_select])}")
+  })
+  
+  
+  output$mainplot <- renderPlot({
+      get_country_data() %>% 
+      ggplot(aes(year, ex_prime)) + 
+      geom_point() + 
+      facet_grid(sex ~ start_age) + 
+      geom_hline(yintercept = 0) +
+      labs(
+        x = "year", 
+        y = "Annual change in ex (years)",
+        title = make_title()
+      ) +
+      geom_path(
+        aes(
[…]

### 2022-06-05 apps/compare_cps_and_bps/app.R
semi working app

@@ -1,0 +1,165 @@
+library(tidyverse)
+library(here)
+library(shiny)
+
+projdir <- here("apps", "compare_cps_and_bps")
+
+country_selection <- c(
+  "England & Wales" = "GBRTENW", 
+  "Scotland"        = "GBR_SCO",
+  "Italy"           = "ITA",
+  "Spain"           = "ESP",
+  "Germany"         = "DEUTSYNTH",
+  "France"          = "FRATNP",
+  "Germany (East)"  = "DEUTE",
+  "Germany (West)"  = "DEUTW"
+)
+
+cp_paths <- read_rds(here(projdir, "data", "cp_paths.rds"))
+seg_models <- read_rds(here(projdir, "data", "segmented_models.rds"))
+country_data <- read_rds(here(projdir, "data", "countries_data.rds"))
+
+
+# Define UI for application that draws a histogram
+ui <- fluidPage(
+
+    # Application title
+    titlePanel("Compare changepoint models"),
+
+    # Sidebar with a slider input for number of bins 
+    sidebarLayout(
+        sidebarPanel(
+          selectInput("code_select", "Select a country", choices = country_selection),
+          checkboxInput("add_obs", "Click to add data points")
+        ),
+
+        # Show a plot of the generated distribution
+        mainPanel(
+          tabsetPanel(id = "maintab",
+            tabPanel("Changepoints",
+               plotOutput("cp_plot")
+            ),
+            tabPanel("Segmented",
+              plotOutput("seg_predictions")
+            ),
+            tabPanel("Both"
+
+                     
+            )
+          )
+        )
+    )
+)
+
+# Define server logic required to draw a histogram
+server <- function(input, output) {
+  # reactives
+  get_country_data <- reactive({
+    country_data %>% 
+      filter(code == input$code_select) %>%
+      rename(start_age = x) %>% 
+      group_by(sex, start_age) %>% 
+      arrange(year) %>% 
+      mutate(ex_prime = ex - lag(ex)) %>% 
+      ungroup() %>% 
+      filter(year >= 1980)
+  })
+  
+  get_relevant_cppaths <- reactive({
+    cp_paths %>% 
+      filter(code == input$code_select) %>% 
+      rename(year = x) 
+  })
+  
+  get_relevant_segmods <- reactive({
+    seg_models %>% 
+      filter(code == input$code_select) 
+  })
+  
+  
+  
+  make_title <- reactive({
[…]

### 2022-06-07 apps/compare_cps_and_bps/app.R
update before deploying

@@ -2,8 +2,10 @@ library(tidyverse)
 library(here)
 library(shiny)
 
-projdir <- here("apps", "compare_cps_and_bps")
+# Define project directories 
+#projdir <- here("apps", "compare_cps_and_bps")
 
+# Define and label countries 
 country_selection <- c(
   "England & Wales" = "GBRTENW", 
   "Scotland"        = "GBR_SCO",
@@ -15,34 +17,52 @@ country_selection <- c(
   "Germany (West)"  = "DEUTW"
 )
 
-cp_paths <- read_rds(here(projdir, "data", "cp_paths.rds"))
-seg_models <- read_rds(here(projdir, "data", "segmented_models.rds"))
-country_data <- read_rds(here(projdir, "data", "countries_data.rds"))
+# Read data 
+# cp_paths <- read_rds(here(projdir, "data", "cp_paths.rds"))
+# seg_models <- read_rds(here(projdir, "data", "segmented_models.rds"))
+# country_data <- read_rds(here(projdir, "data", "countries_data.rds"))
+
+cp_paths <- read_rds("data/cp_paths.rds")
+seg_models <- read_rds("data/segmented_models.rds")
+country_data <- read_rds("data/countries_data.rds")
 
 
-# Define UI for application that draws a histogram
+# Functions (move to source when grown enough)
+
+format_legends <- function(){
+  list(
+    scale_alpha_manual("BIC", values = c("1" = 1.0, "2" = 0.5, "3" = 0.25)),
+      scale_linetype_manual(values = c("1" = "solid", "2" = "dashed", "3" = "dashed")), 
+      scale_size_manual(values = c(1, 1.4, 1.8)), 
+      guides(alpha = "none", linetype = "none")       
+  )
+}
+
+# Define UI
 ui <- fluidPage(
 
     # Application title
     titlePanel("Compare changepoint models"),
 
-    # Sidebar with a slider input for number of bins 
     sidebarLayout(
         sidebarPanel(
           selectInput("code_select", "Select a country", choices = country_selection),
-          checkboxInput("add_obs", "Click to add data points")
+          checkboxInput("add_obs", "Click to add data points", value = TRUE)
         ),
 
         # Show a plot of the generated distribution
         mainPanel(
           tabsetPanel(id = "maintab",
-            tabPanel("Changepoints",
-               plotOutput("cp_plot")
-            ),
             tabPanel("Segmented",
               plotOutput("seg_predictions")
             ),
[…]

### 2022-06-09 draft_working_paper.Rmd
semi working app
knit updated results

@@
 date: "22/02/2022"
-output: word_document
+output:
+  html_document:
+    df_print: paged
+    always_allow_html: yes
+  word_document: default
 ---
@@
   unnest(joined_data)
+
+```
+
+I want to save segmented_breakpoints_models so I can explore it in an app
+
+```{r}
+write_rds(segmented_breakpoints_models, here("apps", "compare_cps_and_bps", "data", "segmented_models.rds"))
 
@@
 
-```{r}
+```{r eval = FALSE}
 
@@
 ```
+# Summary of differences and similarities between segmented approach and changepoint approach
+
+In order to compare the results of the two modelling approaches, an R shiny app was built.
+This app allows each of the countries of interest to be selected, and has three tabs for different displays.
+These tabs are labelled: 'Segmented', which shows the results of the segmented regression approach; 'Changepoints', which shows the result of the changepoint approach; and both, which compares the best performing segmented and changepoint models.
+For each country, the results are shown as a two-by-two array, with each sex a different column and each row a different starting age.
+An option is provided to show the data for the populations, from which the models are derived.
+
+For both the Segmented and Changepoint displays, polylines are shown for the best performing zero, one or two break models.
+Of these three models, the model with the lowest BIC is shown as a thick, solid line of full opacity, and the other two lines are shown as broken lines with some level of transparency.
+This means the polyline corresponding to the best-performing model should be the most prominent of the series of polylines shown.
+
+For England & Wales, the segmented approach identified a two breakpoint model has having the lowest BIC, for all four combinations of sex and starting age.
+These two breakpoint models suggest a moderate level of annual increase throughout the 1980s and 1990s, a further increase throughout the late 1990s/early 2000s, then a consistent and severe decrease in average annual rate of improvement after around 2012.
+The corresponding changepoint approach identified a one changepoint model for both males and females from birth, a two changepoint model for males from age 65 years, and a zero changepoint model for females from age 65.
+Despite these apparent differences between approaches, three of the four changepoint models identify a substantive decrease in the rate of annual improvement around or soon after 2012.
+Comparison of the best-performing changepoint and segmented model shows that, despite differences in the number of breaks identified, the two types of model tend to show similar patterns over time.
+
+For Scotland, the segmented model identifies a two breakpoint model as having the lowest BIC for males and females from age 65, and males from birth.
+Each of these models are qualitatively similar to those for England & Wales, with an increase in the rate of improvement from the late 1990s/early 2000s to around 2012, then a decrease in the rate of improvement thereafter; for males from birth the average annual rate of improvement after around 2013 is estimated to be negative (i.e.
+declining mortality trends rather than stalling mortality trends).
+For females from birth a one breakpoint model is identified, which also estimates a negative average improvement (i.e.
+falling life expectancy) after around 2014.
+
+Unlike with England & Wales, the changepoint approach leads to results which are not highly concordant with those produced using the segmented approach.
+This appears to be because, when the two changepoint model is used, the two points identified can be a single year apart, and  for females from birth and males from age 65, they are.
+Because the data used are for single years, rather than averaged across a number of years, a single point can be substantially higher or lower than the general trend.
+This can be due to factors such as especially mild winters for years with exceptionally high values, or especially cold winters, or hot summers, or especially infectious and pathogenic strains of infectious disease for years with exceptionally low or negative values.
+By fitting two changepoints one year apart, the algorithm is in effect identifying and removing the effect of these outlier years from the estimate of the trends either side.
+For smaller populations like Scotland, these outliers are likely to be more severe than for larger populations like England & Wales, which may explain the identification of such outlier years within the changepoint analysis in Scotland but not England & Wales.
+
+For Italy, a slowdown in gains in life expectancy is identified by the segmented approach for both sexes from birth, and for females from age 65.
+This slowdown was estimated to have occurred around 2006.
+For males from birth, like with England & Wales, a two breakpoint model with a period of faster improvement rates - starting in the late 1990s and ending around 2006 - was also identified.
+For males from age 65 a zero breakpoint model was preferred.
+For the changepoint analysis, as with Scotland, a two changepoint specification was preferred to other specifications for females of both starting ages, and for males from age 65.
+This identified an outlier, of exceptionally fast improvement, in 2004-5, which is adjacent to two years of negative change in 2003-04 and 2005-06.
+For females, the average trend after this outlier was estimated to be slightly below that of the preceding period, providing some indication of a slowdown in life expectancy around 2005.
+
+For Spain, the segmented approach shows a modest slowdown in annual improvements after 2012 or slightly later for females of both starting ages, and males from age 65.
+For males from birth a two breakpoint model is identified, but which identifies a short period in the late 1980s as one of modest decline in life expectancy, then a continued period of improvement thereafter.
+For the changepoint approach, a zero changepoint model was preferred for both females and males from age 65, whereas for life expectancy from birth a two changepoint model, identifying a single year or short period of years as being of negative or negligible improvement in life expectancy during 1984-85 (females), or much of the 1980s (males), was preferred instead.
+When comparing the preferred changepoint with breakpoint models, the breakpoint models were generally able to identify a slowdown in the 2010s, whereas the changepoint models were not.
+
+For Germany (total, synthetic), the preferred segmented models all included a decline in improvement in the latter half of the period considered.
+For females at both starting ages, and for males from birth, this slowdown was identified to have occurred from around 2006/2007, whereas for males from age 65 to have occurred from 2010.
+For males from both starting ages, however, two breakpoint models were preferred, with faster rates of improvement identified from the mid 1990s, and the decline in life expectancy gains largely a return to prior trends more common in the 1980s and first half of the 1990s.
+The results from the changepoint approach are somewhat discordant, through all identify some slowdown in rates of improvement at the end of the series compared with the start.
+
+For France, the segmented approach estimates a slowdown in life expectancy improvements beginning from around 2011 (female, from age 65) to 2013 (female, from birth).
+For males two breakpoint models including a slight period of more rapid improvment in the 2000s are preferred, whereas for females either a single breakpoint model (from age 65) or two breakpoints of successive decline in improvement (from birth, the first breakpoint being in 2011-12) were instead identified.
+The changepoint models indicated a preference for the two changepoint specification, but for all four populations these functioned as outlier detectors, identifying the year 2004-05 as one in which especially rapid improvement occurred (surrounded by two years of especially low improvement); for three of the four changepoint models, a lower rate of average improvement was predicted at the end of the series than at the start, though this slowed rate of improvement was much less than for England & Wales.
+
+## Discussion relating to above
+
+Though there are some important differences between the number, position and magnitude of breakpoints and changepoints produced using the two approaches, these largely appear to be explicable in terms of differences between how the breakpoints/changepoints were identified.
+More specifically, for some populations the changepoint approach identified changepoints that were a single year apart, and correspond to years in which the annual change in values were either especially high or especially low compared with surrounding years.
+In effect, such changepoint models functioned as a kind of outlier removal system, removing a single value from a longer series in order to allow the gradient/intercept to better fit remaining values.
+Although this feature was not the intended aim when producing the changepoint analysis, it is clear from looking at annual changes in life expectancy that some annual changes are outliers; especially large annual improvements can be caused by some combination of particularly favourable climactic and infectious disease conditions (relatively mild summers and winters; strains of seasonal influenza with relatively low infectivity and/or pathogenicity, perhaps due to effective vaccination campaigns), and especially modest or negative annual improvements can be caused by the converse.
+It is important to note that, when looking at unsmoothed annual data, individual years in which life expectancy was lower than in the preceding year are, though infrequent, not uncommon, and may be expected to occur once or more over the full series of years over which the analyses are performed.
+Furthermore, outlier years can be expected to be more extreme, and more frequent, for smaller populations than larger populations, which appears to explain differences between the best changepoint model for Scotland as compared with England & Wales.
+Additionally, because of phenomena including forward mortality displacement and acquired immunity, especially 'bad' and 'good' years should not be expected to be independently distributed, but instead (negatively) predictive of each other.
+More precisely, annual life expectancy series oscillate slightly, rather than are an entirely random process.
+(This can be demonstrated by showing that ARIMA(1,1,0) models tend to fit the data slightly better than ARIMA(0,1,0) models.)
+
+Regardless, the results taken together provide good evidence that England & Wales experienced a substantial decline in rates of annual mortality improvement (i.e.
+stalling) after around 2012, as well as fairly good evidence that Scottish populations may have experienced a decline as or more severe contemporaneously.
+For Scotland, the segmented regression approach implies that life expectancy has fallen, rather than simply stalled, in recent years; however the greater discordance between the results of the segmented and changepoint approaches, and smaller population size for Scotland compared with England & Wales, suggests the estimates for Scotland may be less reliable than those for England & Wales.
+
+For England & Wales, there appears to be fairly good evidence that two apparently 'competing' interpretations of life expectancy trends are both correct: i.e.
+that it is both true that the late 1990s and 2000s were a period of unusually rapid average improvements in life expectancy, *and* that after around 2012 there has been a fairly severe stalling of such annual gains.
+This finding, based on a data series starting 1979-80, is an important contribution to the evidence base supporting the stalling life expectancy hypothesis, because a number of previous analyses used data starting in the late 1990s or early 2000s (EXAMPLES?).
+Without the longer time series considered here, such earlier analyses would not have been able to discount the possibility that any breakpoint identified around 2010/2 or thereabouts represents a return to earlier trends of more modest improvement, rather than the emergence of a new trend of unusually low longevity gains.
+
+The analyses presented here also reveal some similarities and differences between England & Wales and other European nations.
+In particular, we can see that England & Wales was not the only population where the segmented regression approach identified a two breakpoint model as preferable to a zero or one breakpoint model, and where the two breakpoint model identified included both a period of increased, and then a period of decreased, improvement in life expectancy trends as compared with the initial period.
+When looking at trends in life expectancy change from birth, this jump-then-fall pattern is identified in Italy, Germany (combined) and to a lesser extent France for males, though not for females.
+And when looking at comparable trends in conditional life expectancy change from age 65, a jump-then-fall pattern is evident in Spain, France, Germany, again for males only.
+When looking at females, the fall component of this pattern was identified in Germany, Spain, France and Italy, but the jump component seen in males was not; this appears to be the case both for trends in life expectancy from birth, and from age 65.
+
+Though the presence of a fall component - i.e.
+a reduced rate of improvement starting in the late 2000s or early 2010s - was identified in these other European nations as well as England & Wales, in England & Wales the extent of the recent fall appears to be more uniform across population subgroups, more consistently identifiable using the two methods applied here, and more severe than in the other European comparator nations.
+Within Scotland, despite a smaller population and greater discordance between the segmented and changepoint methods, there are indications that life expectancy may be falling, rather than simply 'stalling', with the trend in annual changes in life expectancy from birth being negative for both males and females.
+(It should also be noted that, for females, the French trends from birth are also close to zero.)
+
+In summary, it appears that patterns in life expectancy change over time are *not qualitatively* dissimilar in England & Wales (and Scotland) as compared with other Continental European nations, but are *quantitatively* dissimilar in terms of both the increase in rate of improvement in trends during the late 1990s and 2000s, and the severity of the *fall* in improvement in trends seen since around 2012.
+Put more simply, it appears that what has occurred in England & Wales has also occurred in other European nations, only *more so*.
+This suggests that factors specific to England & Wales (and with more uncertainty, Scotland too) may have exacerbated tendencies -towards both increase and decrease in trends - also present in comparable national populations.
 
@@
 
-```{r}
+```{r eval = FALSE}
 # Start with Scotland, female, 0

### 2022-06-09 apps/compare_cps_and_bps/app.R

@@ -1,6 +1,7 @@
 library(tidyverse)
 library(here)
 library(shiny)
+library(segmented)
 
 # Define project directories 
 #projdir <- here("apps", "compare_cps_and_bps")
@@ -170,7 +171,7 @@ server <- function(input, output) {
       get_relevant_segmods() %>% 
       ggplot(aes(year, ex_prime, group = factor(num_bp), colour = factor(num_bp), size = factor(num_bp), linetype = bic_rank, alpha = bic_rank)) +
       geom_line() +
-      facet_grid(x ~ sex) +
+      facet_grid(sex ~ x) +
       geom_hline(yintercept = 0) +
       labs(
         x = "year", 

### 2022-06-09 apps/download_mastering_shiny/DownloadHandling/app.R

@@ -1,0 +1,40 @@
+#
+# This is a Shiny web application. You can run the application by clicking
+# the 'Run App' button above.
+#
+# Find out more about building applications with Shiny here:
+#
+#    http://shiny.rstudio.com/
+#
+
+library(shiny)
+ui <- fluidPage(
+  selectInput("dataset", "Pick a dataset", ls("package:datasets")),
+  tableOutput("preview"),
+  downloadButton("download", "Download .tsv")
+)
+
+server <- function(input, output, session) {
+  data <- reactive({
+    out <- get(input$dataset, "package:datasets")
+    if (!is.data.frame(out)) {
+      validate(paste0("'", input$dataset, "' is not a data frame"))
+    }
+    out
+  })
+  
+  output$preview <- renderTable({
+    head(data())
+  })
+  
+  output$download <- downloadHandler(
+    filename = function() {
+      paste0(input$dataset, ".tsv")
+    },
+    content = function(file) {
+      vroom::vroom_write(data(), file)
+    }
+  )
+}
+# Run the application 
+shinyApp(ui = ui, server = server)

### 2022-07-17 draft_working_paper.Rmd
mostly adding Netherlands to comparators

@@
    - Simulated/Synthetic Germany
+- Netherlands (`NLD`) - last minute addition from Danny
 
@@
     "GBR_SCO", "Scotland",
-    "DEUTSYNTH", "Synthetic Germany"
+    "DEUTSYNTH", "Synthetic Germany",
+    "NLD", "Netherlands"
   )
@@
   "DEUTE", 
-  "DEUTW"
+  "DEUTW",
+  "NLD"
 )
@@
   left_join(country_code_lookup) %>% 
-  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy"))) %>% 
+  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
   filter(between(year, 1980, 2020)) %>% 
@@
   left_join(country_code_lookup) %>% 
-  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy"))) %>% 
+  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
   select(code, country, x, sex, estimate, std.error) %>% 
@@
 
-The following combinations of country, starting age, and sex indicate statistically signifanct improvements in model fit if using a model which allows the annual changes to vary by decade
+The following combinations of country, starting age, and sex indicate statistically significant improvements in model fit if using a model which allows the annual changes to vary by decade
 
@@
 * East Germany, Males, e65
+* Netherlands, Males, e65
 
@@
   filter(anova_pval < 0.05) %>% 
+  filter(code == "DEUTE") %>% 
   arrange(desc(anova_pval)) %>% 
@@
 These increases throughout the 1990s were greater than a fifth of a year per year on average.
+
+And for Netherlands
+
+```{r}
+all_null_decadal_models %>% 
+  filter(anova_pval < 0.05) %>% 
+  filter(code == "NLD") %>% 
+  arrange(desc(anova_pval)) %>% 
+  select(code, x, sex, model_alt) %>% 
+  mutate(
+    model_results = map(model_alt, broom::tidy)
+  ) %>% 
+  select(-model_alt) %>% 
+  unnest(model_results) %>% 
+  mutate(
+    term  = case_when(
+      term == "(Intercept)" ~ "Baseline (1980s)",
+      term == "factor(decade)1990" ~ "1990s",
+      term == "factor(decade)2000" ~ "2000s",
+      term == "factor(decade)2010" ~ "2010s",
+      TRUE ~ NA_character_
+    )
+  ) %>% 
+  mutate(
+    label = glue::glue("{round(estimate, 3)}{case_when(p.value < 0.01 ~ '***', p.value < 0.05 ~ '**', p.value < 0.10 ~ '*', TRUE ~ '')} ({round(std.error, 3)})")
+  ) %>% 
+  ungroup() %>% 
+  select(
+    x, Term = term, label
+  ) %>% 
+  pivot_wider(values_from = label, names_from = x) %>% 
+  kable(format = "html", table.attr = "style=\"color: black;\"",caption = "Coefficients of decadal differences, Netherlands, males, by starting age (0 = life expectancy at birth; 65 = life expectancy at age 65 years") %>%
+  kableExtra::kable_styling() %>% 
+  footnote(general = "The baseline is the average annual rate of improvement in the 1980s.
+The other terms indicate increments or decrements from this baseline.
+Statistical significance of coefficients: * if p < 0.10, ** if p < 0.05, *** if p < 0.01"
+  )
+```
+
+So, for the Netherlands, males, e65, the 2000s were a period of exceptionally rapid improvement.
 

### 2022-07-19 draft_working_paper.Rmd
save formatted proposed final images

@@
 
+And a simplified version to save
+
+```{r}
+hmd_ex_selected_countries_with_synth %>% 
+  left_join(country_code_lookup) %>% 
+  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
+  filter(!is.na(country)) %>% 
+  filter(between(year, 1980, 2020)) %>% 
+  filter(x == 0) %>% 
+  ggplot(aes(x = year, y = ex, group = sex, colour = sex, linetype = sex)) + 
+  geom_line() +
+  facet_wrap(~country)  +
+  labs(
+    x = "Year",
+    y = "Life Expectancy at birth (years)",
+    title = "Life expectancies at birth for selected nations",
+    subtitle = "1980 to 2020 or latest available year",
+    caption = "Source: Human Mortality Database.\nSynthetic Germany is an 80%/20% weighted average of West Germany and East Germany respectively",
+    colour = "Sex", linetype = "Sex"
+  ) + 
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+
+ggsave("figures/proposed/e0_trend.png", height = 20, width = 20, units = "cm", dpi = 300)
+
+```
+
 ```{r, echo = FALSE, fig.cap = "Life Expectancy at age 65"}
@@
   )
+
+```
+
+I'm going to make this the first proposed figure
+
+```{r}
+hmd_ex_selected_countries_with_synth %>% 
+  left_join(country_code_lookup) %>% 
+  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>%
+  filter(!is.na(country)) %>% 
+  filter(between(year, 1980, 2020)) %>% 
+  filter(x == 65) %>% 
+  ggplot(aes(x = year, y = ex, group = sex, colour = sex, linetype = sex)) + 
+  geom_line() +
+  facet_wrap(~country) +
+  labs(
+    x = "Year",
+    y = "Life Expectancy at age 65 (years)",
+    title = "Life expectancies at age 65 for selected nations",
+    subtitle = "1980 to 2020 or latest available year",
+    caption = "Source: Human Mortality Database.\nSynthetic Germany is an 80%/20% weighted average of West Germany and East Germany respectively",
+    colour = "Sex", linetype = "Sex"
+  ) + 
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+
+ggsave("figures/proposed/e65_trend.png", height = 20, width = 20, units = "cm", dpi = 300)
 ```
@@
 
-### Changes in life expectancy at age 65
-
-```{r, echo = FALSE, warning=FALSE, message = FALSE, fig.height=8, fig.width=8}
+I now want a simplified and slightly better looking version of this for the main manuscript
+
+```{r}
+change_in_ex_selected_countries %>% 
+  filter(x == 0) %>% 
+  left_join(country_code_lookup) %>% 
+  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
+  filter(!is.na(country)) %>% 
+  filter(between(year, 1980, 2020)) %>% 
+  mutate(delta_ex = delta_ex * 52.25) %>% # Convert to weeks
+  ggplot(aes(x = year, y = delta_ex)) + 
+  geom_point() + 
+  stat_smooth(se = FALSE) + 
+  facet_grid(sex~country) + 
+  geom_hline(yintercept = 0) + 
+  scale_y_continuous(breaks = seq(-30, 50, by = 10)) +
+  theme(
+        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
+  ) + 
+  labs(
+    x = "Year",
+    y = "Change in life expectancy from previous year (in weeks)",
+    title = "Annual change in life expectancy at birth, selected countries",
+    subtitle = "Line: nonlinear smoother over the points",
+    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
+  )
+
+ggsave("figures/proposed/che0_smoother.png", height = 15, width = 25, units = "cm", dpi = 300)
+
+```
+
+And now the same but at age 65
+
+```{r}
 change_in_ex_selected_countries %>% 
@@
   left_join(country_code_lookup) %>% 
-  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
+  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
+  filter(!is.na(country)) %>% 
   filter(between(year, 1980, 2020)) %>% 
-  ggplot(aes(x = year, y = delta_ex, colour=sex)) + 
-  geom_point(alpha = 0.2) + 
-  stat_smooth() + 
-  facet_grid(country~sex) + 
+  mutate(delta_ex = delta_ex * 52.25) %>% # Convert to weeks
+  ggplot(aes(x = year, y = delta_ex)) + 
+  geom_point() + 
+  stat_smooth(se = FALSE) + 
+  facet_grid(sex~country) + 
   geom_hline(yintercept = 0) + 
-  theme(strip.text.y = element_text(angle = 0)) + 
+  scale_y_continuous(breaks = seq(-30, 50, by = 10)) +
+  theme(
+        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
+  ) + 
   labs(
     x = "Year",
-    y = "Year-on-year change in life expectancy",
+    y = "Change in life expectancy from previous year (in weeks)",
     title = "Annual change in life expectancy at age 65, selected countries",
-    subtitle = "Blue line: nonlinear smoother",
+    subtitle = "Line: nonlinear smoother over the points",
     caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
   )
+
+ggsave("figures/proposed/che65_smoother.png", height = 15, width = 25, units = "cm", dpi = 300)
+
 ```
@@
   unnest(joined_data)
+
+```
+
+I'm going to just get the predicted values for the best model
+
+```{r}
+best_breakpoint_model_predictions <- 
+  segmented_breakpoints_models %>% 
+    mutate(bic = map_dbl(mdl_outputs, BIC)) %>% 
+    group_by(code, x, sex) %>% 
+    filter(bic == min(bic)) %>% 
+    mutate(predictions = map2(mdl_outputs, data, predict)) %>% 
+    mutate(data_augmented = map2(data, predictions, ~ .x %>% mutate(prediction = .y))) %>% 
+    select(code, x, sex, data_augmented) %>% 
+    ungroup() %>% 
+    unnest(data_augmented)
+
+```
+
+First I want to show the values against predicted
+
+```{r}
+best_breakpoint_model_predictions %>% 
+  left_join(country_code_lookup) %>% 
+  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
+  filter(!is.na(country)) %>% 
+  filter(x == 0) %>% 
+  ggplot(aes(x = year)) + 
+  facet_grid(sex ~ country) + 
+  geom_point(aes(y = ex), alpha = 0.35) + 
+  geom_line(aes(y = prediction), colour = "red") +
+  theme(
+        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
+  ) + 
+  labs(
+    x = "Year",
+    y = "Life expectancy (years)",
+    title = "Predicted and observed life expectancy at birth, selected countries",
+    subtitle = "Line: Best segmented regression model for the country",
+    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
+  )
+
+ggsave("figures/proposed/e0_breakpoint.png", height = 15, width = 25, units = "cm", dpi = 300)
+
+```
+Now, at age 65
+
+```{r}
+best_breakpoint_model_predictions %>% 
+  left_join(country_code_lookup) %>% 
+  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
+  filter(!is.na(country)) %>% 
+  filter(x == 65) %>% 
+  ggplot(aes(x = year)) + 
+  facet_grid(sex ~ country) + 
+  geom_point(aes(y = ex), alpha = 0.35) + 
+  geom_line(aes(y = prediction), colour = "red") +
+  theme(
+        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
+  ) + 
+  labs(
+    x = "Year",
+    y = "Life expectancy (years)",
+    title = "Predicted and observed life expectancy at age 65, selected countries",
+    subtitle = "Line: Best segmented regression model for the country",
+    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
+  )
+
+ggsave("figures/proposed/e65_breakpoint.png", height = 15, width = 25, units = "cm", dpi = 300)
+
+```
+
+And now, the same, but differences
+
+```{r}
+best_breakpoint_model_predictions %>% 
+  left_join(country_code_lookup) %>% 
+  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
+  filter(!is.na(country)) %>% 
+  filter(x == 0) %>% 
+  group_by(country, sex) %>% 
+  arrange(year) %>% 
+  mutate(
+    ch_ex      = 52.25 * (ex - lag(ex)                 ), 
+    ch_ex_pred = 52.25 * (prediction - lag(prediction) )
+  ) %>% 
+  filter(!is.na(ch_ex), !is.na(ch_ex_pred))  %>% 
+  ggplot(aes(year)) + 
+  geom_point(aes(y = ch_ex), alpha = 0.35) +
+  facet_grid(sex ~ country) +
+  geom_hline(yintercept = 0) + 
+  geom_line(aes(y = ch_ex_pred), colour = "red") +
+  theme(
+    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
+  ) + 
+  scale_y_continuous(breaks = seq(-30, 50, by = 10)) +
+  labs(
+    x = "Year",
+    y = "Change in life expectancy from previous year (weeks)",
+    title = "Predicted and observed annual changes in life expectancy at birth, selected countries",
+    subtitle = "Line: Best segmented regression model for the country",
+    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
+  )
+
+ggsave("figures/proposed/ch_e0_breakpoint.png", height = 15, width = 25, units = "cm", dpi = 300)
+
+```
+And finally for age 65
+
+```{r}
+best_breakpoint_model_predictions %>% 
+  left_join(country_code_lookup) %>% 
+  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
+  filter(!is.na(country)) %>% 
+  filter(x == 65) %>% 
+  group_by(country, sex) %>% 
+  arrange(year) %>% 
+  mutate(
+    ch_ex      = 52.25 * (ex - lag(ex)                 ), 
+    ch_ex_pred = 52.25 * (prediction - lag(prediction) )
+  ) %>% 
+  filter(!is.na(ch_ex), !is.na(ch_ex_pred))  %>% 
+  ggplot(aes(year)) + 
+  geom_point(aes(y = ch_ex), alpha = 0.35) +
+  facet_grid(sex ~ country) +
+  geom_hline(yintercept = 0) + 
+  geom_line(aes(y = ch_ex_pred), colour = "red") +
+  theme(
+    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
+  ) + 
+  scale_y_continuous(breaks = seq(-30, 50, by = 10)) +
+  labs(
+    x = "Year",
+    y = "Change in life expectancy from previous year (weeks)",
+    title = "Predicted and observed annual changes in life expectancy at age 65, selected countries",
+    subtitle = "Line: Best segmented regression model for the country",
+    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
+  )
+
+ggsave("figures/proposed/ch_e65_breakpoint.png", height = 15, width = 25, units = "cm", dpi = 300)
 

### 2022-07-20 draft_working_paper.Rmd
add figure of average change

@@
   facet_wrap(~x)
+
+ggsave("figures/proposed/average_improvement1980-2020.png", height = 20, width = 30, units = "cm", dpi = 300)
+
+hmd_ex_selected_countries_with_synth %>% 
+    filter(year >= 1979) %>% 
+    filter(!(code %in% c("DEUTE", "DEUTW", "DEUTNP"))) %>%  # Using only synthetic germany for longer time period
+    group_by(code, x, sex) %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+    nest() %>% 
+    mutate(
+      rwd_model = map(data, ~lm(delta_ex ~ 1, data = . )),
+      ar_model  = map(data, ~lm(delta_ex ~ lag(delta_ex), data = .))
+    ) %>% 
+    mutate(
+      aic_rwd = map_dbl(rwd_model, AIC),
+      aic_ar  = map_dbl(ar_model, AIC)
+    ) %>% 
+    mutate(
+      diff_aic = aic_ar - aic_rwd
+    )  %>% 
+  # filter(diff_aic > 0)
+  mutate(
+    tidied_ar_model = map(ar_model, broom::tidy)
+  ) %>% 
+  unnest(tidied_ar_model) %>% 
+  filter(term == "(Intercept)") %>% 
+  arrange(desc(estimate)) %>%
+  left_join(country_code_lookup) %>% 
+  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>%
+  filter(!is.na(country)) %>% 
+  select(code, country, x, sex, estimate, std.error) %>% 
+  mutate(display_label = glue::glue("{country}_{x}_{sex}")) %>% 
+  mutate(estimate_weeks = estimate * 52.25) %>% 
+  ggplot(aes(estimate_weeks, fct_reorder(display_label, estimate_weeks))) + 
+  geom_point(aes(colour = x), show.legend = FALSE) +
+  expand_limits( x = 0) + 
+  geom_vline(xintercept = 0) + 
+  labs(
+    x = "Average annual life expectancy change (weeks per year)",
+    y = "Country-sex combination",
+    title = "Average annual improvements by country and sex",
+    subtitle = "Range: 1980 to 2020 or nearest available years",
+    caption = "Source: Human Mortality Database"
+  ) + 
+  facet_wrap(~x)
+
+ggsave("figures/proposed/reduced_average_improvement1980-2020.png", height = 20, width = 25, units = "cm", dpi = 300)
 

### 2022-07-22 appendices.Rmd
add appendices folder

---
title: "Appendices"
author: "Jon Minton"
date: '2022-07-21'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Aim

The aim of this document is to produce a series of appendices to support the main paper. The appendices are listed in a slightly arbitrary order and can be amended as required. All material is written as an R markdown document and the code required to produce analyses are presented alongside the results of such analyses.

# Appendix 1: Are annual ex time series random walks with drift (RWD) or do they have autocorrelation?

Section X of the main manuscript noted, from figure 1a and 1b, that there is substantial annual variation in annual changes in life expectancy. After taking the first difference in these series (i.e. subtracting for each value in the series its previous value), the series of remaining difference may be purely random. Alternatively, even after such differencing the series may have some degree of autocorrelation (i.e. the differenced value for one year may be informative as to the differenced value for the next one or more observations in the series), or there may be a moving average component to the series.

Different patterns of variation in the time series can be modelled using different specifications of the ARIMA modelling approach. ARIMA stands for Autoregressive (AR), Integrated (I), Moving Average (MA), and different ARIMA specifications are represented by the shorthand ARIMA(p, d, q). The p term indicates the number of autoregressive terms, the d term the number of times the series have been differenced, and the q term the number of moving average terms.

In this appendix we are primarily interested in comparing the ARIMA(0, 1, 0) specification with the ARIMA(1, 1, 0) specification. The ARIMA(0, 1, 0) is known as Random-Walk-with-Drift (RWD) and is the simpler of the two specifications. ARIMA(1, 1, 0) includes one autoregressive term p, meaning for each value in the series there is some correlation with the previous value in the series. For this autoregressive term p, a positive coefficient may be interpreted as indicating some degree of 'stickiness' in the series (a better-than-average value is more likely than by chance to be followed by a better-than-average value, and vice versa), whereas a negative coefficient interpreted as indicating some degree of 'oscillation' in the series (a better-than-average value is more likely than chance to be followed by a worse-than-average value, and so on).

This appendix section will first compare the two ARIMA model specifications ARIMA(0,1,0) (RWD) and ARIMA(1,1,0) (Autoregressive and integrated) for each of the time series. It will then use the auto.arima function from the fable package to consider a wider range of ARIMA model specifications and identify which model specifications are preferred for which datasets.

## Preparation

First we load the requisite data, packages, and do the required data processing

```{r}
# This loads the required packages (the pacman package must be installed first, using install.packages("pacman"))
pacman::p_load(tidyverse, fable, here)

# The following code calculates e0 and e65 for the countries under consideration, using lifetables previously extracted from the HMD.

hmd_lt <- read_rds(here("data", "lifetables.rds"))

# Labels for codes 
country_code_lookup <- 
  tribble(
    ~code, ~country,
    "DEUTNP", "Germany",
    "DEUTE", "East Germany",
    "DEUTW", "West Germany",
    "ESP", "Spain",
    "FRATNP", "France", 
    "ITA", "Italy",
    "GBRTENW", "England & Wales",
    "GBR_SCO", "Scotland",
    "DEUTSYNTH", "Synthetic Germany",
    "NLD", "Netherlands"
  )

countries_of_interest <- c(
  "GBRTENW",
  "GBR_SCO",
  "GBR_UK",
  "FRATNP",
  "ESP",
  "ITA",
  "DEUTNP",
  "DEUTE", 
  "DEUTW",
  "NLD"
)

source(here("R", "make_synthetic_germany_functions.R"))
source(here("R", "make_pop_selection.R"))

series_of_interest <- 
  hmd_ex_selected_countries_with_synth %>% 
    left_join(country_code_lookup) %>% 
    mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
    filter(!is.na(country)) %>% 
    filter(between(year, 1979, 2020))

series_of_interest
```
Our series contains ex for x = 0 (life expectancy at birth) and x=65 (life expectancy at age 65 years), for each of the countries of interest, including 'Synthetic Germany'.

## Visualisation

The data series look as follows for x = 0

```{r}
series_of_interest %>% 
  filter(x == 0) %>% 
  group_by(country, sex) %>% 
  arrange(year) %>% 
  mutate(delta_ex = ex - lag(ex)) %>% 
  filter(!is.na(delta_ex)) %>% 
  ggplot(aes(x = year, y = delta_ex)) +
  geom_point() + 
  facet_grid(sex ~ country) + 
  geom_hline(yintercept = 0) + 
  theme(
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
  ) + 
  labs(
    x = "Year",
    y = "Change in life expectancy from previous year (in years)",
    title = "Annual change in life expectancy at birth, selected countries",
    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
  )

```
(n.b. the data in this series are shown in years per year, rather than weeks per year as in the main figure)

The equivalent series for e65 is as follows:

```{r}
series_of_interest %>% 
  filter(x == 65) %>% 
  group_by(country, sex) %>% 
  arrange(year) %>% 
  mutate(delta_ex = ex - lag(ex)) %>% 
  filter(!is.na(delta_ex)) %>% 
  ggplot(aes(x = year, y = delta_ex)) +
  geom_point() + 
  facet_grid(sex ~ country) + 
  geom_hline(yintercept = 0) + 
  theme(
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
  ) + 
  labs(
    x = "Year",
    y = "Change in life expectancy from previous year (in years)",
    title = "Annual change in life expectancy at age 65, selected countries",
    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
  )
```

## Calculating and comparing ARIMA(0,1,0) wih ARIMA(1,1,0)

A time series where each observation in the series is simply dependent on the previous value, plus some random variation (which may be negative), can be expressed as an ARIMA(0, 1, 0) model. By contrast, a model where each observation in the series oscillates slightly (i.e. 'worse-than-average' years are more likely than chance to be followed by 'better-than-average' years, and vice versa), is likely to be represented by an ARIMA(1, 1, 0) model, where the coefficient on this first term (called `p`) should be negative rather than positive.

These models can be fit using the `forecast` package, and model fit compared using the AICc metric.

```{r}

fit_arima_model <- function(series, order){
  series %>% 
    pull(ex) %>% 
    as.ts(start = 1979) %>% 
    forecast::Arima(order = order)
}

ts_model_comparisons <- 
  series_of_interest %>% 
    group_by(country, sex, x) %>% 
    nest() %>% 
    mutate(
      arima_010 = map(data, fit_arima_model, order = c(1, 1, 0)),
      arima_110 = map(data, fit_arima_model, order = c(1, 1, 0))
    ) %>% 
  mutate(
    aicc_arima_010 = map_dbl(arima_010, ~summary(.) %>% pluck("aicc")),
    aicc_arima_110 = map_dbl(arima_110, ~summary(.) %>% pluck("aicc"))
  ) %>% 
  mutate(
    which_preferred = if_else(aicc_arima_010 < aicc_arima_110, "Random Walk", "Autocorrelated")
  )

ts_model_comparisons %>% 
  select(country, sex, x, which_preferred)
```
The Autocorrelated specification (ARIMA(1,1,0)) is preferred to RWD (ARIMA(0,1,0)) for all populations.

The following extracts the autocorrelation coefficients and visualises them.

```{r}

get_ar_term_and_se <- function(mdl){
  tibble(
    ar = pluck(mdl, "coef"),
    ar_coef = pluck(mdl, "var.coef")[1,1] %>% sqrt()
  )
}

ts_model_comparisons %>% 
  mutate(mdl_terms = map(arima_110, get_ar_term_and_se)) %>% 
  select(x, sex, country, mdl_terms) %>% 
  unnest_wider(mdl_terms) %>% 
  arrange(ar) %>% 
  ggplot(aes(ar, fct_reorder(country, ar))) + 
  geom_point() + 
  facet_grid(x ~ sex) + 
  geom_vline(xintercept = 0) + 
  labs(x = "Autoregressive coefficient term", y = "Country",
       title ="Autoregressive coefficient term by country, starting age, and sex")

```
The majority of these coefficients are positive, indicating 'stickiness' in the values in the series, rather than oscillation. The exception is for females for conditional life expectancy at age 65, where the coefficients are negative for England & Wales, France, Scotland, Italy, and Spain. This suggests that for older females the life expectancy series tends to 'oscillate' rather than 'stick'.

## Comparing a wider range of ARIMA models

The `auto.arima` function in the `fable` package allows a larger range of ARIMA-type models to be compared. The following code applies this function to each of the populations.

```{r}
tmp <- 
series_of_interest %>% 
  as_tsibble(key = c(sex, x, country), index = year) %>% 
  model(arima = ARIMA(ex ~ pdq(0:3, 1, 0:3))) %>% 
  report() %>% 
  select(sex, x, country, ar_roots, ma_roots ) %>% 
  arrange(country, x, sex)

tmp
```
The length of the vectors ar_roots and ma_roots indicate, respectively, how many ar or ma terms were identified in the best fitting model for the population indicated by sex, x (starting age) and country. For example, for females in England & Wales, from age 0, an ARIMA(1,1,0) model is preferred, whereas for females in Scotland, from age 0, an ARIMA(0, 1, 1) model is preferred.

There are few populations for which the random-walk-with-drift (RWD) model is preferred to more complex models, but also not a single alternative model specification (such as ARIMA(1,1,0)) which is preferred for the majority of populations.

### 2022-07-22 appendices/changepoint_methods.Rmd
add appendices folder

---
title: "Changepoint Approach"
author: "Jon Minton"
date: '2022-07-22'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Aim

The aim of this appendix is to present results for an alternative method of attempting to detect discontinuities in the series of life expectancy changes for different countries.

# Description of approach

* We first fit a 'null' (no change) model to the data. These are models that represent the assumption that no change occurred in trends in life expectancy over time. 
* Then we fit a series on one-break models to the data. These are models that represent the assumption that a single significant change occurred to the series over time. From this series of one-break models, we attempt to find the best one-break model. 
* We then fit a series of two-break models to the data. These are models that represent the assumption that were were two significant changes to the series over time. From this series of two-break models, we attempt to find the best two-break models. 
* We now have three models to compare: the null model ($M_0$), the best one-break model ($M_1$), and the best two-break model ($M_2$). Although both $M_1$ can be compared directly^[By 'compared directly', we mean that one model specification can be expressed as a restricted/constrained version of another model, the unrestricted model, with one or more terms in the unrestricted model set to fixed values, usually zero, in the restricted model. Such models can be compared directly using an F-test] with $M_0$, and $M_2$ can be compared directly with $M_0$, $M_2$ cannot necessarily be compared directly with $M_1$. Instead, the triplet of models ($M_0$, $M_1$ and $M_2$) are compared indirectly using BIC^[BIC stands for Bayesian Information Criterion, like the similar AIC (An Information Criterion or Akaike's Information Criterion), is a penalised model fit score. By penalised this means that the fit of the model to the data (more specifically its log likelihood) is calculated, then a 'penalty' is applied to this score based on the complexity of the model. BIC and AIC differ only according to how the penalty is applied, with BIC tending to penalise more complex models more severely than AIC. This means BIC will tend to be more conservative in selecting models, providing some protection against overfitting. Both AIC and BIC can be used to compare both nested and non-nested models based on the same dataset, unlike the F-test, the Lagrange Multiplier Test, and so on. Measures of model fit should *not* be used to compare models fit to different datasets.^[
For example, it would be wrong to conclude that a model with an $R^2$ or adjusted $R^2$ fit to one dataset of 0.80 is 'better' than a model with an $R^2$ or adjusted $R^2$ of 0.50 fit to a different dataset, even though $R^2$ and its variants are often (mis)interpreted in this way
], which like AIC provides a penalised model fit score. Lower BIC scores indicate better fit to the data, and so, for each dataset, the model ($M_0$, $M_1$ or $M_2$) with the lowest BIC will be selected. 
* For the best of the three models for each dataset, the breakpoint or breakpoints (if any), and the model predictions, will be presented and visualised.

Breaks in the data are estimated by fitting the following model specifications:
    * **$M_0$**: $\frac{de_x}{dt} \sim \alpha$
    * **$M_1$**: $\frac{de_x}{dt} \sim \alpha + \beta T$, where

$$
T = \left\{
    \begin{array}\\
        1 & \mbox{if } \ t \ge \tau \\
        0 & \mbox{otherwise}
    \end{array}
\right.
$$
    * **M_2**: $\frac{de_x}{dt} \sim \alpha + \beta_1 T_1 + \beta_2 T_2$, where

$$
T_1 = \left\{
    \begin{array}\\
        1 & \mbox{if } \tau_1 \le t \lt \tau_2 \\
        0 & \mbox{otherwise}
    \end{array}
\right.
$$

and

$$
T_2 = \left\{
    \begin{array}\\
        1 & \mbox{if } t \ge \tau_2 \\
        0 & \mbox{otherwise}
    \end{array}
\right.
$$
i.e. the selection of $\tau_1$ and $\tau_2$ partitions the model into three sections: $\frac{de_x}{dt} \sim \alpha$ where $t \lt \tau_1$, $\frac{de_x}{dt} \sim \alpha + \beta_1$ where $\tau_1 \le t \lt \tau_2$, and $\frac{de_x}{dt} \sim \alpha + \beta_2$ where $t \ge \tau_2$. As $\tau_1 \lt \tau_2$, only values of $\tau_2$ which are greater than $\tau_1$ are considered. Otherwise all whole integer values of $\tau$ are seached through for both $M_1$ and $M_2$.

The above analyses are performed for each population group, comprising different combinations of country, sex, and starting age.

# Extracting relevant data

```{r}
# load packages
pacman::p_load(here, tidyverse)

# load data 
hmd_lt <- read_rds(here("data", "lifetables.rds"))

# Labels for codes 
country_code_lookup <- 
  tribble(
    ~code, ~country,
    "DEUTNP", "Germany",
    "DEUTE", "East Germany",
    "DEUTW", "West Germany",
    "ESP", "Spain",
    "FRATNP", "France", 
    "ITA", "Italy",
    "GBRTENW", "England & Wales",
    "GBR_SCO", "Scotland",
    "DEUTSYNTH", "Synthetic Germany",
    "NLD", "Netherlands"
  )

countries_of_interest <- c(
  "GBRTENW",
  "GBR_SCO",
  "GBR_UK",
  "FRATNP",
  "ESP",
  "ITA",
  "DEUTNP",
  "DEUTE", 
  "DEUTW",
  "NLD"
)

source(here("R", "make_synthetic_germany_functions.R"))
source(here("R", "make_pop_selection.R"))

change_in_ex_selected_countries <- 
  hmd_ex_selected_countries_with_synth %>% 
    group_by(code, x, sex) %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
    ungroup()

```

Visualise

```{r}
change_in_ex_selected_countries %>% 
  filter(x == 0) %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(!is.na(country)) %>% 
  filter(between(year, 1980, 2020)) %>% 
  ggplot(aes(x = year, y = delta_ex)) + 
  geom_point() + 
  facet_grid(sex~country) + 
  geom_hline(yintercept = 0) + 
  scale_y_continuous(breaks = seq(-30, 50, by = 10)) +
  theme(
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
  ) + 
  labs(
    x = "Year",
    y = "Change in life expectancy from previous year",
    title = "Annual change in life expectancy at birth, selected countries",
    subtitle = "Line: nonlinear smoother over the points",
    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
  )

change_in_ex_selected_countries %>% 
  filter(x == 65) %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(!is.na(country)) %>% 
  filter(between(year, 1980, 2020)) %>% 
  ggplot(aes(x = year, y = delta_ex)) + 
  geom_point() + 
  facet_grid(sex~country) + 
  geom_hline(yintercept = 0) + 
  scale_y_continuous(breaks = seq(-30, 50, by = 10)) +
  theme(
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
  ) + 
  labs(
    x = "Year",
    y = "Change in life expectancy from previous year",
    title = "Annual change in life expectancy at age 65, selected countries",
    subtitle = "Line: nonlinear smoother over the points",
    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
  )

```

The following are functions used to try to identify the best possible breakpoint/changepoint years

```{r}

# Breakpoint functions

run_alt_with_given_tau <- function(tau, df){
  df <- 
    df %>% 
    filter(!is.na(delta_ex)) %>% 
    mutate(
      T_param = ifelse(year < tau, FALSE, TRUE)
    )

  lm(delta_ex ~ T_param, data  = df)
}

get_best_tau_from_optim <- function(df, buffer = 2, what = c("tau", "optim")){
  # We can put in some data validation checks too

  what <- match.arg(what)

  years <- df$year 
  tau_lower <- min(years) + buffer # the lower bound 
  tau_upper <- max(years) - buffer # the upper bound
  tau_start <- (tau_upper + tau_lower) / 2 # start just in the middle 
  # (This shouldn't matter if the algorithm is robust)

  get_result_from_model <- function(par, df, what = c("AIC", "BIC", "model")){
    # Again, can carry out more validation checks 
    what <- match.arg(what)

    tau <- par[["tau"]] # This is how parameters are packed up by optim
    model <-
      df %>% 
      arrange(year) %>% 
      mutate(delta_ex = ex - lag(ex)) %>% 
      filter(!is.na(delta_ex)) %>% 
      mutate(T_param = ifelse(year < tau, FALSE, TRUE)) %>% 
      lm(delta_ex ~ T_param, data = .)
    if (what == "model") {return(model)      }
    if (what == "BIC")   {return(BIC(model)) }
    if (what == "AIC")   {return(AIC(model)) }

    # NULL should never be returned! If it has something's gone wrong!
    NULL
  }

  optim_obj <- 
    optim(
      par = list(tau = tau_start), 
      fn = get_result_from_model, 
      method = "L-BFGS-B",
      lower = tau_lower,
      upper = tau_upper,
      df = df
    )
  # Need to make the result dependent on the what argument

  if (what == "tau") {return(optim_obj[["par"]])}
  else if (what == "optim"){ return(optim_obj) }

  # Again, the following should never be triggered
  NULL
}

get_best_tau_from_gridsearch <- function(df, buffer = 2, what = c("best", "all")){
  # We can put in some data validation checks too

  what <- match.arg(what)

  years <- df$year 
  tau_lower <- min(years) + buffer # the lower bound 
  tau_upper <- max(years) - buffer # the upper bound
  tau_searchrange <- tau_lower:tau_upper
  get_result_from_model <- function(tau, df, what = c("BIC", "AIC", "model")){ #I've changed the order to default to BIC (as more comparable with segmented)
    # Again, can carry out more validation checks 
    what <- match.arg(what)

    model <-
      df %>% 
      arrange(year) %>% 
      mutate(delta_ex = ex - lag(ex)) %>% 
      filter(!is.na(delta_ex)) %>% 
      mutate(T_param = ifelse(year < tau, FALSE, TRUE)) %>% 
      lm(delta_ex ~ T_param, data = .)
    if (what == "model") {return(model)      }
    if (what == "BIC")   {return(BIC(model)) }
    if (what == "AIC")   {return(AIC(model)) }

    # NULL should never be returned! If it has something's gone wrong!
    NULL
  }

  search_df <- tibble(
    tau = tau_searchrange
  ) %>% 
    mutate(bic = map_dbl(tau, get_result_from_model, df = df))

  if (what == "all") {return(search_df)}

  if (what == "best") {
    out <- 
      search_df %>% 
      filter(bic == min(bic)) %>% 
      select(tau, bic)
    return(out) # I've changed this to compare both tau and bic more easily with 2 cp models
  }
  # Again, the following should never be triggered
  NULL
}

# This is a modification of get_best_tau_from_gridsearch, for two taus
get_best_taus_from_gridsearch <- function(df, buffer = 2, what = c("best", "all")){
  # We can put in some data validation checks too

  what <- match.arg(what)

  years <- df$year 
  tau_lower <- min(years) + buffer # the lower bound 
  tau_upper <- max(years) - buffer # the upper bound

  # Now need to list the permutations of taus to consider

  tau_grid <- expand_grid(
    tau1 = tau_lower:tau_upper,
    tau2 = tau_lower:tau_upper
  ) %>% 
    filter(tau2 > tau1) # tau2 should be greater than tau1

  get_result_from_model <- function(tau1, tau2, df, what = c("BIC", "AIC", "model")){ #I've changed the order to default to BIC (as more comparable with segmented)
    # Again, can carry out more validation checks 
    what <- match.arg(what)

    model <-
      df %>% 
      arrange(year) %>% 
      mutate(delta_ex = ex - lag(ex)) %>% 
      filter(!is.na(delta_ex)) %>% 
      mutate(T_param = ifelse(
        year < tau1, 
        "bp0", 
        ifelse(
          year < tau2, "bp1", "bp2"
          )
        )
      ) %>% 
      lm(delta_ex ~ T_param, data = .)
    if (what == "model") {return(model)      }
    if (what == "BIC")   {return(BIC(model)) }
    if (what == "AIC")   {return(AIC(model)) }

    # NULL should never be returned! If it has something's gone wrong!
    NULL
  }

  search_df <- tau_grid %>% 
    mutate(bic = map2_dbl(tau1, tau2, get_result_from_model, df = df))

  if (what == "all") {return(search_df)}

  if (what == "best") {
    out <- 
      search_df %>% 
      filter(bic == min(bic)) %>% 
      select(bic, tau1, tau2)
    return(out)
  }
  # Again, the following should never be triggered
  NULL
}
```

(The optim-based function takes far too long to run, so is not used)

The following calculates the best single breakpoint model for each population

```{r}
changepoint_breakpoint_models <- 
  hmd_ex_selected_countries_with_synth %>% 
  filter(code != "DEUTNP") %>%   
  filter(year >= 1979) %>% 
  group_by(code, x, sex) %>% 
  nest() %>% 
  mutate(
    mdl_outputs = map(data, get_best_tau_from_gridsearch, what = "all")
  )
```
The relationship between proposed `tau` (i.e. changepoint year) and BIC is shown for England & Wales below

```{r}
changepoint_breakpoint_models %>% 
  filter(code == "GBRTENW") %>% 
  unnest(mdl_outputs) %>% 
  group_by(x, sex) %>% 
  mutate(min_bic = bic == min(bic)) %>% 
  ungroup() %>% 
  ggplot(aes(tau, bic, colour = min_bic, shape = min_bic)) + 
  geom_point() + 
  facet_grid(x ~ sex) +
  scale_colour_manual(values = c(`TRUE` = 'red', `FALSE` = "black")) + 
  scale_shape_manual(values = c(`TRUE` = 'triangle', `FALSE` = 'circle'))

```

### 2022-07-22 draft_working_paper.Rmd
add appendices folder

@@
 So, for the Netherlands, males, e65, the 2000s were a period of exceptionally rapid improvement.
-
-## Changepoint and breakpoint analysis.
-
-Though there may be systemic differences in average changes in life expectancy over time, such differences may not be patterned by decade, and so the above model specification may not have been appropriate to detect any such changes.
-Instead of looking for systemic differences by decade, we can instead look at whether there are any systemic and detectable 'breaks' in the trends over time.
-We should also be mindful that, over the period under consideration, there may potentially be no, one, or more distinct breaks in the series.
-
-To investigate this, we use two approaches to identify both whether and how many breaks may exist in the life expectancy trend data.
-
-* Approach one: breakpoint analysis on life expectancy against time 
-* Approach two: changepoint analysis on life expectancy change rates against time.
-
-For both approaches, the underlying rationale is the same:
-
-* We first fit a 'null' (no change) model to the data. These are models that represent the assumption that no change occurred in trends in life expectancy over time. 
-* Then we fit a series on one-break models to the data.
-These are models that represent the assumption that a single significant change occurred to the series over time.
-From this series of one-break models, we attempt to find the best one-break model.
-* We then fit a series of two-break models to the data.
-These are models that represent the assumption that were were two significant changes to the series over time.
-From this series of two-break models, we attempt to find the best two-break models.
-* We now have three models to compare: the null model ($M_0$), the best one-break model ($M_1$), and the best two-break model ($M_2$).
-Although both $M_1$ can be compared directly^[By 'compared directly', we mean that one model specification can be expressed as a restricted/constrained version of another model, the unrestricted model, with one or more terms in the unrestricted model set to fixed values, usually zero, in the restricted model.
-Such models can be compared directly using an F-test] with $M_0$, and $M_2$ can be compared directly with $M_0$, $M_2$ cannot necessarily be compared directly with $M_1$.
-Instead, the triplet of models ($M_0$, $M_1$ and $M_2$) are compared indirectly using BIC^[BIC stands for Bayesian Information Criterion, like the similar AIC (An Information Criterion or Akaike's Information Criterion), is a penalised model fit score.
-By penalised this means that the fit of the model to the data (more specifically its log likelihood) is calculated, then a 'penalty' is applied to this score based on the complexity of the model.
-BIC and AIC differ only according to how the penalty is applied, with BIC tending to penalise more complex models more severely than AIC.
-This means BIC will tend to be more conservative in selecting models, providing some protection against overfitting.
-Both AIC and BIC can be used to compare both nested and non-nested models based on the same dataset, unlike the F-test, the Lagrange Multiplier Test, and so on.
-Measures of model fit should *not* be used to compare models fit to different datasets.^[
-For example, it would be wrong to conclude that a model with an $R^2$ or adjusted $R^2$ fit to one dataset of 0.80 is 'better' than a model with an $R^2$ or adjusted $R^2$ of 0.50 fit to a different dataset, even though $R^2$ and its variants are often (mis)interpreted in this way
-], which like AIC provides a penalised model fit score.
-Lower BIC scores indicate better fit to the data, and so, for each dataset, the model ($M_0$, $M_1$ or $M_2$) with the lowest BIC will be selected.
-* For the best of the three models for each dataset, the breakpoint or breakpoints (if any), and the model predictions, will be presented and visualised.
-
-Breaks in the data are calculated through two approaches:
-
-1.
-For life expectancy against time ($e_x \sim t$), breakpoints are calculated using the $segmented$ function in the $segmented$ package.
-Both one and two breakpoint models are calculated using this package.
-2. For changes in life expectancy against time ($ \frac{de_x}{dt} \sim t$), regressions which allow varying intercepts are calculated. More specifically, the models being compared are: 
-    * **$M_0$**: $\frac{de_x}{dt} \sim \alpha$
-    * **$M_1$**: $\frac{de_x}{dt} \sim \alpha + \beta T$, where
-
-$$
-T = \left\{
-    \begin{array}\\
-        1 & \mbox{if } \ t \ge \tau \\
-        0 & \mbox{otherwise}
-    \end{array}
-\right.
-$$
-    * **M_2**: $\frac{de_x}{dt} \sim \alpha + \beta_1 T_1 + \beta_2 T_2$, where
-
-$$
-T_1 = \left\{
-    \begin{array}\\
-        1 & \mbox{if } \tau_1 \le t \lt \tau_2 \\
-        0 & \mbox{otherwise}
-    \end{array}
-\right.
-$$
-and
-
-$$
-T_2 = \left\{
-    \begin{array}\\
-        1 & \mbox{if } t \ge \tau_2 \\
-        0 & \mbox{otherwise}
-    \end{array}
-\right.
-$$
-i.e.
-the selection of $\tau_1$ and $\tau_2$ partitions the model into three sections: $\frac{de_x}{dt} \sim \alpha$ where $t \lt \tau_1$, $\frac{de_x}{dt} \sim \alpha + \beta_1$ where $\tau_1 \le t \lt \tau_2$, and $\frac{de_x}{dt} \sim \alpha + \beta_2$ where $t \ge \tau_2$.
-As $\tau_1 \lt \tau_2$, only values of $\tau_2$ which are greater than $\tau_1$ are considered.
-Otherwise all whole integer values of $\tau$ are seached through for both $M_1$ and $M_2$.
-
-For any particular population $D_i$, we should generally expect a high degree of concordance between the breakpoint or breakpoints identified by the `segmented` package, and the changepoint or changepoints identified using annual changes in life expectancy against time.
-However, we should not expect the results from the two approaches to be identical, due to differences in the algorithms used to select the best model parameters.
-
-The above analyses are performed for each population group, comprising different combinations of country, sex, and starting age.
 

### 2022-07-23 appendices/synthetic_germany_approach.Rmd
produce synthetic germany appendix

---
title: "Synthetic Germany"
author: "Jon Minton"
date: '2022-07-23'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

# Why a 20% East Germany/80% West Germany weighting was used to produce 'Synthetic Germany'

In the main paper we mention that a Synthetic Germany was produced as a comparator series by using a weighted average of East German and West German life expectancies. The weighting used was 20% East Germany and 80% West Germany.

This appendix describes why and how that particular weighting was arrived at.

# Core problem and approach taken

The core problem is that we do not have a single Germany data series going back to the 1980s, as Germany was not unified at this time. However, we do have series for East Germany and West Germany. In the HMD separate East and West Germany series were still produced even after reunfication, alongside a complete Germany series.

By having separate East, West and complete Germany series for a common range of years, we can produce a synthetic Germany series as follows:

$$
e_x^{syn}(t) = pe_x^{e}(t) + (1-p)e_x^w(t)
$$
Where the superscripts $syn$, $e$ and $w$ indicate synthetic, east and west respectively.

The parameter $p$ is the share of East Germany in the series.

The optimisation challenge is to select a value of $p$ such that the discrepancy between $e_x^{syn}(t)$ and the observed life expectancies for Germany $e_x^{obs}(t)$ is minimised for all values of $t$.

This discrepancy is calculated using Root-mean-squared-error (RMSE)

Our comparator period is 1990-2010 inclusive. This is a period in which all three series exist and so can be compared, and before any possible slowdown effects are likely to exist.

# Packages

```{r}
library(tidyverse)

```

# Data

We start with lifetable data for German populations. We only need e0 and e65, though the code can be adapted for other ages.

```{r}

# If getting from HMD directly we can use a script 
# get_data_from_hmd.R
# as follows
# pacman::p_load(HMDHFDplus)
# source(here("R", "get_data_from_hmd.R"))

# We need life expectancy at age 0 and 65, which we can get from the lifetables

# We will load these lifetables directly (rather than using the script)

# To allow this script to be self-contained we will refer to the github file location directly

hmd_lt <- read_rds("https://github.com/JonMinton/change-in-ex/blob/main/data/lifetables.rds?raw=true")

# We will now just look for German data, and for ages 0 and 65 only

e0_e65 <- 
  hmd_lt  %>% 
    filter(Age %in% c(0, 65)) %>% 
    filter(str_detect(code, "DEU")) %>% 
    select(code, sex, year = Year, x = Age, ex) %>% 
    arrange(year)

```

# Functions

Commented functions written to produce Synthetic Germany are shown and run below

```{r}
# Functions for creating a synthetic Germany and selecting the optimal share of East and West Germany to produce it

# Background: The Human Mortality Database (HMD) includes life expectancy data for Germany (DEUT), East Germany
# (DEUTE) and West Germany (DEUTW). Of course there are no population data for Germany prior to reunification, 
# but there are separate population data maintained for East and West German populations post-reunification.

# This means DEUTE, DEUTW, and DEUT are reported for a number of common years.

# The aim of this code is to produce a 'Synthetic Germany' for years prior to reunification. This 'Synthetic Germany'
# is based on a weighted average of East and West German population data. The weighting between these two populations
# is a parameter to determine based on an objective or loss function.
# A objective/loss function is a function that, given one or more numeric inputs (which can be varied), produces
# a single numeric output. The aim is to select the input that minimises the value returned by the function.

# In this case, the input is the share of a value from the east german series (and so by implication the share
# of West Germany as well), and the output to try to minimise is the root-mean-square-error (RMSE) between the 
# observed life expectancy value from Germany as a whole, and the implied/simulated life expectancy value for 
# a synthetic Germany with the proposed East/West Germany share.

# The following function makes a synthetic germany series with a particular east germany share (p_east)
# The inputs east_germany and west_germany need to be vectors of the the same length. 
# Additionally the proposed value p_east needs to be between 0 and 1.

make_synthetic_population <- function(east_series, west_series, p_east){
  stopifnot("Series are of different lengths" = length(east_series) == length(west_series))
  stopifnot("Proportion out of possible bounds" = between(p_east, 0, 1))

  east_series * p_east + west_series * (1 - p_east)
}

# The following function compares a proposed synthetic germany series against the referenced/observed values
# The series need to be of the same length to allow them to be pairwise compared.

# The arguments to what can be one of "RMSE', 'abs', and 'rel'
#  - The match.arg function checks that the input to this argument is one of the three valid inputs. 
#  - the default argument for what is RMSE

compare_synthetic_to_reference <- function(synthetic, reference, what = c("RMSE", "abs", "rel")){
  stopifnot("Synthetic and reference are different lengths" = length(synthetic) == length(reference))

  what <- match.arg(what)

  if (what == "RMSE"){
    out <- (synthetic - reference)^2 %>% 
      mean() %>% 
      .^(1/2)
    return(out)
  } else if (what == "abs"){
    return(synthetic - reference)
  } else if (what == "rel"){
    out <- (synthetic - reference)/reference
    return(out)
  } else {
    stop("Wrong what argument (which should have been caught earlier")
  }
  NULL
}

# The following function wraps up the two previous functions, including a default proposed p_east share of 0.2 (20%)

# It includes a number of checks that the inputs are of the expected format, and a function within which pulls 
# the columns and rows of interest to populate each series.

compare_series_get_rmse <- function(data, p_east = 0.20,
                                    east_label = "DEUTE", west_label = "DEUTW", ref_label = "DEUTNP",
                                    comp_period = c(1990, 2010))
{

  stopifnot("data is not a dataframe"      = "data.frame" %in% class(e0_e65)         )
  stopifnot("proportion not valid"         = between(p_east, 0, 1)                   )
  stopifnot("comp_period not a range"      = length(comp_period) == 2                )
  stopifnot("comp_period values not valid" = comp_period %>% is.numeric() %>% all()  )

  # Get the series 
  extract_series <- function(data = data, code_label, comp_period){
    data %>% 
      filter(code == code_label) %>% 
      filter(between(year, comp_period[1], comp_period[2])) %>% 
      arrange(year) %>% 
      pull(ex)
  }

  message("getting East series")
  east_series <- extract_series(data, east_label, comp_period)
  message("getting West series")
  west_series <- extract_series(data, west_label, comp_period)
  message("getting reference series")
  ref_series  <- extract_series(data, ref_label,  comp_period)

  stopifnot("East/West series lengths differ" = length(east_series) == length(west_series))
  stopifnot("Ref series length differs from EastWest" = length(ref_series) == length(east_series))

  message("creating synthetic population and calculating RMSE")

  rmse <- make_synthetic_population(east_series, west_series, p_east = p_east) %>% 
    compare_synthetic_to_reference(reference = ref_series)

  stopifnot("rmse failed: not length 1" = length(rmse) == 1)
  stopifnot("rmse failed: not right class" = class(rmse) == "numeric")

  rmse
}

# The following function (with a function inside it!) runs the optim function and packs the inputs 
# in the way that optim expects.

run_optim <- function(data){
  pack_for_optim <- function(par, data){
    p_east <- par["p_east"]
    data %>% compare_series_get_rmse(p_east = p_east)
  }
  optim(
    par = c(p_east = 0.5),
    fn = pack_for_optim, data = data,
    lower = 0, upper = 1,
    method = "L-BFGS-B"
  )
}

```

# Extracting series

The following manually extracts the series for females for x = 0 (for e0)

```{r}
# Comparator period 1990-2010

# this pulls out the series of values for East Germany
east_series <- 
  e0_e65 %>% 
    filter(x == 0) %>% 
    filter(code == "DEUTE") %>% 
    filter(between(year, 1990, 2010)) %>% 
    filter(sex == "female") %>% 
    arrange(year) %>% 
    pull(ex)

# This pulls out out for West Germany 
west_series <- 
  e0_e65 %>% 
  filter(x == 0) %>% 
  filter(code == "DEUTW") %>% 
  filter(between(year, 1990, 2010)) %>% 
  filter(sex == "female") %>% 
  arrange(year) %>% 
  pull(ex)

# And this pulls it out for Germany as a whole
ref_series <-
  e0_e65 %>% 
  filter(x == 0) %>% 
  filter(code == "DEUTNP") %>% 
  filter(between(year, 1990, 2010)) %>% 
  filter(sex == "female") %>% 
  arrange(year) %>% 
  pull(ex)

```

# Gridsearch approach

The parameter whose loss function (RMSE) we want to minimise is called `p_share`. This is the proportion of the East Germany series (and so by implication `1 - p_share` is the proportin of the West Germany Series.

We start with a gridsearch approach, looking at each full percentage of `p_share` between 0 and 1

```{r}
# And this function looks at each percentage value for share of East Germany 
# between 0% and 100%, and pulls out the RMSE given this.

grid_pshare <- 
  tibble(
    p_share = seq(0, 1, by = 0.01)
  ) %>% 
    mutate(
      rmse = map_dbl(
        p_share, 
        ~ make_synthetic_population(east_series, west_series, p_east = .x) %>% 
          compare_synthetic_to_reference(., reference = ref_series))
    )
# The contents of the mutate function are complicated. Probably too complicated. ;) 
# firstly they evoke map_dbl. This is a particular type of map function that forces the output 
# to be a numeric (double) value. This is what we want as RMSE should be a number, and won't be 
# a whole number (integer)

# The first argument map_dbl takes is p_share, this is the contents of the p_share column 
# created alongside the tibble in the previous step.

# The next step is a function, make up of two steps joined together with the pipe operator %>%
# The use of the ~ at the start is tidyverse shorthand for 'function'. 
# The argument which will be varying is p_share. Unfortunately this isn't the first argument to the 
# first function - make_synthetic_population - which it's passed to. 
# By default, the map function, like the pipe operator, will pass to the first argument slot in the 
# function being passed to. To override this, the .x operator is used.
# This explicitly tells the map function where to put the argument of interest (p_share) in this case. 
# (.x is used because there are map2 functions, taking two inputs. The first of these is referred to as 
# .x, and the second as .y. This convention is also followed here for a single argument.)
# The result of the first step (before the pipe) of this operation in map_dbl is then passed to
# another function, compare_synthetic_to_reference, and fits in its first slot (marked with a . in this case)

# This isn't the most clear way of creating the operation. Instead, declaring a single function 
# beforehand would likely be easier to understand. However it's a fairly concise shorthand that works! 
```

We can see the RMSE for each proposed p_east as follows:

```{r}
grid_pshare

```

And we can visualise how the RMSE varies with different `p_share` values as follows:

```{r}
grid_pshare %>% 
  ggplot(aes(p_share, rmse)) + 
  geom_point()

```

So, there's a clear minimal RMSE when around 20% (p = 0.20) is proposed.

Is it exactly 20%? Let's check

```{r}

grid_pshare %>% 
  filter(rmse == min(rmse))

```

Yes, it's at 20% exactly (to the nearest %)

And what does the Synthetic Germany look like when this proposed mix is used?

```{r}

e0_e65 %>% 
  filter(code %in% c("DEUTE", "DEUTW", "DEUTNP")) %>% 
  filter(sex == "female") %>% 
  filter(x == 0) %>% 
  pivot_wider(names_from = code, values_from = ex, values_fill = NA)  %>% 
  arrange(year) %>% 
  mutate(DEUT_SYNTH = 0.2 * DEUTE + (1 - 0.2) * DEUTW) %>% 
  pivot_longer(cols = c("DEUTNP", "DEUTE", "DEUTW", "DEUT_SYNTH"), names_to = "code", values_to = "e0") %>% 
  ggplot(aes(x = year, y = e0, group = code, colour = code)) + 
  geom_line() +
  labs(
    x = "Year",
    y = "Life expectancy at birth",
    title = "Life expectancy for Synthetic Germany (DEUT_SYNTH)\ncompared with Germany (DEUTNP)",
    subtitle = "Females only. DEUTE: East Germany; DEUTW: West Germany"
  )

```

We can see here that over the reference period `DEUT_SYNTH` (red line) matches `DEUTNP` (blue line) quite closely, as we would expect and hope for.

However the above only completes the calculation for one of the four populations of interest: We also need to do this for e65, and for males.

The following code starts to automate the process for these other three populations:

```{r}
# This is a slightly laborious process, however, involving quite a bit of manual copying and 
# pasting of bits of code. For example the above is only for females, and we might want to know 
# the equivalent value for males. We might also want to calculate life expectancy for different ages.

# To start to automate further we can look at where the code above is repetitive, work out the 
# common patterns in the repeated sections of code, and build functions around them. 
# We can also look for where we've hard-coded values, and replace them with objects/references.

# Within the functions developed, a function called get_series() within the compare_series_get_rmse 
# standardises the extraction of series. It looks as follows:

# extract_series <- function(data = data, code_label, comp_period){
#   data %>% 
#     filter(code == code_label) %>% 
#     filter(between(year, comp_period[1], comp_period[2])) %>% 
#     arrange(year) %>% 
#     pull(ex)
# }

# We can use this extract_series function within the broader function to get the RMSE for a given p_east 
# in fewer steps (for us)

e0_e65 %>% 
  filter(x == 0) %>% 
  filter(sex == "female") %>% 
  compare_series_get_rmse(data =., p_east = 0.2)

# This makes it easier to look for another manual combination of starting age (x) and sex

e0_e65 %>% 
  filter(x == 65) %>% 
  filter(sex == "male") %>% 
  compare_series_get_rmse(data =., p_east = 0.2)

# We can also use map_dbl and nesting to pass different datasets to this function

e0_e65 %>% 
  group_by(sex, x) %>% 
  nest() %>% 
  mutate(rmse = map_dbl(data, compare_series_get_rmse, p_east = 0.2))

# We can repeat the grid search approach by adding a third column, p_share, containing 
# a sequence of possible values of the p_east share. We can get all permutations of 
# data and p_share using expand_grid (or expand.grid) as follows:

expand_grid(
  e0_e65 %>% 
    group_by(sex, x) %>% 
    nest(), 
  p_share = seq(0, 1, by = 0.01)
) %>% 
  select(sex, x, p_share, data)

# Then we can pass to the function we used above to get the RMSE, as well as save to a new object

grid_search_pshares <- 
  expand_grid(
    e0_e65 %>% 
      group_by(sex, x) %>% 
      nest(), 
    p_share = seq(0, 1, by = 0.01)
  ) %>% 
  select(sex, x, p_share, data) %>% 
  mutate(rmse = map2_dbl(data, p_share, ~compare_series_get_rmse(data = .x, p_east = .y)))

grid_search_pshares

```

This has now produced a lot of operations! 
What does it show?

```{r}
grid_search_pshares %>% 
  ggplot(aes(x=p_share, y = rmse)) + 
  geom_line() + geom_point() + 
  facet_grid(sex ~ factor(x))
```

So, it looks like approximately the same answer for each of the four groups. Let's check

```{r}
grid_search_pshares %>% 
  group_by(x, sex) %>% 
  filter(rmse == min(rmse))

```
So, to the nearest whole %, the proposed share is 20% except for males at age 65, where it's 18%.

The following code chunk (not run) discusses how the answer could be made more precise by reducing the mesh size on the grid search.

```{r eval=FALSE}

# Using the grid search approach above, we could get more precise answers to the best share by 
# making the grid search 'mesh' smaller. For example, we could include the resolution 10% by changing 
# the by= argument in seq from 0.01 to 0.001.
# i.e. 
# grid_search_pshares <- 
#   expand_grid(
#     e0_e65 %>% 
#       group_by(sex, x) %>% 
#       nest(), 
#     p_share = seq(0, 1, by = 0.001) # change from by=0.01 here
#   ) %>% 
#   select(sex, x, p_share, data) %>% 
#   mutate(rmse = map2_dbl(data, p_share, ~compare_series_get_rmse(data = .x, p_east = .y)))

# But this would take ten times as long to run, as there are ten times as many values to consider.

# We could also narrow the search space, to say between 0.15 and 0.25, and then narrow the mesh size too

# grid_search_pshares <- 
#   expand_grid(
#     e0_e65 %>% 
#       group_by(sex, x) %>% 
#       nest(), 
#     p_share = seq(0.15, 0.25, by = 0.001) # change from by=0.01, and start and end to 0.15 and 0.25 respectively
#   ) %>% 
#   select(sex, x, p_share, data) %>% 
#   mutate(rmse = map2_dbl(data, p_share, ~compare_series_get_rmse(data = .x, p_east = .y)))

# But again, we only know to do this because we first did the coarser search over the full range of values

# And now imagine if we wanted to do this for each individual age, or for many other countries 
# with similar East/West (or North/South) population breakdowns. Or for many different reference periods

```

The following code shows how the above answers can be arrived at using numeric optimisation using the `optim` function.

```{r}
# The grid search approach can quickly become too computationally intensive. That's where numerical 
# optimisation algorithms can be used instead.

# Within R, the standard function to use for numerical optimisation is optim.
# Optim passes a value or values to a function, and tries to get the smallest or largest value 
# out of this function, rejigging the value/values passed to the function multiple times until it can't 
# get a substantially improved (as in minimised/maximised) value from the function. 
#

# To make the optim function work well with the functional programming approach used by the map functions
# I made a small wrapper function for optim:

# run_optim <- function(data){
#   pack_for_optim <- function(par, data){
#     p_east <- par["p_east"]
#     data %>% compare_series_get_rmse(p_east = p_east)
#   }
#   optim(
#     par = c(p_east = 0.5),
#     fn = pack_for_optim, data = data,
#     lower = 0, upper = 1,
#     method = "L-BFGS-B"
#   )
# }

# This function also specifies that a particular method, 'L-BFGS-B', which 
# gives performs bounded optimisation be used, along with the lower and upper bounds

# I've also given the algorithm a starting value of p = 0.5, which we might assume if we had 
# not looked at the data and performed the analysis previously.

# Having done this packing, we now can use optim for each combination of age and sex to get the 
# best estimate for p_east in far fewer computational steps, and with much higher resolution

synth_germany_best_shares_via_optim <-
  e0_e65 %>% 
  group_by(sex, x) %>% 
  nest() %>% 
  mutate(
    optim_outputs = map(data, run_optim)
  ) %>% 
  mutate(
    optim_share = map_dbl(optim_outputs, pluck, "par")
  )

synth_germany_best_shares_via_optim
```

The optim function contains a lot of additional information. For example, we could get the best RMSE by interrogating the 'value' element as follows:

```{r}
synth_germany_best_shares_via_optim %>% 
  mutate(
    rmse_at_optim_pshare = map_dbl(optim_outputs, pluck, "value")
  )

# So, the best value for RMSE was higher (worse) for males at birth, and lowest (better) for 
# males at age 65.

```

# Conclusion

For all populations, a proposed share of around 20% East Germany/80% West Germany appears reasonable, even though slightly different shares are slightly more optimal for some populations.

Rather than use the exact `p_share`s calculated by `optim` as optimal for each of the populations, we decided to use 20% for all populations for ease of explanation within the main text. The precise optim values given above could be used instead, if the 20% assumption were felt to be too coarse.

### 2022-07-23 appendices/life_expectancy_trends.Rmd
produce life expectancy trend graph

---
title: "Life Expectancies for selected populations"
author: "Jon Minton"
date: '2022-07-23'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Aim

This appendix will show the life expectancies for the populations compared in the paper.

# Data

```{r}
library(tidyverse)

# load data 
hmd_lt <- read_rds("https://github.com/JonMinton/change-in-ex/blob/main/data/lifetables.rds?raw=true")

# Labels for codes 
country_code_lookup <- 
  tribble(
    ~code, ~country,
    "DEUTNP", "Germany",
    "DEUTE", "East Germany",
    "DEUTW", "West Germany",
    "ESP", "Spain",
    "FRATNP", "France", 
    "ITA", "Italy",
    "GBRTENW", "England & Wales",
    "GBR_SCO", "Scotland",
    "DEUTSYNTH", "Synthetic Germany",
    "NLD", "Netherlands"
  )

countries_of_interest <- c(
  "GBRTENW",
  "GBR_SCO",
  "GBR_UK",
  "FRATNP",
  "ESP",
  "ITA",
  "DEUTNP",
  "DEUTE", 
  "DEUTW",
  "NLD"
)

source("https://raw.githubusercontent.com/JonMinton/change-in-ex/main/R/make_synthetic_germany_functions.R")

source("https://raw.githubusercontent.com/JonMinton/change-in-ex/main/R/make_pop_selection.R")

```

# Graphs

## Life Expectancy at birth

```{r}
hmd_ex_selected_countries_with_synth %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(!is.na(country)) %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 0) %>% 
  ggplot(aes(x = year, y = ex, group = sex, colour = sex, linetype = sex)) + 
  geom_line() +
  facet_wrap(~country) +
  labs(
    x = "Year",
    y = "Life Expectancy at birth",
    title = "Life expectancies at birth for selected nations",
    subtitle = "1980 to 2020 or latest available year",
    caption = "Source: Human Mortality Database"
  )
```
Equivalently

```{r}
hmd_ex_selected_countries_with_synth %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(!is.na(country)) %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 0) %>% 
  ggplot(aes(x = year, y = ex, group = country, colour = country, linetype = country)) + 
  geom_line() +
  facet_wrap(~sex) +
  labs(
    x = "Year",
    y = "Life Expectancy at birth",
    title = "Life expectancies at birth for selected nations",
    subtitle = "1980 to 2020 or latest available year",
    caption = "Source: Human Mortality Database"
  )

```

## Life expectancy at age 65

```{r}
hmd_ex_selected_countries_with_synth %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(!is.na(country)) %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 65) %>% 
  ggplot(aes(x = year, y = ex, group = sex, colour = sex, linetype = sex)) + 
  geom_line() +
  facet_wrap(~country) +
  labs(
    x = "Year",
    y = "Life Expectancy at age 65",
    title = "Life expectancies at age 65 years  for selected nations",
    subtitle = "1980 to 2020 or latest available year",
    caption = "Source: Human Mortality Database"
  )
```
Equivalently:

```{r}
hmd_ex_selected_countries_with_synth %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(!is.na(country)) %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 65) %>% 
  ggplot(aes(x = year, y = ex, group = country, colour = country, linetype = country)) + 
  geom_line() +
  facet_wrap(~sex) +
  labs(
    x = "Year",
    y = "Life Expectancy at age 65",
    title = "Life expectancies at age 65 years  for selected nations",
    subtitle = "1980 to 2020 or latest available year",
    caption = "Source: Human Mortality Database"
  )
```

### 2022-07-25 appendices/changepoint_methods.Rmd
complete changepoint approach appendix

@@
 The aim of this appendix is to present results for an alternative method of attempting to detect discontinuities in the series of life expectancy changes for different countries.
+The main approach used was segmented regression using the `segmented` package.
+The segmented regressions were fit to the data itself $y ~ t$.
+This alternative approach instead involves fitting data to the first differences $y(t) - y(t-1) ~ t$.
+
+The alternative approach identified some similar breakpoints/changepoints to the main, segmented approach.
+However it differed for many populations in identifying adjacent or near-adjacent changepoints in the data.
+Further inspection of these pairs of changepoints indicated that the approach developed inadvertently operated as a form of outlier detector algorithm, removing from the series extreme values.
+Though extreme values are a genuine phenomenon within mortality change data, they were not the primary purpose of the analysis, and so this approach was not used within the main manuscript.
 
@@
 # load packages
-pacman::p_load(here, tidyverse)
+
+library(tidyverse)
 
 # load data 
-hmd_lt <- read_rds(here("data", "lifetables.rds"))
+hmd_lt <- read_rds("https://github.com/JonMinton/change-in-ex/blob/main/data/lifetables.rds?raw=true")
 
@@
 
-source(here("R", "make_synthetic_germany_functions.R"))
-source(here("R", "make_pop_selection.R"))
+source("https://raw.githubusercontent.com/JonMinton/change-in-ex/main/R/make_synthetic_germany_functions.R")
+
+source("https://raw.githubusercontent.com/JonMinton/change-in-ex/main/R/make_pop_selection.R")
 
@@
     title = "Annual change in life expectancy at birth, selected countries",
-    subtitle = "Line: nonlinear smoother over the points",
     caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
@@
     title = "Annual change in life expectancy at age 65, selected countries",
-    subtitle = "Line: nonlinear smoother over the points",
     caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
@@
 ```
+# Functions
 
@@
 }
+
+# This function was not used in the end. The reason is because the fitness landscape is 
+# difficult to navigate. Instead a gridsearch approach was used
 
@@
 }
+
+# This is the function actually used, for a single changepoint value
 
@@
 
-(The optim-based function takes far too long to run, so is not used)
+# Running the functions
+
+## Single changepoint
 
@@
 ```
+
 The relationship between proposed `tau` (i.e. changepoint year) and BIC is shown for England & Wales below
@@
   geom_point() + 
-  facet_grid(x ~ sex) +
+  facet_wrap(~ sex + x, scales = "free_y") +
   scale_colour_manual(values = c(`TRUE` = 'red', `FALSE` = "black")) + 
-  scale_shape_manual(values = c(`TRUE` = 'triangle', `FALSE` = 'circle'))
+  scale_shape_manual(values = c(`TRUE` = 'triangle', `FALSE` = 'circle')) + 
+  labs(
+    title = "BIC by proposed changepoint year",
+    subtitle = "Red triangle = lowest BIC. Population: England & Wales",
+    x = expression(paste("Proposed changepoint year ", tau)),
+    y = "BIC (lower is better)",
+    caption = "Source: HMD Lifetables"
+  ) + 
+  theme(legend.position = "none")
 
 ```
+
+We can see here that 2012 is consistently identified as having the lowest BIC for each of the four subpopulations, except for males where x = 0, where 2013 is identified instead.
+We can also see that the 'fitness landscape' (i.e.
+how BIC changes as the proposed changepoing years change) is quite uneven, and so simple numerical optimisation algorithms like Newton-Ralphson are likely to get 'stuck' at local rather than the global optima, and so the results of such algorithms may be quite dependent on starting parameters.
+
+## For two changepoints
+
+The following code identifies the best two changepoint model for each population
+
+```{r}
+changepoint_breakpoints_models <- 
+  hmd_ex_selected_countries_with_synth %>% 
+  filter(code != "DEUTNP") %>%   
+  filter(year >= 1979) %>% 
+  group_by(code, x, sex) %>% 
+  nest() %>% 
+  mutate(
+    mdl_outputs = map(data, get_best_taus_from_gridsearch, what = "all")
+  )
+
+```
+
+Whereas for a single changepoint the relationship between BIC and proposed $\tau$ can be visualised as a scatterplot, for two values $\tau_1$ and $\tau_2$ they can be visualised as a heatmap instead.
+Our only known constraint is that $\tau_2 > \tau_1$, meaning within the heatmap only a triangle rather than rectangle of candidate values needs to be considered.
+
+The following shows this heatmap for England & Wales.
+A red dot is added to indicate the pair of candidate values $\{\tau_1, \tau_2\}$ with the lowest BIC.
+As the range of BIC differs with each dataset, the values are normalised for each subpopulations such that the highest value is 1 and the lowest 0.
+
+```{r}
+changepoint_breakpoints_models %>% 
+  filter(code == "GBRTENW") %>% 
+  unnest(mdl_outputs) %>% 
+  group_by(x, sex) %>% 
+  mutate(bic_std = (bic - min(bic)) / (max(bic) - min(bic))) %>% 
+  ggplot(aes(x = tau1, y = tau2, colour = bic_std, fill = bic_std)) + 
+  geom_raster() + 
+  facet_wrap(~ sex + x, ncol = 2) + 
+  scale_color_viridis_c() + 
+  geom_point(colour= 'red', 
+             data = changepoint_breakpoints_models %>% 
+              filter(code == "GBRTENW") %>% 
+              unnest(mdl_outputs) %>% 
+              group_by(x, sex) %>% 
+              mutate(bic_std = (bic - min(bic)) / (max(bic) - min(bic))) %>% 
+              filter(bic_std == min(bic_std))
+  ) + 
+  coord_fixed() +
+  theme(axis.text.x = element_text(angle = 90)) + 
+  labs(
+    x = expression(paste("Proposed ", tau[1])),
+    y = expression(paste("Proposed ", tau[2])),
+    fill = "Standardised BIC",
+    title = "The relationship between BIC and\nproposed changepoint values",
+    subtitle = "Population: England & Wales. Red dot: pair of values with lowest BIC",
+    caption = "Source: HMD lifetables"
+  )
+```
+
+We can see in the above that, for males for x = 65, a $\{\tau_1, \tau_2\}$ pairing of `{2000, 2012}` has been identified.
+For males and females where x = 0, a near-contiguous pairing of `{2009, 2012}` has been identified.
+And for females where x = 65 a contigious pairing of `{2015, 2016}` had the lowest BIC.
+
+Similar identification of contiguous and near-contiguous pairings were also identified for many other countries. For example, the following shows this for France.
+
+```{r}
+changepoint_breakpoints_models %>% 
+  filter(code == "FRATNP") %>% 
+  unnest(mdl_outputs) %>% 
+  group_by(x, sex) %>% 
+  mutate(bic_std = (bic - min(bic)) / (max(bic) - min(bic))) %>% 
+  ggplot(aes(x = tau1, y = tau2, colour = bic_std, fill = bic_std)) + 
+  geom_raster() + 
+  facet_wrap(~ sex + x, ncol = 2) + 
+  scale_color_viridis_c() + 
+  geom_point(colour= 'red', 
+             data = changepoint_breakpoints_models %>% 
+              filter(code == "FRATNP") %>% 
+              unnest(mdl_outputs) %>% 
+              group_by(x, sex) %>% 
+              mutate(bic_std = (bic - min(bic)) / (max(bic) - min(bic))) %>% 
+              filter(bic_std == min(bic_std))
+  ) + 
+  coord_fixed() +
+  theme(axis.text.x = element_text(angle = 90)) + 
+  labs(
+    x = expression(paste("Proposed ", tau[1])),
+    y = expression(paste("Proposed ", tau[2])),
+    fill = "Standardised BIC",
+    title = "The relationship between BIC and\nproposed changepoint values",
+    subtitle = "Population: FRATNP. Red dot: pair of values with lowest BIC",
+    caption = "Source: HMD lifetables"
+  )
+
+```
+For France, for all four subpopulations, a pairing of `{2004, 2005}` was identified. If we look at the series for France, we can see why this pair of years may have been identified
+
+```{r}
+change_in_ex_selected_countries %>% 
+  filter(code == "FRATNP") %>%   
+  filter(year >= 1979) %>% 
+  left_join(country_code_lookup) %>% 
+  mutate(country = factor(country, levels = c("France"))) %>% 
+  filter(!is.na(country)) %>% 
+  filter(between(year, 1980, 2020)) %>% 
+  mutate(is_2004 = year == 2004) %>% 
+  ggplot(aes(x = year, y = delta_ex)) + 
+  geom_point(aes(shape = is_2004, colour = is_2004)) + 
+  facet_grid(sex ~ x) +
+  scale_shape_manual(values =c(`TRUE` = "triangle", `FALSE` = "circle")) + 
+  scale_colour_manual(values =c(`TRUE` = "darkred", `FALSE` = "black")) + 
+  geom_hline(yintercept = 0) + 
+  theme(
+        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
+        legend.position = 'none'
+  ) + 
+  labs(
+    x = "Year",
+    y = "Change in life expectancy from previous year",
+    title = "Annual change in life expectancy in France",
+    subtitle = "Columns: starting age; Red triangle: Year 2004",
+    caption =  "Source: Human Mortality Database"
+  )
+
+```
+
+We can see that for each of this subpopulation, year = 2004 (i.e.
+the change in ex from 2003 to 2004) was much higher than all other years, and so these points are outliers.
+So, it appears the changepoint approach has inadertently 'detected' this outlier.
+
+# Best model for each population
+
+The following code shows the best performing model, from the best zero, one and two changepoint models, for each of the populations.
+
+```{r}
+estimate_null <- function(df){
+  df %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+    filter(!is.na(delta_ex)) %>% 
+    lm(delta_ex ~ 1, data = .)
+}
+
+  hmd_ex_selected_countries_with_synth %>% 
+    filter(code != "DEUTNP") %>%   
+    filter(year >= 1979) %>% 
+    group_by(code, x, sex) %>% 
+    nest() %>% 
+    mutate(
+      mdl_0 = map(data, estimate_null),
+      bic_0    = map_dbl(mdl_0, BIC)
+    ) %>% 
+    select(code, x, sex, mdl_0, bic_0) %>% 
+    left_join(
+      changepoint_breakpoint_models %>% 
+        unnest(mdl_outputs) %>% 
+        group_by(code, x, sex) %>% 
+        filter(bic == min(bic)) %>% 
+        select(code, x, sex, tau1 = tau, bic_1 = bic)  
+    ) %>% 
+    left_join(
+      changepoint_breakpoints_models %>% 
+        unnest(mdl_outputs) %>% 
+        group_by(code, x, sex) %>% 
+        filter(bic == min(bic)) %>% 
+        select(code, x, sex, tau1_2 = tau1, tau2_2 = tau2, bic_2 = bic)
+    ) %>% 
+  select(-mdl_0) %>% 
+  ungroup() # %>% 
+run_tau_1 <- function(tau, df){
+  df %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+    filter(!is.na(delta_ex)) %>% 
+    mutate(T_param = ifelse(year < tau, FALSE, TRUE)) %>% 
+    lm(delta_ex ~ T_param, data = .)
+}
+
+run_tau_2 <- function(tau1, tau2, df){
+  df %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+    filter(!is.na(delta_ex)) %>% 
+    mutate(T_param = ifelse(
+      year < tau1, 
+      "bp0", 
+      ifelse(
+        year < tau2, "bp1", "bp2"
+        )
+      )
+    ) %>% 
+    lm(delta_ex ~ T_param, data = .)
+}
+
+make_path_coords <- function(df, taus = NULL, mdl){
+  stopifnot(
+    "wrong number of coefficients for the model type" = 
+    length(coefficients(mdl)) == length(taus) + 1
+  )
+
+  min_x = min(df$year)
+  max_x = max(df$year)
+
+  coeffs <- coefficients(mdl)
+
+  y0 <- coeffs[1]
+
+  out <- tribble(
+    ~x, ~y,
+    min_x, y0
+  )
+
+  if(length(coeffs) == 1){
+    out <- out %>%
+      bind_rows(
+        tribble(
+          ~x, ~y,
+          max_x, y0
+        )
+      )
+    return(out)
+
+  } else if (length(coeffs) == 2){
+    out <- out %>% 
+      bind_rows(
+        tribble(
+          ~x, ~y,
+          taus[1], y0,
+          taus[1], y0 + coeffs[2],
+          max_x, y0 + coeffs[2]
+        )
+      )
+    return(out)
+
+  } else if (length(coeffs) == 3){
+    out <- out %>% 
+      bind_rows(
+        tribble(
+          ~x, ~y,
+          taus[1], y0,
+          taus[1], y0 + coeffs[2],
+          taus[2], y0 + coeffs[2],
+          taus[2], y0 + coeffs[3],
+          max_x, y0 + coeffs[3]
+
+        )
+      )
+
+    return(out)
+  }
+  return(NULL) # should not be triggered  
+}
+
+# Let's try this with 0cp
+
+paths_0cp <- 
+  hmd_ex_selected_countries_with_synth %>% 
+    filter(code != "DEUTNP") %>%   
+    filter(year >= 1979) %>% 
+    group_by(code, x, sex) %>% 
+    nest() %>% 
+    mutate(
+      mdl_0 = map(data, estimate_null)
+    ) %>% 
+    mutate(bic = map_dbl(mdl_0, BIC)) %>% 
+    mutate(
+      tau_paths = pmap(.l = list(df = data, mdl = mdl_0), make_path_coords)
+    ) %>% 
+  select(code, x, sex, bic_0cp = bic, paths_0cp = tau_paths)
+
+# Now 1cp
+paths_1cp <- 
+  changepoint_breakpoint_models %>% 
+    unnest(mdl_outputs) %>% 
+    group_by(code, x, sex) %>% 
+    filter(bic == min(bic)) %>% 
+    ungroup() %>% 
+    mutate(
+      mdl_with_best_1cp = map2(tau, data, run_tau_1)
+    ) %>% 
+    mutate(
+      tau_paths = pmap(
+        .l = list(
+          df = data, 
+          mdl = mdl_with_best_1cp, 
+          taus = c(tau)
+        ), 
+        make_path_coords
+      )
+    ) %>% 
+    select(
+      code, x, sex, bic_1cp = bic, paths_1cp = tau_paths
+    )
+
+# and now for 2cp
+paths_2cp <- 
+  changepoint_breakpoints_models %>% 
+    unnest(mdl_outputs) %>% 
+    group_by(code, x, sex) %>% 
+    filter(bic == min(bic)) %>% 
+    ungroup() %>% 
+    mutate(mdl_with_best_2cp = pmap(.l = list(tau1=tau1, tau2=tau2, df=data), run_tau_2)) %>%
+    mutate(taus = map2(tau1, tau2, c)) %>% 
+    mutate(
+      tau_paths = pmap(
+        .l = list(
+          df = data, 
+          mdl = mdl_with_best_2cp, 
+          taus = taus
+        ), 
+        make_path_coords
+      )
+    ) %>% 
+    select(code, x, sex, bic_2cp = bic, paths_2cp = tau_paths)
+
+best_cp_paths <- 
+  reduce(list(paths_0cp, paths_1cp, paths_2cp), left_join) %>% 
+    rename(start_age = x) %>% 
+    pivot_longer(cols = starts_with(c("path", "bic")), names_to = c(".value", "path_type"), names_sep = "_") %>% 
+    group_by(code, start_age, sex) %>% 
+    mutate(
+      bic_rank = as.character(rank(bic))
+    ) %>% 
+    ungroup() %>% 
+    unnest(paths) %>% 
+    left_join(country_code_lookup) %>% 
+  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
+  filter(!is.na(country)) %>% 
+  group_by(country, sex, start_age) %>% 
+  filter(bic_rank == 1) 
+```
+
+The best changepoint model (if any), along with the points, are shown for life expectancy at birth below
+
+```{r}
+best_cp_paths %>%
+  filter(start_age == 0) %>% 
+  ggplot(aes(x, y)) + 
+  geom_path(colour = "red") + 
+  facet_grid(sex ~ country ) + 
+  expand_limits(y  = 0) + 
+  geom_hline(yintercept = 0) +
+  theme(
+      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
+      legend.position = 'none'
+  ) +
+  geom_point(
+    aes(year, delta_ex),
+    inherit.aes = FALSE,
+    data = change_in_ex_selected_countries %>% 
+      filter(x == 0) %>% 
+      filter(year >= 1979) %>% 
+      left_join(country_code_lookup) %>% 
+      mutate(
+        country = factor(
+          country, 
+          levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands")
+          )
+        ) %>% 
+        filter(!is.na(country))   %>% 
+        filter(between(year, 1980, 2020))  
+  ) + 
+  labs(
+    x = "Year", 
+    y = "Change in life expectancy since previous year",
+    title = "Best changepoint model fits: Life expectancy at birth",
+    subtitle = "Polyline indicates fit of best performing model",
+    caption = "Source: HMD"
+  )
+
+```
+
+We can see that the best-performing model is a two changepoing model with contiguous years for Scottish females, Spanish females, French females, Italian females, French males and Dutch males.
+For each of these models it is clear that the model in effect 'detected' an outlier of either exceptionally high annual change (France and Italy) or exceptionally low or negative annual change (Scottish males, Spanish females, Dutch males).
+
+For England & Wales (both males and females) and for Scottish females, a single changepoint model was selected, with lower rates of annual life expectancy gain after around 2011 than before.
+For Scottish males this fell to negative values (declining rather than stalling mortality).
+A somewhat similar pattern is observed for Germany, but this appears to be less consistent in term and less severe than for England & Wales.
+
+The equivalent graph for life expectancy from age 65 is shown below:
+
+```{r}
+best_cp_paths %>%
+  filter(start_age == 65) %>% 
+  ggplot(aes(x, y)) + 
+  geom_path(colour = "red") + 
+  facet_grid(sex ~ country ) + 
+  expand_limits(y  = 0) + 
+  geom_hline(yintercept = 0) +
+  theme(
+      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
+      legend.position = 'none'
+  ) +
+  geom_point(
+    aes(year, delta_ex),
+    inherit.aes = FALSE,
+    data = change_in_ex_selected_countries %>% 
+      filter(x == 65) %>% 
+      filter(year >= 1979) %>% 
+      left_join(country_code_lookup) %>% 
+      mutate(
+        country = factor(
+          country, 
+          levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands")
+          )
+        ) %>% 
+        filter(!is.na(country))   %>% 
+        filter(between(year, 1980, 2020))  
+  ) + 
+  labs(
+    x = "Year", 
+    y = "Change in life expectancy since previous year",
+    title = "Best changepoint model fits: Life expectancy at age 65 years",
+    subtitle = "Polyline indicates fit of best performing model",
+    caption = "Source: HMD"
+  )
+
+```
+For changes in conditional life expectancy from age 65 two changepoint models with contiguous years were identified for females in Germany, France and Italy; and for males in Scotland, France, Italy and the Netherlands.
+
+# Summary and conclusion
+
+This appendix has show the implementation and results of applying an alternative method to identifying distinct changes in trends in life expectancy for the selected countries.
+In many cases, a two changepoint model was identified.
+However, often these two changepoint models involved contiguous years, and appear to in effect be operating as outlier detection algorithms, picking out exceptional years from the series.
+
+Although the fact these individual years are outliers is important to note, as it illustrates how varied life expectancy change series can be when single years are used, these single years cannot, by definition, define or constitute a change in trend, which was the purpose of this exercise.
+It is for this reason that the segmented approach was used instead of this approach in the main paper.
+
+The changepoint approach could be adapted to be less 'sensitive' to outlier years by narrowing the search space for $tau_2$ conditional on $tau_1$ so as to be two or more years later.
+(e.g.
+if $tau_1$ were 1990 then candidate values for $tau_2$ would start from 1992 instead of 1991.) However, the choice of how much to narrow down the search space in this way is somewhat arbitrary, so was not adapted further for this paper.

### 2022-07-25 appendices/life_expectancy_trends.Rmd
minor edits

@@
 ```
+
 Equivalently
@@
 ```
+
 Equivalently:

### 2022-07-26 appendices/changepoint_methods.Rmd
minor edits
finally create pdf version of changepoint methods rmd

@@
 date: '2022-07-22'
-output: html_document
+output:
+  pdf_document: default
+  word_document: default
+  html_document: default
 ---
@@
 Breaks in the data are estimated by fitting the following model specifications:
+
 * **$M_0$**: $\frac{de_x}{dt} \sim \alpha$
@@
 T = \left\{
-    \begin{array}\\
-        1 & \mbox{if } \ t \ge \tau \\
+    \begin{array}{cc}
+        1 & \mbox{if   } \  \ t \ge \tau \\
         0 & \mbox{otherwise}
@@
 $$
-    * **M_2**: $\frac{de_x}{dt} \sim \alpha + \beta_1 T_1 + \beta_2 T_2$, where
+
+* **$M_2$**: $\frac{de_x}{dt} \sim \alpha + \beta_1 T_1 + \beta_2 T_2$, where
 
@@
 T_1 = \left\{
-    \begin{array}\\
-        1 & \mbox{if } \tau_1 \le t \lt \tau_2 \\
+    \begin{array}{cc}
+        1 & \mbox{if } \tau_1 \le t < \tau_2 \\
         0 & \mbox{otherwise}
@@
 T_2 = \left\{
-    \begin{array}\\
+    \begin{array}{cc}
         1 & \mbox{if } t \ge \tau_2 \\
@@
 $$
+
 i.e.
-the selection of $\tau_1$ and $\tau_2$ partitions the model into three sections: $\frac{de_x}{dt} \sim \alpha$ where $t \lt \tau_1$, $\frac{de_x}{dt} \sim \alpha + \beta_1$ where $\tau_1 \le t \lt \tau_2$, and $\frac{de_x}{dt} \sim \alpha + \beta_2$ where $t \ge \tau_2$.
-As $\tau_1 \lt \tau_2$, only values of $\tau_2$ which are greater than $\tau_1$ are considered.
+the selection of $\tau_1$ and $\tau_2$ partitions the model into three sections: $\frac{de_x}{dt} \sim \alpha$ where $t < \tau_1$, $\frac{de_x}{dt} \sim \alpha + \beta_1$ where $\tau_1 \le t < \tau_2$, and $\frac{de_x}{dt} \sim \alpha + \beta_2$ where $t \ge \tau_2$.
+As $\tau_1 < \tau_2$, only values of $\tau_2$ which are greater than $\tau_1$ are considered.
 Otherwise all whole integer values of $\tau$ are seached through for both $M_1$ and $M_2$.
@@
 ```
+
 # Functions
@@
 
-```{r}
+```{r cache=TRUE}
 changepoint_breakpoint_models %>% 
@@
 
-```{r}
+```{r cache=TRUE}
 changepoint_breakpoints_models <- 
@@
 ```
+
 For France, for all four subpopulations, a pairing of `{2004, 2005}` was identified. If we look at the series for France, we can see why this pair of years may have been identified
@@
 ```
+
 For changes in conditional life expectancy from age 65 two changepoint models with contiguous years were identified for females in Germany, France and Italy; and for males in Scotland, France, Italy and the Netherlands.
@@
 
-The changepoint approach could be adapted to be less 'sensitive' to outlier years by narrowing the search space for $tau_2$ conditional on $tau_1$ so as to be two or more years later.
+The changepoint approach could be adapted to be less 'sensitive' to outlier years by narrowing the search space for $\tau_2$ conditional on $\tau_1$ so as to be two or more years later.
 (e.g.
-if $tau_1$ were 1990 then candidate values for $tau_2$ would start from 1992 instead of 1991.) However, the choice of how much to narrow down the search space in this way is somewhat arbitrary, so was not adapted further for this paper.
+if $\tau_1$ were 1990 then candidate values for $\tau_2$ would start from 1992 instead of 1991.) However, the choice of how much to narrow down the search space in this way is somewhat arbitrary, so was not adapted further for this paper.

### 2022-07-26 appendices/test.Rmd
finally create pdf version of changepoint methods rmd

--- 2022-07-22 appendices/changepoint_methods.Rmd
@@
 ---
-title: "Changepoint Approach"
+title: "test"
 author: "Jon Minton"
-date: '2022-07-22'
-output: html_document
+date: '2022-07-26'
+output:
+  pdf_document: default
+  html_document: default
 ---
@@
 
+$$\begin{array}{ccc}
+x_{11} & x_{12} & x_{13}\\
+x_{21} & x_{22} & x_{23}
+\end{array}$$
+
 $$
 T = \left\{
-    \begin{array}\\
-        1 & \mbox{if } \ t \ge \tau \\
-        0 & \mbox{otherwise}
-    \end{array}
-\right.
-$$
-    * **M_2**: $\frac{de_x}{dt} \sim \alpha + \beta_1 T_1 + \beta_2 T_2$, where
-
-$$
-T_1 = \left\{
-    \begin{array}\\
-        1 & \mbox{if } \tau_1 \le t \lt \tau_2 \\
+    \begin{array}{cc}
+        1 & \mbox{if   } \  \ t \ge \tau \\
         0 & \mbox{otherwise}
@@
 
-and
+* **$M_2$**: $\frac{de_x}{dt} \sim \alpha + \beta_1 T_1 + \beta_2 T_2$, where
 
 $$
-T_2 = \left\{
-    \begin{array}\\
-        1 & \mbox{if } t \ge \tau_2 \\
+T_1 = \left\{
+    \begin{array}{cc}
+        1 & \mbox{if }  \tau_1  \le t  < \tau_{2}   \\
         0 & \mbox{otherwise}
@@
 $$
-i.e.
+
+<!-- and -->
+
+<!-- $$ -->
+<!-- T_2 = \left\{ -->
+<!--     \begin{array} -->
+<!--         1 & \mbox{if } t \ge \tau_2 \\ -->
+<!--         0 & \mbox{otherwise} -->
+<!--     \end{array} -->
+<!-- \right. -->
+<!-- $$ -->
+
+<!-- i.e.
 the selection of $\tau_1$ and $\tau_2$ partitions the model into three sections: $\frac{de_x}{dt} \sim \alpha$ where $t \lt \tau_1$, $\frac{de_x}{dt} \sim \alpha + \beta_1$ where $\tau_1 \le t \lt \tau_2$, and $\frac{de_x}{dt} \sim \alpha + \beta_2$ where $t \ge \tau_2$.
@@
 Otherwise all whole integer values of $\tau$ are seached through for both $M_1$ and $M_2$.
+-->
 
-The above analyses are performed for each population group, comprising different combinations of country, sex, and starting age.
-
-# Extracting relevant data
-
-```{r}
-# load packages
-pacman::p_load(here, tidyverse)
-
-# load data 
-hmd_lt <- read_rds(here("data", "lifetables.rds"))
-
-# Labels for codes 
-country_code_lookup <- 
-  tribble(
-    ~code, ~country,
-    "DEUTNP", "Germany",
-    "DEUTE", "East Germany",
-    "DEUTW", "West Germany",
-    "ESP", "Spain",
-    "FRATNP", "France", 
-    "ITA", "Italy",
-    "GBRTENW", "England & Wales",
-    "GBR_SCO", "Scotland",
-    "DEUTSYNTH", "Synthetic Germany",
-    "NLD", "Netherlands"
-  )
-
-countries_of_interest <- c(
-  "GBRTENW",
-  "GBR_SCO",
-  "GBR_UK",
-  "FRATNP",
-  "ESP",
-  "ITA",
-  "DEUTNP",
-  "DEUTE", 
-  "DEUTW",
-  "NLD"
-)
-
-source(here("R", "make_synthetic_germany_functions.R"))
-source(here("R", "make_pop_selection.R"))
-
-change_in_ex_selected_countries <- 
-  hmd_ex_selected_countries_with_synth %>% 
-    group_by(code, x, sex) %>% 
-    arrange(year) %>% 
-    mutate(delta_ex = ex - lag(ex)) %>% 
-    ungroup()
-
-```
-
-Visualise
-
-```{r}
-change_in_ex_selected_countries %>% 
-  filter(x == 0) %>% 
-  left_join(country_code_lookup) %>% 
-  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
-  filter(!is.na(country)) %>% 
-  filter(between(year, 1980, 2020)) %>% 
-  ggplot(aes(x = year, y = delta_ex)) + 
-  geom_point() + 
-  facet_grid(sex~country) + 
-  geom_hline(yintercept = 0) + 
-  scale_y_continuous(breaks = seq(-30, 50, by = 10)) +
-  theme(
-        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
-  ) + 
-  labs(
-    x = "Year",
-    y = "Change in life expectancy from previous year",
-    title = "Annual change in life expectancy at birth, selected countries",
-    subtitle = "Line: nonlinear smoother over the points",
-    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
-  )
-
-change_in_ex_selected_countries %>% 
-  filter(x == 65) %>% 
-  left_join(country_code_lookup) %>% 
-  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
-  filter(!is.na(country)) %>% 
-  filter(between(year, 1980, 2020)) %>% 
-  ggplot(aes(x = year, y = delta_ex)) + 
-  geom_point() + 
-  facet_grid(sex~country) + 
-  geom_hline(yintercept = 0) + 
-  scale_y_continuous(breaks = seq(-30, 50, by = 10)) +
-  theme(
-        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
-  ) + 
-  labs(
-    x = "Year",
-    y = "Change in life expectancy from previous year",
-    title = "Annual change in life expectancy at age 65, selected countries",
-    subtitle = "Line: nonlinear smoother over the points",
-    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
-  )
-
-```
-
-The following are functions used to try to identify the best possible breakpoint/changepoint years
-
-```{r}
-
-# Breakpoint functions
-
-run_alt_with_given_tau <- function(tau, df){
-  df <- 
-    df %>% 
-    filter(!is.na(delta_ex)) %>% 
-    mutate(
-      T_param = ifelse(year < tau, FALSE, TRUE)
-    )
-
-  lm(delta_ex ~ T_param, data  = df)
-}
-
-get_best_tau_from_optim <- function(df, buffer = 2, what = c("tau", "optim")){
-  # We can put in some data validation checks too
-
-  what <- match.arg(what)
-
-  years <- df$year 
-  tau_lower <- min(years) + buffer # the lower bound 
-  tau_upper <- max(years) - buffer # the upper bound
-  tau_start <- (tau_upper + tau_lower) / 2 # start just in the middle 
-  # (This shouldn't matter if the algorithm is robust)
-
-  get_result_from_model <- function(par, df, what = c("AIC", "BIC", "model")){
-    # Again, can carry out more validation checks 
-    what <- match.arg(what)
-
-    tau <- par[["tau"]] # This is how parameters are packed up by optim
-    model <-
-      df %>% 
-      arrange(year) %>% 
-      mutate(delta_ex = ex - lag(ex)) %>% 
-      filter(!is.na(delta_ex)) %>% 
-      mutate(T_param = ifelse(year < tau, FALSE, TRUE)) %>% 
-      lm(delta_ex ~ T_param, data = .)
-    if (what == "model") {return(model)      }
-    if (what == "BIC")   {return(BIC(model)) }
-    if (what == "AIC")   {return(AIC(model)) }
-
-    # NULL should never be returned! If it has something's gone wrong!
-    NULL
-  }
-
-  optim_obj <- 
-    optim(
-      par = list(tau = tau_start), 
-      fn = get_result_from_model, 
-      method = "L-BFGS-B",
-      lower = tau_lower,
-      upper = tau_upper,
-      df = df
-    )
-  # Need to make the result dependent on the what argument
-
-  if (what == "tau") {return(optim_obj[["par"]])}
-  else if (what == "optim"){ return(optim_obj) }
-
-  # Again, the following should never be triggered
-  NULL
-}
-
-get_best_tau_from_gridsearch <- function(df, buffer = 2, what = c("best", "all")){
-  # We can put in some data validation checks too
-
-  what <- match.arg(what)
-
-  years <- df$year 
-  tau_lower <- min(years) + buffer # the lower bound 
-  tau_upper <- max(years) - buffer # the upper bound
-  tau_searchrange <- tau_lower:tau_upper
-  get_result_from_model <- function(tau, df, what = c("BIC", "AIC", "model")){ #I've changed the order to default to BIC (as more comparable with segmented)
-    # Again, can carry out more validation checks 
-    what <- match.arg(what)
-
-    model <-
-      df %>% 
-      arrange(year) %>% 
-      mutate(delta_ex = ex - lag(ex)) %>% 
-      filter(!is.na(delta_ex)) %>% 
-      mutate(T_param = ifelse(year < tau, FALSE, TRUE)) %>% 
-      lm(delta_ex ~ T_param, data = .)
-    if (what == "model") {return(model)      }
-    if (what == "BIC")   {return(BIC(model)) }
-    if (what == "AIC")   {return(AIC(model)) }
-
-    # NULL should never be returned! If it has something's gone wrong!
-    NULL
-  }
-
-  search_df <- tibble(
-    tau = tau_searchrange
-  ) %>% 
-    mutate(bic = map_dbl(tau, get_result_from_model, df = df))
-
-  if (what == "all") {return(search_df)}
-
-  if (what == "best") {
-    out <- 
-      search_df %>% 
-      filter(bic == min(bic)) %>% 
-      select(tau, bic)
-    return(out) # I've changed this to compare both tau and bic more easily with 2 cp models
-  }
-  # Again, the following should never be triggered
-  NULL
-}
-
-# This is a modification of get_best_tau_from_gridsearch, for two taus
-get_best_taus_from_gridsearch <- function(df, buffer = 2, what = c("best", "all")){
-  # We can put in some data validation checks too
-
-  what <- match.arg(what)
-
-  years <- df$year 
-  tau_lower <- min(years) + buffer # the lower bound 
-  tau_upper <- max(years) - buffer # the upper bound
-
-  # Now need to list the permutations of taus to consider
-
-  tau_grid <- expand_grid(
-    tau1 = tau_lower:tau_upper,
-    tau2 = tau_lower:tau_upper
-  ) %>% 
-    filter(tau2 > tau1) # tau2 should be greater than tau1
-
-  get_result_from_model <- function(tau1, tau2, df, what = c("BIC", "AIC", "model")){ #I've changed the order to default to BIC (as more comparable with segmented)
-    # Again, can carry out more validation checks 
-    what <- match.arg(what)
-
-    model <-
-      df %>% 
-      arrange(year) %>% 
-      mutate(delta_ex = ex - lag(ex)) %>% 
-      filter(!is.na(delta_ex)) %>% 
-      mutate(T_param = ifelse(
-        year < tau1, 
-        "bp0", 
-        ifelse(
-          year < tau2, "bp1", "bp2"
-          )
-        )
-      ) %>% 
-      lm(delta_ex ~ T_param, data = .)
-    if (what == "model") {return(model)      }
-    if (what == "BIC")   {return(BIC(model)) }
-    if (what == "AIC")   {return(AIC(model)) }
-
-    # NULL should never be returned! If it has something's gone wrong!
-    NULL
-  }
-
-  search_df <- tau_grid %>% 
-    mutate(bic = map2_dbl(tau1, tau2, get_result_from_model, df = df))
-
-  if (what == "all") {return(search_df)}
-
-  if (what == "best") {
-    out <- 
-      search_df %>% 
-      filter(bic == min(bic)) %>% 
-      select(bic, tau1, tau2)
-    return(out)
-  }
-  # Again, the following should never be triggered
-  NULL
-}
-```
-
-(The optim-based function takes far too long to run, so is not used)
-
-The following calculates the best single breakpoint model for each population
-
-```{r}
-changepoint_breakpoint_models <- 
-  hmd_ex_selected_countries_with_synth %>% 
-  filter(code != "DEUTNP") %>%   
-  filter(year >= 1979) %>% 
-  group_by(code, x, sex) %>% 
-  nest() %>% 
-  mutate(
-    mdl_outputs = map(data, get_best_tau_from_gridsearch, what = "all")
-  )
-```
-The relationship between proposed `tau` (i.e. changepoint year) and BIC is shown for England & Wales below
-
-```{r}
-changepoint_breakpoint_models %>% 
-  filter(code == "GBRTENW") %>% 
-  unnest(mdl_outputs) %>% 
-  group_by(x, sex) %>% 
-  mutate(min_bic = bic == min(bic)) %>% 
-  ungroup() %>% 
-  ggplot(aes(tau, bic, colour = min_bic, shape = min_bic)) + 
-  geom_point() + 
-  facet_grid(x ~ sex) +
-  scale_colour_manual(values = c(`TRUE` = 'red', `FALSE` = "black")) + 
-  scale_shape_manual(values = c(`TRUE` = 'triangle', `FALSE` = 'circle'))
-
-```
+<!-- The above analyses are performed for each population group, comprising different combinations of country, sex, and starting age.  -->

### 2022-07-26 appendices/time_series_models.Rmd
finally create pdf version of changepoint methods rmd

--- 2022-07-22 appendices.Rmd
@@
 ---
-title: "Appendices"
+title: "Time Series Models"
 author: "Jon Minton"
@@
 
-## Aim
-
-The aim of this document is to produce a series of appendices to support the main paper.
-The appendices are listed in a slightly arbitrary order and can be amended as required.
-All material is written as an R markdown document and the code required to produce analyses are presented alongside the results of such analyses.
-
-# Appendix 1: Are annual ex time series random walks with drift (RWD) or do they have autocorrelation?
+# Are annual ex time series random walks with drift (RWD) or do they have autocorrelation?
 
@@
 ```
+
 Our series contains ex for x = 0 (life expectancy at birth) and x=65 (life expectancy at age 65 years), for each of the countries of interest, including 'Synthetic Germany'.
@@
 ```
+
 (n.b. the data in this series are shown in years per year, rather than weeks per year as in the main figure)
@@
 ```
+
 The Autocorrelated specification (ARIMA(1,1,0)) is preferred to RWD (ARIMA(0,1,0)) for all populations.
@@
 ```
+
 The majority of these coefficients are positive, indicating 'stickiness' in the values in the series, rather than oscillation.
@@
 ```
+
 The length of the vectors ar_roots and ma_roots indicate, respectively, how many ar or ma terms were identified in the best fitting model for the population indicated by sex, x (starting age) and country.

### 2022-07-26 appendices/life_expectancy_trends.Rmd
add pdf version of trends graphs
add average change to life expectancy trends appendix and render pdf version

@@
 date: '2022-07-23'
-output: html_document
+output:
+  pdf_document: default
+  html_document: default
 ---
@@
 ```
+# Average life expectancy change
+
+The following shows the average change in life expectancy for each population over the full period, and including East Germany and West Germany as separate populations
+
+```{r}
+hmd_ex_selected_countries_with_synth %>% 
+    filter(year >= 1979) %>% 
+    group_by(code, x, sex) %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+    nest() %>% 
+    mutate(
+      rwd_model = map(data, ~lm(delta_ex ~ 1, data = . )),
+      ar_model  = map(data, ~lm(delta_ex ~ lag(delta_ex), data = .))
+    ) %>% 
+    mutate(
+      aic_rwd = map_dbl(rwd_model, AIC),
+      aic_ar  = map_dbl(ar_model, AIC)
+    ) %>% 
+    mutate(
+      diff_aic = aic_ar - aic_rwd
+    )  %>% 
+  # filter(diff_aic > 0)
+  mutate(
+    tidied_ar_model = map(ar_model, broom::tidy)
+  ) %>% 
+  unnest(tidied_ar_model) %>% 
+  filter(term == "(Intercept)") %>% 
+  arrange(desc(estimate)) %>%
+  left_join(country_code_lookup) %>% 
+  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
+  select(code, country, x, sex, estimate, std.error) %>% 
+  mutate(display_label = glue::glue("{country}_{x}_{sex}")) %>% 
+  mutate(estimate_weeks = estimate * 52.25) %>% 
+  ggplot(aes(estimate_weeks, fct_reorder(display_label, estimate_weeks))) + 
+  geom_point(aes(colour = x), show.legend = FALSE) +
+  expand_limits( x = 0) + 
+  geom_vline(xintercept = 0) + 
+  labs(
+    x = "Average annual life expectancy change",
+    y = "Country-sex combination",
+    title = "Average annual improvements by country and sex",
+    subtitle = "Range: 1980 to 2020 or nearest available years",
+    caption = "Source: Human Mortality Database"
+  ) + 
+  facet_wrap(~x)
+
+```
+The figure below shows the same for those populations considered in the main paper.
+
+```{r}
+hmd_ex_selected_countries_with_synth %>% 
+    filter(year >= 1979) %>% 
+    filter(!(code %in% c("DEUTE", "DEUTW", "DEUTNP"))) %>%  # Using only synthetic germany for longer time period
+    group_by(code, x, sex) %>% 
+    arrange(year) %>% 
+    mutate(delta_ex = ex - lag(ex)) %>% 
+    nest() %>% 
+    mutate(
+      rwd_model = map(data, ~lm(delta_ex ~ 1, data = . )),
+      ar_model  = map(data, ~lm(delta_ex ~ lag(delta_ex), data = .))
+    ) %>% 
+    mutate(
+      aic_rwd = map_dbl(rwd_model, AIC),
+      aic_ar  = map_dbl(ar_model, AIC)
+    ) %>% 
+    mutate(
+      diff_aic = aic_ar - aic_rwd
+    )  %>% 
+  # filter(diff_aic > 0)
+  mutate(
+    tidied_ar_model = map(ar_model, broom::tidy)
+  ) %>% 
+  unnest(tidied_ar_model) %>% 
+  filter(term == "(Intercept)") %>% 
+  arrange(desc(estimate)) %>%
+  left_join(country_code_lookup) %>% 
+  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>%
+  filter(!is.na(country)) %>% 
+  select(code, country, x, sex, estimate, std.error) %>% 
+  mutate(display_label = glue::glue("{country}_{x}_{sex}")) %>% 
+  mutate(estimate_weeks = estimate * 52.25) %>% 
+  ggplot(aes(estimate_weeks, fct_reorder(display_label, estimate_weeks))) + 
+  geom_point(aes(colour = x), show.legend = FALSE) +
+  expand_limits( x = 0) + 
+  geom_vline(xintercept = 0) + 
+  labs(
+    x = "Average annual life expectancy change (weeks per year)",
+    y = "Country-sex combination",
+    title = "Average annual improvements by country and sex",
+    subtitle = "Range: 1980 to 2020 or nearest available years",
+    caption = "Source: Human Mortality Database"
+  ) + 
+  facet_wrap(~x)
+
+```

### 2022-07-26 appendices/time_series_models.Rmd
add pdf version of timeseries appendix
add html version of time series appendix
update time series models to include simple linear regression specs, and results from them

@@
 date: '2022-07-21'
-output: html_document
+output:
+  pdf_document: default
+  html_document: default
 ---
@@
 
+## Alternative approach to establishing if series are autocorrelated
+
+An alternative way to determing whether the series are autocorrelated is to compare two linear regression equations: one in which change in life expectancy is regressed against an intercept; another in which the change in life expectancy is additionally regressed against the change in the previous year
+
+```{r}
+
+alt_approach_autocorrelated <- 
+  hmd_ex_selected_countries_with_synth %>% 
+      filter(year >= 1979) %>% 
+      filter(!(code %in% c("DEUTE", "DEUTW", "DEUTNP"))) %>%  # Using only synthetic germany for longer time period
+      group_by(code, x, sex) %>% 
+      arrange(year) %>% 
+      mutate(delta_ex = ex - lag(ex)) %>% 
+      nest() %>% 
+      mutate(
+        rwd_model = map(data, ~lm(delta_ex ~ 1, data = . )),
+        ar_model  = map(data, ~lm(delta_ex ~ lag(delta_ex), data = .))
+      ) %>% 
+      mutate(
+        bic_rwd = map_dbl(rwd_model, BIC),
+        bic_ar  = map_dbl(ar_model, BIC)
+      ) %>% 
+      mutate(
+        diff_bic = bic_ar - bic_rwd
+      ) %>% 
+    mutate(
+      which_preferred = ifelse(diff_bic < 0, "Autocorrelated", "RWD")
+    ) %>% 
+      left_join(country_code_lookup) %>% 
+      mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
+    filter(!is.na(country))
+
+alt_approach_autocorrelated %>% 
+  select(country, x, sex, bic_rwd, bic_ar, which_preferred)
+```
+This approach also identifies most, but not all, of the populations to exhibit some degree of autocorrelation.
+(n.b.
+the more stringent BIC was used instead of AICc in this example, which may explain discrepancies between approaches)
+
+For each of the above, the intercept (average improvement per year) and autocorrelation term can be extracted as follows:
+
+```{r}
+alt_approach_autocorrelated  %>% 
+  left_join(country_code_lookup) %>% 
+  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
+  filter(!is.na(country)) %>% 
+  select(country, x, sex, ar_model) %>% 
+  mutate(tidied_model = map(ar_model, broom::tidy)) %>% 
+  select(-ar_model) %>% 
+  filter(x == 0) %>% 
+  unnest(tidied_model) %>% 
+  ggplot(aes(estimate, reorder(country, estimate))) + 
+  geom_point() + 
+  facet_grid(term~sex) + 
+  geom_vline(xintercept = 0) + 
+  geom_errorbarh(aes(xmin = estimate - 1.96  * std.error, xmax = estimate + 1.96 * std.error))  + 
+  labs(
+    x = "Coefficient value", 
+    y = "Country",
+    title = "Coefficients from autoregressive models\nlife expectancy at birth, selected countries",
+    subtitle = "Horizontal bars indicate 95% confidence intervals",
+    caption = "Source: HMD lifetables"
+
+  )
+
+```
+Here all autoregression coefficients are negative, most of which are statistically significant at p < 0.05.
+
+The following shows the equivalent trends for x = 65
+
+```{r}
+alt_approach_autocorrelated  %>% 
+  left_join(country_code_lookup) %>% 
+  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
+  filter(!is.na(country)) %>% 
+  select(country, x, sex, ar_model) %>% 
+  mutate(tidied_model = map(ar_model, broom::tidy)) %>% 
+  select(-ar_model) %>% 
+  filter(x == 65) %>% 
+  unnest(tidied_model) %>% 
+  ggplot(aes(estimate, reorder(country, estimate))) + 
+  geom_point() + 
+  facet_grid(term~sex) + 
+  geom_vline(xintercept = 0) + 
+  geom_errorbarh(aes(xmin = estimate - 1.96  * std.error, xmax = estimate + 1.96 * std.error))  + 
+  labs(
+    x = "Coefficient value", 
+    y = "Country",
+    title = "Coefficients from autoregressive models\nlife expectancy at age 65, selected countries",
+    subtitle = "Horizontal bars indicate 95% confidence intervals",
+    caption = "Source: HMD lifetables"
+
+  )
+```
+Here the negative correlations appear even stronger, and of larger magnitude in Italy than elsewhere.
+It also appears that the two coefficients are negatively associated, with Netherlands having the lowest average annual improvement, but also the smallest (and non-significant) magnitude of negative autocorrelation; Italy appears the converse.
+
 ## Comparing a wider range of ARIMA models

### 2022-10-01 appendices/changepoint_methods.Rmd
fix typo (changepoing)

@@
 
-We can see that the best-performing model is a two changepoing model with contiguous years for Scottish females, Spanish females, French females, Italian females, French males and Dutch males.
+We can see that the best-performing model is a two changepoint model with contiguous years for Scottish females, Spanish females, French females, Italian females, French males and Dutch males.
 For each of these models it is clear that the model in effect 'detected' an outlier of either exceptionally high annual change (France and Italy) or exceptionally low or negative annual change (Scottish males, Spanish females, Dutch males).

### 2022-10-02 appendices/are_the_series_purely_random.Rmd
correct model specification for arima(0,1,0). Update summary of results accordingly

--- 2022-07-26 appendices/time_series_models.Rmd
@@
 ---
-title: "Time Series Models"
+title: "Appendices"
 author: "Jon Minton"
@@
 
-# Are annual ex time series random walks with drift (RWD) or do they have autocorrelation?
+# Appendix 1: Are annual ex time series random walks with drift (RWD) or do they have autocorrelation?
 
-Section X of the main manuscript noted, from figure 1a and 1b, that there is substantial annual variation in annual changes in life expectancy.
+Our paper noted that there is substantial annual variation in annual changes in life expectancy.
 After taking the first difference in these series (i.e.
@@
     mutate(
-      arima_010 = map(data, fit_arima_model, order = c(1, 1, 0)),
+      arima_010 = map(data, fit_arima_model, order = c(0, 1, 0)), # corrected from c(1, 1, 0)
       arima_110 = map(data, fit_arima_model, order = c(1, 1, 0))
@@
 
-The Autocorrelated specification (ARIMA(1,1,0)) is preferred to RWD (ARIMA(0,1,0)) for all populations.
+For females, the Random Walk with Drift (RWD, ARIMA(0, 1, 0)) specification is preferred for all populations considered except Scottish and German females from age 65.
+
+For males, the Autocorrelated specification (ARIMA(1, 1, 0)) is preferred for all populations for changes in life expectancy from birth (x = 0), whereas from age 65 the RWD specification is preferred for Spain, France, Italy,  and Scotland.
 
@@
 The length of the vectors ar_roots and ma_roots indicate, respectively, how many ar or ma terms were identified in the best fitting model for the population indicated by sex, x (starting age) and country.
-For example, for females in England & Wales, from age 0, an ARIMA(1,1,0) model is preferred, whereas for females in Scotland, from age 0, an ARIMA(0, 1, 1) model is preferred.
+For example, for females in England & Wales, from age 0, an ARIMA(1,1,0) model is preferred, whereas for females in Scotland, from age 0, an ARIMA(0,1,1) model is preferred.
 

### 2022-10-03 appendices/time_series_models.Rmd
correct incorrect arima spec in second document

@@
     mutate(
-      arima_010 = map(data, fit_arima_model, order = c(1, 1, 0)),
+      arima_010 = map(data, fit_arima_model, order = c(0, 1, 0)), # correct from order = c(1, 1, 0)
       arima_110 = map(data, fit_arima_model, order = c(1, 1, 0))
@@
 
-The Autocorrelated specification (ARIMA(1,1,0)) is preferred to RWD (ARIMA(0,1,0)) for all populations.
+For females, the Random Walk with Drift (RWD, ARIMA(0, 1, 0)) specification is preferred for all populations considered except Scottish and German females from age 65.
 
+For males, the Autocorrelated specification (ARIMA(1, 1, 0)) is preferred for all populations for changes in life expectancy from birth (x = 0), whereas from age 65 the RWD specification is preferred for Spain, France, Italy,  and Scotland.
 The following extracts the autocorrelation coefficients and visualises them.

### 2022-10-03 appendices/reviewer_one_responses.Rmd
add appendix for reviewer 1

---
title: "Response to Reviewer 1 Comments"
author: "Jon Minton"
date: "2022-10-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This document presents additional analyses performed in response to comments from Referee 1.

## The queries

- 1) It could prove relevant to verify whether the residuals of the proposed models are white noise. Have you made a model residuals analysis? 
- 2) In addition to the BIC and AIC information criteria, could you compare the proposed models using also AICc
- 3) In view of the results obtained, would it be possible to use an intervention variable to fit the ARIMA model?

# Query 1: Residuals analysis

A residuals analysis was not conducted, but will be within this appendix. As mentioned elsewhere, we have an understanding that, for many populations, the series are better represented by models where each value in the series is autocorrelated with the previous value. However for simplicity no such term is included in the models being compared. This does mean we may expect residuals which are not entirely white noise, however.

To look at the residuals, we first load and process the dataset:

```{r}
# load packages

library(tidyverse)

# load data 
hmd_lt <- read_rds("https://github.com/JonMinton/change-in-ex/blob/main/data/lifetables.rds?raw=true")

# Labels for codes 
country_code_lookup <- 
  tribble(
    ~code, ~country,
    "DEUTNP", "Germany",
    "DEUTE", "East Germany",
    "DEUTW", "West Germany",
    "ESP", "Spain",
    "FRATNP", "France", 
    "ITA", "Italy",
    "GBRTENW", "England & Wales",
    "GBR_SCO", "Scotland",
    "DEUTSYNTH", "Synthetic Germany",
    "NLD", "Netherlands"
  )

countries_of_interest <- c(
  "GBRTENW",
  "GBR_SCO",
  "GBR_UK",
  "FRATNP",
  "ESP",
  "ITA",
  "DEUTNP",
  "DEUTE", 
  "DEUTW",
  "NLD"
)

source("https://raw.githubusercontent.com/JonMinton/change-in-ex/main/R/make_synthetic_germany_functions.R")

source("https://raw.githubusercontent.com/JonMinton/change-in-ex/main/R/make_pop_selection.R")

change_in_ex_selected_countries <- 
  hmd_ex_selected_countries_with_synth %>% 
    group_by(code, x, sex) %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
    ungroup()

```

Then we create and run some convienience functions...

```{r}

estimate_breakpoints_and_pval <- function(df){
  null_mdl <- lm(ex ~ year, data = df) 
  seg_mdl <- segmented::segmented(null_mdl, seg.Z= ~year, psi = 2010)
  seg2_mdl <- segmented::segmented(null_mdl, seg.z= ~year, psi= c(1985, 2010)) # added to test

  list(
    null = null_mdl, 
    seg = seg_mdl,
    seg2 = seg2_mdl
  )
}

segmented_breakpoints_models <- 
  hmd_ex_selected_countries_with_synth %>% 
  filter(code != "DEUTNP") %>%   
  filter(year >= 1979) %>% 
  group_by(code, x, sex) %>% 
  nest() %>% 
  mutate(
    mdl_outputs = map(data, estimate_breakpoints_and_pval)
  ) %>% 
  unnest_longer(mdl_outputs)

# Now let's get the BIC for each model

make_predictions <- function(mdl, dta = tibble(year = 1980:2020)){
  tibble(
    year = dta$year, 
    ex_pred = predict(mdl, newdata = dta)
  )
}

best_model_predictions_descriptions <- 
segmented_breakpoints_models %>% 
  mutate(bic = map_dbl(mdl_outputs, BIC)) %>% 
  group_by(code, x, sex) %>% 
  mutate(rank_bic = rank(bic)) %>%
  filter(rank_bic == 1) %>% 
  mutate(
    best_model = case_when(
      mdl_outputs_id == 'seg2' ~ "Two breakpoints",
      mdl_outputs_id == 'seg'  ~ "One breakpoint", 
      mdl_outputs_id == 'null' ~ "No breakpoints"
    )
  ) %>% 
  mutate(
    first_breakpoint =    map2_dbl(
      mdl_outputs_id, mdl_outputs, 
      function(x, y){
        if (x == 'null'  ) {NA_real_} else {
          y[["psi"]][1,2]
        }
      }                        
      ),
      first_breakpoint_se = map2_dbl(
        mdl_outputs_id, mdl_outputs,
        function(x, y){
        if (x == 'null'  ) {NA_real_} else {
          y[["psi"]][1,3]
        }

        }
      ),
    second_breakpoint = map2_dbl(
      mdl_outputs_id, mdl_outputs,
      function(x, y){
        if (x == 'seg2') {
          y[["psi"]][2,2]
        } else  {
            NA_real_
          }
      }                           
      ),
    second_breakpoint_se = map2_dbl(
      mdl_outputs_id, mdl_outputs,
      function(x, y){
        if (x == 'seg2') {
          y[["psi"]][2,3]
        } else  {
            NA_real_
          }
      }
    )
  ) %>%  # let's add predictions too 
  mutate(
    pred_data = map(mdl_outputs, make_predictions),
    joined_data = map2(data, pred_data, left_join)
  ) %>% 
  select(code, x, sex, joined_data, first_breakpoint:second_breakpoint_se) %>% 
  unnest(joined_data)

```

We can produce predicted values as follows...

```{r}
best_breakpoint_model_predictions <- 
  segmented_breakpoints_models %>% 
    mutate(bic = map_dbl(mdl_outputs, BIC)) %>% 
    group_by(code, x, sex) %>% 
    filter(bic == min(bic)) %>% 
    mutate(predictions = map2(mdl_outputs, data, predict)) %>% 
    mutate(data_augmented = map2(data, predictions, ~ .x %>% mutate(prediction = .y))) %>% 
    select(code, x, sex, data_augmented) %>% 
    ungroup() %>% 
    unnest(data_augmented)

```

The following shows the predicted values (as lines) and observed values (as points) for x = 0:

```{r}
best_breakpoint_model_predictions %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(!is.na(country)) %>% 
  filter(x == 0) %>% 
  ggplot(aes(x = year)) + 
  facet_grid(sex ~ country) + 
  geom_point(aes(y = ex), alpha = 0.35) + 
  geom_line(aes(y = prediction), colour = "red") +
  theme(
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
  ) + 
  labs(
    x = "Year",
    y = "Life expectancy (years)",
    title = "Predicted and observed life expectancy at birth, selected countries",
    subtitle = "Line: Best segmented regression model for the country",
    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
  )

```
And the following figure shows the same for x = 65:

```{r}
best_breakpoint_model_predictions %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(!is.na(country)) %>% 
  filter(x == 65) %>% 
  ggplot(aes(x = year)) + 
  facet_grid(sex ~ country) + 
  geom_point(aes(y = ex), alpha = 0.35) + 
  geom_line(aes(y = prediction), colour = "red") +
  theme(
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
  ) + 
  labs(
    x = "Year",
    y = "Life expectancy (years)",
    title = "Predicted and observed life expectancy at age 65, selected countries",
    subtitle = "Line: Best segmented regression model for the country",
    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
  )

```

The residuals for all populations (ages, sexes, countries) combined is shown in the following figure:

```{r}
best_breakpoint_model_predictions %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(!is.na(country)) %>%
  mutate(
    residual = prediction - ex
  ) %>% 
  ggplot(aes(year, residual)) + 
  geom_point() + 
  geom_hline(yintercept = 0) + 
  stat_smooth() + 
  labs(
    x = "year", y = "Residual",
    title = "Difference between observed and predicted life expectancies",
    subtitle = "Both starting ages (0 and 65), both sexes and all countries",
    caption = "Predictions based on best-performing 0, 1 or 2 breakpoint models as assessed using BIC\n
    A GAM smoother line has been added to indicate any systematic change in residuals over time"
  )

```

And the following figure shows the residuals separated by starting age:

```{r}
best_breakpoint_model_predictions %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(!is.na(country)) %>%
  mutate(
    residual = prediction - ex
  ) %>% 
  ggplot(aes(year, residual)) + 
  geom_point() + 
  facet_wrap(~x) +
  geom_hline(yintercept = 0) + 
  stat_smooth() + 
  labs(
    x = "year", y = "Residual",
    title = "Difference between observed and predicted life expectancies",
    subtitle = "Separated by starting age (0 and 65), both sexes and all countries",
    caption = "Predictions based on best-performing 0, 1 or 2 breakpoint models as assessed using BIC\n
    A GAM smoother line has been added to indicate any systematic change in residuals over time"
  )

```

The following shows residuals separately by sex:

```{r}
best_breakpoint_model_predictions %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(!is.na(country)) %>%
  mutate(
    residual = prediction - ex
  ) %>% 
  ggplot(aes(year, residual)) + 
  geom_point() + 
  facet_wrap(~sex) +
  geom_hline(yintercept = 0) + 
  stat_smooth() + 
  labs(
    x = "year", y = "Residual",
    title = "Difference between observed and predicted life expectancies",
    subtitle = "Separated by sex, both starting ages and all countries",
    caption = "Predictions based on best-performing 0, 1 or 2 breakpoint models as assessed using BIC\n
    A GAM smoother line has been added to indicate any systematic change in residuals over time"
  )

```

## Discussion

Though there are some years where residuals tend to be very high or very low, these tend to be consistent across all countries, despite differences in the preferred model specification for each country. Key examples include 2003 and 2004, and 2014 and 2015. Both pairs of adjacent year comprise one year with systematically high (or low) residuals, followed by another with systematically low (or high) residuals. These are likely to reflect the somewhat oscillatory behaviour of annual mortality, likely driven by phenomena like forward mortality displacement.

# Query 2: Model estimation by AICc

As the reviewer notes, the models and model selection processes make use of BIC to select from a series of non-nested models. Other information criteria could also be used. The reviewer asks about AICc, a variation of AIC adapted for small sample sizes.

BIC is currently used to select between a zero breakpoint model, the best performing one breakpoint model, and the best performing two breakpoint model.

The identification of the best performing one and two breakpoint models is performed within the `segmented::segmented` function, which selects the model which minimises minus log likelihood (See the `fn.obj` argument definition in the `segmented::seg.control` help file). As the number of parameters and observations will not vary when selecting between one and two breakpoint model specifications for a single population, there does not appear to be any advantage to rerunning this stage with a different objective function.

This means AICc should replace BIC as follows. AICc has been calculated using the `AICcmodavg::AICc` package and function.

```{r}

# Now let's get the AICc for each model

compare_aicc_bic <- 
  segmented_breakpoints_models %>% 
    mutate(bic = map_dbl(mdl_outputs, BIC)) %>%
    mutate(aicc = map_dbl(mdl_outputs, AICcmodavg::AICc)) %>% 
    group_by(code, x, sex) %>% 
    mutate(rank_bic = rank(bic)) %>%
    mutate(rank_aicc = rank(aicc)) %>% 
    ungroup() %>% 
    mutate(both_equal = rank_bic == rank_aicc)

compare_aicc_bic

```
The rank order of AICc and BIC is identical for 104 of the 108 rows above. The differences are as follows:

```{r}
compare_aicc_bic %>% 
  filter(!both_equal)

```

For Spain, x=65, females: BIC ranked the 0 breakpoint model 2nd, AICc ranked it 3rd; BIC ranked the 2 breakpoint model 3rd.
For Germany, x = 0, females: BIC ranked the 1 breakpoint model 1st, AICc ranked it 2nd; BIC ranked the 2 breakpoint model second, AICc ranked it 1st.

## Discussion

Adopting AICc in place of BIC would only have led to one difference in terms of the 'best performing model', out of the 36 datasets (comprising country, sex and starting age combination) for which the three classes of model were each estimated and compared. This was for Germany, for life expectancy from birth (x = 0), and for females. For this population AICc indicated a two breakpoint model be preferred; by contrast BIC suggested the one breakpoint model should be preferred.

# Query 3: Intervention variable to fit the ARIMA model

ARIMA models were not used in the analyses presented in the main manuscript, although they are used within one of the appendices.

Both the `forecast::Arima` function and the `stats::arima` function include an argument `xreg`, for 'external regressors'. This means intervention or control variables can be included in ARIMA model specifications. We know from appendix X that ARIMA(1, 1, 0) models (with autoregression) are often preferred to ARIMA(0, 1, 0) (no autoregression) specifications for many populations. We also know from the response to the first reviewer query that the residuals for the years 2003 and 2004, and 2014 and 2015, in particular, appear somewhat incompatible with a white noise error term assumption. Given both of those observations, an ARIMA(1, 1, 0) model specification with at least one external regressor may be reasonable. Examples of such regressors may include number of days where temperatures were either above or below optimal ranges.

### 2022-10-03 draft_working_paper.Rmd
finally create pdf version of changepoint methods rmd
add appendix for reviewer 1

@@
 
+**Reviewer request: Calculation by AICc**
+
+In the manuscript itself, BIC is used twice:
+
+- To select the best performing 1 and 2 breakpoint model
+- To select, for each population, whether the 0, 1 or 2 breakpoint model is preferable.
+
+At both of these stages, the AICc criterion could be used instead. This has already been implemented using the `AICcmodavg::AICc` package and function.
+
+```{r}
+best_breakpoint_model_predictions_aicc <- 
+  segmented_breakpoints_models %>% 
+    mutate(aicc = map_dbl(mdl_outputs, AICcmodavg::AICc)) %>% 
+    group_by(code, x, sex) %>% 
+    filter(aicc == min(aicc)) %>% 
+    mutate(predictions = map2(mdl_outputs, data, predict)) %>% 
+    mutate(data_augmented = map2(data, predictions, ~ .x %>% mutate(prediction = .y))) %>% 
+    select(code, x, sex, data_augmented) %>% 
+    ungroup() %>% 
+    unnest(data_augmented)
+
+```
+
 First I want to show the values against predicted
@@
 
-And now, the same, but differences
+**Reviewer 1 Request: Can the residuals of the models be presented?**
+
+```{r}
+best_breakpoint_model_predictions %>% 
+  left_join(country_code_lookup) %>% 
+  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
+  filter(!is.na(country)) %>%
+  mutate(
+    residual = prediction - ex
+  ) %>% 
+  ggplot(aes(year, residual)) + 
+  geom_point() + 
+  geom_hline(yintercept = 0) + 
+  stat_smooth() + 
+  labs(
+    x = "year", y = "Residual",
+    title = "Difference between observed and predicted life expectancies",
+    subtitle = "Both starting ages (0 and 65), both sexes and all countries",
+    caption = "Predictions based on best-performing 0, 1 or 2 breakpoint models as assessed using BIC\n
+    A GAM smoother line has been added to indicate any systematic change in residuals over time"
+  )
+
+```
+
+The following shows the same but separated by starting age
+
+```{r}
+best_breakpoint_model_predictions %>% 
+  left_join(country_code_lookup) %>% 
+  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
+  filter(!is.na(country)) %>%
+  mutate(
+    residual = prediction - ex
+  ) %>% 
+  ggplot(aes(year, residual)) + 
+  geom_point() + 
+  facet_wrap(~x) +
+  geom_hline(yintercept = 0) + 
+  stat_smooth() + 
+  labs(
+    x = "year", y = "Residual",
+    title = "Difference between observed and predicted life expectancies",
+    subtitle = "Separated by starting age (0 and 65), both sexes and all countries",
+    caption = "Predictions based on best-performing 0, 1 or 2 breakpoint models as assessed using BIC\n
+    A GAM smoother line has been added to indicate any systematic change in residuals over time"
+  )
+
+```
+
+The follow shows differences by sex
+
+```{r}
+best_breakpoint_model_predictions %>% 
+  left_join(country_code_lookup) %>% 
+  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
+  filter(!is.na(country)) %>%
+  mutate(
+    residual = prediction - ex
+  ) %>% 
+  ggplot(aes(year, residual)) + 
+  geom_point() + 
+  facet_wrap(~sex) +
+  geom_hline(yintercept = 0) + 
+  stat_smooth() + 
+  labs(
+    x = "year", y = "Residual",
+    title = "Difference between observed and predicted life expectancies",
+    subtitle = "Separated by sex, both starting ages and all countries",
+    caption = "Predictions based on best-performing 0, 1 or 2 breakpoint models as assessed using BIC\n
+    A GAM smoother line has been added to indicate any systematic change in residuals over time"
+  )
+
+```
+
+### Conclusions
+
+Though there are some years where residuals tend to be very high or very low, these tend to be consistent across all countries, despite differences in the preferred model specification for each country.
+Key examples include 2003 and 2004, and 2014 and 2015.
+Both pairs of adjacent year comprise one year with systematically high (or low) residuals, followed by another with systematically low (or high) residuals.
+These are likely to reflect the somewhat oscillatory behaviour of annual mortality, likely driven by phenomena like forward mortality displacement.
+
+## Differences in annual mortality
+
+The following shows what the best-performing models suggest are the differences in annual mortality from year to year
 
@@
 ```
+
 # Summary of differences and similarities between segmented approach and changepoint approach

### 2022-10-05 draft_working_paper.Rmd
move additional code in response to reviewer 1 to separate appendix

--- 2022-07-22 draft_working_paper.Rmd
@@
 
-And now, the same, but differences
+## Differences in annual mortality
+
+The following shows what the best-performing models suggest are the differences in annual mortality from year to year
 

### 2022-10-28 appendices/cis_for_models.Rmd
create and push new appendix as pdf

---
title: 'Confidence Intervals for loess fits and linear/segmented regression models '
date: "2022-10-18"
output:
  pdf_document: default
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Aim

This appendix shows the 95% confidence intervals for the loess and linear/segmented models presented in the main paper.

# Data preparation

```{r}
# load packages

library(tidyverse)
library(glue)

# load data 
hmd_lt <- read_rds("https://github.com/JonMinton/change-in-ex/blob/main/data/lifetables.rds?raw=true")

# Labels for codes 
country_code_lookup <- 
  tribble(
    ~code, ~country,
    "DEUTNP", "Germany",
    "DEUTE", "East Germany",
    "DEUTW", "West Germany",
    "ESP", "Spain",
    "FRATNP", "France", 
    "ITA", "Italy",
    "GBRTENW", "England & Wales",
    "GBR_SCO", "Scotland",
    "DEUTSYNTH", "Synthetic Germany",
    "NLD", "Netherlands"
  )

countries_of_interest <- c(
  "GBRTENW",
  "GBR_SCO",
  "GBR_UK",
  "FRATNP",
  "ESP",
  "ITA",
  "DEUTNP",
  "DEUTE", 
  "DEUTW",
  "NLD"
)

source("https://raw.githubusercontent.com/JonMinton/change-in-ex/main/R/make_synthetic_germany_functions.R")

source("https://raw.githubusercontent.com/JonMinton/change-in-ex/main/R/make_pop_selection.R")

change_in_ex_selected_countries <- 
  hmd_ex_selected_countries_with_synth %>% 
    group_by(code, x, sex) %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
    ungroup()

```

# LOESS smoother, confidence intervals

The following shows the LOESS smoother line with confidence intervals for annual changes in life expectancy at birth:

```{r}
p_0_loess <- change_in_ex_selected_countries %>% 
  filter(x == 0) %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(!is.na(country)) %>% 
  filter(between(year, 1980, 2020)) %>% 
  mutate(delta_ex = delta_ex * 52.25) %>% # Convert to weeks
  ggplot(aes(x = year, y = delta_ex)) + 
  geom_point() + 
  stat_smooth(se = TRUE) +  # Changed from se = FALSE
  facet_grid(sex~country) + 
  geom_hline(yintercept = 0) + 
  scale_y_continuous(breaks = seq(-30, 50, by = 10)) +
  theme(
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
  ) + 
  labs(
    x = "Year",
    y = "Change in life expectancy from previous year (in weeks)",
    title = "Annual change in life expectancy at birth, selected countries",
    subtitle = "Line: nonlinear smoother over the points",
    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
  )

p_0_loess
```

And the following shows the same for life expectancy from age 65 years

```{r}
p_65_loess <- change_in_ex_selected_countries %>% 
  filter(x == 65) %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(!is.na(country)) %>% 
  filter(between(year, 1980, 2020)) %>% 
  mutate(delta_ex = delta_ex * 52.25) %>% # Convert to weeks
  ggplot(aes(x = year, y = delta_ex)) + 
  geom_point() + 
  stat_smooth(se = TRUE) + # changed from se = FALSE 
  facet_grid(sex~country) + 
  geom_hline(yintercept = 0) + 
  scale_y_continuous(breaks = seq(-30, 50, by = 10)) +
  theme(
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
  ) + 
  labs(
    x = "Year",
    y = "Change in life expectancy from previous year (in weeks)",
    title = "Annual change in life expectancy at age 65, selected countries",
    subtitle = "Line: nonlinear smoother over the points",
    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
  )

p_65_loess
```

We can see that the 95% CI on the smoother includes values below zero for a number of populations.

Within the above, and in the main paper, the LOESS smoother is called in ggplot2's  `stat_smooth` function, which uses slightly different parameterisation than if the `loess` function in the `stats` packages were called directly. For consistency with the main paper we can extract the point estimate, lower and upper CIs, and standard error, directly from the above ggplot objects. The following function does this:

```{r}
extract_loess_ci <- function(p){
  b <- ggplot_build(p)
  panel_lookup <- b$layout$layout

  # get smoother values
  p_loess <- b$data[[2]]

  # Join and simplify

  sm <- p_loess %>% 
    select(x, y, ymin, ymax, se, PANEL) %>% 
    left_join(panel_lookup)

  # find last year for each country

  max_years <- p$data %>% 
    group_by(country, sex) %>% 
    filter(year == max(year)) %>% 
    select(country, sex, max_year = year)

  sm %>% 
    select(x, y, ymin, ymax, se, sex, country) %>% 
    left_join(max_years) %>% 
    filter(x == max_year)
}

```

And the following code applies the function to the two previously created ggplot2 objects

```{r}
loess_lastyear_x0 <- 
extract_loess_ci(p_0_loess) %>% 
  rename(year = x, y_point = y,
         y_lower = ymin, y_upper = ymax, y_se = se) %>% 
  mutate(x = 0)  %>% # x now starting age  
  select(year, x, sex, country, everything(), -max_year)

loess_lastyear_x65 <- 
extract_loess_ci(p_65_loess) %>% 
  rename(year = x, y_point = y,
         y_lower = ymin, y_upper = ymax, y_se = se) %>% 
  mutate(x = 65)  %>% # x now starting age  
  select(year, x, sex, country, everything(), -max_year)

loess_lastyear_both <- 
  bind_rows(loess_lastyear_x0, loess_lastyear_x65)

loess_lastyear_both
```
The estimates are in weeks of increase per year. The following function produces summary statements for each population

```{r}

summarise_loess <- function(df, cntry, dp = 2){
  df <- df %>% filter(country == cntry)

  last_year <- df$year[1]

  y_x0_pt_m <- df %>% filter(x == 0, sex == "male") %>% pull("y_point") %>% round(dp)
  y_x65_pt_m <- df %>% filter(x == 65, sex == "male") %>% pull("y_point") %>% round(dp)
  y_x0_pt_f <- df %>% filter(x == 0, sex == "female") %>% pull("y_point") %>% round(dp)
  y_x65_pt_f <- df %>% filter(x == 65, sex == "female") %>% pull("y_point") %>% round(dp)

  y_x0_lwr_m <- df %>% filter(x == 0, sex == "male") %>% pull("y_lower") %>% round(dp)
  y_x65_lwr_m <- df %>% filter(x == 65, sex == "male") %>% pull("y_lower") %>% round(dp)
  y_x0_lwr_f <- df %>% filter(x == 0, sex == "female") %>% pull("y_lower") %>% round(dp)
  y_x65_lwr_f <- df %>% filter(x == 65, sex == "female") %>% pull("y_lower") %>% round(dp)

  y_x0_upr_m <- df %>% filter(x == 0, sex == "male") %>% pull("y_upper") %>% round(dp)
  y_x65_upr_m <- df %>% filter(x == 65, sex == "male") %>% pull("y_upper") %>% round(dp)
  y_x0_upr_f <- df %>% filter(x == 0, sex == "female") %>% pull("y_upper") %>% round(dp)
  y_x65_upr_f <- df %>% filter(x == 65, sex == "female") %>% pull("y_upper") %>% round(dp)

  y_x0_se_m <- df %>% filter(x == 0, sex == "male") %>% pull("y_se") %>% round(dp)
  y_x65_se_m <- df %>% filter(x == 65, sex == "male") %>% pull("y_se") %>% round(dp)
  y_x0_se_f <- df %>% filter(x == 0, sex == "female") %>% pull("y_se") %>% round(dp)
  y_x65_se_f <- df %>% filter(x == 65, sex == "female") %>% pull("y_se") %>% round(dp)

  pfall_x0_m <- pnorm(0, y_x0_pt_m, y_x0_se_m) %>% round(dp)
  pfall_x65_m <- pnorm(0, y_x65_pt_m, y_x65_se_m) %>% round(dp)
  pfall_x0_f <- pnorm(0, y_x0_pt_f, y_x0_se_f) %>% round(dp)
  pfall_x65_f <- pnorm(0, y_x65_pt_f, y_x65_se_f) %>% round(dp)

  glue::glue("{cntry} in {last_year}: \nFor males, LOESS estimated annual changes of {y_x0_pt_m} (95% CI {y_x0_lwr_m} to {y_x0_upr_m}) weeks/year for life expectancy at birth, and annual changes of {y_x65_pt_m} (95% CI {y_x65_lwr_m} to {y_x65_upr_m}) weeks/year for life expectancy at age 65; the probability that life expectancy fell in the last observed year was {pfall_x0_m} for life expectancy at birth, and {pfall_x65_m} for life expectancy from age 65. \nFor females, LOESS estimated annual changes of {y_x0_pt_f} (95% CI {y_x0_lwr_f} to {y_x0_upr_f}) weeks/year for life expectancy at birth, and annual changes of {y_x65_pt_f} (95% CI {y_x65_lwr_f} to {y_x65_upr_f}) weeks/year for life expectancy at age 65; the probability that life expectancy fell in the last observed year was {pfall_x0_f} for life expectancy at birth, and {pfall_x65_f} for life expectancy from age 65.")    
}
```

The following are descriptive summaries of life expectancy trends in the last observed year, as estimated through the LOESS method:

```{r}
unique(loess_lastyear_both$country) %>% as.character() %>%  
  map(summarise_loess, df = loess_lastyear_both)

```
We can see from the above summaries that the confidence intervals for all populations tend to include negative values, and that the resulting estimated probabilities that the true value in the last observed year (conditional on the model) is below zero, i.e. falling, is substantial (at least 12%) for populations and starting ages considered. For England & Wales, and Scotland (except males from age 65), it is more probable than not that life expectancy fell in the last observed year.

## Linear regression and Breakpoint confidence intervals

```{r}

estimate_breakpoints_and_pval <- function(df){
  null_mdl <- lm(ex ~ year, data = df) 
  seg_mdl <- segmented::segmented(null_mdl, seg.Z= ~year, psi = 2010)
  seg2_mdl <- segmented::segmented(null_mdl, seg.z= ~year, psi= c(1985, 2010)) # added to test

  list(
    null = null_mdl, 
    seg = seg_mdl,
    seg2 = seg2_mdl
  )
}

segmented_breakpoints_models <- 
  hmd_ex_selected_countries_with_synth %>% 
  filter(code != "DEUTNP") %>%   
  filter(year >= 1979) %>% 
  group_by(code, x, sex) %>% 
  nest() %>% 
  mutate(
    mdl_outputs = map(data, estimate_breakpoints_and_pval)
  ) %>% 
  unnest_longer(mdl_outputs)

# Now let's get the BIC for each model

make_predictions <- function(mdl, dta = tibble(year = 1980:2020)){
  tibble(
    year = dta$year, 
    ex_pred = predict(mdl, newdata = dta)
  )
}

best_model_predictions_descriptions <- 
segmented_breakpoints_models %>% 
  mutate(bic = map_dbl(mdl_outputs, BIC)) %>% 
  group_by(code, x, sex) %>% 
  mutate(rank_bic = rank(bic)) %>%
  filter(rank_bic == 1) %>% 
  mutate(
    best_model = case_when(
      mdl_outputs_id == 'seg2' ~ "Two breakpoints",
      mdl_outputs_id == 'seg'  ~ "One breakpoint", 
      mdl_outputs_id == 'null' ~ "No breakpoints"
    )
  ) %>% 
  mutate(
    first_breakpoint =    map2_dbl(
      mdl_outputs_id, mdl_outputs, 
      function(x, y){
        if (x == 'null'  ) {NA_real_} else {
          y[["psi"]][1,2]
        }
      }                        
      ),
      first_breakpoint_se = map2_dbl(
        mdl_outputs_id, mdl_outputs,
        function(x, y){
        if (x == 'null'  ) {NA_real_} else {
          y[["psi"]][1,3]
        }

        }
      ),
    second_breakpoint = map2_dbl(
      mdl_outputs_id, mdl_outputs,
      function(x, y){
        if (x == 'seg2') {
          y[["psi"]][2,2]
        } else  {
            NA_real_
          }
      }                           
      ),
    second_breakpoint_se = map2_dbl(
      mdl_outputs_id, mdl_outputs,
      function(x, y){
        if (x == 'seg2') {
          y[["psi"]][2,3]
        } else  {
            NA_real_
          }
      }
    )
  ) %>%  # let's add predictions too 
  mutate(
    pred_data = map(mdl_outputs, make_predictions),
    joined_data = map2(data, pred_data, left_join)
  ) %>% 
  select(code, x, sex, joined_data, first_breakpoint:second_breakpoint_se) %>% 
  unnest(joined_data)

```

```{r}
best_breakpoint_model_predictions <- 
  segmented_breakpoints_models %>% 
    mutate(bic = map_dbl(mdl_outputs, BIC)) %>% 
    group_by(code, x, sex) %>% 
    filter(bic == min(bic)) %>% 
    mutate(predictions = map2(mdl_outputs, data, predict, interval = "confidence", level = 0.95)) %>% 
    mutate(data_augmented = map2(data, predictions, bind_cols)) %>% 
    select(code, x, sex, data_augmented) %>% 
    ungroup() %>% 
    unnest(data_augmented)

```

```{r}
best_breakpoint_model_predictions %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(!is.na(country)) %>% 
  filter(x == 0) %>% 
  ggplot(aes(x = year)) + 
  facet_grid(sex ~ country) + 
  geom_point(aes(y = ex), alpha = 0.35) + 
  geom_line(aes(y = fit), colour = "red") +
  geom_ribbon(aes(ymin = lwr, ymax = upr), fill = "red", colour = NA, alpha = 0.5) +
  theme(
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
  ) + 
  labs(
    x = "Year",
    y = "Life expectancy (years)",
    title = "Predicted and observed life expectancy at birth, selected countries",
    subtitle = "Line: Best segmented regression model for the country",
    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
  )

```
The confidence intervals are quite hard to see given the data points. In the figure below they are shown without the points

```{r}
best_breakpoint_model_predictions %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(!is.na(country)) %>% 
  filter(x == 0) %>% 
  ggplot(aes(x = year)) + 
  facet_grid(sex ~ country) + 
#  geom_point(aes(y = ex), alpha = 0.35) + 
  geom_line(aes(y = fit), colour = "red") +
  geom_ribbon(aes(ymin = lwr, ymax = upr), fill = "red", colour = NA, alpha = 0.5) +
  theme(
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
  ) + 
  labs(
    x = "Year",
    y = "Life expectancy (years)",
    title = "Predicted and observed life expectancy at birth, selected countries",
    subtitle = "Line: Best segmented regression model for the country",
    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
  )

```
They are repeated, with and then without points, for life expectancy from age 65 below:

```{r}
best_breakpoint_model_predictions %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(!is.na(country)) %>% 
  filter(x == 65) %>% 
  ggplot(aes(x = year)) + 
  facet_grid(sex ~ country) + 
  geom_point(aes(y = ex), alpha = 0.35) + 
  geom_line(aes(y = fit), colour = "red") +
  geom_ribbon(aes(ymin = lwr, ymax = upr), fill = "red", colour = NA, alpha = 0.5) +
  theme(
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
  ) + 
  labs(
    x = "Year",
    y = "Life expectancy (years)",
    title = "Predicted and observed life expectancy at age 65, selected countries",
    subtitle = "Line: Best segmented regression model for the country",
    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
  )

best_breakpoint_model_predictions %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(!is.na(country)) %>% 
  filter(x == 65) %>% 
  ggplot(aes(x = year)) + 
  facet_grid(sex ~ country) + 
#  geom_point(aes(y = ex), alpha = 0.35) + 
  geom_line(aes(y = fit), colour = "red") +
  geom_ribbon(aes(ymin = lwr, ymax = upr), fill = "red", colour = NA, alpha = 0.5) +
  theme(
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
  ) + 
  labs(
    x = "Year",
    y = "Life expectancy (years)",
    title = "Predicted and observed life expectancy at age 65, selected countries",
    subtitle = "Line: Best segmented regression model for the country",
    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
  )

```

```{r}
pred_differenced <- 
  best_breakpoint_model_predictions %>% 
    left_join(country_code_lookup) %>% 
    mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
    filter(!is.na(country)) %>% 
    group_by(country, sex, x) %>% 
    arrange(year) %>% 
    mutate(
      diff_ex = 52.25 * (ex - lag(ex)),
      diff_fit = 52.25 * (fit - lag(fit)),
      adj_upr = 52.25 * (upr - fit),
      adj_lwr = 52.25 * (lwr - fit),
      diff_upr = diff_fit + adj_upr,
      diff_lwr = diff_fit + adj_lwr
    ) %>% 
    filter(!is.na(diff_fit)) 
```

The following shows differences with confidence intervals for life expectancy from birth:

```{r}
pred_differenced %>%
  filter(x == 0) %>% 
  ggplot(aes(year)) + 
  geom_point(aes(y = diff_ex), alpha = 0.35) +
  facet_grid(sex ~ country) +
  geom_hline(yintercept = 0) + 
  geom_line(aes(y = diff_fit), colour = "red") +
  geom_ribbon(aes(ymin = diff_lwr, ymax = diff_upr), colour = NA, fill = "red", alpha = 0.35) +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
  ) + 
  scale_y_continuous(breaks = seq(-30, 50, by = 10)) +
  labs(
    x = "Year",
    y = "Change in life expectancy from previous year (weeks)",
    title = "Predicted and observed annual changes in life expectancy at birth, selected countries",
    subtitle = "Line: Best segmented regression model for the country",
    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
  )

```

And the following shows the same for life expectancy changes from age 65

```{r}
pred_differenced %>%
  filter(x == 65) %>% 
  ggplot(aes(year)) + 
  geom_point(aes(y = diff_ex), alpha = 0.35) +
  facet_grid(sex ~ country) +
  geom_hline(yintercept = 0) + 
  geom_line(aes(y = diff_fit), colour = "red") +
  geom_ribbon(aes(ymin = diff_lwr, ymax = diff_upr), colour = NA, fill = "red", alpha = 0.35) +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
  ) + 
  scale_y_continuous(breaks = seq(-30, 50, by = 10)) +
  labs(
    x = "Year",
    y = "Change in life expectancy from previous year (weeks)",
    title = "Predicted and observed annual changes in life expectancy at age 65, selected countries",
    subtitle = "Line: Best segmented regression model for the country",
    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
  )

```
As with the LOESS model, we can can present the point estimates, 95% CIs, and implied probabilities that the values in the last observed year are below zero.

```{r}
summarise_reg_seg <- function(df, cntry, dp = 2){
  df <- df %>% 
    filter(country == cntry) %>% 
    filter(year == max(year))

  last_year <- max(df$year)

  y_x0_pt_m <- df %>% filter(x == 0, sex == "male") %>% pull("diff_fit") 
  y_x65_pt_m <- df %>% filter(x == 65, sex == "male") %>% pull("diff_fit") 
  y_x0_pt_f <- df %>% filter(x == 0, sex == "female") %>% pull("diff_fit") 
  y_x65_pt_f <- df %>% filter(x == 65, sex == "female") %>% pull("diff_fit")

  y_x0_lwr_m <- df %>% filter(x == 0, sex == "male") %>% pull("diff_lwr") 
  y_x65_lwr_m <- df %>% filter(x == 65, sex == "male") %>% pull("diff_lwr") 
  y_x0_lwr_f <- df %>% filter(x == 0, sex == "female") %>% pull("diff_lwr") 
  y_x65_lwr_f <- df %>% filter(x == 65, sex == "female") %>% pull("diff_lwr")

  y_x0_upr_m <- df %>% filter(x == 0, sex == "male") %>% pull("diff_upr") 
  y_x65_upr_m <- df %>% filter(x == 65, sex == "male") %>% pull("diff_upr") 
  y_x0_upr_f <- df %>% filter(x == 0, sex == "female") %>% pull("diff_upr") 
  y_x65_upr_f <- df %>% filter(x == 65, sex == "female") %>% pull("diff_upr")

  y_x0_se_m <-  c(y_x0_upr_m - y_x0_pt_m, y_x0_pt_m - y_x0_lwr_m) %>% `/`(1.96) %>% mean() %>% round(dp)
  y_x65_se_m <- c(y_x65_upr_m - y_x65_pt_m, y_x65_pt_m - y_x65_lwr_m) %>% `/`(1.96) %>% mean() %>% round(dp)
  y_x0_se_f <-  c(y_x0_upr_f - y_x0_pt_f, y_x0_pt_f - y_x0_lwr_f) %>% `/`(1.96) %>% mean() %>% round(dp)
  y_x65_se_f <- c(y_x65_upr_f - y_x65_pt_f, y_x65_pt_f - y_x65_lwr_f) %>% `/`(1.96) %>% mean() %>% round(dp)

  y_x0_lwr_m <- y_x0_lwr_m %>% round(dp)
  y_x65_lwr_m <- y_x65_lwr_m %>% round(dp) 
  y_x0_lwr_f <- y_x0_lwr_f %>% round(dp)
  y_x65_lwr_f <- y_x65_lwr_f %>% round(dp)

  y_x0_upr_m <- y_x0_upr_m %>% round(dp)
  y_x65_upr_m <- y_x65_upr_m %>% round(dp)
  y_x0_upr_f <- y_x0_upr_f %>% round(dp)
  y_x65_upr_f <- y_x65_upr_f %>% round(dp)

  y_x0_pt_m <- y_x0_pt_m %>% round(dp)
  y_x65_pt_m <- y_x65_pt_m %>% round(dp)
  y_x0_pt_f <- y_x0_pt_f %>% round(dp)
  y_x65_pt_f <- y_x65_pt_f %>% round(dp)

  pfall_x0_m <- pnorm(0, y_x0_pt_m, y_x0_se_m) %>% round(dp)
  pfall_x65_m <- pnorm(0, y_x65_pt_m, y_x65_se_m) %>% round(dp)
  pfall_x0_f <- pnorm(0, y_x0_pt_f, y_x0_se_f) %>% round(dp)
  pfall_x65_f <- pnorm(0, y_x65_pt_f, y_x65_se_f) %>% round(dp)

  glue::glue("{cntry} in {last_year}: \nFor males, segmented/linear estimated annual changes of {y_x0_pt_m} (95% CI {y_x0_lwr_m} to {y_x0_upr_m}) weeks/year for life expectancy at birth, and annual changes of {y_x65_pt_m} (95% CI {y_x65_lwr_m} to {y_x65_upr_m}) weeks/year for life expectancy at age 65; the probability that life expectancy fell in the last observed year was {pfall_x0_m} for life expectancy at birth, and {pfall_x65_m} for life expectancy from age 65. \nFor females, segmented/linear estimated annual changes of {y_x0_pt_f} (95% CI {y_x0_lwr_f} to {y_x0_upr_f}) weeks/year for life expectancy at birth, and annual changes of {y_x65_pt_f} (95% CI {y_x65_lwr_f} to {y_x65_upr_f}) weeks/year for life expectancy at age 65; the probability that life expectancy fell in the last observed year was {pfall_x0_f} for life expectancy at birth, and {pfall_x65_f} for life expectancy from age 65.")    
}

```

As before, we can produce summary statements for each country as follows:

```{r}
unique(pred_differenced$country) %>% as.character() %>%  
  map(summarise_reg_seg , df = pred_differenced)

```

### 2022-10-28 draft_working_paper.Rmd
create and push new appendix as pdf

@@
   mutate(
-    ch_ex      = 52.25 * (ex - lag(ex)                 ), 
+    ch_ex      = 9 (ex - lag(ex)                 ), 
     ch_ex_pred = 52.25 * (prediction - lag(prediction) )

### 2022-11-02 draft_working_paper.Rmd
create tiff versions of figures for ms

@@
 ggsave("figures/proposed/e0_trend.png", height = 20, width = 20, units = "cm", dpi = 300)
+ggsave("figures/proposed/e0_trend.tiff", height = 20, width = 20, units = "cm", dpi = 300)
 
@@
 ggsave("figures/proposed/e65_trend.png", height = 20, width = 20, units = "cm", dpi = 300)
+ggsave("figures/proposed/e65_trend.tiff", height = 20, width = 20, units = "cm", dpi = 300)
 ```
@@
 ggsave("figures/proposed/che0_smoother.png", height = 15, width = 25, units = "cm", dpi = 300)
+ggsave("figures/proposed/che0_smoother.tiff", height = 15, width = 25, units = "cm", dpi = 300)
 
@@
 ggsave("figures/proposed/che65_smoother.png", height = 15, width = 25, units = "cm", dpi = 300)
+ggsave("figures/proposed/che65_smoother.tiff", height = 15, width = 25, units = "cm", dpi = 300)
 
@@
 ggsave("figures/proposed/average_improvement1980-2020.png", height = 20, width = 30, units = "cm", dpi = 300)
+ggsave("figures/proposed/average_improvement1980-2020.tiff", height = 20, width = 30, units = "cm", dpi = 300)
 
@@
 ggsave("figures/proposed/reduced_average_improvement1980-2020.png", height = 20, width = 25, units = "cm", dpi = 300)
+ggsave("figures/proposed/reduced_average_improvement1980-2020.tiff", height = 20, width = 25, units = "cm", dpi = 300)
 
@@
 ggsave("figures/proposed/e0_breakpoint.png", height = 15, width = 25, units = "cm", dpi = 300)
+ggsave("figures/proposed/e0_breakpoint.tiff", height = 15, width = 25, units = "cm", dpi = 300)
 
@@
 ggsave("figures/proposed/e65_breakpoint.png", height = 15, width = 25, units = "cm", dpi = 300)
-
+ggsave("figures/proposed/e65_breakpoint.tiff", height = 15, width = 25, units = "cm", dpi = 300)
 ```
@@
 ggsave("figures/proposed/ch_e0_breakpoint.png", height = 15, width = 25, units = "cm", dpi = 300)
+ggsave("figures/proposed/ch_e0_breakpoint.tiff", height = 15, width = 25, units = "cm", dpi = 300)
 
@@
 ggsave("figures/proposed/ch_e65_breakpoint.png", height = 15, width = 25, units = "cm", dpi = 300)
-
+ggsave("figures/proposed/ch_e65_breakpoint.tiff", height = 15, width = 25, units = "cm", dpi = 300)
 ```

### 2022-11-02 draft_working_paper.Rmd
correct code; create tiffs

@@
   mutate(
-    ch_ex      = 9 (ex - lag(ex)                 ), 
+    ch_ex      = 52.25 * (ex - lag(ex)),             
     ch_ex_pred = 52.25 * (prediction - lag(prediction) )
